---
title: "01_load_data"
output: html_document
date: "2026-01-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data loading, wrangling and cleaning workflow:
1. Load in flux data (parse time stamps, rename variables, add site names)
2. Handle missing/-9999 values - convert to NA
3. Convert half-hourly data to daily
4. Growing season filter
5. Drop NA once before fitting model


1. Load in an eddy flux file

**read_flux function:**
- Reads in one flux csv file 
- Adds column specifying site from the filename
- Works for HH data, see below for daily data
- Outputs half-hourly data

```{r}

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

**This function (read_daily_flux):**
- Same as read_flux
- Reads in daily data and gives daily data

```{r}
read_daily_flux <- function(fname) {
  
   #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{3}[0-9]?", no_ext)
)
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP to datetime and set as index
  df <- df %>%
    mutate(date = ymd(TIMESTAMP)) %>%
    select(-TIMESTAMP) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, site)
  
  return(df)
  
}
```

2. Handle the missing/-9999 values

```{r}

#Specify which variables you want, otherwise it will read dates and show an error

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa")

#Clean dataframe, convert to 'na' if '-9999'

df_clean <- df %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```

3. Convert half-hourly data to hourly
- Added a new section that ensures dry days are not falsely added, it doesn't confuse NA = no data and 0 = true zero rainfall
- After daily conversion, added a mutate chunk that converts NaN back to NA so there is not later confusion

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
      Precip = Precip * SEC_2_HALF_HOUR
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily, with additional Precip section
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(site, date) %>%
    summarise(
      Precip = ifelse(
        sum(!is.na(Precip)) > 0,
        sum(Precip, na.rm = TRUE),
        NA
      ),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      Qle = mean(Qle, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(df)
  
}

#Apply to dataset

df_daily <- convert_to_daily(df_clean)

#Convert NaN ('Not a number') back to NA

df_daily <- df_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

```

5. Growing season filter
Currently:
- Shortwave radiation threshold
- GPP > 0 threshold

```{r}
filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}

#Apply to dataset

growing_season_data <- filter_gs(df)
```

6. Drop NA values

```{r}
df_monthly <- monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle)

```


Growing season:
- Currently a bit too harsh
- Assumes that photosynthesis responds rapidly instantly to radiation
 - growing season is defined purely by light
 
 Ideas:
 - PFT-specific thresholds?
 Would require having a column in each site specifying its PFT
 - Soft GPP threshold instead of GPP > 0 - percentile or 0.5
 - Add temperature, growing season often light and thermal limited
 - Rolling window - sustained activity. When mean GPP over ~7 days is positive?
 - Radiation percentile instead of absolute threshold which may be wrong for some sites/seasons, adapts across sites
 - Soil moisture/precipitation?
 - **Top 3 months of highest GPP during the year**
 







