---
title: "explore_GAM"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in two flux csv sites, fit GAMs training and testing within and out of site

Going to read in two sites (daily data) using the function I made
AU-Tum - evergreen broadleaf forest, mild with no dry season, warm summer
AU-Cum - evergreen broadleaf forest

Aims:
1. Plot average, max and min GPP per year per site

2. Look at VPD across both sites, is there any issues?

3. Fit GAMs
- Which met variables explain the most variation in GPP? VPD, temp, shortwave radiation...Which GAM is the best? One with only VPD, one with only shortwave radiation, one with both? R^2 and RMSE

- Training and testing within site - can a GAM trained on 80% of the data from AU-Tum (e.g., years 1-18) predict GPP on the final 20% (e.g., last 2 years?)

- Training and testing out of site - can a GAM trained on 100% of data from AU-Tum predict GPP of AU-Cum?


```{r}
rm(list = ls())
```


```{r}
library(tidyverse)

read_daily_flux <- function(fname) {
  
   #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{3}[0-9]?", no_ext)
)
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP to datetime and set as index
  df <- df %>%
    mutate(date = ymd(TIMESTAMP)) %>%
    select(-TIMESTAMP) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, site)
  
  return(df)
  
}
```

Read in flux sites:

```{r}
Tum_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Tum_FLUXDATAKIT_FULLSET_DD_2002_2017_2-3.csv")
```

```{r}
Cum_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Cum_FLUXDATAKIT_FULLSET_DD_2013_2018_2-3.csv")
```

I want to filter out growing season using short-wave radiation:

```{r}
define_gs <- function(df, rad_thresh = 100) {
  df <- df %>%
    mutate(
      growing_season = SW_rad > rad_thresh & GPP > 0
    )
  
  return(df)
}
```

Apply to datasets:

```{r}
Tum_data <- define_gs(Tum_data)

Cum_data <- define_gs(Cum_data)
```

Filter out to final subset of the data:

```{r}

Tum_final <- Tum_data %>% filter(growing_season)

Cum_final <- Cum_data %>% filter(growing_season)


```

Confirm:

```{r}
ggplot(Cum_data, aes(date, GPP)) +
  geom_line() +
  geom_point(aes(color = growing_season)) +
  scale_color_manual(values = c("grey70", "darkgreen")) +
  theme_minimal()

```
```{r}
summary(Tum_final$GPP)
```

Filtered by growing season, by shortwave radiation threshold and made sure GPP > 0 

Now we can start to plot GPP per year.

```{r}
library(ggplot2)

ggplot(Cum_final, aes(year, GPP)) +
  geom_point(alpha = 0.6, color = "green") +
  labs(title = "Aus Cum GPP by year") +
  theme_minimal()



```
So now, I want to plot average, max and minimum GPP per year for each site

```{r}
library(dplyr)

#Combine the datasets

dataframe_aus <- bind_rows(Tum_final, Cum_final)

```

Summarise per site per year

```{r}
df_summary <- dataframe_aus %>%
  group_by(site, year) %>%
  summarise(
    GPP_mean = mean(GPP, na.rm = TRUE),
    GPP_max = max(GPP, na.rm = TRUE),
    GPP_min = min(GPP, na.rm = TRUE)
  ) %>%
  
  ungroup()
```


plot

```{r}
library(tidyr)
library(ggplot2)

df_long <- df_summary %>%
  pivot_longer(
    cols = c(GPP_mean, GPP_max, GPP_min),
    names_to = "stat",
    values_to = "GPP"
  )

ggplot(df_long, aes(x = year, y = GPP, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "GPP", x = "Year", color = "Statistic")

```
```{r}
ggplot(df_summary, aes(x = year, group = site)) +
  geom_ribbon(aes(ymin = GPP_min, ymax = GPP_max, fill = site), alpha = 0.2) +
  geom_line(aes(y = GPP_mean, color = site), size = 1) +
  theme_minimal() +
  labs(y = "GPP", x = "Year")

```
Now let's have a look at the VPD

```{r}
summary(Tum_data$VPD)
```

```{r}
summary(Cum_data$VPD)
```

Double check VPD
1. Loaded in daily data, filtered to growing season
now
2. Load in half-hourly data, convert to daily, in theory the VPD range should be exactly the same

```{r}
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{3}[0-9]?", no_ext)
)
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

```{r}
Cum_data_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Cum_FLUXDATAKIT_FULLSET_HH_2013_2018_2-3.csv")
```

```{r}
Tum_data_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Tum_FLUXDATAKIT_FULLSET_HH_2002_2017_2-3.csv")
```

Now convert these datasets to daily:

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      Qle = mean(Qle, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE)
    )
  
  return(df)
  
}
```

```{r}
daily_Tum <- convert_to_daily(Tum_data_HH)

daily_cum <- convert_to_daily(Cum_data_HH)
```
```{r}
summary(daily_Tum$VPD)
```

```{r}
summary(daily_cum$VPD)
```

Now, I am going to try fit GAMs. Using Cum_final and Tum_final. 

Data - daily, growing-season GPP data for each site - Cum_final and Tum_final
Met variables - Tair, VPD, SW_rad

```{r}
#Just gonna drop missing variables if there are any 

Tum_final <- Tum_final %>% drop_na()

Cum_final <- Cum_final %>% drop_na()

```

Create training/testing split:

Temporal split - first 13 years train, last 2 years test

```{r}

train_data <- Tum_final %>% filter(year <= 2013)
test_data <- Tum_final %>% filter(year > 2013)

```

Fit a GAM:

```{r}
library(mgcv)
library(nlme)
```

```{r}
gam_model1 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad), 
                  data = train_data,
                  method = "REML")
```

Predict on test data:

```{r}

predictions_gam1 <- predict(gam_model1, newdata = test_data)

results_gam <- data.frame(
  observed = test_data$GPP,
  predicted = predictions_gam1
)

```

Evaluate performance:

```{r}
rmse_gam <- sqrt(mean((results_gam$observed - results_gam$predicted)^2))
rsq_gam  <- cor(results_gam$observed, results_gam$predicted)^2
nse_gam  <- 1 - sum((results_gam$observed - results_gam$predicted)^2) /
                 sum((results_gam$observed - mean(results_gam$observed))^2)

cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")

```

```{r}
library(ggplot2)

ggplot(results_gam, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")

```
GAM 1 - predictors VPD, Tair, and Shortwave radiation
Results:
RMSE: 2.075 (square of the residual, same as GPP, the lower RMSE the better)
R-squared: 0.6877 (proportion of the variation in the dependent variable that is predictable from the independent variables, model explains 68% of the variability in the dependent variable)
NSE: 0.626

Now lets try train the model on Au-Tum site and then test on the Au-Cum site

```{r}
train_data_2 <- Tum_final
test_data_2 <- Cum_final
```

```{r}
library(mgcv)
library(nlme)

gam_model2 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad), 
                  data = train_data_2,
                  method = "REML")


```

Predict on the test data:

```{r}
predictions_gam2 <- predict(gam_model2, newdata = test_data_2)

results_gam_2 <- data.frame(
  observed = test_data_2$GPP,
  predicted = predictions_gam2
)

```

Evaluate performance

```{r}
rmse_gam2 <- sqrt(mean((results_gam_2$observed - results_gam_2$predicted)^2))
rsq_gam2  <- cor(results_gam_2$observed, results_gam_2$predicted)^2
nse_gam2  <- 1 - sum((results_gam_2$observed - results_gam_2$predicted)^2) /
                 sum((results_gam_2$observed - mean(results_gam_2$observed))^2)

cat("RMSE:", rmse_gam2, "\n")
cat("R-squared:", rsq_gam2, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam2, "\n")

```

```{r}
library(ggplot2)

ggplot(results_gam_2, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")

```
```{r}
summary(train_data_2[, c("Tair", "VPD", "SW_rad")])
summary(test_data_2[, c("Tair", "VPD", "SW_rad")])
```
Trained out of site - not very good at predicting another site.

Today I have done:
- Finished functions reading flux csv, extracting site names
- Started drafting GAMs, trained two GAMS - one within site and one out of site using GPP ~ VPD, Shortwave radiation and Tair

Thursday 22nd Jan:
- Leave-one-out experiment. Load in 3 different sites (forest, grassland, crop?) - run GAM, which variable is the most important for predicting GPP in each ecosystem?
- Does variable importance differ across ecosystems? (Tair, VPD, Shortwave radiation)

Au_Tum = Australian evergreen broadleaf forest

IT_BCi = Italian croplands, Borgo Cioffi 

```{r}

it_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_IT-BCi_FLUXDATAKIT_FULLSET_DD_2004_2020_2-3.csv")

```

Filter out negative GPP and growing season:

```{r}
it_data <- define_gs(it_data)
```

```{r}
it_final <- it_data %>% filter(growing_season)
```

Double check:

```{r}
ggplot(it_data, aes(date, GPP)) +
  geom_line() +
  geom_point(aes(color = growing_season)) +
  scale_color_manual(values = c("grey70", "darkgreen")) +
  theme_minimal()
```

```{r}
summary(it_final$VPD)
```

Not sure about the range of VPD - really need to ask Martin about units of VPD. 

Now, going to fit GAMs for the leave-one-out experiment





