---
title: "explore_GAM"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
```

Read in two flux csv sites, fit GAMs training and testing within and out of site

Going to read in two sites (daily data) using the function I made
AU-Tum - evergreen broadleaf forest, mild with no dry season, warm summer
AU-Cum - evergreen broadleaf forest

Aims:
1. Plot average, max and min GPP per year per site

2. Look at VPD across both sites, is there any issues?

3. Fit GAMs
- Which met variables explain the most variation in GPP? VPD, temp, shortwave radiation...Which GAM is the best? One with only VPD, one with only shortwave radiation, one with both? R^2 and RMSE

- Training and testing within site - can a GAM trained on 80% of the data from AU-Tum (e.g., years 1-18) predict GPP on the final 20% (e.g., last 2 years?)

- Training and testing out of site - can a GAM trained on 100% of data from AU-Tum predict GPP of AU-Cum?


```{r}
rm(list = ls())
```


```{r}
library(tidyverse)

read_daily_flux <- function(fname) {
  
   #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{3}[0-9]?", no_ext)
)
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP to datetime and set as index
  df <- df %>%
    mutate(date = ymd(TIMESTAMP)) %>%
    select(-TIMESTAMP) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, site)
  
  return(df)
  
}
```

Read in flux sites:

```{r}
Tum_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Tum_FLUXDATAKIT_FULLSET_DD_2002_2017_2-3.csv")
```

```{r}
Cumb_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Cum_FLUXDATAKIT_FULLSET_DD_2013_2018_2-3.csv")
```

I want to filter out growing season using short-wave radiation:

```{r}
define_gs <- function(df, rad_thresh = 100) {
  df <- df %>%
    mutate(
      growing_season = SW_rad > rad_thresh & GPP > 0
    )
  
  return(df)
}
```

Apply to datasets:

```{r}
Tum_data <- define_gs(Tum_data)

Cumb_data <- define_gs(Cumb_data)
```

Filter out to final subset of the data:

```{r}

Tum_final <- Tum_data %>% filter(growing_season)

Cumb_final <- Cumb_data %>% filter(growing_season)


```

Confirm:

```{r}
ggplot(Cumb_data, aes(date, GPP)) +
  geom_line() +
  geom_point(aes(color = growing_season)) +
  scale_color_manual(values = c("grey70", "darkgreen")) +
  theme_minimal()

```
```{r}
summary(Tum_final$GPP)
```

Filtered by growing season, by shortwave radiation threshold and made sure GPP > 0 

Looking at some GPP plots for each site:

```{r}
library(ggplot2)

ggplot(Cumb_final, aes(year, GPP)) +
  geom_point(alpha = 0.6, color = "green") +
  labs(title = "Aus Cumb GPP by year") +
  theme_minimal()



```
So now, I want to plot average, max and minimum GPP per year for each site

```{r}
library(dplyr)

#Combine the datasets

dataframe_aus <- bind_rows(Tum_final, Cumb_final)

```

Summarise per site per year

```{r}
df_summary <- dataframe_aus %>%
  group_by(site, year) %>%
  summarise(
    GPP_mean = mean(GPP, na.rm = TRUE),
    GPP_max = max(GPP, na.rm = TRUE),
    GPP_min = min(GPP, na.rm = TRUE)
  ) %>%
  
  ungroup()
```

plot

```{r}
library(tidyr)
library(ggplot2)

df_long <- df_summary %>%
  pivot_longer(
    cols = c(GPP_mean, GPP_max, GPP_min),
    names_to = "stat",
    values_to = "GPP"
  )

ggplot(df_long, aes(x = year, y = GPP, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "GPP", x = "Year", color = "Statistic")

```
```{r}
ggplot(df_summary, aes(x = year, group = site)) +
  geom_ribbon(aes(ymin = GPP_min, ymax = GPP_max, fill = site), alpha = 0.2) +
  geom_line(aes(y = GPP_mean, color = site), size = 1) +
  theme_minimal() +
  labs(y = "GPP", x = "Year")

```
Now let's have a look at the VPD

```{r}
summary(Tum_data$VPD)
```

```{r}
summary(Cumb_data$VPD)
```

Now, I am going to try fit GAMs. Using Cum_final and Tum_final. 

Data - daily, growing-season GPP data for each site - Cum_final and Tum_final
Met variables - Tair, VPD, SW_rad

```{r}
#Just gonna drop missing variables if there are any 

Tum_final <- Tum_final %>% drop_na()

Cumb_final <- Cumb_final %>% drop_na()

```

GAM: Tumbawumba site - training and testing within site

Temporal split - first 13 years train, last 2 years test

```{r}

train_data <- Tum_final %>% filter(year <= 2013)
test_data <- Tum_final %>% filter(year > 2013)

```

Fit a GAM:

```{r}
library(mgcv)
library(nlme)
```

```{r}
gam_model1 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad), 
                  data = train_data,
                  method = "REML")
```

Predict on test data:

```{r}

predictions_gam1 <- predict(gam_model1, newdata = test_data)

results_gam <- data.frame(
  observed = test_data$GPP,
  predicted = predictions_gam1
)

```

Evaluate performance:

```{r}
rmse_gam <- sqrt(mean((results_gam$observed - results_gam$predicted)^2))
rsq_gam  <- cor(results_gam$observed, results_gam$predicted)^2
nse_gam  <- 1 - sum((results_gam$observed - results_gam$predicted)^2) /
                 sum((results_gam$observed - mean(results_gam$observed))^2)

cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")

```

Plot observed vs predicted:

```{r}
library(ggplot2)

ggplot(results_gam, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")

```
GAM 1 - predictors VPD, Tair, and Shortwave radiation

Results:
RMSE: 2.075 (square of the residual, same as GPP, the lower RMSE the better)
R-squared: 0.6877 (proportion of the variation in the dependent variable that is predictable from the independent variables, model explains 68% of the variability in the dependent variable)
NSE: 0.626

Now lets try train the model on Au-Tum site and then test on the Au-Cum site (out-of-site testing)

```{r}
train_data_2 <- Tum_final
test_data_2 <- Cumb_final
```

```{r}
library(mgcv)
library(nlme)

gam_model2 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad), 
                  data = train_data_2,
                  method = "REML")


```

Predict on the test data:

```{r}
predictions_gam2 <- predict(gam_model2, newdata = test_data_2)

results_gam_2 <- data.frame(
  observed = test_data_2$GPP,
  predicted = predictions_gam2
)

```

Evaluate performance

```{r}
rmse_gam2 <- sqrt(mean((results_gam_2$observed - results_gam_2$predicted)^2))
rsq_gam2  <- cor(results_gam_2$observed, results_gam_2$predicted)^2
nse_gam2  <- 1 - sum((results_gam_2$observed - results_gam_2$predicted)^2) /
                 sum((results_gam_2$observed - mean(results_gam_2$observed))^2)

cat("RMSE:", rmse_gam2, "\n")
cat("R-squared:", rsq_gam2, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam2, "\n")

```

```{r}
library(ggplot2)

ggplot(results_gam_2, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")

```
```{r}
summary(train_data_2[, c("Tair", "VPD", "SW_rad")])
summary(test_data_2[, c("Tair", "VPD", "SW_rad")])
```
Trained out of site - not very good at predicting another site.

Why is this?
- Two sites, both evergreen broadleaf forests
- Both not very far from one another?
- Different dominant species?
- Different precipitation?
- Is one managed and one is not?
 
Today I have done:
- Finished functions reading flux csv, extracting site names
- Started drafting GAMs, trained two GAMS - one within site and one out of site using GPP ~ VPD, Shortwave radiation and Tair

Saturday:
- New flux data, load that in and fit some GAMs on that data

```{r}

library(tidyverse)

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

Load in two sites:

AU_Das - Daly River Savanna. Tropical savanna

US_Me3 - Evergreen needleleaf forest, mediterranean mild with dry, warm summers, Oregon

```{r}
sav_data <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

enf_data <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_US-Me3_FLUXNET2015_FULLSET_HH_2004-2009_1-4.csv")

```

Now I am going to convert to daily and filter growing season. 
Note: I wonder if there is a quicker way of doing this?

Improve this growing season filter? 
Should i have different growing season thresholds for different climates? Could I combine shortwave radiation with time of year/season/different for northern and southern hemisphere?

Going to make the growing season filter quicker by taking out the middle bit:

```{r}

filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}

```

```{r}
enf_gs <- filter_gs(enf_data)

sav_gs <- filter_gs(sav_data)
```

Convert to daily:

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
      Precip = Precip * SEC_2_HALF_HOUR
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(site, date) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      Qle = mean(Qle, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(df)
  
}
```

```{r}
enf_final <- convert_to_daily(enf_gs)
sav_final <- convert_to_daily(sav_gs)
```

now we have our final dataframes yippee. 

Drop missing values if there are any:

```{r}
enf_final <- enf_final %>% drop_na()

sav_final <- sav_final %>% drop_na()

```

Load in libraries

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
```

Remove missing value flags (-999)

```{r}
enf_final <- enf_final %>% filter_all(all_vars(. >= -9000) )
```

```{r}
sav_final <- sav_final %>% filter_all(all_vars(. >= -9000))
```

Water stress metric:
- PET = atmospheric demand
- CWD = cumulative water deficit, how much atmospheric demand for water exceeds the available water supply

PET:

```{r}
priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}
```


```{r}
sav_final$pet <- priestley_taylor_pet(sav_final$Rn, sav_final$Tair, sav_final$Pa)
sav_final$cwd <- sav_final$pet - sav_final$ET

```

```{r}
enf_final$pet <- priestley_taylor_pet(enf_final$Rn, enf_final$Tair, enf_final$Pa)
enf_final$cwd <- enf_final$pet - enf_final$ET
```

Train/test split within site
```{r}
idx <- createDataPartition(sav_final$GPP, p=0.8, list=FALSE)
train_data <- sav_final[idx, ]
test_data <- sav_final[-idx, ]

```

Drop any NAs in test_data

```{r}
test_data <- test_data %>% drop_na()
```

7-day moving cumulative sum for cwd
- Accumulated drought, not only the day's stress

```{r}
train_data$cumulative_cwd_7day <- rollapply(train_data$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
test_data$cumulative_cwd_7day <-  rollapply(test_data$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
```

Drop any rows with NA values after rolling sum for CWD

```{r}
train_data <- train_data %>% drop_na()
test_data <- test_data %>% drop_na()
```

Lag the climate variables to add time-lag effects

```{r}
train_data <- train_data %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

test_data <- test_data %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()
```

fit the GAM model for the savanna:

```{r}
savanna_model <- gam(GPP ~ s(Tair) + s(VPD) + s(cumulative_cwd_7day) + 
                   s(Tair_lag3) + s(VPD_lag3), data=train_data, 
                 method = "REML")
```

```{r}
summary(savanna_model)
```
```{r}
predictions_sav <- predict(savanna_model, newdata=test_data)
results_sav <- data.frame(observed=test_data$GPP, Predicted=predictions_sav)
```

Dump some stats:

```{r}
rmse_gam <- sqrt(mean((results_sav$observed - results_sav$Predicted)^2))
rsq_gam <- cor(results_sav$observed, results_sav$Predicted)^2
mean_observed_gam <- mean(results_sav$observed)
nse_gam <- (1 - sum((results_sav$observed - results_sav$Predicted)^2) / 
              sum((results_sav$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

plot:

```{r}
library(ggplot2)

ggplot(results_sav, aes(x = observed, y = Predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")
```

fit the GAM model for the evergreen needleleaf forest:

```{r}
idx2 <- createDataPartition(enf_final$GPP, p=0.8, list=FALSE)
train_data2 <- enf_final[idx, ]
test_data2 <- enf_final[-idx, ]
```

```{r}
test_data2 <- test_data2 %>% drop_na()
```

```{r}
train_data2$cumulative_cwd_7day <- rollapply(train_data2$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
test_data2$cumulative_cwd_7day <-  rollapply(test_data2$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
```

```{r}
train_data2 <- train_data2 %>% drop_na()
test_data2 <- test_data2 %>% drop_na()
```

```{r}
train_data2 <- train_data2 %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

test_data2 <- test_data2 %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()
```

```{r}
enf_model <- gam(GPP ~ s(Tair) + s(VPD) + s(cumulative_cwd_7day) + 
                   s(Tair_lag3) + s(VPD_lag3), data=train_data2, 
                 method = "REML")
```

```{r}
summary(enf_model)
```
```{r}
predictions_enf <- predict(enf_model, newdata=test_data2)
results_enf <- data.frame(observed=test_data2$GPP, Predicted=predictions_enf)
```

```{r}
rmse_gam2 <- sqrt(mean((results_enf$observed - results_enf$Predicted)^2))
rsq_gam2 <- cor(results_enf$observed, results_enf$Predicted)^2
mean_observed_gam2 <- mean(results_enf$observed)
nse_gam2 <- (1 - sum((results_enf$observed - results_enf$Predicted)^2) / 
              sum((results_enf$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam2, "\n")
cat("R-squared:", rsq_gam2, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam2, "\n")
```

```{r}
par(mfrow = c(2, 3))
plot(enf_model, shade = TRUE, main = "ENF")
plot(savanna_model, shade = TRUE, main = "SAV")
```
enf plot:

```{r}
library(ggplot2)

ggplot(results_enf, aes(x = observed, y = Predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Observed GPP", y = "Predicted GPP")
```


Monday 26th Jan:
- Which variables explain the variation in GPP the most?

Got the final dataframes, daily measurements, no missing values, growing season...

Savanna:

```{r}
str(sav_final$date)
```

```{r}
train_data <- sav_final %>% filter(date <= as.POSIXct("2012-01-01"))
test_data <- sav_final %>% filter(date > as.POSIXct("2012-01-01"))
```

Vapour pressure deficit & SW radiation:

```{r}
gam_vpd <- gam(GPP ~ s(VPD), data = train_data, method = "REML")
gam_rad <- gam(GPP ~ s(SW_rad), data = train_data, method = "REML")

pred_vpd <- predict(gam_vpd, newdata = test_data)
pred_swr <- predict(gam_rad, newdata = test_data)

```

results:

```{r}

results_vpd <- data.frame(observed=test_data$GPP, Predicted=pred_vpd)

```

Stats:

```{r}
rmse_vpd <- sqrt(mean((results_vpd$observed - results_vpd$Predicted)^2))
rsq_vpd <- cor(results_vpd$observed, results_vpd$Predicted)^2
nse_vpd  <- 1 - sum((results_vpd$observed - results_vpd$Predicted)^2) /
                 sum((results_vpd$observed - mean(results_vpd$observed))^2)

cat("RMSE:", rmse_vpd, "\n")
cat("R-squared:", rsq_vpd, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_vpd, "\n")
```

Shortwave radiation:

```{r}
results_swr <- data.frame(observed=test_data$GPP, predicted=pred_swr)
```

```{r}
rmse_swr <- sqrt(mean((results_swr$observed - results_swr$predicted)^2))
rsq_swr <- cor(results_swr$observed, results_swr$predicted)^2
nse_swr  <- 1 - sum((results_swr$observed - results_swr$predicted)^2) /
                 sum((results_swr$observed - mean(results_swr$observed))^2)

cat("RMSE:", rmse_swr, "\n")
cat("R-squared:", rsq_swr, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_swr, "\n")
```



