---
title: "explore_GAM"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in two flux csv sites, fit GAMs training and testing within and out of site

Going to read in two sites (daily data) using the function I made
AU-Tum - evergreen broadleaf forest, mild with no dry season, warm summer
AU-Cum - evergreen broadleaf forest

Aims:
1. Plot average, max and min GPP per year per site

2. Look at VPD across both sites, is there any issues?

3. Fit GAMs
- Which met variables explain the most variation in GPP? VPD, temp, shortwave radiation...Which GAM is the best? One with only VPD, one with only shortwave radiation, one with both? R^2 and RMSE

- Training and testing within site - can a GAM trained on 80% of the data from AU-Tum (e.g., years 1-18) predict GPP on the final 20% (e.g., last 2 years?)

- Training and testing out of site - can a GAM trained on 100% of data from AU-Tum predict GPP of AU-Cum?


```{r}
rm(list = ls())
```


```{r}
library(tidyverse)

read_daily_flux <- function(fname) {
  
   #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{3}[0-9]?", no_ext)
)
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP to datetime and set as index
  df <- df %>%
    mutate(date = ymd(TIMESTAMP)) %>%
    select(-TIMESTAMP) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, site)
  
  return(df)
  
}
```

Read in flux sites:

```{r}
Tum_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Tum_FLUXDATAKIT_FULLSET_DD_2002_2017_2-3.csv")
```

```{r}
Cum_data <- read_daily_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUXDATAKIT/FLX_AU-Cum_FLUXDATAKIT_FULLSET_DD_2013_2018_2-3.csv")
```

Now I am going to plot GPP across time