---
title: "explore_ozflux"
output: html_document
date: "2026-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Tue 27th Jan:

Going to look at the North Australian Tropical Transect flux sites.

1. Howard Springs = AU-How (2001-2014)
2. Daly River Pasture = AU-DaP (2007-2013)
3. Daly River Savanna = AU-DaS (2008-2014)
4. Dry River = AU-Dry (2008-2014)
5. Sturt Plains = AU-Stp (2008-2014)
6. Alice Springs = AU-ASM (2010-2014)

These flux sites are located along a precipitation gradient. This rainfall gradient is a 'natural laboratory' with Howard Springs as the wettest and Alice Springs the driest. 

1. Read in flux files:

```{r}
library(tidyverse)

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

```{r}
AU_How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

AU_Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

AU_Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

AU_Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")
```

2. Filter by growing season

```{r}
filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}
```

```{r}
AU_How_gs <- filter_gs(AU_How)
AU_Dap_gs <- filter_gs(AU_Dap)
AU_Das_gs <- filter_gs(AU_Das)
AU_Dry_gs <- filter_gs(AU_Dry)
AU_Stp_gs <- filter_gs(AU_Stp)
AU_Asm_gs <- filter_gs(AU_Asm)

```

3. Drop missing variables

```{r}
How_final <- AU_How_gs %>% drop_na()
Dap_final <- AU_Dap_gs %>% drop_na()
Das_final <- AU_Das_gs %>% drop_na()
Dry_final <- AU_Dry_gs %>% drop_na()
Stp_final <- AU_Stp_gs %>% drop_na()
Asm_final <- AU_Asm_gs %>% drop_na()
```

4. Remove missing flags (-999)

```{r}
How_final <- How_final %>% filter_all(all_vars(. >= -9000) )
Dap_final <- Dap_final %>% filter_all(all_vars(. >= -9000) )
Das_final <- Das_final %>% filter_all(all_vars(. >= -9000) )
Dry_final <- Dry_final %>% filter_all(all_vars(. >= -9000) )
Stp_final <- Stp_final %>% filter_all(all_vars(. >= -9000) )
Asm_final <- Asm_final %>% filter_all(all_vars(. >= -9000) )
```

5. Visualise 

```{r}
#Combine the datasets

library(dplyr)

dataframe_oz <- bind_rows(How_final, Dap_final, Das_final, Dry_final, Stp_final, Asm_final)
```

```{r}
#Summarise by site, by year

df_summary <- dataframe_oz %>%
  group_by(site, year) %>%
  summarise(
    GPP_mean = mean(GPP, na.rm = TRUE),
    GPP_max = max(GPP, na.rm = TRUE),
    GPP_min = min(GPP, na.rm = TRUE)
  ) %>%
  
  ungroup()
```

```{r}
library(tidyr)
library(ggplot2)

df_long <- df_summary %>%
  pivot_longer(
    cols = c(GPP_mean, GPP_max, GPP_min),
    names_to = "stat",
    values_to = "GPP"
  )

ggplot(df_long, aes(x = year, y = GPP, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "GPP", x = "Year", color = "Statistic")
```


```{r}
ggplot(df_summary, aes(x = year, group = site)) +
  geom_ribbon(aes(ymin = GPP_min, ymax = GPP_max, fill = site), alpha = 0.2) +
  geom_line(aes(y = GPP_mean, color = site), size = 1) +
  theme_minimal() +
  labs(y = "GPP", x = "Year")
```

























5. Fit GAM firstly with Tair ~ GPP

- Load in relevant libraries

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
```
- Train/Test within site

```{r}
idx <- createDataPartition(How_final$GPP, p=0.8, list=FALSE)
train_data <- How_final[idx, ]
test_data <- How_final[-idx, ]
```

- Fit GAM

```{r}
tair_model <- gam(GPP ~ s(VPD) + s(SW_rad) + s(Precip), data=train_data, 
                 method = "REML")
```

- Get some stats:

```{r}
summary(tair_model)
```
```{r}

predictions_tair <- predict(tair_model, newdata=test_data)
results_tair <- data.frame(observed=test_data$GPP, predicted=predictions_tair)
```

```{r}
rmse_gam <- sqrt(mean((results_tair$observed - results_tair$predicted)^2))
rsq_gam <- cor(results_tair$observed, results_tair$predicted)^2
mean_observed_gam <- mean(results_tair$observed)
nse_gam <- (1 - sum((results_tair$observed - results_tair$predicted)^2) / 
              sum((results_tair$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```


