---
title: "future_envelope_01"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Future envelope experiment
- Increase my VPD met data by x% for different future scenarios
- Refit RF model and predict GPP

Increase daily VPD data by 15, 30 and 50%
RF is trained on 80% of each site's observed data with (VPD, SW_rad, Tair and cwd_7_day) for each scenario (15, 30, 50%) only VPD is inflated, isolating the VPD effect

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(caret)
library(tidyr)
library(ggplot2)
library(zoo)

# predictors, vpd and scenarios
predictors  <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

standardise <- function(df, vars) {
  df[vars] <- scale(df[vars])
  df
}

#Main function: train on observed, predict on each VPD scenario
run_vpd_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data 
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 100) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  # Standardise using observed data parameters
  scale_params <- df %>% select(all_of(predictors))
  scale_means  <- colMeans(scale_params, na.rm = TRUE)
  scale_sds    <- apply(scale_params, 2, sd, na.rm = TRUE)
  
  scale_with_params <- function(data, means, sds) {
    for (v in names(means)) {
      data[[v]] <- (data[[v]] - means[v]) / sds[v]
    }
    data
  }
  
  # Train RF on observed data
  set.seed(123)
  train_df <- df %>%
    scale_with_params(scale_means, scale_sds)
  
  idx        <- createDataPartition(train_df$GPP, p = 0.8, list = FALSE)
  train_data <- train_df[idx, ]
  
  formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  
  rf_model <- randomForest(formula, data = train_data,
                           ntree = 500, mtry = 3, importance = TRUE)
  
  # Predict GPP for each VPD scenario
  scenario_preds <- map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
    
    scenario_df <- df %>%
      mutate(VPD = VPD * scalar) %>%         # inflate VPD
      scale_with_params(scale_means, scale_sds)  # re-standardise with same params
    
    preds <- predict(rf_model, scenario_df)
    
    data.frame(
      site      = site_name,
      date      = df$date,
      scenario  = label,
      VPD_raw   = df$VPD * scalar,
      GPP_pred  = preds,
      GPP_obs   = df$GPP
    )
  })
  
  scenario_preds
}

#Run across all sites 
vpd_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_vpd_scenario
)
vpd_scenario_results <- Filter(Negate(is.null), vpd_scenario_results)

all_scenarios <- bind_rows(vpd_scenario_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    year     = lubridate::year(date)
  )

#Summary: mean annual GPP per site per scenario
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

# ── Plot 1: Mean annual GPP by scenario, 4 sites per page 
all_sites   <- unique(scenario_annual$site)
n_pages     <- ceiling(length(all_sites) / 4)

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}

# Plot 2: Boxplot of GPP distribution per scenario across all sites 
all_scenarios %>%
  ggplot(aes(x = scenario, y = GPP_pred, fill = scenario)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = scenario_colours) +
  facet_wrap(~ site, scales = "free_y", ncol = 4) +
  labs(
    x    = "VPD scenario",
    y    = expression("Predicted GPP (g C m"^{-2}~"day"^{-1}*")"),
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    strip.background   = element_blank(),
    strip.text         = element_text(face = "bold"),
    legend.position    = "top",
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank()
  )

# Summary table: % change in mean GPP relative to observed 
scenario_summary <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change_15 = (`+15%`  - Observed) / Observed * 100,
    pct_change_30 = (`+30%`  - Observed) / Observed * 100,
    pct_change_50 = (`+50%`  - Observed) / Observed * 100
  )

print(scenario_summary)
```
% change in GPP with each VPD scenario
R-squared for each model

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

#Plot 1: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

scenario_colours <- c(
  "+15%" = "#F28C28",
  "+30%" = "#E06C4B",
  "+50%" = "#8B1A1A"
)

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 2: R² for each site's RF model (observed vs predicted) ───────────────
r2_scenario <- all_scenarios %>%
  filter(scenario == "Observed") %>%      # R² only meaningful on observed fit
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_scenario, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² (observed vs predicted GPP)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))
```
R-squared is really high - model has basically seen all of the data from the training set and then when it is asked to test on higher VPD data, it does well because it has basically seen it before

LOSO: Train on all years except year X, predict X with inflated VPD, repeat for all years and then concatenate the predictions
Gives a more robust idea of model performance

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

#predictors, vpd, etc
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

#Leave-one-year-out RF scenario function 
run_loyo_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year                = year(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 100) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  all_years <- unique(df$year)
  
  if (length(all_years) < 4) {
    warning("Skipping ", site_name, ": insufficient years for LOYO")
    return(NULL)
  }
  
  # Loop over each held-out year
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(year != test_year)
    test_df  <- df %>% filter(year == test_year)
    
    if (nrow(train_df) < 100 || nrow(test_df) < 10) return(NULL)
    
    # Standardise using training data parameters only
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) {
        data[[v]] <- (data[[v]] - means[v]) / sds[v]
      }
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    
    # Train RF on all years except test year
    formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model <- randomForest(formula, data = train_scaled,
                             ntree = 500, mtry = 3, importance = FALSE)
    
    # Predict test year under each VPD scenario
    map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
      
      scenario_test <- test_df %>%
        mutate(VPD = VPD * scalar) %>%
        scale_with_params(scale_means, scale_sds)
      
      preds <- predict(rf_model, scenario_test)
      
      data.frame(
        site      = site_name,
        date      = test_df$date,
        year      = test_year,
        scenario  = label,
        VPD_raw   = test_df$VPD * scalar,
        GPP_pred  = preds,
        GPP_obs   = test_df$GPP
      )
    })
  })
  
  loyo_results
}

#Run across all sites 
loyo_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_loyo_scenario
)
loyo_scenario_results <- Filter(Negate(is.null), loyo_scenario_results)

all_scenarios <- bind_rows(loyo_scenario_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

#PLOTS 

#Plot 1: R² per site (observed scenario only — genuine out-of-sample)
r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))

# Plot 2: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours[c("+15%", "+30%", "+50%")]) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

#Plot 3: Annual mean GPP time series per scenario — paginated 
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

all_sites <- unique(scenario_annual$site)
n_pages   <- ceiling(length(all_sites) / 4)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}
```






