---
title: "future_envelope_01"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Future envelope experiment
- Increase my VPD met data by x% for different future scenarios
- Refit RF model and predict GPP

Increase daily VPD data by 15, 30 and 50%
RF is trained on 80% of each site's observed data with (VPD, SW_rad, Tair and cwd_7_day) for each scenario (15, 30, 50%) only VPD is inflated, isolating the VPD effect

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(caret)
library(tidyr)
library(ggplot2)
library(zoo)

# predictors, vpd and scenarios
predictors  <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

standardise <- function(df, vars) {
  df[vars] <- scale(df[vars])
  df
}

#Main function: train on observed, predict on each VPD scenario
run_vpd_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data 
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 100) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  # Standardise using observed data parameters
  scale_params <- df %>% select(all_of(predictors))
  scale_means  <- colMeans(scale_params, na.rm = TRUE)
  scale_sds    <- apply(scale_params, 2, sd, na.rm = TRUE)
  
  scale_with_params <- function(data, means, sds) {
    for (v in names(means)) {
      data[[v]] <- (data[[v]] - means[v]) / sds[v]
    }
    data
  }
  
  # Train RF on observed data
  set.seed(123)
  train_df <- df %>%
    scale_with_params(scale_means, scale_sds)
  
  idx        <- createDataPartition(train_df$GPP, p = 0.8, list = FALSE)
  train_data <- train_df[idx, ]
  
  formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  
  rf_model <- randomForest(formula, data = train_data,
                           ntree = 500, mtry = 3, importance = TRUE)
  
  # Predict GPP for each VPD scenario
  scenario_preds <- map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
    
    scenario_df <- df %>%
      mutate(VPD = VPD * scalar) %>%         # inflate VPD
      scale_with_params(scale_means, scale_sds)  # re-standardise with same params
    
    preds <- predict(rf_model, scenario_df)
    
    data.frame(
      site      = site_name,
      date      = df$date,
      scenario  = label,
      VPD_raw   = df$VPD * scalar,
      GPP_pred  = preds,
      GPP_obs   = df$GPP
    )
  })
  
  scenario_preds
}

#Run across all sites 
vpd_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_vpd_scenario
)
vpd_scenario_results <- Filter(Negate(is.null), vpd_scenario_results)

all_scenarios <- bind_rows(vpd_scenario_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    year     = lubridate::year(date)
  )

#Summary: mean annual GPP per site per scenario
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

# ── Plot 1: Mean annual GPP by scenario, 4 sites per page 
all_sites   <- unique(scenario_annual$site)
n_pages     <- ceiling(length(all_sites) / 4)

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}

# Plot 2: Boxplot of GPP distribution per scenario across all sites 
all_scenarios %>%
  ggplot(aes(x = scenario, y = GPP_pred, fill = scenario)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = scenario_colours) +
  facet_wrap(~ site, scales = "free_y", ncol = 4) +
  labs(
    x    = "VPD scenario",
    y    = expression("Predicted GPP (g C m"^{-2}~"day"^{-1}*")"),
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    strip.background   = element_blank(),
    strip.text         = element_text(face = "bold"),
    legend.position    = "top",
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank()
  )

# Summary table: % change in mean GPP relative to observed 
scenario_summary <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change_15 = (`+15%`  - Observed) / Observed * 100,
    pct_change_30 = (`+30%`  - Observed) / Observed * 100,
    pct_change_50 = (`+50%`  - Observed) / Observed * 100
  )

print(scenario_summary)
```
% change in GPP with each VPD scenario
R-squared for each model

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

#Plot 1: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

scenario_colours <- c(
  "+15%" = "#F28C28",
  "+30%" = "#E06C4B",
  "+50%" = "#8B1A1A"
)

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 2: R² for each site's RF model (observed vs predicted) ───────────────
r2_scenario <- all_scenarios %>%
  filter(scenario == "Observed") %>%      # R² only meaningful on observed fit
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_scenario, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² (observed vs predicted GPP)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))
```
R-squared is really high - model has basically seen all of the data from the training set and then when it is asked to test on higher VPD data, it does well because it has basically seen it before

LOSO: Train on all years except year X, predict X with inflated VPD, repeat for all years and then concatenate the predictions
Gives a more robust idea of model performance

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

#predictors, vpd, etc
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

#Leave-one-year-out RF scenario function 
run_loyo_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year                = year(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  all_years <- unique(df$year)
  
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years for LOYO")
    return(NULL)
  }
  
  # Loop over each held-out year
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(year != test_year)
    test_df  <- df %>% filter(year == test_year)
    
    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)
    
    # Standardise using training data parameters only
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) {
        data[[v]] <- (data[[v]] - means[v]) / sds[v]
      }
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    
    # Train RF on all years except test year
    formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model <- randomForest(formula, data = train_scaled,
                             ntree = 500, mtry = 3, importance = FALSE)
    
    # Predict test year under each VPD scenario
    map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
      
      scenario_test <- test_df %>%
        mutate(VPD = VPD * scalar) %>%
        scale_with_params(scale_means, scale_sds)
      
      preds <- predict(rf_model, scenario_test)
      
      data.frame(
        site      = site_name,
        date      = test_df$date,
        year      = test_year,
        scenario  = label,
        VPD_raw   = test_df$VPD * scalar,
        GPP_pred  = preds,
        GPP_obs   = test_df$GPP
      )
    })
  })
  
  loyo_results
}

#Run across all sites 
loyo_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_loyo_scenario
)
loyo_scenario_results <- Filter(Negate(is.null), loyo_scenario_results)

all_scenarios <- bind_rows(loyo_scenario_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

#PLOTS 

#Plot 1: R² per site (observed scenario only — genuine out-of-sample)
r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))

# Plot 2: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours[c("+15%", "+30%", "+50%")]) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

#Plot 3: Annual mean GPP time series per scenario — paginated 
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

all_sites <- unique(scenario_annual$site)
n_pages   <- ceiling(length(all_sites) / 4)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}
```

```{r}
plot(sites_daily$Ync$date, sites_daily$Ync$GPP)
```



```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggrepel)

# ── Join R² and % change per site per scenario ────────────────────────────────
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

pct_change_sites <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")))

plot_data <- pct_change_sites %>%
  left_join(r2_sites, by = "site")

# ── Plot: R² vs % change, faceted by scenario ─────────────────────────────────
ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = scenario), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = c(
    "+15%" = "#F28C28",
    "+30%" = "#E06C4B",
    "+50%" = "#8B1A1A"
  )) +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to observed",
    colour = "VPD scenario"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "none",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```
It doesn't seem like predicted GPP decline is an artefact of poor model fit. Even sites where RF struggles to explain GPP variabiliyt are showing similar declines to well-fitted sites.

```{r}
cluster_labels <- site_slopes %>%
  select(site, cluster_multi) %>%
  mutate(
    site          = as.character(site),
    cluster_multi = as.character(cluster_multi)  # strip factor levels before joining
  )

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))
cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

What about if i increase the number of days VPD <90th percentile in the summer? Increase by 15, 30 and 50%?

Train RF on observed summer seasons - both extreme and normal - so just train on summer?
And then test on the 3 inflated VPD summer scenarios?

Replace - swap normal days with extreme days, keeping dataset size constant. 

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(0, 0.15, 0.30, 0.50)   # extra fraction of days that become extreme
scenario_labels <- c("Observed", "+15% extreme days", "+30% extreme days", "+50% extreme days")

scenario_colours <- c(
  "Observed"           = "#4682B4",
  "+15% extreme days"  = "#F28C28",
  "+30% extreme days"  = "#E06C4B",
  "+50% extreme days"  = "#8B1A1A"
)

# Define summer months — adjust if needed per site
is_summer <- function(dates) month(dates) %in% c(12, 1, 2)

run_loyo_summer_extreme <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year = year(date),
      # Use hydrological year so Dec belongs to next year's summer
      hydro_year          = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer              = is_summer(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  # Filter to summer only
  df_summer <- df %>% filter(summer)
  
  if (nrow(df_summer) < 50) {
    warning("Skipping ", site_name, ": insufficient summer data")
    return(NULL)
  }
  
  # Compute site-wide 90th percentile VPD threshold from summer obs
  vpd_90 <- quantile(df_summer$VPD, 0.90, na.rm = TRUE)
  message("  VPD 90th percentile: ", round(vpd_90, 2))
  
  all_years <- unique(df_summer$hydro_year)
  
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }
  
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df_summer %>% filter(hydro_year != test_year)
    test_df  <- df_summer %>% filter(hydro_year == test_year)
    
    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)
    
    # Standardise on training data
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    
    formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model <- randomForest(formula, data = train_scaled,
                             ntree = 500, mtry = 3, importance = FALSE)
    
    # Identify extreme vs normal days in test year
    extreme_days <- test_df %>% filter(VPD >= vpd_90)
    normal_days  <- test_df %>% filter(VPD <  vpd_90)
    
    n_normal   <- nrow(normal_days)
    n_extreme  <- nrow(extreme_days)
    
    map2_dfr(vpd_scalars, scenario_labels, function(extra_frac, label) {
      
      if (extra_frac == 0) {
        # Observed: use test year as-is
        scenario_df <- test_df
        
      } else {
        # How many normal days to replace with (resampled) extreme days
        n_replace <- round(n_normal * extra_frac)
        
        if (n_replace == 0 || n_extreme == 0) {
          scenario_df <- test_df  # fallback if no extreme days exist
        } else {
          # Sample extreme days with replacement to get n_replace days
          extra_extreme <- extreme_days %>%
            slice_sample(n = n_replace, replace = TRUE)
          
          # Remove n_replace normal days (random sample)
          normal_kept <- normal_days %>%
            slice_sample(n = max(0, n_normal - n_replace), replace = FALSE)
          
          scenario_df <- bind_rows(normal_kept, extreme_days, extra_extreme) %>%
            arrange(date)
        }
      }
      
      scenario_scaled <- scale_with_params(scenario_df, scale_means, scale_sds)
      preds <- predict(rf_model, scenario_scaled)
      
      data.frame(
        site          = site_name,
        date          = scenario_df$date,
        year          = test_year,
        scenario      = label,
        VPD_raw       = scenario_df$VPD,
        n_extreme_days = sum(scenario_df$VPD >= vpd_90),
        n_total_days   = nrow(scenario_df),
        GPP_pred      = preds,
        GPP_obs       = scenario_df$GPP   # NA for resampled rows
      )
    })
  })
  
  loyo_results
}

# Run across all sites
loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_summer_extreme)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))
```
```{r}
# R² on observed summer (out-of-sample)
r2_summer <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

# % change in mean summer GPP per scenario
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  pivot_longer(cols = ends_with("days"),
               names_to = "scenario", values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels[-1]),
    site     = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours[scenario_labels[-1]]) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean summer GPP relative to observed",
    fill = "Extreme VPD scenario"
  ) +
  theme_classic() +
  theme(legend.position = "top", axis.text.y = element_text(size = 10))
```

```{r}
all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(
    pct_extreme = mean(n_extreme_days / n_total_days, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = scenario, values_from = pct_extreme)
```

```{r}
r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))
```

