---
title: "future_envelope_01"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Future envelope experiment
- Increase my VPD met data by x% for different future scenarios
- Refit RF model and predict GPP

Increase daily VPD data by 15, 30 and 50%
RF is trained on 80% of each site's observed data with (VPD, SW_rad, Tair and cwd_7_day) for each scenario (15, 30, 50%) only VPD is inflated, isolating the VPD effect

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(caret)
library(tidyr)
library(ggplot2)
library(zoo)

# predictors, vpd and scenarios
predictors  <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

standardise <- function(df, vars) {
  df[vars] <- scale(df[vars])
  df
}

#Main function: train on observed, predict on each VPD scenario
run_vpd_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data 
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 100) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  # Standardise using observed data parameters
  scale_params <- df %>% select(all_of(predictors))
  scale_means  <- colMeans(scale_params, na.rm = TRUE)
  scale_sds    <- apply(scale_params, 2, sd, na.rm = TRUE)
  
  scale_with_params <- function(data, means, sds) {
    for (v in names(means)) {
      data[[v]] <- (data[[v]] - means[v]) / sds[v]
    }
    data
  }
  
  # Train RF on observed data
  set.seed(123)
  train_df <- df %>%
    scale_with_params(scale_means, scale_sds)
  
  idx        <- createDataPartition(train_df$GPP, p = 0.8, list = FALSE)
  train_data <- train_df[idx, ]
  
  formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  
  rf_model <- randomForest(formula, data = train_data,
                           ntree = 500, mtry = 3, importance = TRUE)
  
  # Predict GPP for each VPD scenario
  scenario_preds <- map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
    
    scenario_df <- df %>%
      mutate(VPD = VPD * scalar) %>%         # inflate VPD
      scale_with_params(scale_means, scale_sds)  # re-standardise with same params
    
    preds <- predict(rf_model, scenario_df)
    
    data.frame(
      site      = site_name,
      date      = df$date,
      scenario  = label,
      VPD_raw   = df$VPD * scalar,
      GPP_pred  = preds,
      GPP_obs   = df$GPP
    )
  })
  
  scenario_preds
}

#Run across all sites 
vpd_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_vpd_scenario
)
vpd_scenario_results <- Filter(Negate(is.null), vpd_scenario_results)

all_scenarios <- bind_rows(vpd_scenario_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    year     = lubridate::year(date)
  )

#Summary: mean annual GPP per site per scenario
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

# ── Plot 1: Mean annual GPP by scenario, 4 sites per page 
all_sites   <- unique(scenario_annual$site)
n_pages     <- ceiling(length(all_sites) / 4)

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}

# Plot 2: Boxplot of GPP distribution per scenario across all sites 
all_scenarios %>%
  ggplot(aes(x = scenario, y = GPP_pred, fill = scenario)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = scenario_colours) +
  facet_wrap(~ site, scales = "free_y", ncol = 4) +
  labs(
    x    = "VPD scenario",
    y    = expression("Predicted GPP (g C m"^{-2}~"day"^{-1}*")"),
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    strip.background   = element_blank(),
    strip.text         = element_text(face = "bold"),
    legend.position    = "top",
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank()
  )

# Summary table: % change in mean GPP relative to observed 
scenario_summary <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change_15 = (`+15%`  - Observed) / Observed * 100,
    pct_change_30 = (`+30%`  - Observed) / Observed * 100,
    pct_change_50 = (`+50%`  - Observed) / Observed * 100
  )

print(scenario_summary)
```
% change in GPP with each VPD scenario
R-squared for each model

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

#Plot 1: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

scenario_colours <- c(
  "+15%" = "#F28C28",
  "+30%" = "#E06C4B",
  "+50%" = "#8B1A1A"
)

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 2: R² for each site's RF model (observed vs predicted) ───────────────
r2_scenario <- all_scenarios %>%
  filter(scenario == "Observed") %>%      # R² only meaningful on observed fit
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_scenario, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² (observed vs predicted GPP)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))
```
R-squared is really high - model has basically seen all of the data from the training set and then when it is asked to test on higher VPD data, it does well because it has basically seen it before

LOSO: Train on all years except year X, predict X with inflated VPD, repeat for all years and then concatenate the predictions
Gives a more robust idea of model performance

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

#predictors, vpd, etc
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(1.00, 1.15, 1.30, 1.50)
scenario_labels <- c("Observed", "+15%", "+30%", "+50%")

scenario_colours <- c(
  "Observed" = "#4682B4",
  "+15%"     = "#F28C28",
  "+30%"     = "#E06C4B",
  "+50%"     = "#8B1A1A"
)

#Leave-one-year-out RF scenario function 
run_loyo_scenario <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  # Prepare base data
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year                = year(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  all_years <- unique(df$year)
  
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years for LOYO")
    return(NULL)
  }
  
  # Loop over each held-out year
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(year != test_year)
    test_df  <- df %>% filter(year == test_year)
    
    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)
    
    # Standardise using training data parameters only
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) {
        data[[v]] <- (data[[v]] - means[v]) / sds[v]
      }
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    
    # Train RF on all years except test year
    formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model <- randomForest(formula, data = train_scaled,
                             ntree = 500, mtry = 3, importance = FALSE)
    
    # Predict test year under each VPD scenario
    map2_dfr(vpd_scalars, scenario_labels, function(scalar, label) {
      
      scenario_test <- test_df %>%
        mutate(VPD = VPD * scalar) %>%
        scale_with_params(scale_means, scale_sds)
      
      preds <- predict(rf_model, scenario_test)
      
      data.frame(
        site      = site_name,
        date      = test_df$date,
        year      = test_year,
        scenario  = label,
        VPD_raw   = test_df$VPD * scalar,
        GPP_pred  = preds,
        GPP_obs   = test_df$GPP
      )
    })
  })
  
  loyo_results
}

#Run across all sites 
loyo_scenario_results <- map2(
  sites_daily,
  names(sites_daily),
  run_loyo_scenario
)
loyo_scenario_results <- Filter(Negate(is.null), loyo_scenario_results)

all_scenarios <- bind_rows(loyo_scenario_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

#PLOTS 

#Plot 1: R² per site (observed scenario only — genuine out-of-sample)
r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))

# Plot 2: % change in GPP per scenario per site 
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")),
    site     = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = scenario_colours[c("+15%", "+30%", "+50%")]) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

#Plot 3: Annual mean GPP time series per scenario — paginated 
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

all_sites <- unique(scenario_annual$site)
n_pages   <- ceiling(length(all_sites) / 4)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}
```



```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggrepel)

# ── Join R² and % change per site per scenario ────────────────────────────────
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

pct_change_sites <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15%" = (`+15%` - Observed) / Observed * 100,
    "+30%" = (`+30%` - Observed) / Observed * 100,
    "+50%" = (`+50%` - Observed) / Observed * 100
  ) %>%
  select(site, `+15%`, `+30%`, `+50%`) %>%
  pivot_longer(cols = c(`+15%`, `+30%`, `+50%`),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15%", "+30%", "+50%")))

plot_data <- pct_change_sites %>%
  left_join(r2_sites, by = "site")

# ── Plot: R² vs % change, faceted by scenario ─────────────────────────────────
ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = scenario), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = c(
    "+15%" = "#F28C28",
    "+30%" = "#E06C4B",
    "+50%" = "#8B1A1A"
  )) +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to observed",
    colour = "VPD scenario"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "none",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```
It doesn't seem like predicted GPP decline is an artefact of poor model fit. Even sites where RF struggles to explain GPP variabiliyt are showing similar declines to well-fitted sites.

```{r}
cluster_labels <- site_slopes %>%
  select(site, cluster_multi) %>%
  mutate(
    site          = as.character(site),
    cluster_multi = as.character(cluster_multi)  # strip factor levels before joining
  )

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))
cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

What about if i increase the number of days VPD <90th percentile in the summer? Increase by 15, 30 and 50%?

Train RF on observed data - both extreme and normal 
And then test on the 3 inflated VPD summer scenarios?

Replace - swap normal days with extreme days, keeping dataset size constant. 

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(0, 0.15, 0.30, 0.50)
scenario_labels <- c("Observed", "+15% extreme days", "+30% extreme days", "+50% extreme days")

scenario_colours <- c(
  "Observed"           = "#4682B4",
  "+15% extreme days"  = "#F28C28",
  "+30% extreme days"  = "#E06C4B",
  "+50% extreme days"  = "#8B1A1A"
)

# ── Function definition ───────────────────────────────────────────────────────
run_loyo_summer_extreme <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  vpd_90 <- df %>%
    filter(summer) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)
  
  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }
  
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df    <- df %>% filter(hydro_year != test_year)
    test_summer <- df %>% filter(hydro_year == test_year, summer)
    
    if (nrow(train_df) < 50 || nrow(test_summer) < 10) return(NULL)
    
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)
    
    extreme_days <- test_summer %>% filter(VPD >= vpd_90)
    normal_days  <- test_summer %>% filter(VPD <  vpd_90)
    n_normal     <- nrow(normal_days)
    n_extreme    <- nrow(extreme_days)
    
    map2_dfr(vpd_scalars, scenario_labels, function(extra_frac, label) {
      
      if (extra_frac == 0) {
        scenario_df <- test_summer
      } else {
        n_replace <- round(n_normal * extra_frac)
        
        if (n_replace == 0 || n_extreme == 0) {
          scenario_df <- test_summer
        } else {
          extra_extreme <- extreme_days %>%
            slice_sample(n = n_replace, replace = TRUE)
          
          normal_kept <- normal_days %>%
            slice_sample(n = max(0, n_normal - n_replace), replace = FALSE)
          
          scenario_df <- bind_rows(normal_kept, extreme_days, extra_extreme) %>%
            arrange(date)
        }
      }
      
      scenario_scaled <- scale_with_params(scenario_df, scale_means, scale_sds)
      preds           <- predict(rf_model, scenario_scaled)
      
      data.frame(
        site           = site_name,
        date           = scenario_df$date,
        year           = test_year,
        scenario       = label,
        VPD_raw        = scenario_df$VPD,
        n_extreme_days = sum(scenario_df$VPD >= vpd_90),
        n_total_days   = nrow(scenario_df),
        GPP_pred       = preds,
        GPP_obs        = scenario_df$GPP
      )
    })
  })
  
  loyo_results
}

# ── Run across all sites ──────────────────────────────────────────────────────
loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_summer_extreme)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

# Quick sanity check
message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))
glimpse(all_scenarios)

# ── Plot 1: R² per site ───────────────────────────────────────────────────────
r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample, summer)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10))

# ── Plot 2: % change in summer GPP per scenario per site ─────────────────────
scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15% extreme days",
                                           "+30% extreme days",
                                           "+50% extreme days")),
    site = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = c(
    "+15% extreme days" = "#F28C28",
    "+30% extreme days" = "#E06C4B",
    "+50% extreme days" = "#8B1A1A"
  )) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean summer GPP relative to observed",
    fill = "Extreme VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 3: Annual mean summer GPP time series — paginated ───────────────────
scenario_annual <- all_scenarios %>%
  group_by(site, year, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop")

all_sites <- unique(scenario_annual$site)
n_pages   <- ceiling(length(all_sites) / 4)

for (i in seq_len(n_pages)) {
  site_subset <- all_sites[((i - 1) * 4 + 1):min(i * 4, length(all_sites))]
  
  p <- scenario_annual %>%
    filter(site %in% site_subset) %>%
    mutate(site = factor(site, levels = site_subset)) %>%
    ggplot(aes(x = year, y = mean_GPP, colour = scenario)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    scale_colour_manual(values = scenario_colours) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(
      x      = NULL,
      y      = expression("Mean summer GPP (g C m"^{-2}~"day"^{-1}*")"),
      colour = "VPD scenario"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
  print(p)
}
```

```{r}
library(ggrepel)

cluster_labels <- site_slopes %>%
  select(site, cluster_multi) %>%
  mutate(
    site          = as.character(site),
    cluster_multi = as.character(cluster_multi)  # strip factor levels before joining
  )

# R² from new model
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

# % change per site per scenario
pct_change_sites <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15% extreme days",
                                                 "+30% extreme days",
                                                 "+50% extreme days")))

# Join with cluster labels
cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```
Ok, now I want to train and test on the whole year, not just testing on the summer. 

```{r}
run_loyo_summer_extreme <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  # 90th percentile VPD from summer obs across all years
  vpd_90 <- df %>%
    filter(summer) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)
  
  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }
  
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(hydro_year != test_year)
    
    # ── Full held-out year ────────────────────────────────────────────
    test_nonsummer <- df %>% filter(hydro_year == test_year, !summer)
    test_summer    <- df %>% filter(hydro_year == test_year,  summer)
    
    if (nrow(train_df) < 50 || nrow(test_summer) < 10) return(NULL)
    
    # Standardise on full training set
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)
    
    # Identify extreme vs normal days in test summer
    extreme_days <- test_summer %>% filter(VPD >= vpd_90)
    normal_days  <- test_summer %>% filter(VPD <  vpd_90)
    n_normal     <- nrow(normal_days)
    n_extreme    <- nrow(extreme_days)
    
    map2_dfr(vpd_scalars, scenario_labels, function(extra_frac, label) {
      
      # ── Manipulate summer days only ───────────────────────────────
      if (extra_frac == 0) {
        summer_scenario <- test_summer
        
      } else {
        n_replace <- round(n_normal * extra_frac)
        
        if (n_replace == 0 || n_extreme == 0) {
          summer_scenario <- test_summer
        } else {
          extra_extreme <- extreme_days %>%
            slice_sample(n = n_replace, replace = TRUE)
          
          normal_kept <- normal_days %>%
            slice_sample(n = max(0, n_normal - n_replace), replace = FALSE)
          
          summer_scenario <- bind_rows(normal_kept, extreme_days, extra_extreme) %>%
            arrange(date)
        }
      }
      
      # ── Combine manipulated summer + unchanged non-summer ─────────
      full_test <- bind_rows(test_nonsummer, summer_scenario) %>%
        arrange(date)
      
      full_scaled <- scale_with_params(full_test, scale_means, scale_sds)
      preds       <- predict(rf_model, full_scaled)
      
      data.frame(
        site           = site_name,
        date           = full_test$date,
        year           = test_year,
        scenario       = label,
        summer         = full_test$summer,
        VPD_raw        = full_test$VPD,
        n_extreme_days = sum(full_test$summer & full_test$VPD >= vpd_90),
        n_total_days   = nrow(full_test),
        GPP_pred       = preds,
        GPP_obs        = full_test$GPP
      )
    })
  })
  
  loyo_results
}

# ── Run across all sites ──────────────────────────────────────────────────────
loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_summer_extreme)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggrepel)

# ── Plot 1: R² per site (full year and summer) ────────────────────────────────

r2_fullyear <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order),
         period = "Full year")

r2_summer <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order),
         period = "Summer only")

r2_combined <- bind_rows(r2_fullyear, r2_summer) %>%
  mutate(period = factor(period, levels = c("Full year", "Summer only")))

ggplot(r2_combined, aes(x = site, y = R2, fill = period)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = round(R2, 2)),
            position = position_dodge(width = 0.7),
            hjust = -0.15, size = 2.8, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  scale_fill_manual(values = c("Full year" = "#4682B4", "Summer only" = "#76B7C8")) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)", fill = NULL) +
  theme_classic() +
  theme(
    legend.position  = "top",
    axis.text.y      = element_text(size = 10)
  )

# ── Plot 2: % change in GPP per scenario per site ─────────────────────────────

# Annual GPP change (whole year)
pct_annual <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(period = "Full year")

# Summer GPP change only
pct_summer <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(period = "Summer only")

pct_combined <- bind_rows(pct_annual, pct_summer) %>%
  mutate(
    scenario = factor(scenario, levels = c("+15% extreme days",
                                           "+30% extreme days",
                                           "+50% extreme days")),
    period = factor(period, levels = c("Full year", "Summer only")),
    site   = factor(site, levels = wet_to_dry_order)
  )

ggplot(pct_combined, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = c(
    "+15% extreme days" = "#F28C28",
    "+30% extreme days" = "#E06C4B",
    "+50% extreme days" = "#8B1A1A"
  )) +
  coord_flip() +
  facet_wrap(~ period, ncol = 2) +
  labs(
    x    = NULL,
    y    = "% change in mean GPP relative to observed",
    fill = "Extreme VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position  = "top",
    axis.text.y      = element_text(size = 10),
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )

# ── Plot 3: R² vs % change coloured by cluster group ─────────────────────────

# Use summer R² as most relevant for the scenario
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15% extreme days",
                                                 "+30% extreme days",
                                                 "+50% extreme days")))

cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```



