---
title: "03_site_GAMs"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plots:
- Annual precipitation
- Most/least variable precipitation
- VPD
- Time series of GPP

Plot mean annual precipitation for each site - to find dry - wet gradient of sites

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}

sites_annual <- sites_daily %>%
  map(convert_to_annual)


```

```{r}

library(tidyverse)
library(lubridate)

# Convert daily -> annual and add site column in one go
sites_annual <- sites_daily %>%
  imap(~ convert_to_annual(.x) %>% mutate(site = str_remove(.y, "_HH_daily$")))

# Combine all sites into one big table
annual_precip <- bind_rows(sites_annual)


```

```{r}
#Plot annual precipitation across sites
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE)) %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across Australian Fluxnet sites"
  )


```

Plot variability of precipitation across sites, which sites are the most/least variable in their precipitation?

Add site column to the sites_daily list
```{r}
library(dplyr)
library(purrr)
library(stringr)

sites_daily <- sites_daily %>%
  imap(~ .x %>% mutate(site = str_remove(.y, "_HH_daily$")))

```

Combine all daily dataframes into one big table
```{r}
daily_precip <- bind_rows(sites_daily)
```

Precipitation variability - standard deviation is the absolute variability and coefficient of variation is the SD divided by the mean, so accounts for differences in overall rainfall

```{r}
#creates new object with mean precip, standard deviation and coefficient of variation for each of the sites
precip_variability <- daily_precip %>%
  group_by(site) %>%
  summarise(
    mean_precip = mean(Precip, na.rm = TRUE),
    sd_precip = sd(Precip, na.rm = TRUE),
    cv_precip = sd_precip / mean_precip
  ) %>%
  arrange(desc(cv_precip))  # order from most variable to least

```

```{r}
library(ggplot2)

ggplot(precip_variability, aes(x = reorder(site, cv_precip), y = cv_precip)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Daily precipitation variability (CV)",
    title = "Daily precipitation variability across Australian Fluxnet sites"
  )

```
These are daily data - very noisy day to day variability of rainfall. Could aggregate to monthly/yearly for different time scale?

2. VPD curves:



1. 
GPP ~ VPD curve for all sites:

```{r}
all_sites <- bind_rows(sites_daily)

library(RColorBrewer)
# Define distinct colors for the sites
site_colors <- brewer.pal(n = 9, name = "Set1")  

ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_point(method = "loess", se = TRUE, linewidth = 1.2, span = 0.3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across OzFlux sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  )

```

```{r}
library(dplyr)
library(ggplot2)

gpp_by_vpd <- all_sites %>%
  # create VPD bins
  mutate(VPD_bin = cut(VPD, breaks = seq(0, max(VPD, na.rm=TRUE), by = 0.1))) %>%
  # group by site and the correct bin name
  group_by(site, VPD_bin) %>%
  summarise(
    mean_GPP = mean(GPP, na.rm = TRUE),
    n_points = n(),
    .groups = "drop"
  ) %>%
  # only keep bins with enough points
  filter(n_points >= 10) %>%
  # compute bin midpoint for plotting
  mutate(
    VPD_mid = (as.numeric(sub("\\((.+),.*", "\\1", VPD_bin)) +
               as.numeric(sub(".*,([^]]+)\\]", "\\1", VPD_bin))) / 2
  )


ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_point(alpha = 0.2, size = 1) +   # raw points
  geom_line(
    data = gpp_by_vpd,
    aes(x = VPD_mid, y = mean_GPP, color = site),
    size = 1
  ) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Vapor Pressure Deficit (kPa)",
    y = "GPP (g C m⁻² d⁻¹)",
    title = "Binned GPP vs VPD per site",
    color = "Site"
  )


```
now let's look at a PFT level:

```{r}
#add site column to the datasets
daily_names <- c(
  "How_HH_daily", "Ade_HH_daily", "Das_HH_daily", "Dap_HH_daily",
  "Dry_HH_daily", "Stp_HH_daily", "Asm_HH_daily", "Cum_HH_daily",
  "Emr_HH_daily", "Fog_HH_daily", "Gin_HH_daily", "Gww_HH_daily",
  "Lox_HH_daily", "Rdf_HH_daily", "Rig_HH_daily", "Rob_HH_daily",
  "TTE_HH_daily", "Wac_HH_daily", "Whr_HH_daily", "Wom_HH_daily",
  "Ync_HH_daily", "Tum_HH_daily", "Cpr_HH_daily"
)

# Pull the objects from the global environment into a list
sites_daily_list <- mget(daily_names)

sites_daily_list <- sites_daily_list %>%
  imap(~ .x %>% mutate(site = str_remove(.y, "_HH_daily$")))

list2env(sites_daily_list, envir = .GlobalEnv)

```


```{r}
WSA <- bind_rows(Ade_HH_daily, Gin_HH_daily, How_HH_daily, Rdf_HH_daily) %>% group_by(site)

SAV <- bind_rows(Das_HH_daily, Dry_HH_daily, Gww_HH_daily, Asm_HH_daily, Cpr_HH_daily) %>% group_by(site)

EBF <- bind_rows(Cum_HH_daily, Rob_HH_daily, Wac_HH_daily, Whr_HH_daily, Wom_HH_daily, Tum_HH_daily) %>% group_by(site)

GRA <- bind_rows(Dap_HH_daily, Emr_HH_daily, Rig_HH_daily, Stp_HH_daily, TTE_HH_daily, Ync_HH_daily) %>% group_by(site)

WET <- Fog_HH_daily %>% group_by(site)

DBF <- Lox_HH_daily %>% group_by(site)
```

```{r}
WSA$PFT <- "WSA"
SAV$PFT <- "SAV"
EBF$PFT <- "EBF"
GRA$PFT <- "GRA"
WET$PFT <- "WET"
DBF$PFT <- "DBF"
```

```{r}
PFT_all <- bind_rows(WSA, SAV, EBF, GRA, WET, DBF)
```

```{r}
library(dplyr)
library(ggplot2)

# Create VPD bins of width 0.5 kPa
PFT_all <- PFT_all %>%
  mutate(VPD_bin = cut(
    VPD,
    breaks = seq(floor(min(VPD, na.rm = TRUE)), ceiling(max(VPD, na.rm = TRUE)), by = 0.2),
    include.lowest = TRUE
  ))

# Calculate mean, SE, and number of points per PFT per VPD bin
summary_stats <- PFT_all %>%
  group_by(PFT, VPD_bin) %>%
  summarise(
    mean_GPP = mean(GPP, na.rm = TRUE),
    se_GPP = sd(GPP, na.rm = TRUE) / sqrt(n()),
    n_points = n(),
    .groups = "drop"
  ) %>%
  # Only keep bins with enough points
  filter(n_points >= 10)


# Plot binned means with error bars
ggplot(summary_stats, aes(x = VPD_bin, y = mean_GPP, colour = PFT, group = PFT)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_GPP - se_GPP, ymax = mean_GPP + se_GPP), width = 0.2) +
  geom_line(linewidth = 1) +
  theme_minimal(base_size = 14) +
  labs(
    x = "VPD (0.5 kPa bins)",
    y = "Mean GPP ± SE",
    colour = "PFT",
    title = "Binned GPP–VPD relationships across PFTs"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Map of Australia:

```{r}
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(viridisLite)


# Full site list with lat and long
sites <- tibble(
  site = c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox", "Cpr", "Tum"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283,
           -33.6152, -23.8587, -12.5452, -31.3764, -30.1913, -34.4704, -14.5636,
           -36.6499, -17.1175, -22.287, -37.4259, -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490,
           150.7236, 148.4746, 131.3072, 115.7138, 120.6541, 140.6551, 132.4776,
           145.5759, 145.6301, 133.64, 145.1878, 145.0294, 144.0944, 146.2907, 140.5891, 148.1517),
  wetness_order = NA  # need to assign numbers 1 (wettest) → 21 (driest) if known
)


# Get Australia map
aus <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")


# Plot
ggplot() +
  geom_sf(data = aus, fill = "antiquewhite", color = "black") +
  geom_point(data = sites, aes(x = lon, y = lat, color = wetness_order), size = 4) +
  geom_text(data = sites, aes(x = lon, y = lat, label = site), nudge_y = 0.3, size = 4) +
  scale_color_viridis_c(option = "C", direction = -1, na.value = "grey50", name = "Wet -> Dry") +
  coord_sf(xlim = c(110, 155), ylim = c(-40, -10), expand = FALSE) +  # full map of Australia
  theme_minimal() +
  labs(title = "FluxNet Sites Across Australia",
       x = "Longitude",
       y = "Latitude")

```
```

