---
title: "site_sensitivity"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Straight line slope
- Bin data and take the mean or max GPP value for a clean curve per site
- Identify the VPD bin where GPP is maximum (start point)
- Fit simple linear regression from start point - GPP normalised so slopes are comparable across sites
- Slope per site to compare

```{r}
library(tidyverse)

Ade_binned <- Ade_HH_daily %>%
  mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.5))) %>%
  group_by(vpd_bin) %>%
  summarise(gpp_mean = mean(GPP, na.rm = TRUE),
            vpd_mid = mean(VPD, na.rm = TRUE)) %>%
  mutate(gpp_norm = gpp_mean / max(gpp_mean, na.rm = TRUE)) %>%
  filter(vpd_mid >= vpd_mid[which.max(gpp_norm)])

lm(gpp_norm ~ vpd_mid, data = Ade_binned) %>% coef()
```

```{r}
# fit the model and get predictions
model <- lm(gpp_norm ~ vpd_mid, data = Ade_binned)

Ade_binned <- Ade_binned %>%
  mutate(fitted = predict(model))

# plot
ggplot(Ade_binned, aes(x = vpd_mid, y = gpp_norm)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.4) +  # connect the binned points
  geom_line(aes(y = fitted), colour = "firebrick", linewidth = 1) +
  labs(x = "VPD (kPa)", y = "Normalised GPP",
       title = "Adelaide River - GPP~VPD sensitivity") +
  theme_classic()
```

```{r}
df <- sites_daily[["Fog_HH_daily"]]

# check VPD range first
summary(df$VPD)

# then look at the binned data before any filtering
df %>%
  mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.5))) %>%
  group_by(vpd_bin) %>%
  summarise(
    gpp_max = quantile(GPP, 0.95, na.rm = TRUE),
    vpd_mid = mean(VPD, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  print(n = Inf)
```



```{r}
library(tidyverse)

site_slopes <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  binned <- df %>%
    drop_na(VPD, GPP) %>%
    mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.2))) %>%
    group_by(vpd_bin) %>%
    summarise(
      gpp_max = quantile(GPP, 0.95, na.rm = TRUE),
      vpd_mid = mean(VPD, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(n >= 3) %>%
    mutate(gpp_norm = gpp_max / max(gpp_max, na.rm = TRUE)) %>%
    filter(vpd_mid >= vpd_mid[which.max(gpp_norm)])
  
  # skip sites with too few bins to fit
  if (nrow(binned) < 3) {
    return(tibble(site = site, intercept = NA, slope = NA, n_bins = nrow(binned)))
  }
  
  model <- lm(gpp_norm ~ vpd_mid, data = binned)
  coefs <- coef(model)
  
  tibble(
    site = site,
    intercept = coefs[1],
    slope = coefs[2],
    n_bins = nrow(binned)
  )
})

print(site_slopes)
```

```{r}
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#highest mean LAI to lowest mean LAI
lai_order <- c("Rob", "Tum", "Wom", "Wac", "Dap", "Rig", "How", "Fog", "Dry", "Cum", "Das", "Rdf", "Ade", "Whr", "Gin", "Emr", "Lox", "Stp", "Cpr", "Ync", "Gww", "TTE", "Asm" )

site_slopes <- site_slopes %>%
  mutate(site = str_remove(site, "_HH_daily"))

site_slopes$site <- factor(site_slopes$site, levels = lai_order)


ggplot(site_slopes, aes(x = site, y = slope)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  labs(x = "Site", y = "GPP~VPD Slope", 
       title = "GPP sensitivity to VPD across sites") +
  theme_classic()
```
Rob and Wom unsure 

```{r}
site_slopes <- site_slopes %>%
  mutate(sensitivity_group = case_when(
    slope < -0.4 ~ "high",
    slope < -0.2 ~ "moderate",
    TRUE ~ "low"
  ))

```

Rob - artifact of 3 bins
Wom - interesting as its on the wetter end

High sensitivity - Ade (WSA), Dap (GRA), Fog (WET) (wetter more mesic sites)
Moderate sensitivity - How (WSA), Das (SAV), Dry (SAV), Stp (GRA), Rdf (WSA), Rig (GRA), TTE (GRA), Wac (EBF), Ync (GRA)
Low sensitivity - Cum (SAV), Emr (GRA), Gin (WSA), Gww (SAV), Lox (DBF), Whr (EBF), Wom (EBF), Cpr (SAV), Tum (EBF)

Stomatal strategy gradient? Rather than simply a rainfall gradient? 

Plant functional type does not seem to control VPD sensitivity


```{r}
# first generate binned data and fitted values for all sites
all_binned <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  binned_full <- df %>%
    drop_na(VPD, GPP) %>%
    mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.2))) %>%
    group_by(vpd_bin) %>%
    summarise(
      gpp_max = quantile(GPP, 0.95, na.rm = TRUE),
      vpd_mid = mean(VPD, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(n >= 3) %>%
    mutate(
      gpp_norm = gpp_max / max(gpp_max, na.rm = TRUE),
      site = site,
      descending = vpd_mid >= vpd_mid[which.max(gpp_norm)]
    )
  
  # fit model on descending limb only
  descending <- filter(binned_full, descending)
  
  if (nrow(descending) >= 3) {
    model <- lm(gpp_norm ~ vpd_mid, data = descending)
    binned_full <- binned_full %>%
      mutate(fitted = ifelse(descending, predict(model, newdata = binned_full), NA))
  } else {
    binned_full <- binned_full %>% mutate(fitted = NA)
  }
  
  binned_full
})

# plot
# split sites into groups of 6 for separate panels
site_names <- unique(all_binned$site)
n_per_panel <- 6
site_groups <- split(site_names, ceiling(seq_along(site_names) / n_per_panel))

plots <- map(site_groups, function(sites_subset) {
  all_binned %>%
    filter(site %in% sites_subset) %>%
    ggplot(aes(x = vpd_mid, y = gpp_norm)) +
    geom_line(alpha = 0.3) +
    geom_point(size = 1.5) +
    geom_line(aes(y = fitted), colour = "firebrick", linewidth = 1, na.rm = TRUE) +
    facet_wrap(~ site, scales = "free_x", ncol = 2) +
    labs(x = "VPD (kPa)", y = "Normalised GPP") +
    theme_classic(base_size = 10) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold"))
})

# print all panels
walk(plots, print)
```


