---
title: "site_sensitivity"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Straight line slope
- Bin data and take cloud of GPP values 
- Identify the VPD bin where GPP is maximum (start point)
- Fit simple linear regression from start point - GPP normalised so slopes are comparable across sites
- Slope per site to compare

```{r}
#cap for anomalous GPP

sites_daily <- map(sites_daily, function(df) {
  cap <- quantile(df$GPP, 0.99, na.rm = TRUE)
  df %>% mutate(GPP = ifelse(GPP > cap, NA, GPP))
})
```


```{r}
site_slopes <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  binned <- df %>%
    drop_na(VPD, GPP) %>%
    mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.2))) %>%
    group_by(vpd_bin) %>%
    summarise(
      gpp_max = quantile(GPP, 0.95, na.rm = TRUE),
      vpd_mid = mean(VPD, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(n >= 3) %>%
    mutate(gpp_norm = gpp_max / max(gpp_max, na.rm = TRUE))
  
  vpd_at_peak <- binned$vpd_mid[which.max(binned$gpp_norm)]
  descending  <- binned %>% filter(vpd_mid >= vpd_at_peak)
  
  if (nrow(descending) < 3) {
    return(tibble(site = site, intercept = NA, slope = NA,
                  vpd_at_peak = NA, n_bins = nrow(descending)))
  }
  
  model <- lm(gpp_norm ~ vpd_mid, data = descending)
  coefs <- coef(model)
  
  tibble(
    site        = site,
    intercept   = coefs[1],
    slope       = coefs[2],
    vpd_at_peak = vpd_at_peak,
    n_bins      = nrow(descending)
  )
}) %>%
  mutate(site = str_remove(site, "_HH_daily"))  # this closing }) was also missing


# Step 2: Build point cloud — raw daily points normalised per site
all_clouds <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  
gpp_max_site <- max(df$GPP, na.rm = TRUE)
  
  df %>%
    drop_na(VPD, GPP) %>%
    mutate(
      gpp_norm = GPP / gpp_max_site,
      site     = str_remove(site, "_HH_daily")
    ) %>%
    select(site, VPD, gpp_norm)
})


# Step 3: Build fitted lines from slopes + intercepts
fitted_lines <- site_slopes %>%
  filter(!is.na(slope)) %>%
  rowwise() %>%
  reframe(
    site    = site,
    vpd_mid = seq(vpd_at_peak, 6, by = 0.1),
    fitted  = intercept + slope * vpd_mid
  )


# Step 4: Plot — cloud + slope line per site
site_names  <- unique(all_clouds$site)
site_groups <- split(site_names, ceiling(seq_along(site_names) / 6))

plots <- map(site_groups, function(sites_subset) {
  
  all_clouds %>%
    filter(site %in% sites_subset) %>%
    ggplot(aes(x = VPD, y = gpp_norm)) +
    geom_point(size = 0.4, alpha = 0.2, colour = "grey40") +
    geom_line(data = filter(fitted_lines, site %in% sites_subset),
              aes(x = vpd_mid, y = fitted),
              colour = "firebrick", linewidth = 1) +
    geom_vline(data = filter(site_slopes, site %in% sites_subset),
               aes(xintercept = vpd_at_peak),
               linetype = "dashed", alpha = 0.4) +
    facet_wrap(~ site, scales = "free_x", ncol = 2) +
    coord_cartesian(ylim = c(0, 1.5)) +
    labs(x = "VPD (kPa)", y = "Normalised GPP") +
    theme_classic(base_size = 10) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold"))
})

walk(plots, print)

```

```{r}
# Check GPP distribution per site
sites_daily_df <- bind_rows(sites_daily)

ggplot(sites_daily_df, aes(x = GPP)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  facet_wrap(~ site, scales = "free") +
  theme_bw() +
  labs(title = "GPP distribution by site")

# Also look at the high end specifically
sites_daily_df %>%
  group_by(site) %>%
  summarise(
    max_GPP    = max(GPP, na.rm = TRUE),
    p99_GPP    = quantile(GPP, 0.99, na.rm = TRUE),
    p95_GPP    = quantile(GPP, 0.95, na.rm = TRUE),
    mean_GPP   = mean(GPP, na.rm = TRUE)
  ) %>%
  print(n = Inf)
```

Cluster analysis:
- Based on the GPP ~ VPD response, which sites cluster together?

1. Hierarchical clustering (only slopes)

```{r}
slopes_mat_1 <- site_slopes %>%
  filter(!is.na(slope)) %>%
  column_to_rownames("site") %>%
  select(slope) %>%
  scale()

hclust_fit_1 <- hclust(dist(slopes_mat_1, method = "euclidean"), method = "ward.D2")

site_slopes <- site_slopes %>%
  mutate(cluster_slope = factor(cutree(hclust_fit_1, k = 3)))
```


```{r}
# work out which cluster number corresponds to which sensitivity level
cluster_means <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster_slope = factor(cutree(hclust_fit_1, k = 3))) %>%
  group_by(cluster_slope) %>%
  summarise(mean_slope = mean(slope)) %>%
  arrange(mean_slope)  # most negative = highest sensitivity

# cluster_means will show you the order — then assign labels accordingly
level_order <- cluster_means$cluster_slope  # 1st = most negative = high

site_slopes <- site_slopes %>%
  mutate(
    cluster_slope = factor(cutree(hclust_fit_1, k = 3)),
    cluster_slope = fct_recode(cluster_slope,
      "high"   = as.character(level_order[1]),
      "medium" = as.character(level_order[2]),
      "low"    = as.character(level_order[3])
    )
  )
```


```{r}
site_slopes <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster = factor(cutree(hclust_fit_1, k = 3)))

ggplot(site_slopes, aes(x = site, y = slope, colour = cluster_slope)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  labs(x = "Site", y = "GPP~VPD Slope", colour = "Cluster",
       title = "GPP sensitivity clusters (slope only)") +
  theme_classic()
```

High - Fog
Medium - How, Das, Dap, Dry, Stp, Asm, Rdf, Rig, Wac, Ync
Low - Cum, Emr, Gin, Gww, Whr, Wom, Cpr, Tum

2. Hierarchical clustering 2
- slope
- intercept
- VPD when GPP is at its maximum

```{r}
# Step 1: fit and assign clusters
slopes_mat_2 <- site_slopes %>%
  filter(!is.na(slope)) %>%
  column_to_rownames("site") %>%
  select(slope, intercept, vpd_at_peak) %>%
  scale()

hclust_fit_2 <- hclust(dist(slopes_mat_2, method = "euclidean"), method = "ward.D2")

# Step 2: derive level order from hclust_fit_2 specifically
cluster_means_2 <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster_raw = factor(cutree(hclust_fit_2, k = 3))) %>%
  group_by(cluster_raw) %>%
  summarise(mean_slope = mean(slope)) %>%
  arrange(mean_slope)  # most negative = high sensitivity

level_order_2 <- cluster_means_2$cluster_raw

# Step 3: assign and recode in one clean step
site_slopes <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(
    cluster_multi = factor(cutree(hclust_fit_2, k = 3)),
    cluster_multi = fct_recode(cluster_multi,
      "high"   = as.character(level_order_2[1]),
      "medium" = as.character(level_order_2[2]),
      "low"    = as.character(level_order_2[3])
    ),
    cluster_multi = factor(cluster_multi, levels = c("low", "medium", "high"))
  )

```

```{r}
# Step 4: plot
ggplot(site_slopes %>% 
         mutate(site = factor(site, levels = wet_to_dry_order)),
       aes(x = site, y = slope, colour = cluster_multi)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  scale_colour_manual(values = c("low"    = "steelblue",
                                 "medium" = "orange",
                                 "high"   = "firebrick")) +
  labs(x = "Site (Wet to Dry)", y = "GPP~VPD Slope", colour = "Sensitivity",
       title = "GPP sensitivity clusters") +
  theme_classic()
```

1. Is cluster membership non-random with respect to PFT, precipitation regime, or biome/climate zone?

Add on PFT column
Mean annual precip column
Max annual LAI
Mean annual temp
Biome/climatic classification

```{r}
site_meta_2 <- tibble(
  site = c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", 
           "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "Ync", "Boy"),
  PFT = c("EBF", "WSA", "EBF", "SAV", "WET", "GRA", "WSA", "SAV", "GRA", "EBF", "WSA", "EBF", "WSA", "GRA", "EBF", "SAV", "GRA", "SAV", "SAV", "EBF", "GRA", "WSA"),
  koppen_climate = c("Tropical", "Tropical", "Temperate", "Tropical", "Tropical", "Tropical", "Tropical",
                     "Tropical", "Grassland", "Temperate", "Tropical", "Temperate", "Subtropical", "Grassland", "Temperate", "Grassland", "Temperate", "Desert", "Grassland", "Temperate", "Grassland", "Temperate")
)
```

```{r}
site_climate <- map_dfr(names(sites_daily), function(site) {
  df <- sites_daily[[site]]
  
  annual_precip <- df %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarise(precip_annual = sum(Precip, na.rm = TRUE)) %>%
    pull(precip_annual) %>%
    mean(na.rm = TRUE)
  
  # MAT as mean of annual means
  annual_mat <- df %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarise(tair_annual = mean(Tair, na.rm = TRUE)) %>%
    pull(tair_annual) %>%
    mean(na.rm = TRUE)
  
  tibble(
    site = site,
    MAP  = annual_precip,
    MAT  = annual_mat,
    MAL  = max(df$LAI_filled, na.rm = TRUE)
  )
})

# single clean join — remove any duplicate columns first
site_slopes <- site_slopes %>%
  select(-any_of(c("mat", "map", "mal", "malcc", "MAP", "MAT", "MAL",
                   "lat", "lon", "PFT", "koppen_climate"))) %>%
  left_join(site_meta_2, by = "site") %>%
  left_join(site_meta %>% select(name, lat, lon), by = c("site" = "name")) %>%
  left_join(site_climate, by = "site")
```

Is cluster membership non-random?

Categorical (climate, PFT) = Fisher's exact test

```{r}
# Fisher's exact test (better than chi-square for small n)
fisher.test(table(site_slopes$cluster_multi, site_slopes$PFT), simulate.p.value = TRUE)
fisher.test(table(site_slopes$cluster_multi, site_slopes$koppen_climate), simulate.p.value = TRUE)

# look at the raw counts too
table(site_slopes$cluster_multi, site_slopes$PFT)
table(site_slopes$cluster_multi, site_slopes$koppen_climate)
```



```{r}
# Kruskal-Wallis (non-parametric ANOVA — better than ANOVA for small n)
kruskal.test(MAP ~ cluster_multi, data = site_slopes)
kruskal.test(MAT ~ cluster_multi, data = site_slopes)
kruskal.test(MAL ~ cluster_multi, data = site_slopes)

# If significant, follow up with pairwise Wilcoxon to see which clusters differ
pairwise.wilcox.test(site_slopes$MAP, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
pairwise.wilcox.test(site_slopes$MAT, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
pairwise.wilcox.test(site_slopes$MAL, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
```

```{r}
site_slopes <- site_slopes %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("low", "medium", "high")))

library(ggplot2)

# Boxplots for continuous
site_slopes %>%
  pivot_longer(cols = c(MAP, MAT, MAL), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = cluster_multi, y = value, fill = cluster_multi)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(x = "Sensitivity cluster", y = NULL) +
  theme_classic() +
  theme(legend.position = "none")

# Bar chart for categorical
site_slopes %>%
  ggplot(aes(x = cluster_multi, fill = koppen_climate)) +
  geom_bar(position = "fill") +
  labs(x = "Sensitivity cluster", y = "Proportion") +
  theme_classic()
```

Are nearby sites more similar in slope than distant sites?
Moran's test - whether nearby sites are similar

High LAI but low sensitivity - i believe is Tum, Wac, Wom, Rig (EBF, EBF, EBF, GRA) - They are all grouped together in South Australia


```{r}
cluster_colours <- c(
  "low" = "steelblue",  # steel blue
  "high" = "firebrick",  # raspberry
  "medium" = "orange"   # amber
)

site_slopes %>%
  pivot_longer(cols = c(MAP, MAT, MAL), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = cluster_multi, y = value, fill = cluster_multi)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  scale_fill_manual(values = cluster_colours) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(x = "Sensitivity cluster", y = NULL) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
unique(site_slopes$cluster_multi)
```


```{r}
library(ape)

# build inverse distance weight matrix
coords  <- site_slopes %>% filter(!is.na(slope)) %>% select(lon, lat) %>% as.matrix()
slopes  <- site_slopes %>% filter(!is.na(slope)) %>% pull(slope)

dist_mat    <- as.matrix(dist(coords))
weight_mat  <- 1 / dist_mat
diag(weight_mat) <- 0

# run Moran's I
Moran.I(slopes, weight_mat)
```

```{r}
library(ggplot2)

# map of sites coloured by sensitivity cluster
site_slopes %>%
  filter(!is.na(slope)) %>%
  ggplot(aes(x = lon, y = lat, colour = cluster_multi, size = abs(slope))) +
  geom_point() +
  scale_colour_manual(values = c("low" = "steelblue", "medium" = "orange", "high" = "firebrick")) +
  labs(x = "Longitude", y = "Latitude", colour = "Sensitivity", size = "|Slope|",
       title = "Spatial distribution of GPP~VPD sensitivity") +
  theme_classic()
```
Observed value is above expected and the difference is significant. Sensitivity does cluster spatially. Sites that are geographically close tend to have similar GPP ~ VPD responses.

```{r}
install.packages("ozmaps")
```



```{r}
library(ozmaps)
library(sf)

aus <- ozmap_country

ggplot() +
  geom_sf(data = aus, fill = "grey95", colour = "grey60") +
  geom_point(data = site_slopes,
             aes(x = lon, y = lat, colour = cluster_multi, size = abs(slope)),
             alpha = 0.9) +
  ggrepel::geom_text_repel(data = site_slopes,
                            aes(x = lon, y = lat, label = site), size = 3) +
  scale_colour_manual(values = c("low" = "steelblue",
                                 "medium" = "orange", 
                                 "high" = "firebrick")) +
  labs(x = "Longitude", y = "Latitude",
       colour = "VPD Sensitivity", size = "|Slope|") +
  theme_classic()
```
```{r}
library(plotbiomes)
library(ggplot2)
library(ggrepel)

site_slopes$map_cm <- site_slopes$MAP / 10

whittaker_base_plot() +
  geom_point(data = site_slopes,
             aes(x = MAT, y = map_cm, colour = cluster_multi),
             size = 3) +
  geom_text_repel(data = site_slopes,
                  aes(x = MAT, y = map_cm, label = site),
                  size = 3,
                  max.overlaps = 20) +
  scale_colour_manual(values = c("low"    = "steelblue",
                                 "medium" = "orange",
                                 "high"   = "firebrick"),
                      name = "VPD Sensitivity") +
  labs(x = "Mean Annual Temperature (°C)",
       y = "Mean Annual Precipitation (cm)") +
  theme_classic() +
  theme(panel.grid       = element_blank(),
        panel.background = element_rect(fill = "white", colour = NA),
        plot.background  = element_rect(fill = "white", colour = NA))
```

