---
title: "site_sensitivity"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Straight line slope
- Bin data and take cloud of GPP values 
- Identify the VPD bin where GPP is maximum (start point)
- Fit simple linear regression from start point - GPP normalised so slopes are comparable across sites
- Slope per site to compare


```{r}
site_slopes <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  binned <- df %>%
    drop_na(VPD, GPP) %>%
    mutate(vpd_bin = cut(VPD, breaks = seq(0, 6, by = 0.2))) %>%
    group_by(vpd_bin) %>%
    summarise(
      gpp_max = quantile(GPP, 0.95, na.rm = TRUE),
      vpd_mid = mean(VPD, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(n >= 3) %>%
    mutate(gpp_norm = gpp_max / max(gpp_max, na.rm = TRUE))
  
  vpd_at_peak <- binned$vpd_mid[which.max(binned$gpp_norm)]
  descending  <- binned %>% filter(vpd_mid >= vpd_at_peak)
  
  if (nrow(descending) < 3) {
    return(tibble(site = site, intercept = NA, slope = NA,
                  vpd_at_peak = NA, n_bins = nrow(descending)))
  }
  
  model <- lm(gpp_norm ~ vpd_mid, data = descending)
  coefs <- coef(model)
  
  tibble(
    site        = site,
    intercept   = coefs[1],
    slope       = coefs[2],
    vpd_at_peak = vpd_at_peak,
    n_bins      = nrow(descending)
  )
}) %>%
  mutate(site = str_remove(site, "_HH_daily"))  # this closing }) was also missing


# Step 2: Build point cloud — raw daily points normalised per site
all_clouds <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]]
  
  
gpp_max_site <- max(df$GPP, na.rm = TRUE)
  
  df %>%
    drop_na(VPD, GPP) %>%
    mutate(
      gpp_norm = GPP / gpp_max_site,
      site     = str_remove(site, "_HH_daily")
    ) %>%
    select(site, VPD, gpp_norm)
})


# Step 3: Build fitted lines from slopes + intercepts
fitted_lines <- site_slopes %>%
  filter(!is.na(slope)) %>%
  rowwise() %>%
  reframe(
    site    = site,
    vpd_mid = seq(vpd_at_peak, 6, by = 0.1),
    fitted  = intercept + slope * vpd_mid
  )


# Step 4: Plot — cloud + slope line per site
site_names  <- unique(all_clouds$site)
site_groups <- split(site_names, ceiling(seq_along(site_names) / 6))

plots <- map(site_groups, function(sites_subset) {
  
  all_clouds %>%
    filter(site %in% sites_subset) %>%
    ggplot(aes(x = VPD, y = gpp_norm)) +
    geom_point(size = 0.4, alpha = 0.2, colour = "grey40") +
    geom_line(data = filter(fitted_lines, site %in% sites_subset),
              aes(x = vpd_mid, y = fitted),
              colour = "firebrick", linewidth = 1) +
    geom_vline(data = filter(site_slopes, site %in% sites_subset),
               aes(xintercept = vpd_at_peak),
               linetype = "dashed", alpha = 0.4) +
    facet_wrap(~ site, scales = "free_x", ncol = 2) +
    coord_cartesian(ylim = c(0, 1.5)) +
    labs(x = "VPD (kPa)", y = "Normalised GPP") +
    theme_classic(base_size = 10) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold"))
})

walk(plots, print)
```

Need to note in the results that some of the sites did not have many points

Rob - artifact of 3 bins
Wom - interesting as its on the wetter end

High sensitivity - Ade (WSA), Dap (GRA), Fog (WET) (wetter more mesic sites)
Moderate sensitivity - How (WSA), Das (SAV), Dry (SAV), Stp (GRA), Rdf (WSA), Rig (GRA), TTE (GRA), Wac (EBF), Ync (GRA)
Low sensitivity - Cum (SAV), Emr (GRA), Gin (WSA), Gww (SAV), Lox (DBF), Whr (EBF), Wom (EBF), Cpr (SAV), Tum (EBF)

Stomatal strategy gradient? Rather than simply a rainfall gradient? 

Plant functional type does not seem to control VPD sensitivity


Cluster analysis:
- Based on the GPP ~ VPD response, which sites cluster together?

1. Hierarchical clustering (only slopes)

```{r}
slopes_mat_1 <- site_slopes %>%
  filter(!is.na(slope)) %>%
  column_to_rownames("site") %>%
  select(slope) %>%
  scale()

hclust_fit_1 <- hclust(dist(slopes_mat_1, method = "euclidean"), method = "ward.D2")

site_slopes <- site_slopes %>%
  mutate(cluster_slope = factor(cutree(hclust_fit_1, k = 3)))
```


```{r}
# work out which cluster number corresponds to which sensitivity level
cluster_means <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster_slope = factor(cutree(hclust_fit_1, k = 3))) %>%
  group_by(cluster_slope) %>%
  summarise(mean_slope = mean(slope)) %>%
  arrange(mean_slope)  # most negative = highest sensitivity

# cluster_means will show you the order — then assign labels accordingly
level_order <- cluster_means$cluster_slope  # 1st = most negative = high

site_slopes <- site_slopes %>%
  mutate(
    cluster_slope = factor(cutree(hclust_fit_1, k = 3)),
    cluster_slope = fct_recode(cluster_slope,
      "high"   = as.character(level_order[1]),
      "medium" = as.character(level_order[2]),
      "low"    = as.character(level_order[3])
    )
  )
```


```{r}
site_slopes <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster = factor(cutree(hclust_fit_1, k = 3)))

ggplot(site_slopes, aes(x = site, y = slope, colour = cluster_slope)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  labs(x = "Site", y = "GPP~VPD Slope", colour = "Cluster",
       title = "GPP sensitivity clusters") +
  theme_classic()
```

Cluster 1 - Asm, Gww, Stp, Cum, Emr, Gin, Lox, Rig, Wac, Whr, Wom, Ync, Cpr, Tum
Cluster 2 - Ade, Das, Dap, Dry, Rdf, TTE
Cluster 3 - Rob, Fog 

2. Hierarchical clustering 2
- slope
- intercept
- VPD when GPP is at its maximum

```{r}
slopes_mat_2 <- site_slopes %>%
  filter(!is.na(slope)) %>%
  column_to_rownames("site") %>%
  select(slope, intercept, vpd_at_peak) %>%
  scale()

hclust_fit_2 <- hclust(dist(slopes_mat_2, method = "euclidean"), method = "ward.D2")

site_slopes <- site_slopes %>%
  mutate(cluster_multi = factor(cutree(hclust_fit_2, k = 3)))

site_slopes <- site_slopes %>%
  mutate(
    cluster_multi = factor(cutree(hclust_fit_2, k = 3)),
    cluster_multi = fct_recode(cluster_multi,
      "high"   = "2",
      "medium" = "3",
      "low"    = "1"
    )
  )
```

```{r}
site_slopes <- site_slopes %>%
  filter(!is.na(slope)) %>%
  mutate(cluster = factor(cutree(hclust_fit_2, k = 3)))

ggplot(site_slopes, aes(x = site, y = slope, colour = cluster_multi)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  labs(x = "Site", y = "GPP~VPD Slope", colour = "Cluster",
       title = "GPP sensitivity clusters") +
  theme_classic()
```
Cluster 1 - How (WSA), Stp (GRA), Asm (SAV), Cum (EBF), Emr (GRA), Gin (WSA), Gww (SAV), Lox (DBF), Rig (GRA), Whr (EBF), Wom (EBF), Ync (GRA), Cpr (SAV), Tum (EBF)
Cluster 2 - Ade (WSA), Dap (GRA), Fog (WET), Rob (EBF)
Cluster 3 - Das (SAV), Dry (SAV), Rdf (WSA), TTE (GRA), Wac (EBF)

1. Is cluster membership non-random with respect to PFT, precipitation regime, or biome/climate zone?

Add on PFT column
Mean annual precip column
Max annual LAI
Mean annual temp
Biome/climatic classification

```{r}
site_meta_2 <- tibble(
  site = c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", 
           "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", 
           "TTE", "Ync", "Lox"),
  PFT = c("EBF", "WSA", "EBF", "SAV", "WET", "GRA", "WSA", "SAV", "GRA", "EBF", "WSA", "EBF", "WSA", "GRA", "EBF", "SAV", "GRA", "SAV", "SAV", "EBF", "GRA", "GRA", "DBF"),
  koppen_climate = c("Tropical", "Tropical", "Temperate", "Tropical", "Tropical", "Tropical", "Tropical",
                     "Tropical", "Grassland", "Temperate", "Tropical", "Temperate", "Subtropical", "Grassland", "Temperate", "Grassland", "Temperate", "Desert", "Grassland", "Temperate", "Desert", "Grassland", "Grassland")
)
```

```{r}
site_climate <- map_dfr(names(sites_daily), function(site) {
  df <- sites_daily[[site]]
  
  # annual precip per year then average
  annual_precip <- df %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarise(precip_annual = sum(Precip, na.rm = TRUE)) %>%
    pull(precip_annual) %>%
    mean(na.rm = TRUE)
  
  tibble(
    site = site,
    MAP  = annual_precip,
    MAT  = mean(df$Tair, na.rm = TRUE),
    MAL  = max(df$LAI_4day, na.rm = TRUE)
  )
})

# then join everything onto site_slopes
site_slopes <- site_slopes %>%
  left_join(site_meta_2, by = "site") %>%
  left_join(site_meta, by = c("site" = "name")) %>%
  left_join(site_climate, by = "site")
```

Is cluster membership non-random?

Categorical (climate, PFT) = Fisher's exact test

```{r}
# Fisher's exact test (better than chi-square for small n)
fisher.test(table(site_slopes$cluster_multi, site_slopes$PFT), simulate.p.value = TRUE)
fisher.test(table(site_slopes$cluster_multi, site_slopes$koppen_climate), simulate.p.value = TRUE)

# look at the raw counts too
table(site_slopes$cluster_multi, site_slopes$PFT)
table(site_slopes$cluster_multi, site_slopes$koppen_climate)
```

PFT p = 0.89
PFT does not predict VPD sensitivity. 

Koppen climate p = 0.01
cluster membership is non-random with respect to climate
High sensitivity sites exclusively tropical
Low sensitivity sites dominated by grassland and temperate zones
Medium sensitivity split between desert, tropical and temperate

VPD sensitivity may not be about the vegetation a site has but the climate regime the vegetation operates within. Temperate and grasslands are less sensitive maybe because they already operate close to their water limitation threshold and stomatal responses are more conservative. 

Is tropical vegetation more sensitive to VPD, or is it that tropical climates produce more extreme VPD excursions that push GPP into declines more frequently? 



Continuous (mean annual precipitation, mean annual temperature and max LAI)

```{r}
# Kruskal-Wallis (non-parametric ANOVA — better than ANOVA for small n)
kruskal.test(MAP ~ cluster_multi, data = site_slopes)
kruskal.test(MAT ~ cluster_multi, data = site_slopes)
kruskal.test(MAL ~ cluster_multi, data = site_slopes)

# If significant, follow up with pairwise Wilcoxon to see which clusters differ
pairwise.wilcox.test(site_slopes$MAP, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
pairwise.wilcox.test(site_slopes$MAT, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
pairwise.wilcox.test(site_slopes$MAL, site_slopes$cluster_multi, p.adjust.method = "bonferroni")
```
Mean annual precipitation (p = 0.022)
High sensitivity sites have significantly higher mean annual precipitation than low sensitivity sites. 
Wetter sites are more sensitive to VPD
Plants in high rainfall sites are adapted to higher water availability and may have less conservative stomatal regulation

```{r}
site_slopes <- site_slopes %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("low", "medium", "high")))

library(ggplot2)

# Boxplots for continuous
site_slopes %>%
  pivot_longer(cols = c(MAP, MAT, MAL), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = cluster_multi, y = value, fill = cluster_multi)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(x = "Sensitivity cluster", y = NULL) +
  theme_classic() +
  theme(legend.position = "none")

# Bar chart for categorical
site_slopes %>%
  ggplot(aes(x = cluster_multi, fill = koppen_climate)) +
  geom_bar(position = "fill") +
  labs(x = "Sensitivity cluster", y = "Proportion") +
  theme_classic()
```






2. Simple regression of slope ~ climate variable (LAI, precip, mean VPD regime) and a multiple regression

3. Variance partitioning 


4. Spatial scale - Moran's test, is sensitivity a site-level phenomenon? 
If this test is significant, sensitivity clusters spatially - climate, regional constraint. However, if not it is more likely a site/vegetation strategy.


