---
title: "explore_ozflux"
output: html_document
date: "2026-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Tue 27th Jan:

Going to look at the North Australian Tropical Transect flux sites.

1. Howard Springs = AU-How (2001-2014) 13 years
2. Daly River Pasture = AU-DaP (2007-2013) 6 years
3. Daly River Savanna = AU-DaS (2008-2014) 6 years
4. Dry River = AU-Dry (2008-2014) 6 years
5. Sturt Plains = AU-Stp (2008-2014) 6 years
6. Alice Springs = AU-ASM (2010-2014) 4 years

These flux sites are located along a precipitation gradient. This rainfall gradient is a 'natural laboratory' with Howard Springs as the wettest and Alice Springs the driest. 

1. Read in flux files:

```{r}
library(tidyverse)

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

```{r}
AU_How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

AU_Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

AU_Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

AU_Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")
```

2. Filter by growing season

```{r}
filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}
```

```{r}
AU_How_gs <- filter_gs(AU_How)
AU_Dap_gs <- filter_gs(AU_Dap)
AU_Das_gs <- filter_gs(AU_Das)
AU_Dry_gs <- filter_gs(AU_Dry)
AU_Stp_gs <- filter_gs(AU_Stp)
AU_Asm_gs <- filter_gs(AU_Asm)

```

3. Drop missing variables

```{r}
How_final <- AU_How_gs %>% drop_na()
Dap_final <- AU_Dap_gs %>% drop_na()
Das_final <- AU_Das_gs %>% drop_na()
Dry_final <- AU_Dry_gs %>% drop_na()
Stp_final <- AU_Stp_gs %>% drop_na()
Asm_final <- AU_Asm_gs %>% drop_na()
```

4. Remove missing flags (-999)

```{r}
How_final <- How_final %>% filter_all(all_vars(. >= -9000) )
Dap_final <- Dap_final %>% filter_all(all_vars(. >= -9000) )
Das_final <- Das_final %>% filter_all(all_vars(. >= -9000) )
Dry_final <- Dry_final %>% filter_all(all_vars(. >= -9000) )
Stp_final <- Stp_final %>% filter_all(all_vars(. >= -9000) )
Asm_final <- Asm_final %>% filter_all(all_vars(. >= -9000) )
```
5. Visualise 

```{r}
#Combine the datasets

library(dplyr)

dataframe_oz <- bind_rows(How_final, Dap_final, Das_final, Dry_final, Stp_final, Asm_final)
```

```{r}
#Summarise by site, by year

df_summary <- dataframe_oz %>%
  group_by(site, year) %>%
  summarise(
    GPP_mean = mean(GPP, na.rm = TRUE),
    GPP_max = max(GPP, na.rm = TRUE),
    GPP_min = min(GPP, na.rm = TRUE)
  ) %>%
  
  ungroup()
```

```{r}
library(tidyr)
library(ggplot2)

df_long <- df_summary %>%
  pivot_longer(
    cols = c(GPP_mean, GPP_max, GPP_min),
    names_to = "stat",
    values_to = "GPP"
  )

ggplot(df_long, aes(x = year, y = GPP, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "GPP", x = "Year", color = "Statistic")
```


```{r}
ggplot(df_summary, aes(x = year, group = site)) +
  geom_ribbon(aes(ymin = GPP_min, ymax = GPP_max, fill = site), alpha = 0.2) +
  geom_line(aes(y = GPP_mean, color = site), size = 1) +
  theme_minimal() +
  labs(y = "GPP", x = "Year")
```

Question: Which site along the NATT is the most/least predictable?
"Along a precipitation gradient (wet -> dry), how well can atmospheric drivers predict GPP?

5. Fit GAM firstly with Tair ~ GPP

- Load in relevant libraries

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
```

- Train/Test within site

```{r}
idx <- createDataPartition(Asm_final$GPP, p=0.8, list=FALSE)
train_data <- Asm_final[idx, ]
test_data <- Asm_final[-idx, ]
```

- Fit GAM

```{r}
How_model <- gam(GPP ~ s(VPD) + s(SW_rad) + s(Precip) + s(Tair), data=train_data, 
                 method = "REML")
```

- Get some stats:

```{r}
summary(How_model)
```
```{r}

predictions_how <- predict(How_model, newdata=test_data)
results_how <- data.frame(observed=test_data$GPP, predicted=predictions_how)
```

```{r}
rmse_gam <- sqrt(mean((results_how$observed - results_how$predicted)^2))
rsq_gam <- cor(results_how$observed, results_how$predicted)^2
mean_observed_gam <- mean(results_how$observed)
nse_gam <- (1 - sum((results_how$observed - results_how$predicted)^2) / 
              sum((results_how$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

Comparing the R-squared from each of the models, AU-DaP is the most predictable site and AU-Stp is the least predictable site.

When comparing RMSE, AU-Asm has the lowest RMSE and AU-Dap has the highest RMSE. 

Wed 28th Jan:
- Improve GAMs for each site, what makes these sites more predictable?
- GPP predictability time of day
- GPP predictability by year 
- Post drought and post-VPD extreme events

1. Add water stress component

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
```

```{r}
priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}
```

Alice Springs: ASM

```{r}
Asm_final$pet <- priestley_taylor_pet(Asm_final$Rn, Asm_final$Tair, Asm_final$Pa)
Asm_final$cwd <- Asm_final$pet - Asm_final$Qle
```

Sturt Plains: Stp

```{r}
Stp_final$pet <- priestley_taylor_pet(Stp_final$Rn, Stp_final$Tair, Stp_final$Pa)
Stp_final$cwd <- Stp_final$pet - Stp_final$Qle
```

Dry River: Dry

```{r}
Dry_final$pet <- priestley_taylor_pet(Dry_final$Rn, Dry_final$Tair, Dry_final$Pa)
Dry_final$cwd <- Dry_final$pet - Dry_final$Qle
```

Daly River Savanna: DaS

```{r}
Das_final$pet <- priestley_taylor_pet(Das_final$Rn, Das_final$Tair, Das_final$Pa)
Das_final$cwd <- Das_final$pet - Das_final$Qle
```

Daly River Pasture: DaP

```{r}
Dap_final$pet <- priestley_taylor_pet(Dap_final$Rn, Dap_final$Tair, Dap_final$Pa)
Dap_final$cwd <- Dap_final$pet - Dap_final$Qle
```

Howard Springs: How

```{r}
How_final$pet <- priestley_taylor_pet(How_final$Rn, How_final$Tair, How_final$Pa)
How_final$cwd <- How_final$pet - How_final$Qle
```

Split by training and testing:

```{r}
idx <- createDataPartition(How_final$GPP, p=0.8, list=FALSE)
train_data <- How_final[idx, ]
test_data <- How_final[-idx, ]
```

Drop n/a's
```{r}
test_data <- test_data %>% drop_na()
```

7-day moving cumulative sum for cwd
- Accumulated drought, not only the day's stress

```{r}

train_data$cumulative_cwd_7day <- rollapply(train_data$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
test_data$cumulative_cwd_7day <-  rollapply(test_data$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
```

Drop any NAs after this

```{r}
train_data <- train_data %>% drop_na()
test_data <- test_data %>% drop_na()
```

Lag the climate variables

```{r}
train_data <- train_data %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

test_data <- test_data %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()
```

Fit model

```{r}
model <- gam(GPP ~ s(Tair) + s(VPD) + s(cumulative_cwd_7day) + 
                   s(Tair_lag3) + s(VPD_lag3), data=train_data, 
                 method = "REML")

predictions <- predict(model, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")


```

When we add cwd, and lag climate variables, predictability improves for AU-Asm, AU-Stp quite significantly. Improves AU-Dry and AU-Dap a little. However, R-squared is smaller with cwd for AU-Das and AU-How. 

Likely because we move from a water-limited climate (ASM) to an energy-limited climate/mesic environment and so soil water doesn't predict GPP as strongly. 

Standardise model evaluation across sites:
I want to store model results in a table from each site

1. train + test data for each site

```{r}
idx <- createDataPartition(Asm_final$GPP, p=0.8, list=FALSE)
Asm_train <- Asm_final[idx, ]
Asm_test <- Asm_final[-idx, ]

idx <- createDataPartition(Stp_final$GPP, p=0.8, list=FALSE)
Stp_train <- Stp_final[idx, ]
Stp_test <- Stp_final[-idx, ]

idx <- createDataPartition(Dry_final$GPP, p=0.8, list=FALSE)
Dry_train <- Dry_final[idx, ]
Dry_test <- Dry_final[-idx, ]

idx <- createDataPartition(Das_final$GPP, p=0.8, list=FALSE)
Das_train <- Das_final[idx, ]
Das_test <- Das_final[-idx, ]

idx <- createDataPartition(Dap_final$GPP, p=0.8, list=FALSE)
Dap_train <- Dap_final[idx, ]
Dap_test <- Dap_final[-idx, ]

idx <- createDataPartition(How_final$GPP, p=0.8, list=FALSE)
How_train <- How_final[idx, ]
How_test <- How_final[-idx, ]

```

Lag climate variables...

```{r}
Asm_train$cumulative_cwd_7day <- rollapply(Asm_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Asm_test$cumulative_cwd_7day <-  rollapply(Asm_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Stp_train$cumulative_cwd_7day <- rollapply(Stp_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Stp_test$cumulative_cwd_7day <-  rollapply(Stp_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dry_train$cumulative_cwd_7day <- rollapply(Dry_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dry_test$cumulative_cwd_7day <-  rollapply(Dry_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Das_train$cumulative_cwd_7day <- rollapply(Das_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Das_test$cumulative_cwd_7day <-  rollapply(Das_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dap_train$cumulative_cwd_7day <- rollapply(Dap_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dap_test$cumulative_cwd_7day <-  rollapply(Dap_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

How_train$cumulative_cwd_7day <- rollapply(How_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
How_test$cumulative_cwd_7day <-  rollapply(How_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

```

```{r}
climate_lags <- function(df, lag_n = 3) {
  df %>%
    mutate(
      Tair_lag = lag(Tair, lag_n),
      VPD_lag = lag(VPD, lag_n)
    )
}
```

Apply to data:

```{r}

Asm_train <- climate_lags(Asm_train, lag_n = 3)
Asm_test <- climate_lags(Asm_test, lag_n = 3)

Stp_train <- climate_lags(Stp_train, lag_n = 3)
Stp_test <- climate_lags(Stp_test, lag_n = 3)

Dry_train <- climate_lags(Dry_train, lag_n = 3)
Dry_test <- climate_lags(Dry_test, lag_n = 3)

Das_train <- climate_lags(Das_train, lag_n = 3)
Das_test <- climate_lags(Das_test, lag_n = 3)

Dap_train <- climate_lags(Dap_train, lag_n = 3)
Dap_test <- climate_lags(Dap_test, lag_n = 3)

How_train <- climate_lags(How_train, lag_n = 3)
How_test <- climate_lags(How_test, lag_n = 3)



```

Remove NAs:

```{r}
Asm_train <- Asm_train %>% drop_na()
Asm_test <- Asm_test %>% drop_na()

Stp_train <- Stp_train %>% drop_na()
Stp_test <- Stp_test %>% drop_na()

Dry_train <- Dry_train %>% drop_na()
Dry_test <- Dry_test %>% drop_na()

Das_train <- Das_train %>% drop_na()
Das_test <- Das_test %>% drop_na()

Dap_train <- Dap_train %>% drop_na()
Dap_test <-Dap_test %>% drop_na()

How_train <- How_train %>% drop_na()
How_test <-How_test %>% drop_na()

```

Function to fit GAM (one site)

```{r}
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(cumulative_cwd_7day) +
          s(Tair_lag) + s(VPD_lag),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

```

Run function for each site, combine results. This saves the R-squared and RMSE results into a table called 'site_metrics'

```{r}
site_metrics <- bind_rows(
  fit_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "AU-ASM"),
  fit_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "AU-Stp"),
  fit_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "AU-DaP"),
  fit_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "AU-DaS"),
  fit_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "AU-Dry"),
  fit_gam_and_metrics(How_train, How_test) %>% mutate(site = "AU-How")
)

```

Now, going to try plot R-squared:

```{r}
ggplot(site_metrics, aes(x = reorder(site, R2), y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites"
  ) +
  theme_minimal()

```

Drier sites comparatively are easier to predict. Why?
- This GAM used Tair, VPD, cwd and lagTair and lagVPD
- Drier sites are more limited by water, whereas the mesic sites (e.g., AU-How) are more energy limited/not as limited by water.

So what happens now if we look at drought/VPD events?
1. How does drought affect GPP across sites? Are there differences?
2. What are the legacy effects of drought across these sites?
3. Can a GAM trained on non-drought years be able to predict GPP during a drought year or a dry period?

```{r}
library(tidyr)
library(ggplot2)

df_long <- df_summary %>%
  pivot_longer(
    cols = c(GPP_mean, GPP_max, GPP_min),
    names_to = "stat",
    values_to = "GPP"
  )

ggplot(df_long, aes(x = year, y = GPP, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "GPP", x = "Year", color = "Statistic") +
  theme(legend.position = "none")
```

Let's see if this matches VPD, Precip, temp

```{r}
climate_summary <- dataframe_oz %>%
  group_by(site, year) %>%
  summarise(
    VPD_mean = mean(VPD, na.rm = TRUE),
    VPD_max = max(VPD, na.rm = TRUE),
    VPD_min = min(VPD, na.rm = TRUE),
    Precip_mean = mean(Precip, na.rm = TRUE),
    Precip_max = max(Precip, na.rm = TRUE),
    Precip_min = min(Precip, na.rm = TRUE),
    Tair_mean = mean(Tair, na.rm = TRUE),
    Tair_max = max(Tair, na.rm = TRUE),
    Tair_min = min(Tair, na.rm = TRUE)
  ) %>%
  
  ungroup()
```

```{r}

VPD_df <- climate_summary %>%
  pivot_longer(
    cols = c(VPD_mean, VPD_max, VPD_min),
    names_to = "stat",
    values_to = "VPD"
  )

Tair_df <- climate_summary %>%
  pivot_longer(
    cols = c(Tair_mean, Tair_max, Tair_min),
    names_to = "stat",
    values_to = "Tair"
  )

Precip_df <- climate_summary %>%
  pivot_longer(
    cols = c(Precip_mean, Precip_max, Precip_min),
    names_to = "stat",
    values_to = "Precip"
  )

ggplot(VPD_df, aes(x = year, y = VPD, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "VPD", x = "Year", color = "Statistic") +
  theme(legend.position = "none")

ggplot(Tair_df, aes(x = year, y = Tair, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "Air temp °C", x = "Year", color = "Statistic") +
  theme(legend.position = "none")

ggplot(Precip_df, aes(x = year, y = Precip, color = stat)) +
  geom_line() +
  geom_point() +
  facet_wrap(~site) +
  theme_minimal() +
  labs(y = "Precipitation (mm)", x = "Year", color = "Statistic") +
  theme(legend.position = "none")



```
Identifying a drought event:

Indicators of drought
- High VPD
- High Tair
- Low Precip
- Low GPP



