---
title: "long_term_drought"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Long term sites (How and Tum) - both 13 years
- Are Droughts and extreme VPD events becoming more unpredictable?


1. First going to try years - next time I will do dry season/summers

```{r}
library(tidyverse)
library(mgcv)
library(lubridate)

long_term_sites <- c("How", "Tum")

#Define extreme years per site
#drought event - bottom 20th percentile annual precip
#extreme VPD - top 20th percentile annual mean VPD

extreme_years <- map_dfr(long_term_sites, function(site) {
  df <- sites_daily[[site]] %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarise(
      precip_annual = sum(Precip, na.rm = TRUE), 
      vpd_annual = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      site = site, 
      drought_year = precip_annual < quantile(precip_annual, 0.2, na.rm = TRUE),
      extreme_vpd = vpd_annual > quantile(vpd_annual, 0.8, na.rm = TRUE)
    )
})

print(extreme_years)

```

Fit GAM per year per site and extract R-squared 

```{r}

annual_r2 <- map_dfr(long_term_sites, function(site) {
  df <- sites_daily[[site]] %>%
    mutate(year = year(date)) %>%
    drop_na(GPP, Tair, SW_rad, Precip, VPD)
  
  map_dfr(unique(df$year), function(yr) {
    df_yr <- df %>% filter(year == yr)
    
  if (nrow(df_yr) <30) return(NULL)
      
      r2 <- tryCatch({
        mod <- gam(GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD),
                   data = df_yr, method = "REML")
        preds <- predict(mod)
        cor(df_yr$GPP, preds, use = "complete.obs")^2
      }, error = function(e) NA_real_)
    
    tibble(site = site, year = yr, r2 = r2)
  })
  }) %>% 
  left_join(extreme_years, by = c("site", "year"))


```

Is R-squared declining over time?

Spearman correlation between year and R-squared per site

```{r}

annual_r2 %>%
  group_by(site) %>%
  summarise(
    trend_rho = cor.test(year, r2, method = "spearman")$estimate,
    trend_p   = cor.test(year, r2, method = "spearman")$p.value,
    .groups = "drop"
  )

```
```{r}
# Step 4: Is R² lower during drought / extreme VPD years?
# per site separately since n is small

map(long_term_sites, function(site) {
  df_site <- annual_r2 %>% filter(site == !!site)
  
  cat("\n---", site, "---\n")
  cat("Drought years R²:", round(mean(df_site$r2[df_site$drought_year], na.rm = TRUE), 3),
      "vs normal:", round(mean(df_site$r2[!df_site$drought_year], na.rm = TRUE), 3), "\n")
  cat("Extreme VPD R²:", round(mean(df_site$r2[df_site$extreme_vpd], na.rm = TRUE), 3),
      "vs normal:", round(mean(df_site$r2[!df_site$extreme_vpd], na.rm = TRUE), 3), "\n")
  
  # Wilcoxon only if enough years in each group
  if (sum(df_site$drought_year, na.rm = TRUE) >= 2) {
    w <- wilcox.test(r2 ~ drought_year, data = df_site)
    cat("Drought Wilcoxon p =", round(w$p.value, 3), "\n")
  }
  if (sum(df_site$extreme_vpd, na.rm = TRUE) >= 2) {
    w <- wilcox.test(r2 ~ extreme_vpd, data = df_site)
    cat("Extreme VPD Wilcoxon p =", round(w$p.value, 3), "\n")
  }
})
```

```{r}
# Step 5: Visualise — R² over time with extreme years highlighted
annual_r2 %>%
  mutate(event = case_when(
    drought_year & extreme_vpd ~ "Drought + Extreme VPD",
    drought_year               ~ "Drought",
    extreme_vpd                ~ "Extreme VPD",
    TRUE                       ~ "Normal"
  )) %>%
  ggplot(aes(x = year, y = r2, colour = event, size = event)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", colour = "grey40",
              linetype = "dashed", se = TRUE, linewidth = 0.8) +
  scale_colour_manual(values = c(
    "Normal"               = "grey60",
    "Drought"              = "steelblue",
    "Extreme VPD"          = "orange",
    "Drought + Extreme VPD" = "firebrick"
  )) +
  scale_size_manual(values = c(
    "Normal" = 2, "Drought" = 3, "Extreme VPD" = 3, "Drought + Extreme VPD" = 4
  )) +
  facet_wrap(~ site, ncol = 1) +
  labs(x = "Year", y = expression(R^2),
       colour = "Year type", size = "Year type",
       title = "Is model predictability declining over time?",
       subtitle = "Dashed line = linear trend; coloured points = extreme years") +
  theme_classic() +
  theme(legend.position = "bottom")
```
How = Drought years actually show slightly higher R-squared than normal years. Extreme VPD years show significantly lower R-squared (p=0.022). At How it is not water deficit but actually VPD that causes the model to struggle. 

Tum = extreme VPD years show higher R-squared though not significantly. Drought years show slightly lower R-squared but not significantly. 

2. Seasonal extremes 

Need to define which sites have wet/dry seasons and temperate summer/winter climates 

