---
title: "Untitled"
output: html_document
date: "2026-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Site analysis:
- Which site has the most predictable GPP?
- What variables are the most important drivers of GPP across bioclimatic gradients? (structural, physiology, climate)
- How well do models generalize to new sites?

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

Sites:
- Howard Springs
- Adelaide River
- Daly Pasture
- Daly savanna
- Dry river
- Sturt plains
- Alice springs


```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

```

Convert -9999 to NA
```{r}
flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_HH <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_HH <- Ade_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_HH <- Das_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_HH <- Dap_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_HH <- Dry_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_HH <- Stp_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_HH <- Asm_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


Filter missing data before daily conversion

```{r}
filter_low_coverage_days <- function(df, min_frac = 0.9, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```

```{r}
How_HH <- filter_low_coverage_days(How_HH)
Ade_HH <- filter_low_coverage_days(Ade_HH)
Das_HH <- filter_low_coverage_days(Das_HH)
Dap_HH <- filter_low_coverage_days(Dap_HH)
Dry_HH <- filter_low_coverage_days(Dry_HH)
Stp_HH <- filter_low_coverage_days(Stp_HH)
Asm_HH <- filter_low_coverage_days(Asm_HH)
```

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

```{r}
How_daily <- convert_to_daily(How_HH)
Ade_daily <- convert_to_daily(Ade_HH)
Das_daily <- convert_to_daily(Das_HH)
Dap_daily <- convert_to_daily(Dap_HH)
Dry_daily <- convert_to_daily(Dry_HH)
Stp_daily <- convert_to_daily(Stp_HH)
Asm_daily <- convert_to_daily(Asm_HH)
```

Convert NaN back to NA

```{r}
#Convert NaN ('Not a number') back to NA

How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

```


```{r}
plot(How_daily$VPD, How_daily$GPP)
```
Look at annual precipitation:

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}
```

```{r}
How_annual <- convert_to_annual(How_daily)
Asm_annual <- convert_to_annual(Asm_daily)
```

```{r}
summary(How_annual$Precip)
summary(Asm_annual$Precip)
```
Which sites are the most/least predictable?
- Climate variables - Temp, VPD, Precip, SW radiation, water stress component, lagged climate variables
- Physiology variables - Leaf chlorophyll content, vcmax

Along a precipitation gradient, how well can atmospheric drivers and physiology predict GPP?

Adding in water stress component for sites:


```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

```

```{r}
Asm_daily$pet <- priestley_taylor_pet(Asm_daily$Rn, Asm_daily$Tair, Asm_daily$Pa)
Asm_daily$cwd <- Asm_daily$pet - Asm_daily$Qle

How_daily$pet <- priestley_taylor_pet(How_daily$Rn, How_daily$Tair, How_daily$Pa)
How_daily$cwd <- How_daily$pet - How_daily$Qle

Das_daily$pet <- priestley_taylor_pet(Das_daily$Rn, Das_daily$Tair, Das_daily$Pa)
Das_daily$cwd <- Das_daily$pet - Das_daily$Qle

Dap_daily$pet <- priestley_taylor_pet(Dap_daily$Rn, Dap_daily$Tair, Dap_daily$Pa)
Dap_daily$cwd <- Dap_daily$pet - Dap_daily$Qle

Dry_daily$pet <- priestley_taylor_pet(Dry_daily$Rn, Dry_daily$Tair, Dry_daily$Pa)
Dry_daily$cwd <- Dry_daily$pet - Dry_daily$Qle

Ade_daily$pet <- priestley_taylor_pet(Ade_daily$Rn, Ade_daily$Tair, Ade_daily$Pa)
Ade_daily$cwd <- Ade_daily$pet - Ade_daily$Qle

Stp_daily$pet <- priestley_taylor_pet(Stp_daily$Rn, Stp_daily$Tair, Stp_daily$Pa)
Stp_daily$cwd <- Stp_daily$pet - Stp_daily$Qle
```

Load in vcmax and LLC:

Add in doy and year again:

```{r}
How_daily <- How_daily %>%
  mutate(
    year = lubridate::year(date),
    doy  = lubridate::yday(date)
  )

Dap_daily <- Dap_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Das_daily <- Das_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
    )

Dry_daily <- Dry_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Stp_daily <- Stp_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Asm_daily <- Asm_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )
```

Add in vcmax and LCC:
```{r}
# -----------------------------
# Libraries
# -----------------------------
library(terra)
library(ncdf4)
library(dplyr)

# -----------------------------
# File paths
# -----------------------------
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# -----------------------------
# Helper: Convert lat/lon to NetCDF indices
# -----------------------------
latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180. / nlines  
  line <- round((90. - lat) / lat_res)
  
  lon_res <- 360. / nsamples
  sample <- round((lon + 180.) / lon_res)
  
  # keep indices in bounds
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  list(line = line, sample = sample)
}

# -----------------------------
# Helper: Extract Vcmax from raster
# -----------------------------
get_vcmax <- function(lat, lon, raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(raster, coords)
  
  if (is.null(vcmax) || all(is.na(vcmax))) {
    stop("Vcmax extraction failed for coordinates: ", lat, ", ", lon)
  }
  
  as.numeric(vcmax[1,2])
}

# -----------------------------
# Helper: Extract daily LCC from NetCDF
# -----------------------------
get_daily_LCC <- function(nc_file, lat, lon) {
  nc <- nc_open(nc_file)
  
  coords <- latlon_to_line_sample(lat, lon)
  
  pixel_time_series <- ncvar_get(
    nc, "chl_all",
    start = c(coords$line, coords$sample, 1, 1),
    count = c(1, 1, -1, -1)
  )
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel_time_series[pixel_time_series == fillvalue] <- NA
  nc_close(nc)
  
  LCC <- pixel_time_series / 100   # convert units
  weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
  
  weeks <- 1:length(weekly_climatology)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  
  daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y
  return(daily_climatology)
}

# -----------------------------
# Main function: Attach Vcmax + LCC
# -----------------------------
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc_file) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  message("Vcmax extracted: ", round(vcmax_val, 2))
  
  daily_LCC <- get_daily_LCC(lcc_nc_file, lat, lon)
  
  # Repeat daily LCC to match df rows
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df <- df %>%
    mutate(
      vcmax = vcmax_val,
      LCC_daily = daily_LCC_full
    )
  
  return(df)
}


# -----------------------------
# Site metadata table
# -----------------------------
sites <- tibble(
  name = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490),
  df   = list(How_daily, Ade_daily, Das_daily, Dap_daily, Dry_daily, Stp_daily, Asm_daily)
)

# -----------------------------
# Apply function to all sites
# -----------------------------
sites <- sites %>%
  rowwise() %>%
  mutate(
    df_with_data = list(attach_site_data(df, lat, lon, vcmax_raster, lcc_nc))
  )

# -----------------------------
# Extract updated dataframes
# -----------------------------
How_daily <- sites$df_with_data[[1]]
Ade_daily <- sites$df_with_data[[2]]
Das_daily <- sites$df_with_data[[3]]
Dap_daily <- sites$df_with_data[[4]]
Dry_daily <- sites$df_with_data[[5]]
Stp_daily <- sites$df_with_data[[6]]
Asm_daily <- sites$df_with_data[[7]]

```

Now, let's fit some models yaaaaaaay

Filter by growing season:
```{r}

filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}

#Apply to dataset

How_gs <- filter_gs(How_daily)
Ade_gs <- filter_gs(Ade_daily)
Das_gs <- filter_gs(Das_daily)
Dap_gs <- filter_gs(Dap_daily)
Dry_gs <- filter_gs(Dry_daily)
Stp_gs <- filter_gs(Stp_daily)
Asm_gs <- filter_gs(Asm_daily)

```

Drop NA values

```{r}
How_gs <- How_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Ade_gs <- Ade_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Das_gs <- Das_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Dap_gs <- Dap_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Dry_gs <- Dry_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Stp_gs <- Stp_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Asm_gs <- Asm_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

```


Diagnostic plots:

```{r}
library(ggplot2)
library(GGally)
library(dplyr)

vars <- c("GPP", "VPD", "Tair", "Precip", "Pa", "Qle", "cwd", "pet", "vcmax", "LCC_daily", "ET", "Rn")

diagnostic_plots <- function(df, site_name) {
  
  df <- df %>% select(all_of(vars))
  
  # 1. Histograms for each variable
  hist_plots <- lapply(names(df), function(v) {
    ggplot(df, aes_string(x = v)) +
      geom_histogram(bins = 30, fill = "steelblue", color = "black") +
      theme_minimal() +
      labs(title = paste(site_name, "-", v))
  })
  
  # Print histograms (one by one)
  for (p in hist_plots) print(p)
  
  # 2. Scatterplots of GPP vs predictors
  scatter_plots <- lapply(names(df)[-1], function(v) {
    ggplot(df, aes_string(x = v, y = "GPP")) +
      geom_point(alpha = 0.5) +
      geom_smooth(method = "loess", color = "red") +
      theme_minimal() +
      labs(title = paste(site_name, "GPP vs", v))
  })
  
  for (p in scatter_plots) print(p)
  
  # 3. Correlation matrix
  cor_mat <- cor(df, use = "complete.obs")
  print(site_name)
  print(round(cor_mat, 2))
  
  return(list(histograms = hist_plots, scatter = scatter_plots, correlation = cor_mat))
}

How_diag <- diagnostic_plots(How_gs, "How")
Ade_diag <- diagnostic_plots(Ade_gs, "Ade")
Das_diag <- diagnostic_plots(Das_gs, "Das")
Dap_diag <- diagnostic_plots(Dap_gs, "Dap")
Dry_diag <- diagnostic_plots(Dry_gs, "Dry")
Stp_diag <- diagnostic_plots(Stp_gs, "Stp")
Asm_diag <- diagnostic_plots(Asm_gs, "Asm")




```
VPD ~ GPP plot

```{r}
library(dplyr)
library(ggplot2)

# Add a 'site' column to each dataframe
How_gs  <- How_gs  %>% mutate(site = "How")
Ade_gs  <- Ade_gs  %>% mutate(site = "Ade")
Das_gs  <- Das_gs  %>% mutate(site = "Das")
Dap_gs  <- Dap_gs  %>% mutate(site = "Dap")
Dry_gs  <- Dry_gs  %>% mutate(site = "Dry")
Stp_gs  <- Stp_gs  %>% mutate(site = "Stp")
Asm_gs  <- Asm_gs  %>% mutate(site = "Asm")

# Combine all sites
all_sites <- bind_rows(How_gs, Ade_gs, Das_gs, Dap_gs, Dry_gs, Stp_gs, Asm_gs)

# Set wet → dry order
all_sites$site <- factor(all_sites$site,
                         levels = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"))

```

```{r}

library(RColorBrewer)
# Define 7 distinct colors for the sites
site_colors <- brewer.pal(n = 7, name = "Set1")  # Set1 is bright and categorical


ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across FluxNet sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )

ggplot(all_sites, aes(x = Tair, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to air temperature across FluxNet sites",
    x = "Tair (°C)",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )

ggplot(all_sites, aes(x = Precip, y = GPP, color = site)) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE, size = 1.2) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to Precipitation across FluxNet sites",
    x = "Precipitation (mm/day)",
    y = "GPP (g C m-2 d-1)"
  )


ggplot(all_sites, aes(x = pet, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to pet across FluxNet sites",
    x = "pet",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )
```
Interestingly, Das and Dap respond differently to VPD. Das is a savanna ecosystem and Dap is a grasslands/wetlands ecosystem. 

As VPD increases, variability increases, likely because there are fewer data recorded when VPD is extremely high. 

Let's visualise the sites on a map:

```{r}
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
```


```{r}
library(dplyr)

# Site coordinates + name + rough "wetness" ordering
sites <- tibble(
  site = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490),
  wetness_order = 1:7  # 1 = wettest (How), 7 = driest (Asm)
)

library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

# Get Australia map
aus <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

ggplot() +
  geom_sf(data = aus, fill = "antiquewhite", color = "black") +
  geom_point(data = sites, aes(x = lon, y = lat, color = factor(wetness_order)), size = 4) +
  geom_text(data = sites, aes(x = lon, y = lat, label = site), nudge_y = 0.3, size = 4) +
  scale_color_viridis_d(option = "C", direction = -1, name = "Wet → Dry") +
  coord_sf(xlim = c(130, 135), ylim = c(-23, -12), expand = FALSE) +  # zoom to your region
  theme_minimal() +
  labs(title = "FluxNet Sites in Northern Australia",
       x = "Longitude",
       y = "Latitude")


```
Hierarchical GAM framework:
- Climate only models (Tair, VPD, precip, cwd, SW_rad, + lags)
- Physiology only models (LCC)
- Full model (climate + physiology)

```{r}

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# Split data in test and training 
idx <- createDataPartition(How_gs$GPP, p=0.8, list=FALSE)
How_train <- How_gs[idx, ]
How_test <- How_gs[-idx, ]

idx <- createDataPartition(Ade_gs$GPP, p=0.8, list=FALSE)
Ade_train <- Ade_gs[idx, ]
Ade_test <- Ade_gs[-idx, ]

idx <- createDataPartition(Das_gs$GPP, p=0.8, list=FALSE)
Das_train <- Das_gs[idx, ]
Das_test <- Das_gs[-idx, ]

idx <- createDataPartition(Dap_gs$GPP, p=0.8, list=FALSE)
Dap_train <- Dap_gs[idx, ]
Dap_test <- Dap_gs[-idx, ]

idx <- createDataPartition(Dry_gs$GPP, p=0.8, list=FALSE)
Dry_train <- Dry_gs[idx, ]
Dry_test <- Dry_gs[-idx, ]

idx <- createDataPartition(Stp_gs$GPP, p=0.8, list=FALSE)
Stp_train <- Stp_gs[idx, ]
Stp_test <- Stp_gs[-idx, ]

idx <- createDataPartition(Asm_gs$GPP, p=0.8, list=FALSE)
Asm_train <- Asm_gs[idx, ]
Asm_test <- Asm_gs[-idx, ]


# Drop any rows with NA values in test data
How_test <- How_test %>% drop_na()
Ade_test <- Ade_test %>% drop_na()
Das_test <- Das_test %>% drop_na()
Dap_test <- Dap_test %>% drop_na()
Dry_test <- Dry_test %>% drop_na()
Stp_test <- Stp_test %>% drop_na()
Asm_test <- Asm_test %>% drop_na()

# Create a 7-day moving cumulative sum for CWD
How_train$cumulative_cwd_7day <- rollapply(How_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
How_test$cumulative_cwd_7day <-  rollapply(How_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)


Ade_train$cumulative_cwd_7day <- rollapply(Ade_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Ade_test$cumulative_cwd_7day <-  rollapply(Ade_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Das_train$cumulative_cwd_7day <- rollapply(Das_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Das_test$cumulative_cwd_7day <-  rollapply(Das_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dap_train$cumulative_cwd_7day <- rollapply(Dap_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dap_test$cumulative_cwd_7day <-  rollapply(Dap_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dry_train$cumulative_cwd_7day <- rollapply(Dry_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dry_test$cumulative_cwd_7day <-  rollapply(Dry_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Stp_train$cumulative_cwd_7day <- rollapply(Stp_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Stp_test$cumulative_cwd_7day <-  rollapply(Stp_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Asm_train$cumulative_cwd_7day <- rollapply(Asm_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Asm_test$cumulative_cwd_7day <-  rollapply(Asm_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

# Drop any rows with NA values after the rolling sum calculations
How_train <- How_train%>% drop_na()
How_test <- How_test %>% drop_na()

Ade_train <- Ade_train%>% drop_na()
Ade_test <- Ade_test %>% drop_na()

Das_train <- Das_train%>% drop_na()
Das_test <- Das_test %>% drop_na()

Dap_train <- Dap_train%>% drop_na()
Dap_test <- Dap_test %>% drop_na()

Dry_train <- Dry_train%>% drop_na()
Dry_test <- Dry_test %>% drop_na()

Stp_train <- Stp_train%>% drop_na()
Stp_test <- Stp_test %>% drop_na()

Asm_train <- Asm_train%>% drop_na()
Asm_test <- Asm_test %>% drop_na()

# Lag the variables to add time-lag effects
How_train <- How_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

How_test <- How_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()



Ade_train <- Ade_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Ade_test <- Ade_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Das_train <- Das_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Das_test <- Das_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dap_train <- Dap_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dap_test <- Dap_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dry_train <- Dry_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dry_test <- Dry_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Stp_train <- Stp_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Stp_test <- Stp_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Asm_train <- Asm_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Asm_test <- Asm_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

```

1. Full climate model for each site:

```{r}
climate_How <- gam(GPP ~ s(Tair) + s(VPD) + s(Precip) +s(SW_rad) + s(cumulative_cwd_7day) + 
                   s(Tair_lag3) + s(VPD_lag3), data=How_train, 
                 method = "REML")

summary(climate_How)


predictions_gam <- predict(climate_How, newdata=How_test)
results_gam <- data.frame(Observed=How_test$GPP, Predicted=predictions_gam)


# Dumps some stats
rmse_gam <- sqrt(mean((results_gam$Observed - results_gam$Predicted)^2))
rsq_gam <- cor(results_gam$Observed, results_gam$Predicted)^2
mean_observed_gam <- mean(results_gam$Observed)
nse_gam <- (1 - sum((results_gam$Observed - results_gam$Predicted)^2) / 
              sum((results_gam$Observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")

```
Function to fit GAM (one site)
```{r}
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) +s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

```

Run function for each site, combine results. Saves R-squared and RMSE 

```{r}
climate_metrics <- bind_rows(
  fit_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  fit_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  fit_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  fit_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  fit_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  fit_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  fit_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)
```

```{r}
# Define wet → dry order
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

climate_metrics$site <- factor(climate_metrics$site, levels = wet_dry_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()

```

2. Physiology only model (LCC)

```{r}
lcc_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

```

```{r}
lcc_metrics <- bind_rows(
  lcc_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  lcc_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  lcc_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  lcc_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  lcc_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  lcc_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  lcc_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)
```

```{r}
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

lcc_metrics$site <- factor(lcc_metrics$site, levels = wet_dry_order)

ggplot(lcc_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()

```
3. Full climate + LCC model

```{r}
full_model_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) +s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}
```

```{r}
full_metrics <- bind_rows(
  full_model_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  full_model_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  full_model_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  full_model_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  full_model_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  full_model_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  full_model_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)

```

```{r}
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

full_metrics$site <- factor(full_metrics$site, levels = wet_dry_order)

ggplot(full_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()
```
Compare models:

```{r}

library(dplyr)

gam_comparison <- climate_metrics %>%
  select(site, R2) %>%
  rename(R2_climate = R2) %>%
  
  inner_join(
    lcc_metrics %>%
      select(site, R2) %>%
      rename(R2_LCC = R2),
    by = "site"
  ) %>%
  
  inner_join(
    full_metrics %>%
      select(site, R2) %>%
      rename(R2_full = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_LCC = R2_full - R2_climate
  )


wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison

```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_LCC)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet → Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of LCC across OzFlux sites"
  )

```
Inclusion of leaf chlorophyll content increased model explanatory power for Stp, Dry, Dap, Das and How. However, it decreased explanatory power in Ade and Asm. 

Adelaide river is a wet site, chlorophyll content may likely be constant. Productivity at Ade is climate-driven rather than chlorophyll-limited.
Alice springs is a dry site (lowest precipitation site), LCC may remain low most of the time, GPP is limited by water availability than LCC. 

But, Howard springs is wetter than Adelaide river? Both are woody savanna ecosystems. 

```{r}
summary(How_daily$LCC_daily)
```

```{r}
summary(Ade_daily$LCC_daily)
```

Both AU-How and AU-Ade have very similar LCC distributions and similar precipitation. Howard springs is slightly wetter. Both are woody savannas. 
Differences in canopy structure could determine whether LCC still captures meaningful variation.
Dominant species may differ
LAI - if Adelaide river had more dense canopy (high LAI), this could be why. The canopy is more dense, some leaves are blocked by light, so LCC may not matter. In Howard springs, there may be a lower LAI - each leaf gets more light, so differences in LCC matter for productivity. So by adding LCC, the model improves because it explains variation. 

Inclusion of LCC improved explanatory power of the models for the intermediate precipitation gradients. 
Wet sites - LCC might be saturated, at dry sites LCC might be strongly constrained by stress. 
Intermediate sites - large variation in LCC. 

**Now I want to compare how VPD improves predictability**

1. Climate model (no VPD)

```{r}

no_vpd_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) +s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

c_metrics <- bind_rows(
  no_vpd_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  no_vpd_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  no_vpd_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  no_vpd_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  no_vpd_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  no_vpd_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  no_vpd_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)

```

```{r}
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

c_metrics$site <- factor(c_metrics$site, levels = wet_dry_order)

ggplot(c_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()
```
2. VPD only

```{r}
vpd_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(VPD) + s(VPD_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

vpd_metrics <- bind_rows(
  vpd_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  vpd_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  vpd_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  vpd_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  vpd_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  vpd_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  vpd_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)

```

```{r}
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

vpd_metrics$site <- factor(vpd_metrics$site, levels = wet_dry_order)

ggplot(vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()
```

3. Full model

```{r}
full_model_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) +s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

full_metrics <- bind_rows(
  full_model_gam_and_metrics(Asm_train, Asm_test) %>% mutate(site = "Asm"),
  full_model_gam_and_metrics(Stp_train, Stp_test) %>% mutate(site = "Stp"),
  full_model_gam_and_metrics(Dap_train, Dap_test) %>% mutate(site = "Dap"),
  full_model_gam_and_metrics(Das_train, Das_test) %>% mutate(site = "Das"),
  full_model_gam_and_metrics(Dry_train, Dry_test) %>% mutate(site = "Dry"),
  full_model_gam_and_metrics(How_train, How_test) %>% mutate(site = "How"),
  full_model_gam_and_metrics(Ade_train, Ade_test) %>% mutate(site = "Ade")
  
)
```

```{r}
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")

full_metrics$site <- factor(full_metrics$site, levels = wet_dry_order)

ggplot(full_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Wet → Dry)"
  ) +
  theme_minimal()
```

Compare the models:

```{r}
# Add a column to identify model type
no_vpd_metrics <- c_metrics %>% mutate(Model = "No VPD")
vpd_metrics    <- vpd_metrics %>% mutate(Model = "VPD Only")
full_metrics   <- full_metrics %>% mutate(Model = "Full Model")

# Combine into one data frame
all_metrics <- bind_rows(no_vpd_metrics, vpd_metrics, full_metrics)

# Ensure sites are ordered from wet to dry
wet_dry_order <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm")
all_metrics$site <- factor(all_metrics$site, levels = wet_dry_order)

# View combined table
all_metrics

```
```{r}
library(ggplot2)

ggplot(all_metrics, aes(x = site, y = R2, fill = Model)) +
  geom_col(position = "dodge") +   # side-by-side bars
  coord_flip() +                   # horizontal bars
  scale_fill_manual(values = c("No VPD" = "violet", "VPD Only" = "skyblue", "Full Model" = "forestgreen")) +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Comparison of GPP Predictability Across OzFlux Sites",
    fill = "Model"
  ) +
  theme_minimal(base_size = 14)

```
```{r}
# Merge R2 from no VPD and full model
gam_comparison <- no_vpd_metrics %>%
  select(site, R2_no_VPD = R2) %>%
  left_join(full_metrics %>% select(site, R2_with_VPD = R2), by = "site") %>%
  mutate(delta_R2_VPD = R2_with_VPD - R2_no_VPD)

ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet → Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )


```
Now, I want to fit a random forest to determine which variables are most important across sites:

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

standardise <- function(df, vars_to_scale) {
  df_scaled <- df
  df_scaled[vars_to_scale] <- scale(df[vars_to_scale])
  
  return(df_scaled)
}

###
# Random forest
###

# Rescale the training and test data predictors
predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day", "Precip", "LCC_daily" )  
How_train <- standardise(How_train, predictors)
How_test <- standardise(How_test, predictors)

Ade_train <- standardise(Ade_train, predictors)
Ade_test <- standardise(Ade_test, predictors)

Das_train <- standardise(Das_train, predictors)
Das_test <- standardise(Das_test, predictors)

Dap_train <- standardise(Dap_train, predictors)
Dap_test <- standardise(Dap_test, predictors)

Dry_train <- standardise(Dry_train, predictors)
Dry_test <- standardise(Dry_test, predictors)

Stp_train <- standardise(Stp_train, predictors)
Stp_test <- standardise(Stp_test, predictors)

Asm_train <- standardise(Asm_train, predictors)
Asm_test <- standardise(Asm_test, predictors)



# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=How_train, ntree=500, 
                              mtry=2)
How_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=How_train, 
                      ntree=500, mtry=3)
print(How_model)

# Predict on the testing data
predictions <- predict(How_model, How_test)
results <- data.frame(Observed=How_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(How_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(How_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(How_model$mse[length(How_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Ade_train, ntree=500, 
                              mtry=2)
Ade_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Ade_train, 
                      ntree=500, mtry=3)
print(Ade_model)

# Predict on the testing data
predictions <- predict(Ade_model, Ade_test)
results <- data.frame(Observed=Ade_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Ade_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Ade_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Ade_model$mse[length(Ade_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Das_train, ntree=500, 
                              mtry=2)
Das_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Das_train, 
                      ntree=500, mtry=3)
print(Das_model)

# Predict on the testing data
predictions <- predict(Das_model, Das_test)
results <- data.frame(Observed=Das_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Das_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Das_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Das_model$mse[length(Das_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Dap_train, ntree=500, 
                              mtry=2)
Dap_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Dap_train, 
                      ntree=500, mtry=3)
print(Dap_model)

# Predict on the testing data
predictions <- predict(Dap_model, Dap_test)
results <- data.frame(Observed=Dap_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Dap_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Dap_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Dap_model$mse[length(Dap_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Dry_train, ntree=500, 
                              mtry=2)
Dry_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Dry_train, 
                      ntree=500, mtry=3)
print(Dry_model)

# Predict on the testing data
predictions <- predict(Dry_model, Dry_test)
results <- data.frame(Observed=Dry_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Dry_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Dry_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Dry_model$mse[length(Dry_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Stp_train, ntree=500, 
                              mtry=2)
Stp_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Stp_train, 
                      ntree=500, mtry=3)
print(Stp_model)

# Predict on the testing data
predictions <- predict(Stp_model, Stp_test)
results <- data.frame(Observed=Stp_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Stp_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Stp_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Stp_model$mse[length(Stp_model$mse)])  
```
```{r}
# Train the random forest model
simpler_model <- randomForest(GPP ~ Tair + SW_rad, data=Asm_train, ntree=500, 
                              mtry=2)
Asm_model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + Precip, 
                      data=Asm_train, 
                      ntree=500, mtry=3)
print(Asm_model)

# Predict on the testing data
predictions <- predict(Asm_model, Asm_test)
results <- data.frame(Observed=Asm_test$GPP, Predicted=predictions)

# Dumps some stats
rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
rsq <- cor(results$Observed, results$Predicted)^2
mean_observed <- mean(results$Observed)
nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
          sum((results$Observed - mean_observed)^2))

cat("RMSE:", rmse, "\n")
cat("R-squared:", rsq, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse, "\n")

ggplot(results, aes(x=Observed, y=Predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color="red") +
  theme_minimal() +
  labs(x="Observed GPP", y="Predicted GPP")


importance_values <- importance(Asm_model)
print(importance_values)

# the higher the score, the more important the variable is in predicting GPP
varImpPlot(Asm_model, main = "")

# the model with the lower Out-of-Bag (OOB) MSE is better.
print(simpler_model$mse[length(simpler_model$mse)])  
print(Asm_model$mse[length(Asm_model$mse)])  
```

