---
title: "Untitled"
output: html_document
date: "2026-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Site analysis:
- Which site has the most predictable GPP?
- What variables are the most important drivers of GPP across bioclimatic gradients? (structural, physiology, climate)
- How well do models generalize to new sites?

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

Sites:
- Howard Springs
- Adelaide River
- Daly Pasture
- Daly savanna
- Dry river
- Sturt plains
- Alice springs


```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

```

Convert -9999 to NA
```{r}
flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_HH <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_HH <- Ade_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_HH <- Das_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_HH <- Dap_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_HH <- Dry_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_HH <- Stp_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_HH <- Asm_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


Filter missing data before daily conversion

```{r}
filter_low_coverage_days <- function(df, min_frac = 0.9, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```

```{r}
How_HH <- filter_low_coverage_days(How_HH)
Ade_HH <- filter_low_coverage_days(Ade_HH)
Das_HH <- filter_low_coverage_days(Das_HH)
Dap_HH <- filter_low_coverage_days(Dap_HH)
Dry_HH <- filter_low_coverage_days(Dry_HH)
Stp_HH <- filter_low_coverage_days(Stp_HH)
Asm_HH <- filter_low_coverage_days(Asm_HH)
```

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

```{r}
How_daily <- convert_to_daily(How_HH)
Ade_daily <- convert_to_daily(Ade_HH)
Das_daily <- convert_to_daily(Das_HH)
Dap_daily <- convert_to_daily(Dap_HH)
Dry_daily <- convert_to_daily(Dry_HH)
Stp_daily <- convert_to_daily(Stp_HH)
Asm_daily <- convert_to_daily(Asm_HH)
```

Convert NaN back to NA

```{r}
#Convert NaN ('Not a number') back to NA

How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

```


```{r}
plot(How_daily$VPD, How_daily$GPP)
```
Look at annual precipitation:

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}
```

```{r}
How_annual <- convert_to_annual(How_daily)
Asm_annual <- convert_to_annual(Asm_daily)
```

```{r}
summary(How_annual$Precip)
summary(Asm_annual$Precip)
```
Which sites are the most/least predictable?
- Climate variables - Temp, VPD, Precip, SW radiation, water stress component, lagged climate variables
- Physiology variables - Leaf chlorophyll content, vcmax

Along a precipitation gradient, how well can atmospheric drivers and physiology predict GPP?

Adding in water stress component for sites:


```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

```

```{r}
Asm_daily$pet <- priestley_taylor_pet(Asm_daily$Rn, Asm_daily$Tair, Asm_daily$Pa)
Asm_daily$cwd <- Asm_daily$pet - Asm_daily$Qle

How_daily$pet <- priestley_taylor_pet(How_daily$Rn, How_daily$Tair, How_daily$Pa)
How_daily$cwd <- How_daily$pet - How_daily$Qle

Das_daily$pet <- priestley_taylor_pet(Das_daily$Rn, Das_daily$Tair, Das_daily$Pa)
Das_daily$cwd <- Das_daily$pet - Das_daily$Qle

Dap_daily$pet <- priestley_taylor_pet(Dap_daily$Rn, Dap_daily$Tair, Dap_daily$Pa)
Dap_daily$cwd <- Dap_daily$pet - Dap_daily$Qle

Dry_daily$pet <- priestley_taylor_pet(Dry_daily$Rn, Dry_daily$Tair, Dry_daily$Pa)
Dry_daily$cwd <- Dry_daily$pet - Dry_daily$Qle

Ade_daily$pet <- priestley_taylor_pet(Ade_daily$Rn, Ade_daily$Tair, Ade_daily$Pa)
Ade_daily$cwd <- Ade_daily$pet - Ade_daily$Qle

Stp_daily$pet <- priestley_taylor_pet(Stp_daily$Rn, Stp_daily$Tair, Stp_daily$Pa)
Stp_daily$cwd <- Stp_daily$pet - Stp_daily$Qle
```

Load in vcmax and LLC:

Add in doy and year again:

```{r}
How_daily <- How_daily %>%
  mutate(
    year = lubridate::year(date),
    doy  = lubridate::yday(date)
  )

Dap_daily <- Dap_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Das_daily <- Das_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
    )

Dry_daily <- Dry_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Stp_daily <- Stp_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Asm_daily <- Asm_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )
```

Add in vcmax and LCC:
```{r}
# -----------------------------
# Libraries
# -----------------------------
library(terra)
library(ncdf4)
library(dplyr)

# -----------------------------
# File paths
# -----------------------------
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# -----------------------------
# Helper: Convert lat/lon to NetCDF indices
# -----------------------------
latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180. / nlines  
  line <- round((90. - lat) / lat_res)
  
  lon_res <- 360. / nsamples
  sample <- round((lon + 180.) / lon_res)
  
  # keep indices in bounds
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  list(line = line, sample = sample)
}

# -----------------------------
# Helper: Extract Vcmax from raster
# -----------------------------
get_vcmax <- function(lat, lon, raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(raster, coords)
  
  if (is.null(vcmax) || all(is.na(vcmax))) {
    stop("Vcmax extraction failed for coordinates: ", lat, ", ", lon)
  }
  
  as.numeric(vcmax[1,2])
}

# -----------------------------
# Helper: Extract daily LCC from NetCDF
# -----------------------------
get_daily_LCC <- function(nc_file, lat, lon) {
  nc <- nc_open(nc_file)
  
  coords <- latlon_to_line_sample(lat, lon)
  
  pixel_time_series <- ncvar_get(
    nc, "chl_all",
    start = c(coords$line, coords$sample, 1, 1),
    count = c(1, 1, -1, -1)
  )
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel_time_series[pixel_time_series == fillvalue] <- NA
  nc_close(nc)
  
  LCC <- pixel_time_series / 100   # convert units
  weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
  
  weeks <- 1:length(weekly_climatology)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  
  daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y
  return(daily_climatology)
}

# -----------------------------
# Main function: Attach Vcmax + LCC
# -----------------------------
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc_file) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  message("Vcmax extracted: ", round(vcmax_val, 2))
  
  daily_LCC <- get_daily_LCC(lcc_nc_file, lat, lon)
  
  # Repeat daily LCC to match df rows
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df <- df %>%
    mutate(
      vcmax = vcmax_val,
      LCC_daily = daily_LCC_full
    )
  
  return(df)
}


# -----------------------------
# Site metadata table
# -----------------------------
sites <- tibble(
  name = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490),
  df   = list(How_daily, Ade_daily, Das_daily, Dap_daily, Dry_daily, Stp_daily, Asm_daily)
)

# -----------------------------
# Apply function to all sites
# -----------------------------
sites <- sites %>%
  rowwise() %>%
  mutate(
    df_with_data = list(attach_site_data(df, lat, lon, vcmax_raster, lcc_nc))
  )

# -----------------------------
# Extract updated dataframes
# -----------------------------
How_daily <- sites$df_with_data[[1]]
Ade_daily <- sites$df_with_data[[2]]
Das_daily <- sites$df_with_data[[3]]
Dap_daily <- sites$df_with_data[[4]]
Dry_daily <- sites$df_with_data[[5]]
Stp_daily <- sites$df_with_data[[6]]
Asm_daily <- sites$df_with_data[[7]]

```

Now, let's fit some models yaaaaaaay

Filter by growing season:
```{r}

filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}

#Apply to dataset

How_gs <- filter_gs(How_daily)
Ade_gs <- filter_gs(Ade_daily)
Das_gs <- filter_gs(Das_daily)
Dap_gs <- filter_gs(Dap_daily)
Dry_gs <- filter_gs(Dry_daily)
Stp_gs <- filter_gs(Stp_daily)
Asm_gs <- filter_gs(Asm_daily)

```

Drop NA values

```{r}
How_gs <- How_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Ade_gs <- Ade_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Das_gs <- Das_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Dap_gs <- Dap_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Dry_gs <- Dry_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Stp_gs <- Stp_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

Asm_gs <- Asm_gs %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Qle, cwd, pet, vcmax, LCC_daily, ET, Rn)

```


Diagnostic plots:

```{r}
library(ggplot2)
library(GGally)
library(dplyr)

vars <- c("GPP", "VPD", "Tair", "Precip", "Pa", "Qle", "cwd", "pet", "vcmax", "LCC_daily", "ET", "Rn")

diagnostic_plots <- function(df, site_name) {
  
  df <- df %>% select(all_of(vars))
  
  # 1. Histograms for each variable
  hist_plots <- lapply(names(df), function(v) {
    ggplot(df, aes_string(x = v)) +
      geom_histogram(bins = 30, fill = "steelblue", color = "black") +
      theme_minimal() +
      labs(title = paste(site_name, "-", v))
  })
  
  # Print histograms (one by one)
  for (p in hist_plots) print(p)
  
  # 2. Scatterplots of GPP vs predictors
  scatter_plots <- lapply(names(df)[-1], function(v) {
    ggplot(df, aes_string(x = v, y = "GPP")) +
      geom_point(alpha = 0.5) +
      geom_smooth(method = "loess", color = "red") +
      theme_minimal() +
      labs(title = paste(site_name, "GPP vs", v))
  })
  
  for (p in scatter_plots) print(p)
  
  # 3. Correlation matrix
  cor_mat <- cor(df, use = "complete.obs")
  print(site_name)
  print(round(cor_mat, 2))
  
  return(list(histograms = hist_plots, scatter = scatter_plots, correlation = cor_mat))
}

How_diag <- diagnostic_plots(How_gs, "How")
Ade_diag <- diagnostic_plots(Ade_gs, "Ade")
Das_diag <- diagnostic_plots(Das_gs, "Das")
Dap_diag <- diagnostic_plots(Dap_gs, "Dap")
Dry_diag <- diagnostic_plots(Dry_gs, "Dry")
Stp_diag <- diagnostic_plots(Stp_gs, "Stp")
Asm_diag <- diagnostic_plots(Asm_gs, "Asm")




```
VPD ~ GPP plot

```{r}
library(dplyr)
library(ggplot2)

# Add a 'site' column to each dataframe
How_gs  <- How_gs  %>% mutate(site = "How")
Ade_gs  <- Ade_gs  %>% mutate(site = "Ade")
Das_gs  <- Das_gs  %>% mutate(site = "Das")
Dap_gs  <- Dap_gs  %>% mutate(site = "Dap")
Dry_gs  <- Dry_gs  %>% mutate(site = "Dry")
Stp_gs  <- Stp_gs  %>% mutate(site = "Stp")
Asm_gs  <- Asm_gs  %>% mutate(site = "Asm")

# Combine all sites
all_sites <- bind_rows(How_gs, Ade_gs, Das_gs, Dap_gs, Dry_gs, Stp_gs, Asm_gs)

# Set wet → dry order
all_sites$site <- factor(all_sites$site,
                         levels = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"))

```

```{r}

library(RColorBrewer)
# Define 7 distinct colors for the sites
site_colors <- brewer.pal(n = 7, name = "Set1")  # Set1 is bright and categorical


ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across FluxNet sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )

ggplot(all_sites, aes(x = Tair, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to air temperature across FluxNet sites",
    x = "Tair (°C)",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )

ggplot(all_sites, aes(x = Precip, y = GPP, color = site)) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE, size = 1.2) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to Precipitation across FluxNet sites",
    x = "Precipitation (mm/day)",
    y = "GPP (g C m-2 d-1)"
  )


ggplot(all_sites, aes(x = pet, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2, span = 0.3) +
  scale_color_manual(values = site_colors, name = "Site (Wet → Dry)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to pet across FluxNet sites",
    x = "pet",
    y = "GPP (g C m-2 d-1)"
  ) +
  theme(
    legend.position = "right"
  )
```
Interestingly, Das and Dap respond differently to VPD. Das is a savanna ecosystem and Dap is a grasslands/wetlands ecosystem. 

As VPD increases, variability increases, likely because there are fewer data recorded when VPD is extremely high. 

Let's visualise the sites on a map:

```{r}
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
```


```{r}
library(dplyr)

# Site coordinates + name + rough "wetness" ordering
sites <- tibble(
  site = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490),
  wetness_order = 1:7  # 1 = wettest (How), 7 = driest (Asm)
)

library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

# Get Australia map
aus <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

ggplot() +
  geom_sf(data = aus, fill = "antiquewhite", color = "black") +
  geom_point(data = sites, aes(x = lon, y = lat, color = factor(wetness_order)), size = 4) +
  geom_text(data = sites, aes(x = lon, y = lat, label = site), nudge_y = 0.3, size = 4) +
  scale_color_viridis_d(option = "C", direction = -1, name = "Wet → Dry") +
  coord_sf(xlim = c(130, 135), ylim = c(-23, -12), expand = FALSE) +  # zoom to your region
  theme_minimal() +
  labs(title = "FluxNet Sites in Northern Australia",
       x = "Longitude",
       y = "Latitude")


```
Hierarchical GAM framework:
- Climate only models (Tair, VPD, precip, cwd, SW_rad, + lags)
- Physiology only models (LCC)
- Full model (climate + physiology)

```{r}

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# Split data in test and training 
idx <- createDataPartition(How_gs$GPP, p=0.8, list=FALSE)
How_train <- How_gs[idx, ]
How_test <- How_gs[-idx, ]

idx <- createDataPartition(Ade_gs$GPP, p=0.8, list=FALSE)
Ade_train <- Ade_gs[idx, ]
Ade_test <- Ade_gs[-idx, ]

idx <- createDataPartition(Das_gs$GPP, p=0.8, list=FALSE)
Das_train <- Das_gs[idx, ]
Das_test <- Das_gs[-idx, ]

idx <- createDataPartition(Dap_gs$GPP, p=0.8, list=FALSE)
Dap_train <- Dap_gs[idx, ]
Dap_test <- Dap_gs[-idx, ]

idx <- createDataPartition(Dry_gs$GPP, p=0.8, list=FALSE)
Dry_train <- Dry_gs[idx, ]
Dry_test <- Dry_gs[-idx, ]

idx <- createDataPartition(Stp_gs$GPP, p=0.8, list=FALSE)
Stp_train <- Stp_gs[idx, ]
Stp_test <- Stp_gs[-idx, ]

idx <- createDataPartition(Asm_gs$GPP, p=0.8, list=FALSE)
Asm_train <- Asm_gs[idx, ]
Asm_test <- Asm_gs[-idx, ]


# Drop any rows with NA values in test data
How_test <- How_test %>% drop_na()
Ade_test <- Ade_test %>% drop_na()
Das_test <- Das_test %>% drop_na()
Dap_test <- Dap_test %>% drop_na()
Dry_test <- Dry_test %>% drop_na()
Stp_test <- Stp_test %>% drop_na()
Asm_test <- Asm_test %>% drop_na()

# Create a 7-day moving cumulative sum for CWD
How_train$cumulative_cwd_7day <- rollapply(How_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
How_test$cumulative_cwd_7day <-  rollapply(How_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)


Ade_train$cumulative_cwd_7day <- rollapply(Ade_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Ade_test$cumulative_cwd_7day <-  rollapply(Ade_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Das_train$cumulative_cwd_7day <- rollapply(Das_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Das_test$cumulative_cwd_7day <-  rollapply(Das_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dap_train$cumulative_cwd_7day <- rollapply(Dap_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dap_test$cumulative_cwd_7day <-  rollapply(Dap_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Dry_train$cumulative_cwd_7day <- rollapply(Dry_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Dry_test$cumulative_cwd_7day <-  rollapply(Dry_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Stp_train$cumulative_cwd_7day <- rollapply(Stp_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Stp_test$cumulative_cwd_7day <-  rollapply(Stp_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

Asm_train$cumulative_cwd_7day <- rollapply(Asm_train$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)
Asm_test$cumulative_cwd_7day <-  rollapply(Asm_test$cwd, width=7, FUN=sum, 
                                            align="right", fill=NA, na.rm=TRUE)

# Drop any rows with NA values after the rolling sum calculations
How_train <- How_train%>% drop_na()
How_test <- How_test %>% drop_na()

Ade_train <- Ade_train%>% drop_na()
Ade_test <- Ade_test %>% drop_na()

Das_train <- Das_train%>% drop_na()
Das_test <- Das_test %>% drop_na()

Dap_train <- Dap_train%>% drop_na()
Dap_test <- Dap_test %>% drop_na()

Dry_train <- Dry_train%>% drop_na()
Dry_test <- Dry_test %>% drop_na()

Stp_train <- Stp_train%>% drop_na()
Stp_test <- Stp_test %>% drop_na()

Asm_train <- Asm_train%>% drop_na()
Asm_test <- Asm_test %>% drop_na()

# Lag the variables to add time-lag effects
How_train <- How_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

How_test <- How_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()



Ade_train <- Ade_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Ade_test <- Ade_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Das_train <- Das_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Das_test <- Das_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dap_train <- Dap_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dap_test <- Dap_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dry_train <- Dry_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Dry_test <- Dry_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Stp_train <- Stp_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Stp_test <- Stp_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Asm_train <- Asm_train %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

Asm_test <- Asm_test %>%
  mutate(
    Tair_lag3 = lag(Tair, 3),
    VPD_lag3 = lag(VPD, 3)
  ) %>%
  drop_na()

```

1. Full climate model for each site:

```{r}
climate_How <- gam(GPP ~ s(Tair) + s(VPD) + s(Precip) +s(SW_rad) + s(cumulative_cwd_7day) + 
                   s(Tair_lag3) + s(VPD_lag3), data=How_train, 
                 method = "REML")

summary(climate_How)


predictions_gam <- predict(climate_How, newdata=How_test)
results_gam <- data.frame(Observed=How_test$GPP, Predicted=predictions_gam)


# Dumps some stats
rmse_gam <- sqrt(mean((results_gam$Observed - results_gam$Predicted)^2))
rsq_gam <- cor(results_gam$Observed, results_gam$Predicted)^2
mean_observed_gam <- mean(results_gam$Observed)
nse_gam <- (1 - sum((results_gam$Observed - results_gam$Predicted)^2) / 
              sum((results_gam$Observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")

```
