---
title: "02_site_analysis"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Which Australian flux site is the most/least predictable?
- R-squared from model
- Change in R-squared when VPD is added to GAM
- Can VPD explain residual variation?
- VPD curves for sites/PFT groups?

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```
```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cpr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cpr_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cum_FLUXNET2015_FULLSET_HH_2012-2014_2-4.csv")

Emr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Emr_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Fog_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Fog_FLUXNET2015_FULLSET_HH_2006-2008_1-4.csv")

Gin_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Gin_FLUXNET2015_FULLSET_HH_2011-2014_1-4.csv")

Gww_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-GWW_FLUXNET2015_FULLSET_HH_2013-2014_1-4.csv")

Lox_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Lox_FLUXNET2015_FULLSET_HH_2008-2009_1-4.csv")

Rdf_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-RDF_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Rig_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rig_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Rob_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rob_FLUXNET2015_FULLSET_HH_2014-2014_1-4.csv")

TTE_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-TTE_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Wac_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wac_FLUXNET2015_FULLSET_HH_2005-2008_1-4.csv")

Whr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Whr_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Wom_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wom_FLUXNET2015_FULLSET_HH_2010-2014_1-4.csv")

Ync_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ync_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")


```

```{r}
Tum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/hourly_2/FLX_AU-Tum_FLUXNET2015_FULLSET_HR_2001-2014_2-4.csv")
```

Remove -9999 values

```{r}
flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_HH <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_HH <- Ade_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_HH <- Das_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_HH <- Dap_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_HH <- Dry_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_HH <- Stp_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_HH <- Asm_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Cum_HH <- Cum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Emr_HH <- Emr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Fog_HH <- Fog_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gin_HH <- Gin_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gww_HH <- Gww_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Lox_HH <- Lox_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rdf_HH <- Rdf_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rig_HH <- Rig_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rob_HH <- Rob_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

TTE_HH <- TTE_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wac_HH <- Wac_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Whr_HH <- Whr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wom_HH <- Wom_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ync_HH <- Ync_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Tum_HH <- Tum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
Tum_HH <- Tum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
filter_low_coverage_days <- function(df, min_frac = 0.9, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```

```{r}
How_HH <- filter_low_coverage_days(How_HH)
Ade_HH <- filter_low_coverage_days(Ade_HH)
Das_HH <- filter_low_coverage_days(Das_HH)
Dap_HH <- filter_low_coverage_days(Dap_HH)
Dry_HH <- filter_low_coverage_days(Dry_HH)
Stp_HH <- filter_low_coverage_days(Stp_HH)
Asm_HH <- filter_low_coverage_days(Asm_HH)
Cum_HH <- filter_low_coverage_days(Cum_HH)
Emr_HH <- filter_low_coverage_days(Emr_HH)
Fog_HH <- filter_low_coverage_days(Fog_HH)
Gin_HH <- filter_low_coverage_days(Gin_HH)
Gww_HH <- filter_low_coverage_days(Gww_HH)
Lox_HH <- filter_low_coverage_days(Lox_HH)
Rdf_HH <- filter_low_coverage_days(Rdf_HH)
Rig_HH <- filter_low_coverage_days(Rig_HH)
Rob_HH <- filter_low_coverage_days(Rob_HH)
TTE_HH <- filter_low_coverage_days(TTE_HH)
Wac_HH <- filter_low_coverage_days(Wac_HH)
Wom_HH <- filter_low_coverage_days(Wom_HH)
Whr_HH <- filter_low_coverage_days(Whr_HH)
Ync_HH <- filter_low_coverage_days(Ync_HH)
Tum_HH <- filter_low_coverage_days(Tum_HH)
```

```{r}
Tum_HH <- filter_low_coverage_days(Tum_HH)
```


Convert to daily:

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

Convert hourly to daily (AU_Tum)
```{r}
library(tidyverse)
library(lubridate)

convert_hourly_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HOUR <- 3600.
  WM2_TO_KG_M2_S <- 1.0 / (2.45 * 1E6)
  PA_2_KPA <- 0.001
  HPA_2_KPA <- 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HOUR,
      VPD = VPD * HPA_2_KPA
    )

  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df)
}

```

```{r}
How_daily <- convert_to_daily(How_HH)
Ade_daily <- convert_to_daily(Ade_HH)
Das_daily <- convert_to_daily(Das_HH)
Dap_daily <- convert_to_daily(Dap_HH)
Dry_daily <- convert_to_daily(Dry_HH)
Stp_daily <- convert_to_daily(Stp_HH)
Asm_daily <- convert_to_daily(Asm_HH)
Cum_daily <- convert_to_daily(Cum_HH)
Emr_daily <- convert_to_daily(Emr_HH)
Fog_daily <- convert_to_daily(Fog_HH)
Gin_daily <- convert_to_daily(Gin_HH)
Gww_daily <- convert_to_daily(Gww_HH)
Lox_daily <- convert_to_daily(Lox_HH)
Rdf_daily <- convert_to_daily(Rdf_HH)
Rig_daily <- convert_to_daily(Rig_HH)
Rob_daily <- convert_to_daily(Rob_HH)
TTE_daily <- convert_to_daily(TTE_HH)
Wac_daily <- convert_to_daily(Wac_HH)
Whr_daily <- convert_to_daily(Whr_HH)
Wom_daily <- convert_to_daily(Wom_HH)
Ync_daily <- convert_to_daily(Ync_HH)
Tum_daily <- convert_hourly_to_daily(Tum_HH)
```

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Cum_daily <- Cum_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Emr_daily <- Emr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Fog_daily <- Fog_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gin_daily <- Gin_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gww_daily <- Gww_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Lox_daily <- Lox_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rdf_daily <- Rdf_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rig_daily <- Rig_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rob_daily <- Rob_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

TTE_daily <- TTE_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wac_daily <- Wac_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Whr_daily <- Whr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wom_daily <- Wom_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ync_daily <- Ync_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

```


```{r}
plot(Wac_daily$VPD)
```
Plot annual precipitation for each site - to find dry - wet gradient of sites

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}

How_a <- convert_to_annual(How_daily)
Ade_a <- convert_to_annual(Ade_daily)
Das_a <- convert_to_annual(Das_daily)
Dap_a <- convert_to_annual(Dap_daily)
Dry_a <- convert_to_annual(Dry_daily)
Stp_a <- convert_to_annual(Stp_daily)
Asm_a <- convert_to_annual(Asm_daily)
Cum_a <- convert_to_annual(Cum_daily)
Emr_a <- convert_to_annual(Emr_daily)
Fog_a <- convert_to_annual(Fog_daily)
Gin_a <- convert_to_annual(Gin_daily)
Gww_a <- convert_to_annual(Gww_daily)
Lox_a <- convert_to_annual(Lox_daily)
Rdf_a <- convert_to_annual(Rdf_daily)
Rig_a <- convert_to_annual(Rig_daily)
Rob_a <- convert_to_annual(Rob_daily)
TTE_a <- convert_to_annual(TTE_daily)
Wac_a <- convert_to_annual(Wac_daily)
Whr_a <- convert_to_annual(Whr_daily)
Wom_a <- convert_to_annual(Wom_daily)
Ync_a <- convert_to_annual(Ync_daily)

```

```{r}
library(tidyverse)

annual_precip <- bind_rows(
  How_a %>% mutate(site = "How"),
  Ade_a %>% mutate(site = "Ade"),
  Das_a %>% mutate(site = "Das"),
  Dap_a %>% mutate(site = "Dap"),
  Dry_a %>% mutate(site = "Dry"),
  Stp_a %>% mutate(site = "Stp"),
  Asm_a %>% mutate(site = "Asm"),
  Cum_a %>% mutate(site = "Cum"),
  Emr_a %>% mutate(site = "Emr"),
  Fog_a %>% mutate(site = "Fog"),
  Gin_a %>% mutate(site = "Gin"),
  Gww_a %>% mutate(site = "Gww"),
  Lox_a %>% mutate(site = "Lox"),
  Rdf_a %>% mutate(site = "Rdf"),
  Rig_a %>% mutate(site = "Rig"),
  Rob_a %>% mutate(site = "Rob"),
  TTE_a %>% mutate(site = "TTE"),
  Wac_a %>% mutate(site = "Wac"),
  Whr_a %>% mutate(site = "Whr"),
  Wom_a %>% mutate(site = "Wom"),
  Ync_a %>% mutate(site = "Ync")
)

```

```{r}
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across OzFlux sites"
  )

```
```{r}
#growing season - this time i am keeping the top 75% productive periods per site because hard threshold is not fair across different ecosystems

library(dplyr)

filter_gs <- function(df, gpp_quantile = 0.25, swrad_thresh = 20) {

  df %>%
    # remove nighttime
    filter(SW_rad > swrad_thresh) %>%
    
    # global GPP filtering
    filter(GPP > quantile(GPP, gpp_quantile, na.rm = TRUE))
}


#Apply to dataset

How_gs <- filter_gs(How_daily)
Ade_gs <- filter_gs(Ade_daily)
Das_gs <- filter_gs(Das_daily)
Dap_gs <- filter_gs(Dap_daily)
Dry_gs <- filter_gs(Dry_daily)
Stp_gs <- filter_gs(Stp_daily)
Asm_gs <- filter_gs(Asm_daily)
Cum_gs <- filter_gs(Cum_daily)
Emr_gs <- filter_gs(Emr_daily)
Fog_gs <- filter_gs(Fog_daily)
Gin_gs <- filter_gs(Gin_daily)
Gww_gs <- filter_gs(Gww_daily)
Lox_gs <- filter_gs(Lox_daily)
Rdf_gs <- filter_gs(Rdf_daily)
Rig_gs <- filter_gs(Rig_daily)
Rob_gs <- filter_gs(Rob_daily)
TTE_gs <- filter_gs(TTE_daily)
Wac_gs <- filter_gs(Wac_daily)
Wom_gs <- filter_gs(Wom_daily)
Ync_gs <- filter_gs(Ync_daily)
Whr_gs <- filter_gs(Whr_daily)
```

1. 
GPP ~ VPD curve for all sites:

```{r}
library(dplyr)
library(ggplot2)

# Add a 'site' column to each dataframe
How_gs  <- How_gs  %>% mutate(site = "How")
Ade_gs  <- Ade_gs  %>% mutate(site = "Ade")
Das_gs  <- Das_gs  %>% mutate(site = "Das")
Dap_gs  <- Dap_gs  %>% mutate(site = "Dap")
Dry_gs  <- Dry_gs  %>% mutate(site = "Dry")
Stp_gs  <- Stp_gs  %>% mutate(site = "Stp")
Asm_gs  <- Asm_gs  %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")

# Combine all sites
all_sites <- bind_rows(How_gs, Ade_gs, Das_gs, Dap_gs, Dry_gs, Stp_gs, Asm_gs, Cum_gs, Emr_gs, Fog_gs, Gin_gs, Gww_gs, Lox_gs, Rdf_gs, Rig_gs, Rob_gs, TTE_gs, Wac_gs, Wom_gs, Ync_gs, Whr_gs)

# Set wet → dry order
all_sites$site <- factor(all_sites$site,
                         levels = c("Wac", "Rob", "How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp","Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox"))
```

```{r}
library(RColorBrewer)
# Define 7 distinct colors for the sites
site_colors <- brewer.pal(n = 9, name = "Set1")  # Set1 is bright and categorical

ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.2, span = 0.3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across OzFlux sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  )

```
Overall, we can see that the drier sites are less sensitive to higher VPD. 

2. 
Full GAM (climate + LCC + VPD) for each site
Plot sites r-squared by same wetness 
- Load in pet/cwd and LCC

Load in water stress component
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

Asm_daily$pet <- priestley_taylor_pet(Asm_daily$Rn, Asm_daily$Tair, Asm_daily$Pa)
Asm_daily$cwd <- Asm_daily$pet - Asm_daily$Qle

How_daily$pet <- priestley_taylor_pet(How_daily$Rn, How_daily$Tair, How_daily$Pa)
How_daily$cwd <- How_daily$pet - How_daily$Qle

Das_daily$pet <- priestley_taylor_pet(Das_daily$Rn, Das_daily$Tair, Das_daily$Pa)
Das_daily$cwd <- Das_daily$pet - Das_daily$Qle

Dap_daily$pet <- priestley_taylor_pet(Dap_daily$Rn, Dap_daily$Tair, Dap_daily$Pa)
Dap_daily$cwd <- Dap_daily$pet - Dap_daily$Qle

Dry_daily$pet <- priestley_taylor_pet(Dry_daily$Rn, Dry_daily$Tair, Dry_daily$Pa)
Dry_daily$cwd <- Dry_daily$pet - Dry_daily$Qle

Ade_daily$pet <- priestley_taylor_pet(Ade_daily$Rn, Ade_daily$Tair, Ade_daily$Pa)
Ade_daily$cwd <- Ade_daily$pet - Ade_daily$Qle

Stp_daily$pet <- priestley_taylor_pet(Stp_daily$Rn, Stp_daily$Tair, Stp_daily$Pa)
Stp_daily$cwd <- Stp_daily$pet - Stp_daily$Qle

Cum_daily$pet <- priestley_taylor_pet(Cum_daily$Rn, Cum_daily$Tair, Cum_daily$Pa)
Cum_daily$cwd <- Cum_daily$pet - Cum_daily$Qle

Emr_daily$pet <- priestley_taylor_pet(Emr_daily$Rn, Emr_daily$Tair, Emr_daily$Pa)
Emr_daily$cwd <- Emr_daily$pet - Emr_daily$Qle

Fog_daily$pet <- priestley_taylor_pet(Fog_daily$Rn, Fog_daily$Tair, Fog_daily$Pa)
Fog_daily$cwd <- Fog_daily$pet - Fog_daily$Qle

Gin_daily$pet <- priestley_taylor_pet(Gin_daily$Rn, Gin_daily$Tair, Gin_daily$Pa)
Gin_daily$cwd <- Gin_daily$pet - Gin_daily$Qle

Gww_daily$pet <- priestley_taylor_pet(Gww_daily$Rn, Gww_daily$Tair, Gww_daily$Pa)
Gww_daily$cwd <- Gww_daily$pet - Gww_daily$Qle

Lox_daily$pet <- priestley_taylor_pet(Lox_daily$Rn, Lox_daily$Tair, Lox_daily$Pa)
Lox_daily$cwd <- Lox_daily$pet - Lox_daily$Qle

Rdf_daily$pet <- priestley_taylor_pet(Rdf_daily$Rn, Rdf_daily$Tair, Rdf_daily$Pa)
Rdf_daily$cwd <- Rdf_daily$pet - Rdf_daily$Qle

Rig_daily$pet <- priestley_taylor_pet(Rig_daily$Rn, Rig_daily$Tair, Rig_daily$Pa)
Rig_daily$cwd <- Rig_daily$pet - Rig_daily$Qle

Rob_daily$pet <- priestley_taylor_pet(Rob_daily$Rn, Rob_daily$Tair, Rob_daily$Pa)
Rob_daily$cwd <- Rob_daily$pet - Rob_daily$Qle

TTE_daily$pet <- priestley_taylor_pet(TTE_daily$Rn, TTE_daily$Tair, TTE_daily$Pa)
TTE_daily$cwd <- TTE_daily$pet - TTE_daily$Qle

Wac_daily$pet <- priestley_taylor_pet(Wac_daily$Rn, Wac_daily$Tair, Wac_daily$Pa)
Wac_daily$cwd <- Wac_daily$pet - Wac_daily$Qle

Wom_daily$pet <- priestley_taylor_pet(Wom_daily$Rn, Wom_daily$Tair, Wom_daily$Pa)
Wom_daily$cwd <- Wom_daily$pet - Wom_daily$Qle

Ync_daily$pet <- priestley_taylor_pet(Ync_daily$Rn, Ync_daily$Tair, Ync_daily$Pa)
Ync_daily$cwd <- Ync_daily$pet - Ync_daily$Qle

Whr_daily$pet <- priestley_taylor_pet(Whr_daily$Rn, Whr_daily$Tair, Whr_daily$Pa)
Whr_daily$cwd <- Whr_daily$pet - Whr_daily$Qle

```

```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

# Apply to each dataset
How_daily <- add_year_doy(How_daily)
Dap_daily <- add_year_doy(Dap_daily)
Das_daily <- add_year_doy(Das_daily)
Dry_daily <- add_year_doy(Dry_daily)
Stp_daily <- add_year_doy(Stp_daily)
Asm_daily <- add_year_doy(Asm_daily)
Cum_daily <- add_year_doy(Cum_daily)
Emr_daily <- add_year_doy(Emr_daily)
Fog_daily <- add_year_doy(Fog_daily)
Gin_daily <- add_year_doy(Gin_daily)
Gww_daily <- add_year_doy(Gww_daily)
Lox_daily <- add_year_doy(Lox_daily)
Rdf_daily <- add_year_doy(Rdf_daily)
Rig_daily <- add_year_doy(Rig_daily)
Rob_daily <- add_year_doy(Rob_daily)
TTE_daily <- add_year_doy(TTE_daily)
Wac_daily <- add_year_doy(Wac_daily)
Wom_daily <- add_year_doy(Wom_daily)
Ync_daily <- add_year_doy(Ync_daily)
Whr_daily <- add_year_doy(Whr_daily)


```


```{r}
# -----------------------------
# Libraries
# -----------------------------
library(terra)
library(ncdf4)
library(dplyr)
library(lubridate)
library(purrr)

# -----------------------------
# File paths
# -----------------------------
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# -----------------------------
# Helper: Convert lat/lon to NetCDF indices
# -----------------------------
latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180 / nlines  
  line <- round((90 - lat) / lat_res)
  
  lon_res <- 360 / nsamples
  sample <- round((lon + 180) / lon_res)
  
  # clamp to valid range
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  list(line = line, sample = sample)
}

# -----------------------------
# Helper: Extract Vcmax from raster
# -----------------------------
get_vcmax <- function(lat, lon, raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(raster, coords)
  
  if (is.null(vcmax) || all(is.na(vcmax))) {
    stop("Vcmax extraction failed for coordinates: ", lat, ", ", lon)
  }
  
  as.numeric(vcmax[1,2])
}

# -----------------------------
# Helper: Extract daily LCC from NetCDF
# -----------------------------
get_daily_LCC <- function(nc_file, lat, lon) {
  nc <- nc_open(nc_file)
  coords <- latlon_to_line_sample(lat, lon)
  
  # Extract time series for the pixel
  pixel_time_series <- ncvar_get(
    nc, "chl_all",
    start = c(coords$line, coords$sample, 1, 1),
    count = c(1, 1, -1, -1)
  )
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel_time_series[pixel_time_series == fillvalue] <- NA
  nc_close(nc)
  
  # Convert units
  LCC <- pixel_time_series / 100
  
  # Weekly climatology -> daily
  weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
  weeks <- 1:length(weekly_climatology)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  
  daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y
  return(daily_climatology)
}

# -----------------------------
# Main function: attach site data
# -----------------------------
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc_file) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  daily_LCC <- get_daily_LCC(lcc_nc_file, lat, lon)
  
  # Repeat daily LCC to match number of rows in df
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df %>%
    mutate(
      vcmax = vcmax_val,
      LCC_daily = daily_LCC_full,
      year = year(date),
      doy  = yday(date)
    )
}

# -----------------------------
# Site metadata table
# -----------------------------
sites <- tibble(
  name = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr", "Fog",
           "Gin", "Gww", "Lox", "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr", "Wom", "Ync"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283, -33.6152, -23.8587,
           -12.5452, -31.3764, -30.1913, -34.4704, -14.5636, -36.6499, -17.1175, -22.287, -37.4259,
           -36.6732, -37.4222, -34.9893),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 120.6541, 140.6551, 132.4776, 145.5759, 145.6301, 133.64, 145.1878,
           145.0294, 144.0944, 146.2907),
  df   = list(How_daily, Ade_daily, Das_daily, Dap_daily, Dry_daily, Stp_daily, Asm_daily, Cum_daily,
              Emr_daily, Fog_daily, Gin_daily, Gww_daily, Lox_daily, Rdf_daily, Rig_daily, Rob_daily,
              TTE_daily, Wac_daily, Whr_daily, Wom_daily, Ync_daily)
)

# -----------------------------
# Apply function to all sites
# -----------------------------
sites <- sites %>%
  mutate(
    df_with_data = pmap(list(df, lat, lon),
                        ~ attach_site_data(..1, ..2, ..3, vcmax_raster, lcc_nc))
  )

# -----------------------------
# Extract updated dataframes to global environment
# -----------------------------
list2env(setNames(sites$df_with_data, sites$name), envir = .GlobalEnv)

```


```{r}
library(dplyr)

filter_gs <- function(df, gpp_quantile = 0.25, swrad_thresh = 20) {

  df %>%
    # remove nighttime
    filter(SW_rad > swrad_thresh) %>%
    
    # global GPP filtering
    filter(GPP > quantile(GPP, gpp_quantile, na.rm = TRUE))
}


#Apply to dataset

How_gs <- filter_gs(How)
Ade_gs <- filter_gs(Ade)
Das_gs <- filter_gs(Das)
Dap_gs <- filter_gs(Dap)
Dry_gs <- filter_gs(Dry)
Stp_gs <- filter_gs(Stp)
Asm_gs <- filter_gs(Asm)
Cum_gs <- filter_gs(Cum)
Emr_gs <- filter_gs(Emr)
Fog_gs <- filter_gs(Fog)
Gin_gs <- filter_gs(Gin)
Gww_gs <- filter_gs(Gww)
Lox_gs <- filter_gs(Lox)
Rdf_gs <- filter_gs(Rdf)
Rig_gs <- filter_gs(Rig)
Rob_gs <- filter_gs(Rob)
TTE_gs <- filter_gs(TTE)
Wac_gs <- filter_gs(Wac)
Wom_gs <- filter_gs(Wom)
Ync_gs <- filter_gs(Ync)
Whr_gs <- filter_gs(Whr)
```

```{r}
install.packages("viridisLite")
```


```{r}
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(viridisLite)# for color scale

# -----------------------------
# Full site list
# -----------------------------
sites <- tibble(
  site = c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283,
           -33.6152, -23.8587, -12.5452, -31.3764, -30.1913, -34.4704, -14.5636,
           -36.6499, -17.1175, -22.287, -37.4259, -36.6732, -37.4222, -34.9893),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490,
           150.7236, 148.4746, 131.3072, 115.7138, 120.6541, 140.6551, 132.4776,
           145.5759, 145.6301, 133.64, 145.1878, 145.0294, 144.0944, 146.2907),
  wetness_order = 1:21  # optional: assign numbers 1 (wettest) → 21 (driest) if known
)

# -----------------------------
# Get Australia map
# -----------------------------
aus <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

# -----------------------------
# Plot
# -----------------------------
ggplot() +
  geom_sf(data = aus, fill = "antiquewhite", color = "black") +
  geom_point(data = sites, aes(x = lon, y = lat, color = wetness_order), size = 4) +
  geom_text(data = sites, aes(x = lon, y = lat, label = site), nudge_y = 0.3, size = 4) +
  scale_color_viridis_c(option = "C", direction = -1, na.value = "grey50", name = "Wet → Dry") +
  coord_sf(xlim = c(110, 155), ylim = c(-40, -10), expand = FALSE) +  # full map of Australia
  theme_minimal() +
  labs(title = "FluxNet Sites Across Australia",
       x = "Longitude",
       y = "Latitude")

```
1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# -----------------------------
# List all sites
# -----------------------------
all_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync")

# -----------------------------
# Split, rolling CWD, lagged variables
# -----------------------------
for(site in all_sites){
  
  df <- get(paste0(site, "_gs"))
  
  # Split into train/test
  idx <- createDataPartition(df$GPP, p=0.8, list=FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  # Drop NA in test
  test <- test %>% drop_na()
  
  # 7-day rolling cumulative CWD
  train$cumulative_cwd_7day <- rollapply(train$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  
  # Drop NA after rolling
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  # Add 3-day lag variables
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  # Save back to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"),  test,  envir = .GlobalEnv)
}

```


```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
climate_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```



```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox")

climate_metrics$site <- factor(climate_metrics$site, levels = wet_dry_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Dry -> Wet)"
  ) +
  theme_minimal()

```
Range of R-squared across OzFlux sites. GAM used with VPD, Tair, Precip, cumulative cwd, VPD lag, Tair lag, leaf chlorophyll content and shortwave radiation. Only thing missing is LAI (leaf area index) which could improve model fit for some of the sites.


No of observations per site:
```{r}
length(unique(How_train$GPP))
length(unique(Ade_train$GPP))
length(unique(Das_train$GPP))
length(unique(Dap_train$GPP))
length(unique(Dry_train$GPP))
length(unique(Stp_train$GPP))
length(unique(Asm_train$GPP))
length(unique(Cum_train$GPP))
length(unique(Emr_train$GPP))
length(unique(Fog_train$GPP))
length(unique(Gin_train$GPP))
length(unique(Gww_train$GPP))
length(unique(Lox_train$GPP))
length(unique(Rdf_train$GPP))
length(unique(Rig_train$GPP))
length(unique(Rob_train$GPP))
length(unique(TTE_train$GPP))
length(unique(Wac_train$GPP))
length(unique(Whr_train$GPP))
length(unique(Wom_train$GPP))
length(unique(Ync_train$GPP))


```
```{r}
# Define wet → dry order
obs_order <- c("How", "Das", "Stp", "Dry", "Dap", "Ync", "Asm", "Rob", "Whr",
  "Gww", "Wom", "Fog", "Gin", "Wac", "Emr", "Cum", "Lox", "Rig",
  "Ade", "TTE", "Rdf")


climate_metrics$site <- factor(climate_metrics$site, levels = obs_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (least to most number of obs)"
  ) +
  theme_minimal()
```
Now let's look at how VPD improves model fit:

1. Model without VPD
2. VPD only model
3. Full model with VPD

Model without VPD
```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
no_vpd_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox")

no_vpd_metrics$site <- factor(no_vpd_metrics$site, levels = wet_dry_order)

ggplot(no_vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Dry -> Wet)"
  ) +
  theme_minimal()
```
VPD only model

```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(VPD) + s(VPD_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
vpd_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)
```

```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox")

vpd_metrics$site <- factor(vpd_metrics$site, levels = wet_dry_order)

ggplot(vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Dry -> Wet)"
  ) +
  theme_minimal()
```
Full model with VPD

```{r}
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
full_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)
```

compare models:
```{r}
library(dplyr)

gam_comparison <- no_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    vpd_metrics %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  inner_join(
    full_metrics %>%
      select(site, R2) %>%
      rename(R2_full = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_full - R2_no_vpd
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet → Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```



Separate into PFTs/groups/precip gradient:
1. VPD only model and full model without VPD 
2. Compare with previous model of climate + VPD + physiology
3. Added predictive value of VPD?
