---
title: "02_site_analysis"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Which Australian flux site is the most/least predictable?
- R-squared from model
- Change in R-squared when VPD is added to GAM
- Can VPD explain residual variation?
- VPD curves for sites/PFT groups?

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cpr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cpr_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cum_FLUXNET2015_FULLSET_HH_2012-2014_2-4.csv")

Emr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Emr_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Fog_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Fog_FLUXNET2015_FULLSET_HH_2006-2008_1-4.csv")

Gin_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Gin_FLUXNET2015_FULLSET_HH_2011-2014_1-4.csv")

Gww_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-GWW_FLUXNET2015_FULLSET_HH_2013-2014_1-4.csv")

Lox_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Lox_FLUXNET2015_FULLSET_HH_2008-2009_1-4.csv")

Rdf_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-RDF_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Rig_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rig_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Rob_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rob_FLUXNET2015_FULLSET_HH_2014-2014_1-4.csv")

TTE_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-TTE_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Wac_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wac_FLUXNET2015_FULLSET_HH_2005-2008_1-4.csv")

Whr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Whr_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Wom_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wom_FLUXNET2015_FULLSET_HH_2010-2014_1-4.csv")

Ync_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ync_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Cpr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cpr_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

```

Hourly AU-Tum
```{r}
Tum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/hourly_2/FLX_AU-Tum_FLUXNET2015_FULLSET_HR_2001-2014_2-4.csv")
```

Remove -9999 values

```{r}
sites_HH <- list(
  How = How_HH,
  Ade = Ade_HH,
  Das = Das_HH,
  Dap = Dap_HH,
  Tum = Tum_HH, 
  Cpr = Cpr_HH, 
  Dry = Dry_HH, 
  Stp = Stp_HH, 
  Asm = Asm_HH, 
  Cum = Cum_HH, 
  Emr = Emr_HH, 
  Fog = Fog_HH, 
  Gin = Gin_HH, 
  Gww = Gww_HH, 
  Lox = Lox_HH, 
  Rdf = Rdf_HH, 
  Rig = Rig_HH, 
  Rob = Rob_HH, 
  TTE = TTE_HH, 
  Wac = Wac_HH, 
  Whr = Whr_HH, 
  Wom = Wom_HH,
  Ync = Ync_HH
)
```



```{r}
flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_HH <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_HH <- Ade_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_HH <- Das_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_HH <- Dap_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_HH <- Dry_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_HH <- Stp_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_HH <- Asm_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Cum_HH <- Cum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Emr_HH <- Emr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Fog_HH <- Fog_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gin_HH <- Gin_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gww_HH <- Gww_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Lox_HH <- Lox_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rdf_HH <- Rdf_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rig_HH <- Rig_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rob_HH <- Rob_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

TTE_HH <- TTE_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wac_HH <- Wac_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Whr_HH <- Whr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wom_HH <- Wom_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ync_HH <- Ync_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Tum_HH <- Tum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Cpr_HH <- Cpr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```

Filter low coverage days (half-hourly daya)
```{r}
filter_low_coverage_days <- function(df, min_frac = 0.9, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```

Filter low coverage days (hourly data, AU-Tum)
```{r}
filter_low_coverage_days_2 <- function(df, min_frac = 0.9, steps_per_day = 24) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```


```{r}
How_HH <- filter_low_coverage_days(How_HH)
Ade_HH <- filter_low_coverage_days(Ade_HH)
Das_HH <- filter_low_coverage_days(Das_HH)
Dap_HH <- filter_low_coverage_days(Dap_HH)
Dry_HH <- filter_low_coverage_days(Dry_HH)
Stp_HH <- filter_low_coverage_days(Stp_HH)
Asm_HH <- filter_low_coverage_days(Asm_HH)
Cum_HH <- filter_low_coverage_days(Cum_HH)
Emr_HH <- filter_low_coverage_days(Emr_HH)
Fog_HH <- filter_low_coverage_days(Fog_HH)
Gin_HH <- filter_low_coverage_days(Gin_HH)
Gww_HH <- filter_low_coverage_days(Gww_HH)
Lox_HH <- filter_low_coverage_days(Lox_HH)
Rdf_HH <- filter_low_coverage_days(Rdf_HH)
Rig_HH <- filter_low_coverage_days(Rig_HH)
Rob_HH <- filter_low_coverage_days(Rob_HH)
TTE_HH <- filter_low_coverage_days(TTE_HH)
Wac_HH <- filter_low_coverage_days(Wac_HH)
Wom_HH <- filter_low_coverage_days(Wom_HH)
Whr_HH <- filter_low_coverage_days(Whr_HH)
Ync_HH <- filter_low_coverage_days(Ync_HH)
Tum_HH <- filter_low_coverage_days_2(Tum_HH)
Cpr_HH <- filter_low_coverage_days(Cpr_HH)
```


Convert to daily:

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

Convert hourly to daily (AU_Tum)
```{r}
library(tidyverse)
library(lubridate)

convert_hourly_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HOUR <- 3600.
  WM2_TO_KG_M2_S <- 1.0 / (2.45 * 1E6)
  PA_2_KPA <- 0.001
  HPA_2_KPA <- 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HOUR,
      VPD = VPD * HPA_2_KPA
    )

  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df)
}

```

```{r}
How_daily <- convert_to_daily(How_HH)
Ade_daily <- convert_to_daily(Ade_HH)
Das_daily <- convert_to_daily(Das_HH)
Dap_daily <- convert_to_daily(Dap_HH)
Dry_daily <- convert_to_daily(Dry_HH)
Stp_daily <- convert_to_daily(Stp_HH)
Asm_daily <- convert_to_daily(Asm_HH)
Cum_daily <- convert_to_daily(Cum_HH)
Emr_daily <- convert_to_daily(Emr_HH)
Fog_daily <- convert_to_daily(Fog_HH)
Gin_daily <- convert_to_daily(Gin_HH)
Gww_daily <- convert_to_daily(Gww_HH)
Lox_daily <- convert_to_daily(Lox_HH)
Rdf_daily <- convert_to_daily(Rdf_HH)
Rig_daily <- convert_to_daily(Rig_HH)
Rob_daily <- convert_to_daily(Rob_HH)
TTE_daily <- convert_to_daily(TTE_HH)
Wac_daily <- convert_to_daily(Wac_HH)
Whr_daily <- convert_to_daily(Whr_HH)
Wom_daily <- convert_to_daily(Wom_HH)
Ync_daily <- convert_to_daily(Ync_HH)
Cpr_daily <- convert_to_daily(Cpr_HH)
Tum_daily <- convert_hourly_to_daily(Tum_HH)
```

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Cum_daily <- Cum_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Emr_daily <- Emr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Fog_daily <- Fog_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gin_daily <- Gin_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gww_daily <- Gww_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Lox_daily <- Lox_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rdf_daily <- Rdf_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rig_daily <- Rig_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rob_daily <- Rob_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

TTE_daily <- TTE_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wac_daily <- Wac_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Whr_daily <- Whr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wom_daily <- Wom_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ync_daily <- Ync_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Cpr_daily <- Cpr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Tum_daily <- Tum_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))
```


```{r}
plot(Cpr_daily$VPD)
```
Plot annual precipitation for each site - to find dry - wet gradient of sites

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}

How_a <- convert_to_annual(How_daily)
Ade_a <- convert_to_annual(Ade_daily)
Das_a <- convert_to_annual(Das_daily)
Dap_a <- convert_to_annual(Dap_daily)
Dry_a <- convert_to_annual(Dry_daily)
Stp_a <- convert_to_annual(Stp_daily)
Asm_a <- convert_to_annual(Asm_daily)
Cum_a <- convert_to_annual(Cum_daily)
Emr_a <- convert_to_annual(Emr_daily)
Fog_a <- convert_to_annual(Fog_daily)
Gin_a <- convert_to_annual(Gin_daily)
Gww_a <- convert_to_annual(Gww_daily)
Lox_a <- convert_to_annual(Lox_daily)
Rdf_a <- convert_to_annual(Rdf_daily)
Rig_a <- convert_to_annual(Rig_daily)
Rob_a <- convert_to_annual(Rob_daily)
TTE_a <- convert_to_annual(TTE_daily)
Wac_a <- convert_to_annual(Wac_daily)
Whr_a <- convert_to_annual(Whr_daily)
Wom_a <- convert_to_annual(Wom_daily)
Ync_a <- convert_to_annual(Ync_daily)
Tum_a <- convert_to_annual(Tum_daily)
Cpr_a <- convert_to_annual(Cpr_daily)

```

```{r}
library(tidyverse)

annual_precip <- bind_rows(
  How_a %>% mutate(site = "How"),
  Ade_a %>% mutate(site = "Ade"),
  Das_a %>% mutate(site = "Das"),
  Dap_a %>% mutate(site = "Dap"),
  Dry_a %>% mutate(site = "Dry"),
  Stp_a %>% mutate(site = "Stp"),
  Asm_a %>% mutate(site = "Asm"),
  Cum_a %>% mutate(site = "Cum"),
  Emr_a %>% mutate(site = "Emr"),
  Fog_a %>% mutate(site = "Fog"),
  Gin_a %>% mutate(site = "Gin"),
  Gww_a %>% mutate(site = "Gww"),
  Lox_a %>% mutate(site = "Lox"),
  Rdf_a %>% mutate(site = "Rdf"),
  Rig_a %>% mutate(site = "Rig"),
  Rob_a %>% mutate(site = "Rob"),
  TTE_a %>% mutate(site = "TTE"),
  Wac_a %>% mutate(site = "Wac"),
  Whr_a %>% mutate(site = "Whr"),
  Wom_a %>% mutate(site = "Wom"),
  Ync_a %>% mutate(site = "Ync"),
  Tum_a %>% mutate(site = "Tum"),
  Cpr_a %>% mutate(site = "Cpr")
)

```

```{r}
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across Australian Fluxnet sites"
  )

```


```{r}
#growing season - for now, it is going to be above a certain temp threshold 5°C and above a short-wave radiation threshold

filter_gs <- function(df) {
  df %>%
    filter(
      Tair > 5,
      SW_rad > 50
    )
}


```

#Apply to dataset
```{r}

How_gs <- filter_gs(How_daily)
Ade_gs <- filter_gs(Ade_daily)
Das_gs <- filter_gs(Das_daily)
Dap_gs <- filter_gs(Dap_daily)
Dry_gs <- filter_gs(Dry_daily)
Stp_gs <- filter_gs(Stp_daily)
Asm_gs <- filter_gs(Asm_daily)
Cum_gs <- filter_gs(Cum_daily)
Emr_gs <- filter_gs(Emr_daily)
Fog_gs <- filter_gs(Fog_daily)
Gin_gs <- filter_gs(Gin_daily)
Gww_gs <- filter_gs(Gww_daily)
Lox_gs <- filter_gs(Lox_daily)
Rdf_gs <- filter_gs(Rdf_daily)
Rig_gs <- filter_gs(Rig_daily)
Rob_gs <- filter_gs(Rob_daily)
TTE_gs <- filter_gs(TTE_daily)
Wac_gs <- filter_gs(Wac_daily)
Wom_gs <- filter_gs(Wom_daily)
Ync_gs <- filter_gs(Ync_daily)
Whr_gs <- filter_gs(Whr_daily)
Tum_gs <- filter_gs(Tum_daily)
Cpr_gs <- filter_gs(Cpr_daily)
```

1. 
GPP ~ VPD curve for all sites:

```{r}
library(dplyr)
library(ggplot2)

# Add a 'site' column to each dataframe
How_gs  <- How_gs  %>% mutate(site = "How")
Ade_gs  <- Ade_gs  %>% mutate(site = "Ade")
Das_gs  <- Das_gs  %>% mutate(site = "Das")
Dap_gs  <- Dap_gs  %>% mutate(site = "Dap")
Dry_gs  <- Dry_gs  %>% mutate(site = "Dry")
Stp_gs  <- Stp_gs  %>% mutate(site = "Stp")
Asm_gs  <- Asm_gs  %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")
Tum_gs <- Tum_gs %>% mutate(site = "Tum")
Cpr_gs <- Cpr_gs %>% mutate(site = "Cpr")

# Combine all sites
all_sites <- bind_rows(How_gs, Ade_gs, Das_gs, Dap_gs, Dry_gs, Stp_gs, Asm_gs, Cum_gs, Emr_gs, Fog_gs, Gin_gs, Gww_gs, Lox_gs, Rdf_gs, Rig_gs, Rob_gs, TTE_gs, Wac_gs, Wom_gs, Ync_gs, Whr_gs, Tum_gs, Cpr_gs)

# Set wet → dry order
all_sites$site <- factor(all_sites$site,
                         levels = c("Wac", "Rob", "How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp","Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox", "Tum", "Cpr"))
```

Loess smooth lines: (ok but not great, may not show true curves )
```{r}
library(RColorBrewer)
# Define distinct colors for the sites
site_colors <- brewer.pal(n = 9, name = "Set1")  # Set1 is bright and categorical

ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.2, span = 0.3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across OzFlux sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  )

```

Instead, I am going to bin VPD and calculate mean GPP and Standard error for each VPD bin - and then plot that
```{r}
#Can increase to larger bins later (0.1, 0.2, 0.5...)
all_sites$VPD_bin <- cut(all_sites$VPD, breaks = seq(0, max(all_sites$VPD, na.rm = TRUE), by = 0.5))
```

Calculate mean GPP and standard error
```{r}
library(dplyr)

binned_data <- all_sites %>%
  group_by(site, VPD_bin) %>%
  summarise(
    mean_GPP = mean(GPP, na.rm = TRUE),
    sd_GPP = sd(GPP, na.rm = TRUE),
    n = n(),
    se_GPP = sd_GPP / sqrt(n)
  ) %>%
  ungroup()

```

```{r}
ggplot(binned_data, aes(x = as.numeric(VPD_bin), y = mean_GPP, color = site)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_GPP - se_GPP, ymax = mean_GPP + se_GPP), width = 0.05) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Binned GPP response to VPD across OzFlux sites",
    x = "VPD bin (kPa)",
    y = "Mean GPP (g C m-2 d-1)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Quite difficult to make out any sort of pattern. I think drier sites are less sensitive to changes in VPD compared to wet sites. 

Let's separate sites into PFT and then plot their VPD curves:

```{r}
WSA <- bind_rows(Ade_gs, Gin_gs, How_gs, Rdf_gs) %>% group_by(site)

SAV <- bind_rows(Das_gs, Dry_gs, Gww_gs, Asm_gs, Cpr_gs) %>% group_by(site)

EBF <- bind_rows(Cum_gs, Rob_gs, Wac_gs, Whr_gs, Wom_gs, Tum_gs) %>% group_by(site)

GRA <- bind_rows(Dap_gs, Emr_gs, Rig_gs, Stp_gs, TTE_gs, Ync_gs) %>% group_by(site)

WET <- Fog_gs %>% group_by(site)

DBF <- Lox_gs %>% group_by(site)
```

```{r}
WSA$PFT <- "WSA"
SAV$PFT <- "SAV"
EBF$PFT <- "EBF"
GRA$PFT <- "GRA"
WET$PFT <- "WET"
DBF$PFT <- "DBF"
```

```{r}
PFT_all <- bind_rows(WSA, SAV, EBF, GRA, WET, DBF)
```

```{r}
library(dplyr)
library(ggplot2)

# Create VPD bins of width 0.5 kPa
PFT_all <- PFT_all %>%
  mutate(VPD_bin = cut(VPD,
                       breaks = seq(floor(min(VPD, na.rm = TRUE)), 
                                    ceiling(max(VPD, na.rm = TRUE)), 
                                    by = 0.5),
                       include.lowest = TRUE))

# Calculate mean and SE of GPP per PFT per VPD bin
summary_stats <- PFT_all %>%
  group_by(PFT, VPD_bin) %>%
  summarise(
    mean_GPP = mean(GPP, na.rm = TRUE),
    se_GPP = sd(GPP, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot binned means with error bars
ggplot(summary_stats, aes(x = VPD_bin, y = mean_GPP, colour = PFT, group = PFT)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_GPP - se_GPP, ymax = mean_GPP + se_GPP), width = 0.2) +
  geom_line(linewidth = 1) +
  theme_minimal(base_size = 14) +
  labs(
    x = "VPD (0.5 kPa bins)",
    y = "Mean GPP ± SE",
    colour = "PFT",
    title = "Binned GPP–VPD relationships across PFTs"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
2. 
How does predictabilty vary across sites in Australia?
Full GAM (climate + LCC + VPD) for each site
Plot sites r-squared by same wetness 
- Load in pet/cwd and LCC

Load in water stress component
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

Asm_daily$pet <- priestley_taylor_pet(Asm_daily$Rn, Asm_daily$Tair, Asm_daily$Pa)
Asm_daily$cwd <- Asm_daily$pet - Asm_daily$ET

How_daily$pet <- priestley_taylor_pet(How_daily$Rn, How_daily$Tair, How_daily$Pa)
How_daily$cwd <- How_daily$pet - How_daily$ET

Das_daily$pet <- priestley_taylor_pet(Das_daily$Rn, Das_daily$Tair, Das_daily$Pa)
Das_daily$cwd <- Das_daily$pet - Das_daily$ET

Dap_daily$pet <- priestley_taylor_pet(Dap_daily$Rn, Dap_daily$Tair, Dap_daily$Pa)
Dap_daily$cwd <- Dap_daily$pet - Dap_daily$ET

Dry_daily$pet <- priestley_taylor_pet(Dry_daily$Rn, Dry_daily$Tair, Dry_daily$Pa)
Dry_daily$cwd <- Dry_daily$pet - Dry_daily$ET

Ade_daily$pet <- priestley_taylor_pet(Ade_daily$Rn, Ade_daily$Tair, Ade_daily$Pa)
Ade_daily$cwd <- Ade_daily$pet - Ade_daily$ET

Stp_daily$pet <- priestley_taylor_pet(Stp_daily$Rn, Stp_daily$Tair, Stp_daily$Pa)
Stp_daily$cwd <- Stp_daily$pet - Stp_daily$ET

Cum_daily$pet <- priestley_taylor_pet(Cum_daily$Rn, Cum_daily$Tair, Cum_daily$Pa)
Cum_daily$cwd <- Cum_daily$pet - Cum_daily$ET

Emr_daily$pet <- priestley_taylor_pet(Emr_daily$Rn, Emr_daily$Tair, Emr_daily$Pa)
Emr_daily$cwd <- Emr_daily$pet - Emr_daily$ET

Fog_daily$pet <- priestley_taylor_pet(Fog_daily$Rn, Fog_daily$Tair, Fog_daily$Pa)
Fog_daily$cwd <- Fog_daily$pet - Fog_daily$ET

Gin_daily$pet <- priestley_taylor_pet(Gin_daily$Rn, Gin_daily$Tair, Gin_daily$Pa)
Gin_daily$cwd <- Gin_daily$pet - Gin_daily$ET

Gww_daily$pet <- priestley_taylor_pet(Gww_daily$Rn, Gww_daily$Tair, Gww_daily$Pa)
Gww_daily$cwd <- Gww_daily$pet - Gww_daily$ET

Lox_daily$pet <- priestley_taylor_pet(Lox_daily$Rn, Lox_daily$Tair, Lox_daily$Pa)
Lox_daily$cwd <- Lox_daily$pet - Lox_daily$ET

Rdf_daily$pet <- priestley_taylor_pet(Rdf_daily$Rn, Rdf_daily$Tair, Rdf_daily$Pa)
Rdf_daily$cwd <- Rdf_daily$pet - Rdf_daily$ET

Rig_daily$pet <- priestley_taylor_pet(Rig_daily$Rn, Rig_daily$Tair, Rig_daily$Pa)
Rig_daily$cwd <- Rig_daily$pet - Rig_daily$ET

Rob_daily$pet <- priestley_taylor_pet(Rob_daily$Rn, Rob_daily$Tair, Rob_daily$Pa)
Rob_daily$cwd <- Rob_daily$pet - Rob_daily$ET

TTE_daily$pet <- priestley_taylor_pet(TTE_daily$Rn, TTE_daily$Tair, TTE_daily$Pa)
TTE_daily$cwd <- TTE_daily$pet - TTE_daily$ET

Wac_daily$pet <- priestley_taylor_pet(Wac_daily$Rn, Wac_daily$Tair, Wac_daily$Pa)
Wac_daily$cwd <- Wac_daily$pet - Wac_daily$ET

Wom_daily$pet <- priestley_taylor_pet(Wom_daily$Rn, Wom_daily$Tair, Wom_daily$Pa)
Wom_daily$cwd <- Wom_daily$pet - Wom_daily$ET

Ync_daily$pet <- priestley_taylor_pet(Ync_daily$Rn, Ync_daily$Tair, Ync_daily$Pa)
Ync_daily$cwd <- Ync_daily$pet - Ync_daily$ET

Whr_daily$pet <- priestley_taylor_pet(Whr_daily$Rn, Whr_daily$Tair, Whr_daily$Pa)
Whr_daily$cwd <- Whr_daily$pet - Whr_daily$ET

Cpr_daily$pet <- priestley_taylor_pet(Cpr_daily$Rn, Cpr_daily$Tair, Cpr_daily$Pa)
Cpr_daily$cwd <- Cpr_daily$pet - Cpr_daily$ET

Tum_daily$pet <- priestley_taylor_pet(Tum_daily$Rn, Tum_daily$Tair, Tum_daily$Pa)
Tum_daily$cwd <- Tum_daily$pet - Tum_daily$ET

```

```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

# Apply to each dataset
How_daily <- add_year_doy(How_daily)
Dap_daily <- add_year_doy(Dap_daily)
Das_daily <- add_year_doy(Das_daily)
Dry_daily <- add_year_doy(Dry_daily)
Stp_daily <- add_year_doy(Stp_daily)
Asm_daily <- add_year_doy(Asm_daily)
Cum_daily <- add_year_doy(Cum_daily)
Emr_daily <- add_year_doy(Emr_daily)
Fog_daily <- add_year_doy(Fog_daily)
Gin_daily <- add_year_doy(Gin_daily)
Gww_daily <- add_year_doy(Gww_daily)
Lox_daily <- add_year_doy(Lox_daily)
Rdf_daily <- add_year_doy(Rdf_daily)
Rig_daily <- add_year_doy(Rig_daily)
Rob_daily <- add_year_doy(Rob_daily)
TTE_daily <- add_year_doy(TTE_daily)
Wac_daily <- add_year_doy(Wac_daily)
Wom_daily <- add_year_doy(Wom_daily)
Ync_daily <- add_year_doy(Ync_daily)
Whr_daily <- add_year_doy(Whr_daily)
Tum_daily <- add_year_doy(Tum_daily)
Cpr_daily <- add_year_doy(Cpr_daily)
```


```{r}
# -----------------------------
# Libraries
# -----------------------------
library(terra)
library(ncdf4)
library(dplyr)
library(lubridate)
library(purrr)

# -----------------------------
# File paths
# -----------------------------
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# -----------------------------
# Helper: Convert lat/lon to NetCDF indices
# -----------------------------
latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180 / nlines  
  line <- round((90 - lat) / lat_res)
  
  lon_res <- 360 / nsamples
  sample <- round((lon + 180) / lon_res)
  
  # clamp to valid range
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  list(line = line, sample = sample)
}

# -----------------------------
# Helper: Extract Vcmax from raster
# -----------------------------
get_vcmax <- function(lat, lon, raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(raster, coords)
  
  if (is.null(vcmax) || all(is.na(vcmax))) {
    stop("Vcmax extraction failed for coordinates: ", lat, ", ", lon)
  }
  
  as.numeric(vcmax[1,2])
}

# -----------------------------
# Helper: Extract daily LCC from NetCDF
# -----------------------------
get_daily_LCC <- function(nc_file, lat, lon) {
  nc <- nc_open(nc_file)
  coords <- latlon_to_line_sample(lat, lon)
  
  # Extract time series for the pixel
  pixel_time_series <- ncvar_get(
    nc, "chl_all",
    start = c(coords$line, coords$sample, 1, 1),
    count = c(1, 1, -1, -1)
  )
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel_time_series[pixel_time_series == fillvalue] <- NA
  nc_close(nc)
  
  # Convert units
  LCC <- pixel_time_series / 100
  
  # Weekly climatology -> daily
  weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
  weeks <- 1:length(weekly_climatology)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  
  daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y
  return(daily_climatology)
}

# -----------------------------
# Main function: attach site data
# -----------------------------
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc_file) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  daily_LCC <- get_daily_LCC(lcc_nc_file, lat, lon)
  
  # Repeat daily LCC to match number of rows in df
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df %>%
    mutate(
      vcmax = vcmax_val,
      LCC_daily = daily_LCC_full,
      year = year(date),
      doy  = yday(date)
    )
}

# -----------------------------
# Site metadata table
# -----------------------------
sites <- tibble(
  name = c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr", "Fog",
           "Gin", "Gww", "Lox", "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr", "Wom", "Ync", "Cpr", "Tum"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283, -33.6152, -23.8587,
           -12.5452, -31.3764, -30.1913, -34.4704, -14.5636, -36.6499, -17.1175, -22.287, -37.4259,
           -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 120.6541, 140.6551, 132.4776, 145.5759, 145.6301, 133.64, 145.1878,
           145.0294, 144.0944, 146.2907, 140.5891, 148.1517),
  df   = list(How_daily, Ade_daily, Das_daily, Dap_daily, Dry_daily, Stp_daily, Asm_daily, Cum_daily,
              Emr_daily, Fog_daily, Gin_daily, Gww_daily, Lox_daily, Rdf_daily, Rig_daily, Rob_daily,
              TTE_daily, Wac_daily, Whr_daily, Wom_daily, Ync_daily, Cpr_daily, Tum_daily)
)

# -----------------------------
# Apply function to all sites
# -----------------------------
sites <- sites %>%
  mutate(
    df_with_data = pmap(list(df, lat, lon),
                        ~ attach_site_data(..1, ..2, ..3, vcmax_raster, lcc_nc))
  )

# -----------------------------
# Extract updated dataframes to global environment
# -----------------------------
list2env(setNames(sites$df_with_data, sites$name), envir = .GlobalEnv)

```
sites called "Cpr" and "Tum" in global environment after that function

```{r}
#For now, growing season filter - Temp, SW rad and GPP > 0, but this needs some work
library(dplyr)

filter_gs <- function(df) {
  df %>%
    filter(
      Tair > 5,
      SW_rad > 50, 
      GPP > 0
    )
}

#Apply to dataset

How_gs <- filter_gs(How)
Ade_gs <- filter_gs(Ade)
Das_gs <- filter_gs(Das)
Dap_gs <- filter_gs(Dap)
Dry_gs <- filter_gs(Dry)
Stp_gs <- filter_gs(Stp)
Asm_gs <- filter_gs(Asm)
Cum_gs <- filter_gs(Cum)
Emr_gs <- filter_gs(Emr)
Fog_gs <- filter_gs(Fog)
Gin_gs <- filter_gs(Gin)
Gww_gs <- filter_gs(Gww)
Lox_gs <- filter_gs(Lox)
Rdf_gs <- filter_gs(Rdf)
Rig_gs <- filter_gs(Rig)
Rob_gs <- filter_gs(Rob)
TTE_gs <- filter_gs(TTE)
Wac_gs <- filter_gs(Wac)
Wom_gs <- filter_gs(Wom)
Ync_gs <- filter_gs(Ync)
Whr_gs <- filter_gs(Whr)
Cpr_gs <- filter_gs(Cpr)
Tum_gs <- filter_gs(Tum)
```

```{r}
install.packages("viridisLite")
```


```{r}
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(viridisLite)# for color scale

# -----------------------------
# Full site list
# -----------------------------
sites <- tibble(
  site = c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox", "Cpr", "Tum"),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283,
           -33.6152, -23.8587, -12.5452, -31.3764, -30.1913, -34.4704, -14.5636,
           -36.6499, -17.1175, -22.287, -37.4259, -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490,
           150.7236, 148.4746, 131.3072, 115.7138, 120.6541, 140.6551, 132.4776,
           145.5759, 145.6301, 133.64, 145.1878, 145.0294, 144.0944, 146.2907, 140.5891, 148.1517),
  wetness_order = NA  # optional: assign numbers 1 (wettest) → 21 (driest) if known
)

# -----------------------------
# Get Australia map
# -----------------------------
aus <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

# -----------------------------
# Plot
# -----------------------------
ggplot() +
  geom_sf(data = aus, fill = "antiquewhite", color = "black") +
  geom_point(data = sites, aes(x = lon, y = lat, color = wetness_order), size = 4) +
  geom_text(data = sites, aes(x = lon, y = lat, label = site), nudge_y = 0.3, size = 4) +
  scale_color_viridis_c(option = "C", direction = -1, na.value = "grey50", name = "Wet -> Dry") +
  coord_sf(xlim = c(110, 155), ylim = c(-40, -10), expand = FALSE) +  # full map of Australia
  theme_minimal() +
  labs(title = "FluxNet Sites Across Australia",
       x = "Longitude",
       y = "Latitude")

```
1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# -----------------------------
# List all sites
# -----------------------------
all_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# -----------------------------
# Split, rolling CWD, lagged variables
# -----------------------------
for(site in all_sites){
  
  df <- get(paste0(site, "_gs"))
  
  # Split into train/test
  idx <- createDataPartition(df$GPP, p=0.8, list=FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  # Drop NA in test
  test <- test %>% drop_na()
  
  # 7-day rolling cumulative CWD
  train$cumulative_cwd_7day <- rollapply(train$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  
  # Drop NA after rolling
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  # Add 3-day lag variables
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  # Save back to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"),  test,  envir = .GlobalEnv)
}

```


```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
climate_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```



```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

climate_metrics$site <- factor(climate_metrics$site, levels = wet_dry_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Dry -> Wet)"
  ) +
  theme_minimal()

```
Range of R-squared across OzFlux sites. GAM used with VPD, Tair, Precip, cumulative cwd, VPD lag, Tair lag, leaf chlorophyll content and shortwave radiation. Only thing missing is LAI (leaf area index) which could improve model fit for some of the sites.

Just wanted to double check there wasn't underlying effect of no. of observations:
No of observations per site:
```{r}
length(unique(How_train$GPP))
length(unique(Ade_train$GPP))
length(unique(Das_train$GPP))
length(unique(Dap_train$GPP))
length(unique(Dry_train$GPP))
length(unique(Stp_train$GPP))
length(unique(Asm_train$GPP))
length(unique(Cum_train$GPP))
length(unique(Emr_train$GPP))
length(unique(Fog_train$GPP))
length(unique(Gin_train$GPP))
length(unique(Gww_train$GPP))
length(unique(Lox_train$GPP))
length(unique(Rdf_train$GPP))
length(unique(Rig_train$GPP))
length(unique(Rob_train$GPP))
length(unique(TTE_train$GPP))
length(unique(Wac_train$GPP))
length(unique(Whr_train$GPP))
length(unique(Wom_train$GPP))
length(unique(Ync_train$GPP))


```
```{r}
# Define wet → dry order
obs_order <- c("How", "Das", "Stp", "Dry", "Dap", "Ync", "Asm", "Rob", "Whr",
  "Gww", "Wom", "Fog", "Gin", "Wac", "Emr", "Cum", "Lox", "Rig",
  "Ade", "TTE", "Rdf")


climate_metrics$site <- factor(climate_metrics$site, levels = obs_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (least to most number of obs)"
  ) +
  theme_minimal()
```

Now let's look at how VPD improves model fit:

1. Model without VPD
2. VPD only model
3. Full model with VPD

Model without VPD
```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
no_vpd_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

no_vpd_metrics$site <- factor(no_vpd_metrics$site, levels = wet_dry_order)

ggplot(no_vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (no VPD) (Dry -> Wet)"
  ) +
  theme_minimal()
```
VPD only model

```{r}
# -----------------------------
# Function to fit GAM and calculate metrics
# -----------------------------
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(VPD) + s(VPD_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
vpd_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)
```

```{r}
# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

vpd_metrics$site <- factor(vpd_metrics$site, levels = wet_dry_order)

ggplot(vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (VPD only) (Dry -> Wet)"
  ) +
  theme_minimal()
```
Full model with VPD

```{r}
fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
          s(Tair_lag3) + s(VPD_lag3) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}

# -----------------------------
# Run GAM for all sites and combine results
# -----------------------------
full_metrics <- bind_rows(
  lapply(all_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

# Define wet → dry order
wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

full_metrics$site <- factor(full_metrics$site, levels = wet_dry_order)

ggplot(full_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (Dry -> Wet)"
  ) +
  theme_minimal()

```

compare models:
```{r}
library(dplyr)

gam_comparison <- no_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    vpd_metrics %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  inner_join(
    full_metrics %>%
      select(site, R2) %>%
      rename(R2_full = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_full - R2_no_vpd
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```

Not sure if there is any pattern here. 
TTE - interestingly VPD reduces site predictability. 
*Explore this further*. - think this could be interesting
Could be interesting to add in LAI to this, so there can be some sort of mechanistic interpretation of these results. 

Separate sites into PFT to visualize trends:

WSA - Ade, Gin, How, Rdf
SAV - DaS, Dry, Gww, Asm, Cpr
EBF - Cum, Rob, Wac, Whr, Wom, Tum
GRA - Dap, Emr, Rig, Stp, TTE, Ync
WET - Fog
DBF - Lox

```{r}
Ade_gs <- Ade_gs %>% mutate(site = "Ade")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
How_gs <- How_gs %>% mutate(site = "How")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Das_gs <- Das_gs %>% mutate(site = "Das")
Dry_gs <- Dry_gs %>% mutate(site = "Dry")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Asm_gs <- Asm_gs %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Dap_gs <- Dap_gs %>% mutate(site = "Dap")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Stp_gs <- Stp_gs %>% mutate(site = "Stp")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Tum_gs <- Tum_gs %>% mutate(site = "Tum")
Cpr_gs <- Cpr_gs %>% mutate(site = "Cpr")
```

```{r}
WSA <- bind_rows(Ade_gs, Gin_gs, How_gs, Rdf_gs) %>% group_by(site)

SAV <- bind_rows(Das_gs, Dry_gs, Gww_gs, Asm_gs, Cpr_gs) %>% group_by(site)

EBF <- bind_rows(Cum_gs, Rob_gs, Wac_gs, Whr_gs, Wom_gs, Tum_gs) %>% group_by(site)

GRA <- bind_rows(Dap_gs, Emr_gs, Rig_gs, Stp_gs, TTE_gs, Ync_gs) %>% group_by(site)

WET <- Fog_gs %>% group_by(site)

DBF <- Lox_gs %>% group_by(site)

```

```{r}
WSA$PFT <- "WSA"
SAV$PFT <- "SAV"
EBF$PFT <- "EBF"
GRA$PFT <- "GRA"
WET$PFT <- "WET"
DBF$PFT <- "DBF"

```

```{r}
PFT_all <- bind_rows(WSA, SAV, EBF, GRA, WET, DBF)

```


```{r}
library(ggplot2)

ggplot(PFT_all, aes(x = VPD, y = GPP, colour = PFT)) +

  geom_point(alpha = 0.05) +

  geom_smooth(method = "gam",
              formula = y ~ s(x, k = 10),
              se = FALSE,
              linewidth = 1.2) +

  theme_minimal(base_size = 14) +

  labs(
    x = "VPD",
    y = "GPP",
    colour = "PFT",
    title = "GPP–VPD relationships across PFTs"
  )

```
Mean functional response of PFT to atmospheric drying using GAM to characterise non-linear responses of GPP to VPD. 

Need to bin VPD because these lines look too smooth


Fit GAM for each PFT
- No VPD
- VPD only
- Full with VPD

What is the additive predictive value of VPD across plant functional groups?

```{r}
library(tidyverse)
library(mgcv)
library(zoo)
library(caret)

PFT_all <- c("GRA", "EBF", "SAV", "WSA", "WET", "DBF")

```

```{r}
fit_gam_models <- function(train_data, test_data) {
  
  # ---- Model formulas ----
  formula_no_vpd   <- GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
                      s(Tair_lag3) + s(LCC_daily)
  
  formula_vpd_only <- GPP ~ s(VPD)
  
  formula_full_vpd <- GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(cumulative_cwd_7day) +
                      s(Tair_lag3) + s(LCC_daily) + s(VPD)
  
  # ---- Fit models ----
  gam_no_vpd   <- gam(formula_no_vpd, data = train_data, method = "REML")
  gam_vpd_only <- gam(formula_vpd_only, data = train_data, method = "REML")
  gam_full_vpd <- gam(formula_full_vpd, data = train_data, method = "REML")
  
  # ---- Predictions ----
  preds_no_vpd   <- predict(gam_no_vpd, newdata = test_data)
  preds_vpd_only <- predict(gam_vpd_only, newdata = test_data)
  preds_full_vpd <- predict(gam_full_vpd, newdata = test_data)
  
  # ---- Calculate metrics ----
  tibble(
    Model = c("No_VPD", "VPD_only", "Full_VPD"),
    RMSE  = c(
      sqrt(mean((test_data$GPP - preds_no_vpd)^2)),
      sqrt(mean((test_data$GPP - preds_vpd_only)^2)),
      sqrt(mean((test_data$GPP - preds_full_vpd)^2))
    ),
    R2    = c(
      cor(test_data$GPP, preds_no_vpd)^2,
      cor(test_data$GPP, preds_vpd_only)^2,
      cor(test_data$GPP, preds_full_vpd)^2
    )
  )
}

```

```{r}
all_metrics <- map_dfr(PFT_all, function(PFT) {
  
  train_data <- get(paste0(PFT, "_train"))
  test_data  <- get(paste0(PFT, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(PFT = PFT)
})

```

```{r}
all_metrics %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```
```{r}
library(tidyr)
library(dplyr)

vpd_additive <- all_metrics %>%
  pivot_wider(names_from = Model, values_from = c(R2, RMSE)) %>%
  mutate(
    R2_gain_VPD = R2_Full_VPD - R2_No_VPD,
    RMSE_drop_VPD = RMSE_No_VPD - RMSE_Full_VPD
  )

```

```{r}
library(ggplot2)

ggplot(vpd_additive, aes(x = PFT, y = R2_gain_VPD, fill = PFT)) +
  geom_col() +
  theme_minimal(base_size = 14) +
  labs(
    y = "R² gain from VPD",
    x = "Plant Functional Type",
    title = "Additive predictive value of VPD on GPP"
  )

```

```{r}
ggplot(all_metrics %>% filter(Model %in% c("No_VPD", "Full_VPD")),
       aes(x = PFT, y = R2, colour = Model)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 14) +
  labs(
    y = "R²",
    x = "Plant Functional Type",
    title = "Predictive value of VPD across PFTs"
  )

```
```{r}
full_model_only <- all_metrics %>%
  filter(Model == "Full_VPD")
```

```{r}
library(ggplot2)

ggplot(full_model_only, aes(x = PFT, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Plant Functional Type",
    y = expression(R^2),
    title = "Full GAM predictability of GPP across PFTs"
  )

```
Predictability decreases when sites are grouped by PFT. 
- PFTs are heterogenous in their GPP-VPD sensitivty
- Plant functional groups do not behave as single functional units 

Predictive performance decreased when sites were aggregated by plant functional type, reflecting the substantial within-PFT heterogeneity in GPP responses to atmospheric and hydrological drivers. This suggests that PFT-level grouping conceals site-specific controls.

This supports emeging literature that PFTs are too broad - model fit was particularly low compared to site-specific GAMs. 

What if we grouped sites based on their annual precipitation?

Separate sites into very wet, wet, dry, and very dry? Is this better than separating into PFT in terms of predictability?

```{r}
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across Australian Fluxnet sites"
  )
```
23 sites 
4 groups, very wet, wet, dry and very dry
Grouped into 4 groups based on quantile to make it even (although a bit arbitrary)

```{r}
site_precip_class <- annual_summary %>%
  mutate(
    precip_quantile = ntile(annual_summary, 4),
    precip_class = case_when(
      precip_quantile == 1 ~ "Very wet",
      precip_quantile == 2 ~ "Wet",
      precip_quantile == 3 ~ "Dry",
      precip_quantile == 4 ~ "Very Dry"
    )
  )

```

```{r}
Very_wet <- bind_rows(Wac, Rob, How, Das, Fog, Rdf)
Wet <- bind_rows(Dap, Ade, Tum, Wom, Dry, Stp)
Dry <- bind_rows(Ync, Emr, Gin, Cum, Rig, Asm)
Very_dry <- bind_rows(Whr, Gww, Cpr, TTE, Lox)
```

```{r}
precip_groups <- c("Very_dry", "Dry", "Wet", "Very_wet")
```


```{r}
for(group in precip_groups){
  
  df <- get(group)
  
  idx <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  test <- test %>% drop_na()
  
  train$cumulative_cwd_7day <- rollapply(train$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  assign(paste0(group, "_train"), train, envir = .GlobalEnv)
  assign(paste0(group, "_test"),  test,  envir = .GlobalEnv)
}


```

Fit GAMs for each precip class
```{r}
all_metrics_precip <- map_dfr(precip_groups, function(group) {
  
  train_data <- get(paste0(group, "_train"))
  test_data  <- get(paste0(group, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(precip_class = group)
})

```

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```

```{r}
all_metrics_precip %>%
  filter(Model == "Full_VPD") %>%
  ggplot(aes(x = precip_class, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation regime",
    y = expression(R^2),
    title = "Full GAM predictability across precipitation classes"
  )

```
VPD additive value

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  select(precip_class, Model, R2) %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

ggplot(vpd_additive_precip,
       aes(x = precip_class, y = VPD_additive)) +
  geom_col(fill = "firebrick") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation class",
    y = expression(Delta*R^2~"(VPD contribution)"),
    title = "Additive contribution of VPD to GPP predictability"
  )

```
The addition of VPD improved model fit for all precipitation classes, more in dry classes than in wet classes. 

VPD improves predictability more on a wider, climate scale than at an individual site or PFT level scale. PFTs assume all forests/savannas/woody savannas behave similarly, however, within PFTs VPD-GPP relationships are heterogenous.

Precipitation class performs nearly as well as site-specific GAMS, highlighting the importance of hydrological footprint on ecosystem functioning and response. 


