---
title: "02_site_analysis"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Which Australian flux site is the most/least predictable?
- R-squared from model
- Change in R-squared when VPD is added to GAM
- Can VPD explain residual variation?
- VPD curves for sites/PFT groups?

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```
```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cpr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cpr_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cum_FLUXNET2015_FULLSET_HH_2012-2014_2-4.csv")

Emr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Emr_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Fog_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Fog_FLUXNET2015_FULLSET_HH_2006-2008_1-4.csv")

Gin_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Gin_FLUXNET2015_FULLSET_HH_2011-2014_1-4.csv")

Gww_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-GWW_FLUXNET2015_FULLSET_HH_2013-2014_1-4.csv")

Lox_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Lox_FLUXNET2015_FULLSET_HH_2008-2009_1-4.csv")

Rdf_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-RDF_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Rig_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rig_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Rob_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rob_FLUXNET2015_FULLSET_HH_2014-2014_1-4.csv")

TTE_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-TTE_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Wac_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wac_FLUXNET2015_FULLSET_HH_2005-2008_1-4.csv")

Whr_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Whr_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Wom_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wom_FLUXNET2015_FULLSET_HH_2010-2014_1-4.csv")

Ync_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ync_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")


```

```{r}
Tum_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/hourly_2/FLX_AU-Tum_FLUXNET2015_FULLSET_HR_2001-2014_2-4.csv")
```

Remove -9999 values

```{r}
flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_HH <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_HH <- Ade_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_HH <- Das_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_HH <- Dap_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_HH <- Dry_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_HH <- Stp_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_HH <- Asm_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Cum_HH <- Cum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Emr_HH <- Emr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Fog_HH <- Fog_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gin_HH <- Gin_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Gww_HH <- Gww_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Lox_HH <- Lox_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rdf_HH <- Rdf_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rig_HH <- Rig_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Rob_HH <- Rob_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

TTE_HH <- TTE_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wac_HH <- Wac_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Whr_HH <- Whr_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Wom_HH <- Wom_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ync_HH <- Ync_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Tum_HH <- Tum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
Tum_HH <- Tum_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
filter_low_coverage_days <- function(df, min_frac = 0.9, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    mutate(n_steps = n()) %>%
    ungroup() %>%
    filter(n_steps >= min_steps) %>%
    select(-n_steps)
}
```

```{r}
How_HH <- filter_low_coverage_days(How_HH)
Ade_HH <- filter_low_coverage_days(Ade_HH)
Das_HH <- filter_low_coverage_days(Das_HH)
Dap_HH <- filter_low_coverage_days(Dap_HH)
Dry_HH <- filter_low_coverage_days(Dry_HH)
Stp_HH <- filter_low_coverage_days(Stp_HH)
Asm_HH <- filter_low_coverage_days(Asm_HH)
Cum_HH <- filter_low_coverage_days(Cum_HH)
Emr_HH <- filter_low_coverage_days(Emr_HH)
Fog_HH <- filter_low_coverage_days(Fog_HH)
Gin_HH <- filter_low_coverage_days(Gin_HH)
Gww_HH <- filter_low_coverage_days(Gww_HH)
Lox_HH <- filter_low_coverage_days(Lox_HH)
Rdf_HH <- filter_low_coverage_days(Rdf_HH)
Rig_HH <- filter_low_coverage_days(Rig_HH)
Rob_HH <- filter_low_coverage_days(Rob_HH)
TTE_HH <- filter_low_coverage_days(TTE_HH)
Wac_HH <- filter_low_coverage_days(Wac_HH)
Wom_HH <- filter_low_coverage_days(Wom_HH)
Whr_HH <- filter_low_coverage_days(Whr_HH)
Ync_HH <- filter_low_coverage_days(Ync_HH)
Tum_HH <- filter_low_coverage_days(Tum_HH)
```

```{r}
Tum_HH <- filter_low_coverage_days(Tum_HH)
```


Convert to daily:

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

Convert hourly to daily (AU_Tum)
```{r}
library(tidyverse)
library(lubridate)

convert_hourly_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HOUR <- 3600.
  WM2_TO_KG_M2_S <- 1.0 / (2.45 * 1E6)
  PA_2_KPA <- 0.001
  HPA_2_KPA <- 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HOUR,
      VPD = VPD * HPA_2_KPA
    )

  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df)
}

```

```{r}
How_daily <- convert_to_daily(How_HH)
Ade_daily <- convert_to_daily(Ade_HH)
Das_daily <- convert_to_daily(Das_HH)
Dap_daily <- convert_to_daily(Dap_HH)
Dry_daily <- convert_to_daily(Dry_HH)
Stp_daily <- convert_to_daily(Stp_HH)
Asm_daily <- convert_to_daily(Asm_HH)
Cum_daily <- convert_to_daily(Cum_HH)
Emr_daily <- convert_to_daily(Emr_HH)
Fog_daily <- convert_to_daily(Fog_HH)
Gin_daily <- convert_to_daily(Gin_HH)
Gww_daily <- convert_to_daily(Gww_HH)
Lox_daily <- convert_to_daily(Lox_HH)
Rdf_daily <- convert_to_daily(Rdf_HH)
Rig_daily <- convert_to_daily(Rig_HH)
Rob_daily <- convert_to_daily(Rob_HH)
TTE_daily <- convert_to_daily(TTE_HH)
Wac_daily <- convert_to_daily(Wac_HH)
Whr_daily <- convert_to_daily(Whr_HH)
Wom_daily <- convert_to_daily(Wom_HH)
Ync_daily <- convert_to_daily(Ync_HH)
Tum_daily <- convert_hourly_to_daily(Tum_HH)
```

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Cum_daily <- Cum_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Emr_daily <- Emr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Fog_daily <- Fog_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gin_daily <- Gin_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Gww_daily <- Gww_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Lox_daily <- Lox_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rdf_daily <- Rdf_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rig_daily <- Rig_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Rob_daily <- Rob_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

TTE_daily <- TTE_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wac_daily <- Wac_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Whr_daily <- Whr_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Wom_daily <- Wom_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ync_daily <- Ync_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

```


```{r}
plot(Wac_daily$VPD)
```
Plot annual precipitation for each site - to find dry - wet gradient of sites

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

  df_annual <- df_daily %>%
    group_by(year) %>%
    summarise(
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df_annual)
}

How_a <- convert_to_annual(How_daily)
Ade_a <- convert_to_annual(Ade_daily)
Das_a <- convert_to_annual(Das_daily)
Dap_a <- convert_to_annual(Dap_daily)
Dry_a <- convert_to_annual(Dry_daily)
Stp_a <- convert_to_annual(Stp_daily)
Asm_a <- convert_to_annual(Asm_daily)
Cum_a <- convert_to_annual(Cum_daily)
Emr_a <- convert_to_annual(Emr_daily)
Fog_a <- convert_to_annual(Fog_daily)
Gin_a <- convert_to_annual(Gin_daily)
Gww_a <- convert_to_annual(Gww_daily)
Lox_a <- convert_to_annual(Lox_daily)
Rdf_a <- convert_to_annual(Rdf_daily)
Rig_a <- convert_to_annual(Rig_daily)
Rob_a <- convert_to_annual(Rob_daily)
TTE_a <- convert_to_annual(TTE_daily)
Wac_a <- convert_to_annual(Wac_daily)
Whr_a <- convert_to_annual(Whr_daily)
Wom_a <- convert_to_annual(Wom_daily)
Ync_a <- convert_to_annual(Ync_daily)

```

```{r}
library(tidyverse)

annual_precip <- bind_rows(
  How_a %>% mutate(site = "How"),
  Ade_a %>% mutate(site = "Ade"),
  Das_a %>% mutate(site = "Das"),
  Dap_a %>% mutate(site = "Dap"),
  Dry_a %>% mutate(site = "Dry"),
  Stp_a %>% mutate(site = "Stp"),
  Asm_a %>% mutate(site = "Asm"),
  Cum_a %>% mutate(site = "Cum"),
  Emr_a %>% mutate(site = "Emr"),
  Fog_a %>% mutate(site = "Fog"),
  Gin_a %>% mutate(site = "Gin"),
  Gww_a %>% mutate(site = "Gww"),
  Lox_a %>% mutate(site = "Lox"),
  Rdf_a %>% mutate(site = "Rdf"),
  Rig_a %>% mutate(site = "Rig"),
  Rob_a %>% mutate(site = "Rob"),
  TTE_a %>% mutate(site = "TTE"),
  Wac_a %>% mutate(site = "Wac"),
  Whr_a %>% mutate(site = "Whr"),
  Wom_a %>% mutate(site = "Wom"),
  Ync_a %>% mutate(site = "Ync")
)

```

```{r}
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across OzFlux sites"
  )

```
```{r}
#growing season - this time i am keeping the top 75% productive periods per site because hard threshold is not fair across different ecosystems

library(dplyr)

filter_gs <- function(df, gpp_quantile = 0.25, swrad_thresh = 20) {

  df %>%
    # remove nighttime
    filter(SW_rad > swrad_thresh) %>%
    
    # global GPP filtering
    filter(GPP > quantile(GPP, gpp_quantile, na.rm = TRUE))
}


#Apply to dataset

How_gs <- filter_gs(How_daily)
Ade_gs <- filter_gs(Ade_daily)
Das_gs <- filter_gs(Das_daily)
Dap_gs <- filter_gs(Dap_daily)
Dry_gs <- filter_gs(Dry_daily)
Stp_gs <- filter_gs(Stp_daily)
Asm_gs <- filter_gs(Asm_daily)
Cum_gs <- filter_gs(Cum_daily)
Emr_gs <- filter_gs(Emr_daily)
Fog_gs <- filter_gs(Fog_daily)
Gin_gs <- filter_gs(Gin_daily)
Gww_gs <- filter_gs(Gww_daily)
Lox_gs <- filter_gs(Lox_daily)
Rdf_gs <- filter_gs(Rdf_daily)
Rig_gs <- filter_gs(Rig_daily)
Rob_gs <- filter_gs(Rob_daily)
TTE_gs <- filter_gs(TTE_daily)
Wac_gs <- filter_gs(Wac_daily)
Wom_gs <- filter_gs(Wom_daily)
Ync_gs <- filter_gs(Ync_daily)
Whr_gs <- filter_gs(Whr_daily)
```

1. 
GPP ~ VPD curve for all sites:

```{r}
library(dplyr)
library(ggplot2)

# Add a 'site' column to each dataframe
How_gs  <- How_gs  %>% mutate(site = "How")
Ade_gs  <- Ade_gs  %>% mutate(site = "Ade")
Das_gs  <- Das_gs  %>% mutate(site = "Das")
Dap_gs  <- Dap_gs  %>% mutate(site = "Dap")
Dry_gs  <- Dry_gs  %>% mutate(site = "Dry")
Stp_gs  <- Stp_gs  %>% mutate(site = "Stp")
Asm_gs  <- Asm_gs  %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")

# Combine all sites
all_sites <- bind_rows(How_gs, Ade_gs, Das_gs, Dap_gs, Dry_gs, Stp_gs, Asm_gs, Cum_gs, Emr_gs, Fog_gs, Gin_gs, Gww_gs, Lox_gs, Rdf_gs, Rig_gs, Rob_gs, TTE_gs, Wac_gs, Wom_gs, Ync_gs, Whr_gs)

# Set wet → dry order
all_sites$site <- factor(all_sites$site,
                         levels = c("Wac", "Rob", "How", "Das", "Fog", "Rdf", "Dap", "Ade", "Wom", "Dry", "Stp","Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "TTE", "Lox"))
```

```{r}
library(RColorBrewer)
# Define 7 distinct colors for the sites
site_colors <- brewer.pal(n = 9, name = "Set1")  # Set1 is bright and categorical

ggplot(all_sites, aes(x = VPD, y = GPP, color = site)) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.2, span = 0.3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "GPP response to VPD across OzFlux sites",
    x = "VPD (kPa)",
    y = "GPP (g C m-2 d-1)"
  )

```
Overall, we can see that the drier sites are less sensitive to higher VPD. 

2. 
Full GAM (climate + LCC + VPD) for each site
Plot sites r-squared by same wetness 
- Load in pet/cwd and LCC













Separate into PFTs/groups/precip gradient:
1. VPD only model and full model without VPD 
2. Compare with previous model of climate + VPD + physiology
3. Added predictive value of VPD?
