---
title: "02_site_analysis"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated data cleaning pipeline:
1. Load in flux data
2. store in sites_clean
3. Remove -9999 values
4. Filter low coverage days
5. Convert to daily
6. NaN back to NA
7. Growing season filter
8. Add in pet and cwd 

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_F,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      VPD_QC = VPD_F_QC
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "VPD_QC")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

Load in 21 sites across Aus
```{r}
How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Emr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Emr_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Fog <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Fog_FLUXNET2015_FULLSET_HH_2006-2008_1-4.csv")

Gin <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Gin_FLUXNET2015_FULLSET_HH_2011-2014_1-4.csv")

Gww <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-GWW_FLUXNET2015_FULLSET_HH_2013-2014_1-4.csv")

Rdf <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-RDF_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Rig <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rig_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Wac <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wac_FLUXNET2015_FULLSET_HH_2005-2008_1-4.csv")

Whr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Whr_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Ync <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ync_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Ade <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Boy <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/TERN/TERN_AU-Boy_FLUXNET_FLUXMET_HH_2017-2024_v1.3_r1.csv")

Wom <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/TERN/TERN_AU-Wom_FLUXNET_FLUXMET_HH_2010-2024_v1.3_r1.csv")

Cpr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/TERN/TERN_AU-Cpr_FLUXNET_FLUXMET_HH_2010-2024_v1.3_r1.csv")

Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/TERN/TERN_AU-ASM_FLUXNET_FLUXMET_HH_2010-2024_v1.3_r1.csv")

Cum <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/TERN/TERN_AU-Cum_FLUXNET_FLUXMET_HH_2014-2024_v1.3_r1.csv")

Tum <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/hourly_2/FLX_AU-Tum_FLUXNET2015_FULLSET_HR_2001-2014_2-4.csv") #remember AU-Tum is hourly

```

Remove -9999 values:

```{r}
#make a function that removes -9999 values to NA

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD", "VPD_QC")

flux_nas <- function(df, flux_vars, na_vals = c(-9999, -9999.9)) {
  df %>%
    mutate(across(
      all_of(flux_vars),
      ~replace(., . %in% na_vals, NA)
    ))
}
```


```{r}
#names of datasets I have loaded in so R knows what I'm yapping about

site_names <- c("How", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr", "Fog", "Gin", "Gww", "Rdf", "Rig", "Wac", "Whr", "Wom", "Ync", "Cpr", "Tum", "Ade", "Boy")

#pulls the objects from the environment - sites is a list of dataframes
sites <- mget(site_names)

#sites_clean is another list of dataframes, every -9999 has been replaced by NA, only work with site_clean now
sites_clean <- map(sites, ~flux_nas(.x, flux_vars))

```


Filter low coverage days:

```{r}

library(dplyr)
library(lubridate)
library(purrr)

filter_low_coverage_days <- function(df, min_frac = 0.8, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    filter(n() >= min_steps) %>%
    ungroup()
}

sites_clean <- sites_clean %>%
  imap(~ filter_low_coverage_days(.x, steps_per_day = if (.y=="Tum") 24 else 48))

#Basically saying if it is Tum, it is hourly, so should be 24 not 48. There might be a quicker way to do this, but this is what it will be for now. For reproducibility, just plonk in the known hourly sites into this part. 
  
```

So all the cleaned AU dataframes are in the object sites_clean. The external site in the environment is the raw data

Convert to daily:

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  #I have added in GPP_n_valid - basically tracks how many values are valid (should be 48 or 24 with Tum), if a day has a lot of missing GPP, the daily total sum may be smaller than it actually is, producing an underestimated GPP
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP_n_valid = sum(!is.na(GPP)),
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

Convert hourly to daily (AU_Tum)
```{r}
library(tidyverse)
library(lubridate)

convert_hourly_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HOUR <- 3600.
  WM2_TO_KG_M2_S <- 1.0 / (2.45 * 1E6)
  PA_2_KPA <- 0.001
  HPA_2_KPA <- 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HOUR,
      VPD = VPD * HPA_2_KPA
    )

  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP_n_valid = sum(!is.na(GPP)),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df)
}

```

```{r}
#so basically, apply hourly to daily if it equals Tum, but if not apply half-hourly to daily conversion. Save in a list called sites_daily

sites_daily <- sites_clean %>%
  imap(~ if (.y == "Tum") {
    convert_hourly_to_daily(.x)
  } else {
    convert_to_daily(.x)
  })
```


```{r}
#convert everything back to NA

sites_daily <- sites_daily %>%
  map(~ .x %>% mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .))))

```

Growing season:

```{r}
#growing season - for now, it is going to be above 0 GPP

filter_gs <- function(df) {
  df %>%
    filter(
      GPP > 0
    )
}

sites_daily <- sites_daily %>%
  map(filter_gs)
```

Add in water stress component (pet and cwd) 
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

sites_daily <- sites_daily %>%
  map(~ .x %>%
        mutate(
          pet = priestley_taylor_pet(Rn, Tair, Pa),
          cwd = pet - ET
        ))
 
```

```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

sites_daily <- sites_daily %>%
  map(add_year_doy)
```


Add in LAI data

```{r}
library(tidyverse)
library(data.table)
library(zoo)

sites_daily <- imap(sites_daily, function(df, name) {
  df %>% mutate(site = name)
})
```

```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/site_LAI_MCD15A3H_2000_2024_3.csv") %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))
```

```{r}
lai_data <- lai_data %>%
  filter(site != "AU-Rob")
```


```{r}
cat("LAI site names:\n")
print(unique(lai_data$site))
cat("\nFlux site names:\n")
print(names(sites_daily))
```

```{r}
sites_daily <- map(sites_daily, function(df) {
  df %>% mutate(date = as.Date(date))
})
```

Add site name back in
```{r}
sites_daily <- imap(sites_daily, function(df, name) {
  df %>% mutate(site = name)
})
```


```{r}
# Step 5: Exact join first to preserve NAs between 4-day obs
sites_daily <- map(sites_daily, function(df) {
  site_name <- df$site[1]
  
  lai_site <- lai_data %>%
    filter(site == site_name) %>%
    select(date, mean) %>%
    rename(LAI_4day = mean) %>%
    arrange(date)
  
  if (nrow(lai_site) == 0) {
    warning(paste("No LAI data found for site:", site_name))
    return(df %>% mutate(LAI_4day = NA_real_, LAI_filled = NA_real_))
  }
  
  # Exact join - non-matching dates get NA, which is what we want
  left_join(df, lai_site, by = "date")
})

# Step 6: Interpolate between the 4-day obs
sites_daily <- map(sites_daily, function(df) {
  df %>%
    arrange(date) %>%
    mutate(LAI_filled = na.approx(LAI_4day, x = date, maxgap = 8, na.rm = FALSE))
})
  
```

```{r}
# Step 7: Check how many valid LAI days each site has
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  n_valid <- sum(!is.na(df$LAI_filled))
  n_total <- nrow(df)
  cat(sprintf("%s - valid LAI days: %d / %d\n", name, n_valid, n_total))
})

# Step 8: Visual check - line = interpolated, red dots = raw 4-day obs
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  plot(df$date, df$LAI_filled,
       main  = name,
       xlab  = "Date", 
       ylab  = "LAI",
       type  = "l", 
       col   = "darkgreen",
       ylim  = range(c(df$LAI_filled, df$LAI_4day), na.rm = TRUE))
  points(df$date, df$LAI_4day, col = "red", pch = 16, cex = 0.5)
  legend("topright", 
         legend = c("Interpolated", "Raw 4-day"),
         col    = c("darkgreen", "red"), 
         lty    = c(1, NA), 
         pch    = c(NA, 16))
})
```

```{r}
# get max LAI per site from sites_daily
canopy_order <- map_dfr(sites_daily, function(df) {
  tibble(
    site = unique(df$site),
    max_LAI = max(df$LAI_4day, na.rm = TRUE)
  )
}) %>%
  arrange(desc(max_LAI))

print(canopy_order)
```

```{r}
ggplot(canopy_order, aes(x = reorder(site, desc(max_LAI)), y = max_LAI)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(x = "Site", y = "Mean LAI") +
  theme_classic()
```


