---
title: "02_site_analysis"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated data cleaning pipeline:
1. Load in flux data
2. store in sites_clean
3. Remove -9999 values
4. Filter low coverage days
5. Convert to daily
6. NaN back to NA
7. Growing season filter

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

Load in 23 sites across Aus
```{r}
How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Cum <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cum_FLUXNET2015_FULLSET_HH_2012-2014_2-4.csv")

Emr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Emr_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Fog <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Fog_FLUXNET2015_FULLSET_HH_2006-2008_1-4.csv")

Gin <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Gin_FLUXNET2015_FULLSET_HH_2011-2014_1-4.csv")

Gww <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-GWW_FLUXNET2015_FULLSET_HH_2013-2014_1-4.csv")

Lox <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Lox_FLUXNET2015_FULLSET_HH_2008-2009_1-4.csv")

Rdf <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-RDF_FLUXNET2015_FULLSET_HH_2011-2013_1-4.csv")

Rig <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rig_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Rob <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Rob_FLUXNET2015_FULLSET_HH_2014-2014_1-4.csv")

TTE <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-TTE_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Wac <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wac_FLUXNET2015_FULLSET_HH_2005-2008_1-4.csv")

Whr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Whr_FLUXNET2015_FULLSET_HH_2011-2014_2-4.csv")

Wom <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Wom_FLUXNET2015_FULLSET_HH_2010-2014_1-4.csv")

Ync <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ync_FLUXNET2015_FULLSET_HH_2012-2014_1-4.csv")

Cpr <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Cpr_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

Tum <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/hourly_2/FLX_AU-Tum_FLUXNET2015_FULLSET_HR_2001-2014_2-4.csv") #remember AU-Tum is hourly

```

Remove -9999 values:

```{r}
#make a function that removes -9999 values to NA

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

flux_nas <- function(df, flux_vars, na_vals = c(-9999, -9999.9)) {
  df %>%
    mutate(across(
      all_of(flux_vars),
      ~replace(., . %in% na_vals, NA)
    ))
}
```


```{r}
#names of datasets I have loaded in so R knows what I'm yapping about

site_names <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr", "Fog", "Gin", "Gww", "Lox", "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr", "Wom", "Ync", "Cpr", "Tum")

#pulls the objects from the environment - sites is a list of dataframes
sites <- mget(site_names)

#sites_clean is another list of dataframes, every -9999 has been replaced by NA, only work with site_clean now
sites_clean <- map(sites, ~flux_nas(.x, flux_vars))

```
Filter low coverage days:

```{r}

library(dplyr)
library(lubridate)
library(purrr)

filter_low_coverage_days <- function(df, min_frac = 0.8, steps_per_day = 48) {

  min_steps <- ceiling(min_frac * steps_per_day)

  df %>%
    group_by(date = floor_date(date, "day")) %>%
    filter(n() >= min_steps) %>%
    ungroup()
}

sites_clean <- sites_clean %>%
  imap(~ filter_low_coverage_days(.x, steps_per_day = if (.y=="Tum") 24 else 48))

#Basically saying if it is Tum, it is hourly, so should be 24 not 48. There might be a quicker way to do this, but this is what it will be for now. For reproducibility, just plonk in the known hourly sites into this part. 
  
```

So all the cleaned AU dataframes are in the object sites_clean. The external site in the environment is the raw data

Convert to daily:

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )


  # Resample to daily
  #I have added in GPP_n_valid - basically tracks how many values are valid (should be 48 or 24 with Tum), if a day has a lot of missing GPP, the daily total sum may be smaller than it actually is, producing an underestimated GPP
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP_n_valid = sum(!is.na(GPP)),
      GPP = sum(GPP, na.rm=TRUE),
      ET = sum(ET, na.rm=TRUE),
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )

  return(df)
}
```

Convert hourly to daily (AU_Tum)
```{r}
library(tidyverse)
library(lubridate)

convert_hourly_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HOUR <- 3600.
  WM2_TO_KG_M2_S <- 1.0 / (2.45 * 1E6)
  PA_2_KPA <- 0.001
  HPA_2_KPA <- 0.1

  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HOUR,
      VPD = VPD * HPA_2_KPA
    )

  # Resample to daily
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(date) %>%
    summarise(
      GPP_n_valid = sum(!is.na(GPP)),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE),
      Precip = sum(Precip, na.rm = TRUE),
      Qle = mean(Qle, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE),
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      VPD = mean(VPD, na.rm = TRUE),
      .groups = "drop"
    )

  return(df)
}

```

```{r}
#so basically, apply hourly to daily if it equals Tum, but if not apply half-hourly to daily conversion. Save in a list called sites_daily

sites_daily <- sites_clean %>%
  imap(~ if (.y == "Tum") {
    convert_hourly_to_daily(.x)
  } else {
    convert_to_daily(.x)
  })
```

```{r}
plot(sites_daily$Gww$GPP)
```

```{r}
#convert everything back to NA

sites_daily <- sites_daily %>%
  map(~ .x %>% mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .))))

```

Growing season:

```{r}
#growing season - for now, it is going to be above a certain temp threshold 5°C and above a short-wave radiation threshold

filter_gs <- function(df) {
  df %>%
    filter(
      GPP >= 0
    )
}

sites_daily <- sites_daily %>%
  map(filter_gs)
```

Add in water stress component (pet and cwd) and leaf chlorophyll content (LCC)
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

sites_daily <- sites_daily %>%
  map(~ .x %>%
        mutate(
          pet = priestley_taylor_pet(Rn, Tair, Pa),
          cwd = pet - ET
        ))
 
```

```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

sites_daily <- sites_daily %>%
  map(add_year_doy)
```

```{r}
library(dplyr)
library(lubridate)
library(purrr)
library(terra)
library(ncdf4)

gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# --- 1. Functions to get VCmax and daily LCC ---
get_vcmax <- function(lat, lon, vcmax_raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(vcmax_raster, coords)
  as.numeric(vcmax[1,2])
}

latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180 / nlines  
  line <- round((90 - lat) / lat_res)
  lon_res <- 360 / nsamples
  sample <- round((lon + 180) / lon_res)
  list(
    line   = pmin(pmax(line, 1), nlines),
    sample = pmin(pmax(sample, 1), nsamples)
  )
}

get_daily_LCC <- function(lcc_nc, lat, lon) {
  nc <- nc_open(lcc_nc)
  coords <- latlon_to_line_sample(lat, lon)
  
  pixel <- ncvar_get(nc, "chl_all",
                     start = c(coords$line, coords$sample, 1, 1),
                     count = c(1, 1, -1, -1))
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel[pixel == fillvalue] <- NA
  nc_close(nc)
  
  LCC <- pixel / 100
  weekly_clim <- rowMeans(LCC, na.rm = TRUE)
  
  weeks <- 1:length(weekly_clim)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  approx(weeks, weekly_clim, xout = daily_weeks)$y
}

# --- 2. Function to attach VCmax and LCC to a site dataframe ---
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc) {
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  daily_LCC <- get_daily_LCC(lcc_nc, lat, lon)  # ~365 values
  
  df %>%
    mutate(
      vcmax = vcmax_val,
      LCC_daily = daily_LCC[yday(date)],  # index by actual DOY
      year = year(date),
      doy = yday(date)
    )
}

# --- 3. Map over sites_daily using site metadata ---
site_meta <- tibble(
  name = names(sites_daily),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283, -33.6152, -23.8587,
           -12.5452, -31.3764, -30.1913, -34.4704, -14.5636, -36.6499, -17.1175, -22.287, -37.4259,
           -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 120.6541, 140.6551, 132.4776, 145.5759, 145.6301, 133.64, 145.1878,
           145.0294, 144.0944, 146.2907, 140.5891, 148.1517)
)

sites_daily <- pmap(
  list(sites_daily, site_meta$lat, site_meta$lon),
  function(df, lat, lon) attach_site_data(df, lat, lon, vcmax_raster, lcc_nc)
)

```

```{r}
list2env(
  setNames(sites_daily, paste0(names(sites_daily), "_daily")),
  envir = .GlobalEnv
)
```




