---
title: "experiment"
output: html_document
date: "2026-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Highlighting issue with precipitation in Howard Springs. 
Note: VPD looks fine at the end but precipitation still shows really high values

Load in half-hourly flux site:

```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```

Half-hourly Howard Springs

```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")
```

Summarise raw half-hourly data:

```{r}
summary(How_HH$Precip)
plot(How_HH$Precip)
```

```{r}
summary(How_HH$VPD)
plot(How_HH$VPD)
```

```{r}
summary(How_HH$Tair)
```

```{r}
summary(How_HH$Rn)
```

Remove -9999 values

```{r}

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_1 <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
summary(How_1$Precip)
quantile(How_1$Precip, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$VPD)
quantile(How_1$VPD, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$Tair)
quantile(How_1$Tair, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$Qle)
quantile(How_1$Qle, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$GPP)
quantile(How_1$GPP, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$SW_rad)
quantile(How_1$SW_rad, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$Rn)
quantile(How_1$Rn, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

summary(How_1$Pa)
quantile(How_1$Pa, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)
```
Table so easier to see:
```{r}
vars <- c("Precip", "VPD", "Tair", "Qle", "GPP", "SW_rad", "Rn", "Pa")

summary_table <- do.call(rbind, lapply(vars, function(v) {
  
  x <- How_1[[v]]
  
  c(
    Min    = min(x, na.rm = TRUE),
    Q25    = quantile(x, 0.25, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Mean   = mean(x, na.rm = TRUE),
    Q75    = quantile(x, 0.75, na.rm = TRUE),
    Max    = max(x, na.rm = TRUE),
    Q90    = quantile(x, 0.90, na.rm = TRUE),
    Q99    = quantile(x, 0.99, na.rm = TRUE)
  )
}))

summary_table <- as.data.frame(summary_table)
summary_table$Variable <- vars

summary_table <- summary_table[, c("Variable", setdiff(names(summary_table), "Variable"))]

summary_table

```
Assuming there is instrument failure for precipitation, remove 99th percentile values. Controlling for obvious outliers before converting to daily. 

```{r}
library(dplyr)

precip_thresh <- quantile(How_1$Precip, 0.99, na.rm = TRUE)

How_clean <- How_1 %>%
  filter(Precip <= precip_thresh)

```

```{r}
plot(How_clean$Precip)
```
Now, it looks a lot more reasonable?

Convert cleaned data to daily:

```{r}
 ####
  ##  Convert the 30 min to daily ... someone should add something to check 
  ##. if the data is hourly ...
  ##
  ##
  ##  author: Martin De Kauwe
  ##  date: 27th February, 2025
  ##  email: mdekauwe@gmail.com
  ####
  library(tidyverse)
  
  convert_to_daily <- function(df) {
  
    # Conversion constants
    UMOL_TO_MOL <- 1E-6
    MOL_C_TO_GRAMS_C <- 12.0
    SEC_2_HLFHOUR <- 1800.
    WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
    PA_2_KPA = 0.001
    HPA_2_KPA = 0.1
    
    df <- df %>%
      mutate(
        GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
        Precip = Precip * SEC_2_HLFHOUR,
        ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
        VPD = VPD * HPA_2_KPA
      )
    
    df$date <- ymd_hms(df$date)
    
    # Resample to daily 
    df <- df %>%
      mutate(date = floor_date(date, "day")) %>%  
      group_by(date) %>%
      summarise(
        GPP = sum(GPP, na.rm=TRUE), 
        ET = sum(ET, na.rm=TRUE),  
        Precip = sum(Precip, na.rm=TRUE),
        Qle = mean(Qle, na.rm=TRUE),
        Tair = mean(Tair, na.rm=TRUE),
        SW_rad = mean(SW_rad, na.rm=TRUE),
        Rn = mean(Rn, na.rm=TRUE),
        Pa = mean(Pa, na.rm=TRUE),
        VPD = mean(VPD, na.rm=TRUE),
        .groups = "drop"
      )
    
    return(df)
  }
```

```{r}
How_daily <- convert_to_daily(How_clean)
```
```{r}
plot(How_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm/day", main = "Howard springs")
```
One outlier

```{r}
# 99th percentile threshold
daily_thresh <- quantile(How_daily$Precip, 0.99, na.rm = TRUE)
daily_thresh

```

Remove 99th percentile, see what happens:

```{r}
How_daily_clean <- How_daily %>%
  mutate(
    Precip = ifelse(Precip > daily_thresh, NA, Precip)
  )
```

```{r}
plot(How_daily_clean$Precip, ylab = "precipitation mm/day")
```
Still looks really large for mm/day?

```{r}
plot(How_daily$VPD)
```
VPD looks normal
