---
title: "GAM_analysis"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

How does predictability vary across ecosystems in Australia, how does VPD affect predictability?

Workflow:
- Model (climate)
- Model (climate + VPD)
- Model (climate + VPD + lags)
- Model (climate + VPD + lags + LCC)


Across different scales:
- Sites
- PFTs
- Mean annual precipitation gradient
- Daily precipitation variability

1. Firstly start off training and testing within site
2. Train outside of site - how well do these models generalise
3. Hold some sites back - maybe 1 from each PFT/Precip regime

Load in water stress component
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

sites_daily <- sites_daily %>%
  map(~ .x %>%
        mutate(
          pet = priestley_taylor_pet(Rn, Tair, Pa),
          cwd = pet - ET
        ))
 
```

put updated sites back into global environment

```{r}
list2env(sites_daily, envir = .GlobalEnv)

```

Add in day of year and year again for the LCC data:
```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

sites_daily <- sites_daily %>%
  map(add_year_doy)

```

Now add that back into GE:
```{r}
list2env(sites_daily, envir = .GlobalEnv)
```
```{r}
library(dplyr)
library(lubridate)
library(purrr)
library(terra)
library(ncdf4)

# --- 1. Functions to get VCmax and daily LCC ---
get_vcmax <- function(lat, lon, raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(raster, coords)
  as.numeric(vcmax[1,2])
}

latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180 / nlines  
  line <- round((90 - lat) / lat_res)
  lon_res <- 360 / nsamples
  sample <- round((lon + 180) / lon_res)
  list(
    line   = pmin(pmax(line, 1), nlines),
    sample = pmin(pmax(sample, 1), nsamples)
  )
}

get_daily_LCC <- function(nc_file, lat, lon) {
  nc <- nc_open(nc_file)
  coords <- latlon_to_line_sample(lat, lon)
  
  pixel <- ncvar_get(nc, "chl_all",
                     start = c(coords$line, coords$sample, 1, 1),
                     count = c(1, 1, -1, -1))
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel[pixel == fillvalue] <- NA
  nc_close(nc)
  
  LCC <- pixel / 100
  weekly_clim <- rowMeans(LCC, na.rm = TRUE)
  
  weeks <- 1:length(weekly_clim)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  approx(weeks, weekly_clim, xout = daily_weeks)$y
}

# --- 2. Function to attach VCmax and LCC to a site dataframe ---
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc_file) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  daily_LCC <- get_daily_LCC(lcc_nc_file, lat, lon)
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df %>%
    mutate(
      vcmax      = vcmax_val,
      LCC_daily  = daily_LCC_full,
      year       = year(date),
      doy        = yday(date)
    )
}

# --- 3. Map over sites_daily using site metadata ---
site_meta <- tibble(
  name = names(sites_daily),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283, -33.6152, -23.8587,
           -12.5452, -31.3764, -30.1913, -34.4704, -14.5636, -36.6499, -17.1175, -22.287, -37.4259,
           -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 120.6541, 140.6551, 132.4776, 145.5759, 145.6301, 133.64, 145.1878,
           145.0294, 144.0944, 146.2907, 140.5891, 148.1517)
)

sites_daily <- pmap(
  list(sites_daily, site_meta$lat, site_meta$lon),
  function(df, lat, lon) attach_site_data(df, lat, lon, vcmax_raster, lcc_nc)
)

```

```{r}
# Push back to global environment
list2env(sites_daily, envir = .GlobalEnv)

```

Everything loaded in properly.



1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility


# List all sites
aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")


# Split, rolling CWD, lagged variables
for(site in aus_sites){
  
  df <- get(paste0(site, "_HH_daily"))
  
  # Split into train/test
  idx <- createDataPartition(df$GPP, p=0.8, list=FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  # Drop NA in test
  test <- test %>% drop_na()
  
  # 7-day rolling cumulative CWD
  train$cumulative_cwd_7day <- rollapply(train$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width=7, FUN=sum, 
                                         align="right", fill=NA, na.rm=TRUE)
  
  # Drop NA after rolling
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  # Add 7-day lag variables
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 7),
      VPD_lag3  = lag(VPD, 7)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  # Save back to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"),  test,  envir = .GlobalEnv)
}

```


```{r}

#Model 1 - climate only

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```



```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_metrics$site <- factor(climate_metrics$site, levels = precip_variability_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (high cv to low cv)"
  ) +
  theme_minimal()

```

Seems to be trend of declining R-squared from high precipitation coefficient of variation to low coefficient of variation. 

```{r}
#Model 2 - climate and VPD

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_vpd_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_vpd_metrics$site <- factor(climate_vpd_metrics$site, levels = wet_dry_order)

ggplot(climate_vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```
```{r}
#compare the no VPD and climate + VPD model

library(dplyr)

gam_comparison <- climate_metrics %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    climate_vpd_metrics %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_VPD - R2_no_vpd
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```
```{r}
#Model 3 - climate and VPD + lags

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag3) + s(Tair_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

lag_metrics$site <- factor(lag_metrics$site, levels = wet_dry_order)

ggplot(lag_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```
```{r}

library(dplyr)

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```
```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of climate lags across OzFlux sites"
  )

```
```{r}
#Model 4 - climate and VPD + lags and cumulative water deficit

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag3) + s(Tair_lag3) + s(cumulative_cwd_7day),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

cwd_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

cwd_metrics$site <- factor(cwd_metrics$site, levels = wet_dry_order)

ggplot(cwd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```
Now need to do additive value of LCC
- GAMs across PFTs
- GAMs across precipitation gradients

----------------------------------------------------------------------------------------


Not sure if there is any pattern here. 
TTE - interestingly VPD reduces site predictability. 
*Explore this further*. - think this could be interesting
Could be interesting to add in LAI to this, so there can be some sort of mechanistic interpretation of these results. 

2. Separate sites into PFT to visualize trends:

WSA (woody savanna) - Ade, Gin, How, Rdf
SAV (savanna) - DaS, Dry, Gww, Asm, Cpr
EBF (evergreen broadleaf forest) - Cum, Rob, Wac, Whr, Wom, Tum
GRA (grassland) - Dap, Emr, Rig, Stp, TTE, Ync
WET (permanent wetland) - Fog
DBF (deciduous broadleaf forest) - Lox

```{r}
#Add in site identifier to each site (did this earlier, need to improve this)

Ade_gs <- Ade_gs %>% mutate(site = "Ade")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
How_gs <- How_gs %>% mutate(site = "How")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Das_gs <- Das_gs %>% mutate(site = "Das")
Dry_gs <- Dry_gs %>% mutate(site = "Dry")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Asm_gs <- Asm_gs %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Dap_gs <- Dap_gs %>% mutate(site = "Dap")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Stp_gs <- Stp_gs %>% mutate(site = "Stp")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Tum_gs <- Tum_gs %>% mutate(site = "Tum")
Cpr_gs <- Cpr_gs %>% mutate(site = "Cpr")
```

```{r}
WSA <- bind_rows(Ade_gs, Gin_gs, How_gs, Rdf_gs) %>% group_by(site)

SAV <- bind_rows(Das_gs, Dry_gs, Gww_gs, Asm_gs, Cpr_gs) %>% group_by(site)

EBF <- bind_rows(Cum_gs, Rob_gs, Wac_gs, Whr_gs, Wom_gs, Tum_gs) %>% group_by(site)

GRA <- bind_rows(Dap_gs, Emr_gs, Rig_gs, Stp_gs, TTE_gs, Ync_gs) %>% group_by(site)

WET <- Fog_gs %>% group_by(site)

DBF <- Lox_gs %>% group_by(site)

```

```{r}
WSA$PFT <- "WSA"
SAV$PFT <- "SAV"
EBF$PFT <- "EBF"
GRA$PFT <- "GRA"
WET$PFT <- "WET"
DBF$PFT <- "DBF"

```

```{r}
PFT_all <- bind_rows(WSA, SAV, EBF, GRA, WET, DBF)

```


```{r}
library(ggplot2)

ggplot(PFT_all, aes(x = VPD, y = GPP, colour = PFT)) +

  geom_point(alpha = 0.05) +

  geom_smooth(method = "gam",
              formula = y ~ s(x, k = 10),
              se = FALSE,
              linewidth = 1.2) +

  theme_minimal(base_size = 14) +

  labs(
    x = "VPD",
    y = "GPP",
    colour = "PFT",
    title = "GPP–VPD relationships across PFTs"
  )

```
Mean functional response of PFT to atmospheric drying using GAM to characterise non-linear responses of GPP to VPD. 

**Need to bin VPD because these lines look too smooth**


Fit GAM for each PFT
1. No VPD
2. VPD only
3. Full with VPD

What is the additive predictive value of VPD across plant functional groups in Australia?


```{r}
PFT_groups <- c("WSA", "SAV", "EBF", "GRA", "WET", "DBF")

```

Create train/test within each PFT

```{r}
for(group in PFT_groups){
  
  df <- get(group)
  
  idx <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  test <- test %>% drop_na()
  
  train$cumulative_cwd_7day <- rollapply(train$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  assign(paste0(group, "_train"), train, envir = .GlobalEnv)
  assign(paste0(group, "_test"),  test,  envir = .GlobalEnv)
}


```

```{r}
PFT_all <- c("GRA", "EBF", "SAV", "WSA", "WET", "DBF")

all_metrics <- map_dfr(PFT_all, function(PFT) {
  
  train_data <- get(paste0(PFT, "_train"))
  test_data  <- get(paste0(PFT, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(PFT = PFT)
})

```



```{r}
all_metrics %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```

```{r}
library(tidyr)
library(dplyr)

vpd_additive <- all_metrics %>%
  pivot_wider(names_from = Model, values_from = c(R2, RMSE)) %>%
  mutate(
    R2_gain_VPD = R2_Full_VPD - R2_No_VPD,
    RMSE_drop_VPD = RMSE_No_VPD - RMSE_Full_VPD
  )

```

```{r}
library(ggplot2)

ggplot(vpd_additive, aes(x = PFT, y = R2_gain_VPD, fill = PFT)) +
  geom_col() +
  theme_minimal(base_size = 14) +
  labs(
    y = "R² gain from VPD",
    x = "Plant Functional Type",
    title = "Additive predictive value of VPD on GPP"
  )

```

```{r}
ggplot(all_metrics %>% filter(Model %in% c("No_VPD", "Full_VPD")),
       aes(x = PFT, y = R2, colour = Model)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 14) +
  labs(
    y = "R²",
    x = "Plant Functional Type",
    title = "Predictive value of VPD across PFTs"
  )

```
```{r}
full_model_only <- all_metrics %>%
  filter(Model == "Full_VPD")
```

```{r}
library(ggplot2)

ggplot(full_model_only, aes(x = PFT, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Plant Functional Type",
    y = expression(R^2),
    title = "Full GAM predictability of GPP across PFTs"
  )

```
**Predictability decreases when sites are grouped by PFT**
- PFTs are heterogeneous in their GPP-VPD sensitivity
- Plant functional groups do not behave as single functional units 

Predictive performance decreased when sites were aggregated by plant functional type, reflecting the substantial within-PFT heterogeneity in GPP responses to atmospheric and hydrological drivers. This suggests that PFT-level grouping conceals site-specific controls.

This supports emerging literature that PFTs are perhaps too broad - model fit was particularly low compared to site-specific GAMs. 

What if we grouped sites based on their annual precipitation?

Separate sites into very wet, wet, dry, and very dry? Is this better than separating into PFT in terms of predictability?

```{r}
#Look at annual precip again
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across Australian Fluxnet sites"
  )
```
23 sites 
4 groups, very wet, wet, dry and very dry
Grouped into 4 groups based on quantile to make it even (although a bit arbitrary), could improve this by finding reference in the literature for what is considered 'wet', 'dry' etc

```{r}
#Separate into quantiles of annual precipitation
site_precip_class <- annual_summary %>%
  mutate(
    precip_quantile = ntile(annual_summary, 4),
    precip_class = case_when(
      precip_quantile == 1 ~ "Very wet",
      precip_quantile == 2 ~ "Wet",
      precip_quantile == 3 ~ "Dry",
      precip_quantile == 4 ~ "Very Dry"
    )
  )

```

```{r}
Very_wet <- bind_rows(Wac, Rob, How, Das, Fog, Rdf)
Wet <- bind_rows(Dap, Ade, Tum, Wom, Dry, Stp)
Dry <- bind_rows(Ync, Emr, Gin, Cum, Rig, Asm)
Very_dry <- bind_rows(Whr, Gww, Cpr, TTE, Lox)
```

```{r}
precip_groups <- c("Very_dry", "Dry", "Wet", "Very_wet")
```


```{r}
#Add in cumulative water deficit and lagged climate variables for each precipitation class
for(group in precip_groups){
  
  df <- get(group)
  
  idx <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  test <- test %>% drop_na()
  
  train$cumulative_cwd_7day <- rollapply(train$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  assign(paste0(group, "_train"), train, envir = .GlobalEnv)
  assign(paste0(group, "_test"),  test,  envir = .GlobalEnv)
}


```

Fit GAMs for each precip class
```{r}
all_metrics_precip <- map_dfr(precip_groups, function(group) {
  
  train_data <- get(paste0(group, "_train"))
  test_data  <- get(paste0(group, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(precip_class = group)
})

```

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```

```{r}
all_metrics_precip %>%
  filter(Model == "Full_VPD") %>%
  ggplot(aes(x = precip_class, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation regime",
    y = expression(R^2),
    title = "Full GAM predictability across precipitation classes"
  )

```
Full model predictability is better for precipitation classes than for plant functional groups, however, it is not as good as site-level sensitivity. 

VPD additive value:

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  select(precip_class, Model, R2) %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

ggplot(vpd_additive_precip,
       aes(x = precip_class, y = VPD_additive)) +
  geom_col(fill = "firebrick") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation class",
    y = expression(Delta*R^2~"(VPD contribution)"),
    title = "Additive contribution of VPD to GPP predictability"
  )

```
The addition of VPD improved model fit for all precipitation classes, more in dry classes than in wet classes. 

VPD improves predictability more on a wider, climate scale than at an individual site or PFT level scale. PFTs assume all forests/savannas/woody savannas behave similarly, however, within PFTs VPD-GPP relationships are heterogenous.

Precipitation class performs nearly as well as site-specific GAMS, highlighting the importance of hydrological footprint on ecosystem functioning and response. 

Summary:
- VPD needs to be binned before plotting
- site, PFT and precipitation class predictability across Australia
- PFT models perform worse than precip class models 
- VPD has greater additional value in precip class models than PFT and site-level models


