---
title: "analysis_sites_1"
output: html_document
date: "2026-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list = ls())
```

How does predictability change across bioclimatic gradients? (OzFlux)
- which sites are the most predictable?
- which variables are the most/least important?

```{r}

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

Read in sites:
- Howard springs
- Adelaide river
- Daly River
- Daly Savanna
- Dry river
- Sturt plains
- Alice springs

```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

```

Have a look at some variables now
```{r}
plot(Dry_HH$Precip)
```

Convert -9999 to NA:

```{r}
#Specify which variables you want, otherwise it will read dates and show an error

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_1 <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_1 <- Ade_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_1 <- Asm_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_1 <- Dap_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_1 <- Das_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_1 <- Dry_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_1 <- Stp_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


```{r}
plot(Dry_1$Precip)
```

Before converting to daily, remove extreme outliers as they skew the data:

99% percentile of rainfall data for half-hourly (kg m-2 s-1) mm2 s-1
```{r}
quantile(Dry_daily$Precip, probs = c(0.90, 0.95, 0.99, 0.999), na.rm = TRUE)


```

1.5mm s-1 half-hourly

```{r}
threshold <- quantile(Dry_1$Precip, 0.999, na.rm = TRUE)
```

```{r}
Dry_1 <- Dry_1 %>%
  mutate(
    Precip_flag = Precip > threshold,      # flag extreme spikes
    Precip_clean = ifelse(Precip > threshold, NA, Precip)  # set outliers to NA
  )

```

```{r}
# Number of flagged points
sum(Dry_1$Precip_flag, na.rm = TRUE)

# Plot cleaned vs original
library(ggplot2)
ggplot(Dry_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  theme_minimal()

```


```{r}
# Create a new dataframe with Precip below the threshold
Dry_clean <- Dry_1 %>%
  filter(Precip <= threshold | is.na(Precip))

# Optional: check how many rows were removed
removed <- nrow(Dry_1) - nrow(Dry_clean)
cat("Removed", removed, "rows above threshold.\n")

```

```{r}
plot(Dry_clean$Precip)
```







We can convert these sites to daily data

```{r}
####
##  Convert the 30 min to daily ... someone should add something to check 
##. if the data is hourly ...
##
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####
library(tidyverse)

convert_to_daily <- function(df) {

  # Conversion constants
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HLFHOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
      Precip = Precip * SEC_2_HLFHOUR,
      ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
      VPD = VPD * HPA_2_KPA
    )
  
  df$date <- ymd_hms(df$date)
  
  # Resample to daily 
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%  
    group_by(date) %>%
    summarise(
      GPP = sum(GPP, na.rm=TRUE), 
      ET = sum(ET, na.rm=TRUE),  
      Precip = sum(Precip, na.rm=TRUE),
      Qle = mean(Qle, na.rm=TRUE),
      Tair = mean(Tair, na.rm=TRUE),
      SW_rad = mean(SW_rad, na.rm=TRUE),
      Rn = mean(Rn, na.rm=TRUE),
      Pa = mean(Pa, na.rm=TRUE),
      VPD = mean(VPD, na.rm=TRUE),
      .groups = "drop"
    )
  
  return(df)
}
```



```{r}
How_daily <- convert_to_daily(How_1)
Ade_daily <- convert_to_daily(Ade_1)
Das_daily <- convert_to_daily(Das_1)
Dap_daily <- convert_to_daily(Dap_1)
Dry_daily <- convert_to_daily(Dry_1)
Asm_daily <- convert_to_daily(Asm_1)
Stp_daily <- convert_to_daily(Stp_1)
```


```{r}
summary(Ade_daily$Precip)
```

Convert NAN back to NA:

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <-Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))
```

```{r}
plot(Dry_1$Precip)
```

Major outliers in precipitation:

```{r}
library(ggplot2)

ggplot(How_1, aes(x = log(Precip))) +
  geom_histogram(binwidth = 0.1) +
  scale_x_log10() +  # optional for skewed data
  theme_minimal()

```

```{r}
ggplot(Asm_daily, aes(x = date, y = Precip)) +
  geom_line() +
  theme_minimal()

```
```{r}
plot(Asm_daily$Precip)
```




