---
title: "analysis_sites_1"
output: html_document
date: "2026-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Script: found out the precipitation conversion was not correct. Now fixed. 

```{r}
rm(list = ls())
```

1. Read in Flux data
```{r}
library(tidyverse)
read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      Qle = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

Read in sites:
- Howard springs
- Adelaide river
- Daly River
- Daly Savanna
- Dry river
- Sturt plains
- Alice springs

```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")

Ade_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Ade_FLUXNET2015_FULLSET_HH_2007-2009_1-4.csv")

Dap_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

Das_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Dry_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

Stp_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

Asm_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")

```

```{r}
How_HH <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/Data/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv")
```

```{r}
summary(How_HH$Precip)
```

```{r}
max(How_HH$Precip, na.rm=TRUE)
max(How_HH$VPD, na.rm=TRUE)

```

```{r}
flux_vars <- c("GPP","Qle","Tair","SW_rad","Rn","Precip","Pa","VPD")

How_1 <- How_HH %>%
  mutate(across(all_of(flux_vars),
                ~ ifelse(abs(.) > 1000, NA, .)))

```

```{r}
summary(How_1$VPD)
summary(How_1$Precip)

```
```{r}
plot(How_1$Precip)
```



3. Plotting raw data: Half-hourly (kg m-2 s-1)

```{r}
plot(How_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Howard springs")
plot(Ade_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Adelaide river")
plot(Das_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Daly savanna")
plot(Dap_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Daly pasture")
plot(Dry_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Dry river")
plot(Stp_HH$Preci, ylab = "precipitation kg m-2 s-1", main = "Sturt plains")
plot(Asm_HH$Precip, ylab = "precipitation kg m-2 s-1", main = "Alice springs")
```
So, these large values are extremely unlikely. 


4. Convert -9999 values to NA:

```{r}
#Specify which variables you want, otherwise it will read dates and show an error

flux_vars <- c("GPP", "Qle", "Tair", "SW_rad", "Rn", "Precip", "Pa", "VPD")

#Clean dataframe, convert to 'na' if '-9999'

How_1 <- How_HH %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Ade_1 <- Ade_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Asm_1 <- Asm_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dap_1 <- Dap_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Das_1 <- Das_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Dry_1 <- Dry_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

Stp_1 <- Stp_HH %>%
   mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```

Plotting precipitation after removing -9999

```{r}
plot(Asm_1$Precip, ylab = "Precipitation kg m-2 s-1")
```
12 kg m-2 s-1 would be 43,200 mm/hour - which is impossible?

```{r}
plot(How_1$VPD)
```


Now, I am going to remove extreme outliers before converting to daily data.

99% percentile of rainfall data for half-hourly (kg m-2 s-1) mm2 s-1
```{r}
quantile(Asm_1$Precip, probs = c(0.90, 0.95, 0.99, 0.999), na.rm = TRUE)
```

```{r}
summary(How_HH$Precip)
quantile(How_HH$Precip, probs = c(0.5, 0.9, 0.99), na.rm = TRUE)

```
If in the 99th percentile, say outlier is TRUE

```{r}
How_thresh <- quantile(How_1$Precip, 0.99, na.rm = TRUE)
Ade_thresh <- quantile(Ade_1$Precip, 0.99, na.rm = TRUE)
Dap_thresh <- quantile(Dap_1$Precip, 0.99, na.rm = TRUE)
Das_thresh <- quantile(Das_1$Precip, 0.99, na.rm = TRUE)
Dry_thresh <- quantile(Dry_1$Precip, 0.99, na.rm = TRUE)
Stp_thresh <- quantile(Stp_1$Precip, 0.99, na.rm = TRUE)
Asm_thresh <- quantile(Asm_1$Precip, 0.99, na.rm = TRUE)
```

Apply to datasets

```{r}
How_1 <- How_1 %>%
  mutate(
    Precip_flag = Precip > How_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > How_thresh, NA, Precip)  # set outliers to NA
  )

Ade_1 <- Ade_1 %>%
  mutate(
    Precip_flag = Precip > Ade_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Ade_thresh, NA, Precip)  # set outliers to NA
  )

Dap_1 <- Dap_1 %>%
  mutate(
    Precip_flag = Precip > Dap_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Dap_thresh, NA, Precip)  # set outliers to NA
  )

Das_1 <- Das_1 %>%
  mutate(
    Precip_flag = Precip > Das_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Das_thresh, NA, Precip)  # set outliers to NA
  )

Dry_1 <- Dry_1 %>%
  mutate(
    Precip_flag = Precip > Dry_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Dry_thresh, NA, Precip)  # set outliers to NA
  )

Stp_1 <- Stp_1 %>%
  mutate(
    Precip_flag = Precip > Stp_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Stp_thresh, NA, Precip)  # set outliers to NA
  )

Asm_1 <- Asm_1 %>%
  mutate(
    Precip_flag = Precip > Asm_thresh,      # flag extreme spikes
    Precip_clean = ifelse(Precip > Asm_thresh, NA, Precip)  # set outliers to NA
  )


```

```{r}
# Number of flagged points
sum(How_1$Precip_flag, na.rm = TRUE)

# Plot cleaned vs original
library(ggplot2)
ggplot(Dry_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Dry River")
  theme_minimal()
  
  library(ggplot2)
ggplot(How_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Howard Springs")
  theme_minimal()

  library(ggplot2)
ggplot(Das_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Daly Savanna")
  theme_minimal()
  
  library(ggplot2)
ggplot(Dap_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Dsly Pasture")
  theme_minimal()
  
  library(ggplot2)
ggplot(Ade_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Adelaide River")
  theme_minimal()
  
  library(ggplot2)
ggplot(Asm_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Alice springs")
  theme_minimal()
  
  library(ggplot2)
ggplot(Stp_1, aes(x = date)) +
  geom_line(aes(y = Precip), color = "red") +
  labs(y = "Precipitation kg m-2 s-1", x = "Date") +
  geom_line(aes(y = Precip_clean), color = "blue") +
  ggtitle("Sturt plains")
  theme_minimal()
```


```{r}
# Create a new dataframe with Precip below the threshold
How_clean <- How_1 %>%
  filter(Precip <= How_thresh | is.na(Precip_clean))

Ade_clean <- Ade_1 %>%
  filter(Precip <= Ade_thresh | is.na(Precip_clean))

Dap_clean <- Dap_1 %>%
  filter(Precip <= Dap_thresh | is.na(Precip_clean))

Das_clean <- Das_1 %>%
  filter(Precip <= Das_thresh | is.na(Precip_clean))

Dry_clean <- Dry_1 %>%
  filter(Precip <= Dry_thresh | is.na(Precip_clean))

Stp_clean <- Stp_1 %>%
  filter(Precip <= Stp_thresh | is.na(Precip_clean))

Asm_clean <- Asm_1 %>%
  filter(Precip <= Asm_thresh | is.na(Precip_clean))
```

Plot the cleaned, half-hourly precipitation data

```{r}
plot(Stp_clean$Precip, ylab = "Precipitation kg m-2 s-1", main = "Sturt plains")
plot(How_clean$Precip,  ylab = "Precipitation kg m-2 s-1", main = "Howard springs")
plot(Ade_clean$Precip,  ylab = "Precipitation kg m-2 s-1", main = "Adelaide river")
plot(Das_clean$Precip,  ylab = "Precipitation kg m-2 s-1", main = "Daly savanna")
plot(Dap_clean$Precip,  ylab = "Precipitation kg m-2 s-1", main = "daly pasture")
plot(Asm_clean$Precip,  ylab = "Precipitation kg m-2 s-1", main = "Alice springs")
```
Now convert to daily with the cleaned data:

```{r}
  ####
  ##  Convert the 30 min to daily ... someone should add something to check 
  ##. if the data is hourly ...
  ##
  ##
  ##  author: Martin De Kauwe
  ##  date: 27th February, 2025
  ##  email: mdekauwe@gmail.com
  ####
  library(tidyverse)
  
  convert_to_daily <- function(df) {
  
    # Conversion constants
    UMOL_TO_MOL <- 1E-6
    MOL_C_TO_GRAMS_C <- 12.0
    SEC_2_HLFHOUR <- 1800.
    WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
    PA_2_KPA = 0.001
    HPA_2_KPA = 0.1
    
    df <- df %>%
      mutate(
        GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
        Precip = Precip * SEC_2_HLFHOUR,
        ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
        VPD = VPD * HPA_2_KPA
      )
    
    df$date <- ymd_hms(df$date)
    
    # Resample to daily 
    df <- df %>%
      mutate(date = floor_date(date, "day")) %>%  
      group_by(date) %>%
      summarise(
        GPP = sum(GPP, na.rm=TRUE), 
        ET = sum(ET, na.rm=TRUE),  
        Precip = sum(Precip, na.rm=TRUE),
        Qle = mean(Qle, na.rm=TRUE),
        Tair = mean(Tair, na.rm=TRUE),
        SW_rad = mean(SW_rad, na.rm=TRUE),
        Rn = mean(Rn, na.rm=TRUE),
        Pa = mean(Pa, na.rm=TRUE),
        VPD = mean(VPD, na.rm=TRUE),
        .groups = "drop"
      )
    
    return(df)
  }
```

Convert to daily with cleaned data:

```{r}
How_daily <- convert_to_daily(How_clean)
Ade_daily <- convert_to_daily(Ade_clean)
Das_daily <- convert_to_daily(Das_clean)
Dap_daily <- convert_to_daily(Dap_clean)
Dry_daily <- convert_to_daily(Dry_clean)
Asm_daily <- convert_to_daily(Asm_clean)
Stp_daily <- convert_to_daily(Stp_clean)
```

```{r}
plot(How_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Howard springs")
plot(Ade_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Adelaide river")
plot(Dap_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Daly pasture")
plot(Das_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Daly savanna")
plot(Stp_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Sturt plains")
plot(Dry_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Dry River")
plot(Asm_daily$Precip, ylab = "Precipitation kg m-2 d-1 / mm day", main = "Alice springs")
```



Convert NAN back to NA:

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Ade_daily <- Ade_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Das_daily <-Das_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))
```

```{r}
plot(Asm_daily$Precip)
plot(How_daily$Precip)
plot(Asm_daily$Precip)
plot(Dap_daily$Precip)
plot(Das_daily$Precip)
plot(Ade_daily$Precip)
```
```{r}
 library(tidyverse)
  
  convert_to_daily_2 <- function(df) {
  
    # Conversion constants
    UMOL_TO_MOL <- 1E-6
    MOL_C_TO_GRAMS_C <- 12.0
    SEC_2_HLFHOUR <- 1800.
    WM2_TO_KG_M2_S <- 1.0 / ( 2.45 * 1E6 )
    PA_2_KPA = 0.001
    HPA_2_KPA = 0.1
    
    df <- df %>%
      mutate(
        GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HLFHOUR,
        Precip = Precip,
        ET = Qle * WM2_TO_KG_M2_S * SEC_2_HLFHOUR,
        VPD = VPD * HPA_2_KPA
      )
    
    df$date <- ymd_hms(df$date)
    
    # Resample to daily 
    df <- df %>%
      mutate(date = floor_date(date, "day")) %>%  
      group_by(date) %>%
      summarise(
        GPP = sum(GPP, na.rm=TRUE), 
        ET = sum(ET, na.rm=TRUE),  
        Precip = sum(Precip, na.rm=TRUE),
        Qle = mean(Qle, na.rm=TRUE),
        Tair = mean(Tair, na.rm=TRUE),
        SW_rad = mean(SW_rad, na.rm=TRUE),
        Rn = mean(Rn, na.rm=TRUE),
        Pa = mean(Pa, na.rm=TRUE),
        VPD = mean(VPD, na.rm=TRUE),
        .groups = "drop"
      )
    
    return(df)
  }
```

```{r}

How_daily2 <- convert_to_daily_2(How_clean)
Ade_daily2 <- convert_to_daily_2(Ade_clean)
Das_daily2 <- convert_to_daily_2(Das_clean)
Dap_daily2 <- convert_to_daily_2(Dap_clean)
Dry_daily2 <- convert_to_daily_2(Dry_clean)
Asm_daily2 <- convert_to_daily_2(Asm_clean)
Stp_daily2 <- convert_to_daily_2(Stp_clean)

```
```{r}
plot(How_daily2$Precip, ylab = "precipitation (mm/day)", main = "Howard springs")
plot(Ade_daily2$Precip, ylab = "precipitation (mm/day)", main = "Adelaide river")
plot(Das_daily2$Precip, ylab = "precipitation (mm/day", main = "Daly savanna")
plot(Dap_daily2$Precip, ylab = "precipitation (mm/day)", main = "Daly pasture")
plot(Dry_daily2$Precip, ylab = "Precipitation (mm/day)", main = "Dry River")
plot(Stp_daily2$Precip, ylab = "precipitation (mm/day", main = "Sturt plains")
plot(Asm_daily2$Precip, ylab = "precipitation (mm/day", main = "Alice springs")

```

```{r}
any(duplicated(Das_clean$date))

```






