---
title: "moisture_stress_contribution"
output: html_document
date: "2026-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Contribution of moisture stress** 
- Train on non-moisture stress period and test on moisture stress period - the difference (residual) between the observed and predicted GPP is the effect that soil moisture stress has on predictability

1. Define wet and dry seasons (precipitation threshold)
2. Predictors - VPD, Tair, SW rad and LAI
3. Train on unstressed period 
4. Predict on dry season and calculate residuals
Check model actually fits wel, a low R-squared means the model isn't capturing GPP even under good conditions, so the dry season residuals won't actually be meaningful

Wet/dry threshold
- 90 day rolling precipitation
- 70/30 split - top 70% unstressed and bottom 30% stressed periods

```{r}
library(mgcv)
library(zoo)
library(tidyverse)

stress_results <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]] %>%
    arrange(date) %>%
    drop_na(GPP, VPD, Tair, SW_rad, LAI_filled) %>%
    mutate(
      precip_90day = rollsum(Precip, k = 90, fill = NA, align = "right")
    )
  
  # site-specific percentile thresholds
  wet_thresh <- quantile(df$precip_90day, 0.70, na.rm = TRUE)
  dry_thresh  <- quantile(df$precip_90day, 0.30, na.rm = TRUE)
  
  df <- df %>%
    mutate(season = case_when(
      precip_90day >= wet_thresh ~ "wet",
      precip_90day <= dry_thresh ~ "dry",
      TRUE ~ "transition"
    ))
  
  wet <- df %>% filter(season == "wet")
  dry <- df %>% filter(season == "dry")
  
  if (nrow(wet) < 50 | nrow(dry) < 10) return(NULL)
  
  # fit GAM on wet season
  model <- gam(GPP ~ s(VPD) + s(Tair) + s(SW_rad) + s(LAI_filled),
               data = wet,
               method = "REML")
  
  r2 <- summary(model)$r.sq
  
  # predict onto dry season
  dry %>%
    mutate(
      site = str_remove(site, "_HH_daily"),
      GPP_predicted = predict(model, newdata = dry),
      GPP_residual = GPP - GPP_predicted,
      stress_index = GPP_residual / GPP_predicted
    ) %>%
    group_by(site) %>%
    summarise(
      mean_residual = mean(GPP_residual, na.rm = TRUE),
      mean_stress_index = mean(stress_index, na.rm = TRUE),
      wet_season_r2 = r2,
      n_wet_days = nrow(wet),
      n_dry_days = nrow(dry),
      wet_thresh_mm = wet_thresh,
      dry_thresh_mm = dry_thresh,
      .groups = "drop"
    )
})

```

stress_results
- mean_residual - average difference between observed and predicted GPP in the dry season (moisture stress signal)
- mean_stress_index - normalised by predicted GPP so comparable across sites
- wet_season_r2 - how well the GAM fit during the wet season

2. Add in R-squared for dry season and I want to plot the GAMs:
```{r}
library(mgcv)
library(zoo)
library(tidyverse)

# store daily predictions separately
all_predictions <- map_dfr(names(sites_daily), function(site) {
  
  df <- sites_daily[[site]] %>%
    arrange(date) %>%
    drop_na(GPP, VPD, Tair, SW_rad, LAI_filled) %>%
    mutate(precip_90day = rollsum(Precip, k = 90, fill = NA, align = "right"))
  
  wet_thresh <- quantile(df$precip_90day, 0.70, na.rm = TRUE)
  dry_thresh  <- quantile(df$precip_90day, 0.30, na.rm = TRUE)
  
  df <- df %>%
    mutate(season = case_when(
      precip_90day >= wet_thresh ~ "wet",
      precip_90day <= dry_thresh ~ "dry",
      TRUE ~ "transition"
    ))
  
  wet <- df %>% filter(season == "wet")
  dry <- df %>% filter(season == "dry")
  
  if (nrow(wet) < 50 | nrow(dry) < 10) return(NULL)
  
  model <- gam(GPP ~ s(VPD) + s(Tair) + s(SW_rad) + s(LAI_filled),
               data = wet, method = "REML")
  
  # fitted values on wet season
  wet_out <- wet %>%
    mutate(
      site = str_remove(site, "_HH_daily"),
      GPP_predicted = fitted(model),
      season = "wet"
    )
  
  # predicted values on dry season
  dry_out <- dry %>%
    mutate(
      site = str_remove(site, "_HH_daily"),
      GPP_predicted = predict(model, newdata = dry),
      season = "dry"
    )
  
  bind_rows(wet_out, dry_out) %>%
    select(site, date, season, GPP, GPP_predicted)
})

# now calculate summaries including dry season R²
stress_results <- all_predictions %>%
  group_by(site, season) %>%
  summarise(
    ss_res = sum((GPP - GPP_predicted)^2, na.rm = TRUE),
    ss_tot = sum((GPP - mean(GPP, na.rm = TRUE))^2, na.rm = TRUE),
    r2 = 1 - (ss_res / ss_tot),
    mean_residual = mean(GPP - GPP_predicted, na.rm = TRUE),
    mean_stress_index = mean((GPP - GPP_predicted) / GPP_predicted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = season, 
              values_from = c(r2, mean_residual, mean_stress_index),
              names_glue = "{.value}_{season}")
```

```{r}
site_names <- unique(all_predictions$site)
n_per_panel <- 6
site_groups <- split(site_names, ceiling(seq_along(site_names) / n_per_panel))

plots <- map(site_groups, function(sites_subset) {
  all_predictions %>%
    filter(site %in% sites_subset) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = GPP), colour = "grey50", alpha = 0.6, linewidth = 0.3) +
    geom_line(aes(y = GPP_predicted, colour = season), linewidth = 0.5) +
    scale_colour_manual(values = c("wet" = "steelblue", "dry" = "firebrick")) +
    facet_wrap(~ site, scales = "free_y", ncol = 2) +
    labs(x = NULL, y = "GPP", colour = "Season",
         title = "Observed vs GAM-predicted GPP") +
    theme_classic(base_size = 9) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold"))
})

walk(plots, print)
```

