---
title: "01_driver_importance"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Subsection 1: VPD across Australia
- **Driver importance across sites**
- GAMs - predictability
- GPP~VPD responses

Fit Random Forests to determine the importance of explanatory variables of GPP across Australian flux
- Compare random forests with GAM analysis

```{r}
####
##  Fit a random forest to predict GPP, here we're testing the use of lags
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

#standardise function
standardise <- function(df, vars_to_scale) {
  df_scaled <- df
  df_scaled[vars_to_scale] <- scale(df[vars_to_scale])
  
  return(df_scaled)
}

#Function to run RF

run_rf_for_site <- function(df, site_name) {
  
  #Drop NAs
  df <- df %>% drop_na()
  
  #check you have enough data
   if (nrow(df) < 100) {
    warning(paste("Site", site_name, "has insufficient data"))
    return(NULL)
   }
  
  #lagged variables before splitting the data
   df <- df %>%
    arrange(date) %>%  # Ensure data is sorted by date
    mutate(
      # 7-day cumulative CWD
      cumulative_cwd_7day = rollapply(cwd, width=7, FUN=sum, 
                                      align="right", fill=NA, na.rm=TRUE),
      # Temperature lags
      Tair_lag1 = lag(Tair, 1),
      Tair_lag2 = lag(Tair, 2),
      Tair_lag3 = lag(Tair, 3),
      # VPD lags
      VPD_lag1 = lag(VPD, 1),
      VPD_lag2 = lag(VPD, 2),
      VPD_lag3 = lag(VPD, 3)
    ) %>%
    drop_na()
   
   #Now split into train/test
   set.seed(123)
  idx <- createDataPartition(df$GPP, p=0.8, list=FALSE)
  train_data <- df[idx, ]
  test_data <- df[-idx, ]
  
  # Standardise predictors
  predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day",
                  "Tair_lag1", "Tair_lag2", "Tair_lag3", 
                  "VPD_lag1", "VPD_lag2", "VPD_lag3")
  
  train_data <- standardise(train_data, predictors)
  test_data <- standardise(test_data, predictors)
  
  #train random forest
   model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + 
                          Tair_lag1 + Tair_lag2 + Tair_lag3 + 
                          VPD_lag1 + VPD_lag2 + VPD_lag3, 
                        data=train_data, ntree=500, mtry=3)
   
   #now predict
    predictions <- predict(model, test_data)
  results <- data.frame(Observed=test_data$GPP, Predicted=predictions)
  
  #calculate metrics
  rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
  rsq <- cor(results$Observed, results$Predicted)^2
  mean_observed <- mean(results$Observed)
  nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
            sum((results$Observed - mean_observed)^2))
  
  #get variable importance
  
importance_df <- data.frame(
    variable = rownames(importance(model)),
    importance = importance(model)[,1]
  )

#return results
return(list(
    site = site_name,
    model = model,
    metrics = data.frame(site=site_name, RMSE=rmse, R2=rsq, NSE=nse),
    importance = importance_df %>% mutate(site = site_name),
    predictions = results %>% mutate(site = site_name)
  ))

}

# Apply to all sites in the list
rf_results <- map2(sites_daily, names(sites_daily), run_rf_for_site)

# Extract metrics for all sites
all_metrics <- map_df(rf_results, ~.x$metrics)
print(all_metrics)

# Extract importance for all sites
all_importance <- map_df(rf_results, ~.x$importance)

```

```{r}
site_names <- unique(all_importance$site)
n_groups <- ceiling(length(site_names) / 6)

for(i in 1:n_groups) {
  start_idx <- (i-1)*6 + 1
  end_idx <- min(i*6, length(site_names))
  sites_subset <- site_names[start_idx:end_idx]
  
  p <- ggplot(all_importance %>% filter(site %in% sites_subset), 
              aes(x=reorder(variable, importance), y=importance)) +
    geom_col(fill="steelblue") +
    facet_wrap(~site, scales="free_x", ncol=3) +
    coord_flip() +
    theme_minimal() +
    theme(
      strip.text = element_text(size=10),
      axis.text = element_text(size=8)
    ) +
    labs(title=paste("Variable Importance - Group", i), 
         x="Variable", y="Importance")
  
  print(p)
  ggsave(paste0("importance_group_", i, ".png"), width=12, height=8, dpi=300)
}
```

```{r}
install.packages("tidytext")
```

Random forest:
Predictors - VPD lag1, lag2, lag3, cwd 7 day, VPD, SW_rad, Tair, Tair lag 1, Tair lag 2, Tair lag 3


```{r}
# Loop through each site and create individual plots
for(site_name in unique(all_importance$site)) {
  
  plot_data <- all_importance %>%
    filter(site == site_name) %>%
    arrange(desc(importance)) %>%
    mutate(rank = row_number(),
           variable = paste0(rank, ". ", variable))
  
  p <- ggplot(plot_data, aes(x=importance, y=reorder(variable, importance))) +
    geom_col(fill="steelblue") +
    theme_minimal() +
    theme(text = element_text(size=12)) +
    labs(title=paste("Variable Importance -", site_name),
         x="Importance (%IncMSE)", 
         y="")
  
  print(p)
  ggsave(paste0("importance_", site_name, ".png"), width=8, height=6, dpi=300)
}
```
Drier sites
- cwd in top predictor, water-limited environments

VPD lags more important at intermediate sites

Wet sites:
- SW rad and VPD dominate

EBF consistently show SW radiation as their most important predictor with VPD 2-3, water abundant so energy and atmospheric demand control photosynthesis

Fog (wettest site) - only VPD lags matter

Grasslands = strong CWD control + temp lag effects
Savannas - mixed cwd and vpd
EBF - consistently SW radiation then VPD then cwd
WSA - mixed
DBF - SW radiaion




