---
title: "summer_extremes_analysis"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1) No. of extreme summer VPD days over time in my long-term sites
2) Average annual summer VPD over time in my long-term sites

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)

long_term_sites <- c("How", "Tum", "Wom", "Asm", "Cpr")

# ── Prepare data 
# Combine long-term sites and flag summer (DJF) days
# Summer year assigned to the January/February year so Dec is grouped forward
lt_daily <- map_dfr(long_term_sites, function(s) {
  sites_daily[[s]] %>%
    mutate(
      site        = s,
      date        = as.Date(date),
      month       = month(date),
      year        = year(date),
      summer_year = if_else(month == 12, year + 1L, year)
    )
}) %>%
  filter(month %in% c(12, 1, 2))

# ── Compute 90th percentile VPD threshold per site (across all summers) 
vpd_thresholds <- lt_daily %>%
  group_by(site) %>%
  summarise(vpd_90th = quantile(VPD, 0.90, na.rm = TRUE), .groups = "drop")

# ── Plot 1: Number of extreme VPD days per summer per site
extreme_days <- lt_daily %>%
  left_join(vpd_thresholds, by = "site") %>%
  group_by(site, summer_year) %>%
  summarise(
    n_extreme = sum(VPD > vpd_90th, na.rm = TRUE),
    .groups   = "drop"
  )

ggplot(extreme_days, aes(x = summer_year, y = n_extreme)) +
  geom_col(fill = "#E06C4B", alpha = 0.8, width = 0.8) +
  facet_wrap(~ site, scales = "free_y", ncol = 2) +
  labs(
    x = NULL,
    y = "Number of days VPD > 90th percentile"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )

# ── Plot 2: Mean summer VPD per year per site 
summer_vpd <- lt_daily %>%
  group_by(site, summer_year) %>%
  summarise(
    mean_vpd = mean(VPD, na.rm = TRUE),
    .groups  = "drop"
  )

ggplot(summer_vpd, aes(x = summer_year, y = mean_vpd)) +
  geom_line(colour = "grey60", linewidth = 0.5) +
  geom_point(colour = "#E06C4B", size = 2) +
  facet_wrap(~ site, scales = "free_y", ncol = 2) +
  labs(
    x = NULL,
    y = "Mean summer VPD (kPa)"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)

long_term_sites <- c("How", "Tum", "Wom", "Asm", "Cpr")

# ── Prepare data ──────────────────────────────────────────────────────────────
lt_daily <- map_dfr(long_term_sites, function(s) {
  sites_daily[[s]] %>%
    mutate(
      site        = s,
      date        = as.Date(date),
      month       = month(date),
      year        = year(date),
      summer_year = if_else(month == 12, year + 1L, year)
    )
}) %>%
  filter(month %in% c(12, 1, 2))

# ── 90th percentile threshold per site ───────────────────────────────────────
vpd_thresholds <- lt_daily %>%
  group_by(site) %>%
  summarise(vpd_90th = quantile(VPD, 0.90, na.rm = TRUE), .groups = "drop")

# ── Summarised data ───────────────────────────────────────────────────────────
extreme_days <- lt_daily %>%
  left_join(vpd_thresholds, by = "site") %>%
  group_by(site, summer_year) %>%
  summarise(n_extreme = sum(VPD > vpd_90th, na.rm = TRUE), .groups = "drop")

summer_vpd <- lt_daily %>%
  group_by(site, summer_year) %>%
  summarise(mean_vpd = mean(VPD, na.rm = TRUE), .groups = "drop")

# ── Plot 1: Extreme VPD days — one plot per site ──────────────────────────────
walk(long_term_sites, function(s) {
  p <- extreme_days %>%
    filter(site == s) %>%
    ggplot(aes(x = summer_year, y = n_extreme)) +
    geom_col(fill = "#E06C4B", alpha = 0.8, width = 0.8) +
    labs(
      title = s,
      x     = NULL,
      y     = "Days VPD > 90th percentile"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(face = "bold")
    )
  print(p)
})

# ── Plot 2: Mean summer VPD — one plot per site ───────────────────────────────
walk(long_term_sites, function(s) {
  p <- summer_vpd %>%
    filter(site == s) %>%
    ggplot(aes(x = summer_year, y = mean_vpd)) +
    geom_line(colour = "grey60", linewidth = 0.5) +
    geom_point(colour = "#E06C4B", size = 2) +
    labs(
      title = s,
      x     = NULL,
      y     = "Mean summer VPD (kPa)"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(face = "bold")
    )
  print(p)
})
```
```{r}
# Check how many valid VPD observations exist per site per summer
lt_daily %>%
  group_by(site, summer_year) %>%
  summarise(
    n_obs     = sum(!is.na(VPD)),
    n_extreme = sum(VPD > vpd_thresholds$vpd_90th[vpd_thresholds$site == first(site)], na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  filter(n_extreme == 0) %>%
  print(n = Inf)
```
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)

long_term_sites <- c("How", "Tum", "Wom", "Asm", "Cpr")

# ── Prepare summer daily data ─────────────────────────────────────────────────
lt_daily <- map_dfr(long_term_sites, function(s) {
  sites_daily[[s]] %>%
    mutate(
      site        = s,
      date        = as.Date(date),
      month       = month(date),
      year        = year(date),
      summer_year = if_else(month == 12, year + 1L, year)
    )
}) %>%
  filter(month %in% c(12, 1, 2))

# ── Flag extreme days using 90th percentile per site ─────────────────────────
lt_daily <- lt_daily %>%
  group_by(site) %>%
  mutate(
    vpd_90th = quantile(VPD, 0.90, na.rm = TRUE),
    extreme  = VPD > vpd_90th
  ) %>%
  ungroup()

# ── Filter to summers with sufficient data (>= 30 valid VPD obs) ─────────────
valid_summers <- lt_daily %>%
  group_by(site, summer_year) %>%
  summarise(n_obs = sum(!is.na(VPD)), .groups = "drop") %>%
  filter(n_obs >= 10)

```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)

long_term_sites <- c("How", "Tum", "Wom", "Asm", "Cpr")

# ── Site groups ───────────────────────────────────────────────────────────────
group1 <- c("How", "Tum")
group2 <- c("Wom", "Asm", "Cpr")

# ── Plot function — takes a dataset and site group ────────────────────────────
plot_n_extreme <- function(data, sites) {
  data %>%
    filter(site %in% sites) %>%
    mutate(site = factor(site, levels = sites)) %>%
    ggplot(aes(x = summer_year, y = n_extreme)) +
    geom_line(colour = "grey60", linewidth = 0.5) +
    geom_point(size = 1.5, colour = "#E06C4B") +
    facet_wrap(~ site, ncol = 1, scales = "free_y") +
    labs(x = NULL, y = "No. days VPD > 90th percentile") +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "none"
    )
}

plot_max_extreme <- function(data, sites) {
  data %>%
    filter(site %in% sites) %>%
    mutate(site = factor(site, levels = sites)) %>%
    ggplot(aes(x = summer_year, y = max_extreme_vpd)) +
    geom_line(colour = "grey60", linewidth = 0.5) +
    geom_point(size = 1.5, colour = "#E06C4B") +
    facet_wrap(~ site, ncol = 1, scales = "free_y") +
    labs(x = NULL, y = "Maximum VPD on extreme days (kPa)") +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "none"
    )
}

```

```{r}
```{r}
# ── Update the summary to include both max and mean ───────────────────────────
max_extreme_summer <- lt_daily %>%
  inner_join(valid_summers, by = c("site", "summer_year")) %>%
  filter(extreme) %>%
  group_by(site, summer_year) %>%
  summarise(
    max_extreme_vpd  = max(VPD, na.rm = TRUE),
    mean_extreme_vpd = mean(VPD, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(is.finite(max_extreme_vpd)) %>%
  pivot_longer(cols = c(max_extreme_vpd, mean_extreme_vpd),
               names_to = "metric", values_to = "vpd") %>%
  mutate(
    metric = recode(metric,
                    "max_extreme_vpd"  = "Maximum",
                    "mean_extreme_vpd" = "Mean"),
    site = factor(site, levels = long_term_sites)
  )

# ── Update plot function ───────────────────────────────────────────────────────
plot_max_extreme <- function(data, sites) {
  data %>%
    filter(site %in% sites) %>%
    mutate(site = factor(site, levels = sites)) %>%
    ggplot(aes(x = summer_year, y = vpd, colour = metric)) +
    geom_line(linewidth = 0.5) +
    geom_point(size = 1.5) +
    scale_colour_manual(values = c("Maximum" = "#E06C4B", "Mean" = "#4682B4")) +
    facet_wrap(~ site, ncol = 1, scales = "free_y") +
    coord_cartesian(ylim = c(0, NA)) +
    labs(x = NULL, y = "VPD on extreme days (kPa)", colour = NULL) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold"),
      legend.position  = "top"
    )
}
```


```{r}
# ── Print Plot 1 (n extreme days) ─────────────────────────────────────────────
print(plot_n_extreme(n_extreme_summer, group1))
print(plot_n_extreme(n_extreme_summer, group2))

# ── Print Plot 2 (max VPD on extreme days) ────────────────────────────────────
print(plot_max_extreme(max_extreme_summer, group1))
print(plot_max_extreme(max_extreme_summer, group2))

```

```{r}
Wom %>%
  mutate(VPD = if_else(VPD == -9999, NA_real_, VPD))

plot(Wom$date, Wom$VPD, ylim = c(0, max(Wom$VPD[Wom$VPD > 0], na.rm = TRUE)))
```
```{r}
# Check annual stats to see where it goes wrong
Wom %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(
    mean_VPD = mean(VPD, na.rm = TRUE),
    max_VPD  = max(VPD, na.rm = TRUE),
    n_obs    = sum(!is.na(VPD))
  ) %>%
  print(n = Inf)
```
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)
library(patchwork)

long_term_sites <- c("How", "Tum", "Wom", "Asm", "Cpr")
group1 <- c("How", "Tum")
group2 <- c("Wom", "Asm", "Cpr")

# ── Prepare summer daily data ─────────────────────────────────────────────────
lt_daily <- map_dfr(long_term_sites, function(s) {
  sites_daily[[s]] %>%
    mutate(
      site        = s,
      date        = as.Date(date),
      month       = month(date),
      year        = year(date),
      summer_year = if_else(month == 12, year + 1L, year)
    )
})

# ── Mean annual VPD per site per year (whole year, not just summer) ───────────
annual_vpd <- lt_daily %>%
  group_by(site, year) %>%
  summarise(mean_vpd = mean(VPD, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(mean_vpd))

# ── Plot function ─────────────────────────────────────────────────────────────
plot_annual_vpd <- function(data, sites) {
  data %>%
    filter(site %in% sites) %>%
    mutate(site = factor(site, levels = sites)) %>%
    ggplot(aes(x = year, y = mean_vpd)) +
    geom_line(colour = "grey60", linewidth = 0.5) +
    geom_point(colour = "#4682B4", size = 2) +
    geom_smooth(method = "lm", se = TRUE,
                colour = "#E06C4B", fill = "#E06C4B",
                alpha = 0.15, linewidth = 0.7, linetype = "dashed") +
    facet_wrap(~ site, ncol = 1, scales = "free_y") +
    labs(x = NULL, y = "Mean annual VPD (kPa)") +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold")
    )
}

# ── Print both groups ─────────────────────────────────────────────────────────
print(plot_annual_vpd(annual_vpd, group1))
print(plot_annual_vpd(annual_vpd, group2))
```

