---
title: "precip_regime"
output: html_document
date: "2026-02-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Not sure if there is any pattern here. 
TTE - interestingly VPD reduces site predictability. 
*Explore this further*. - think this could be interesting
Could be interesting to add in LAI to this, so there can be some sort of mechanistic interpretation of these results. 

2. Separate sites into PFT to visualize trends:

WSA (woody savanna) - Ade, Gin, How, Rdf
SAV (savanna) - DaS, Dry, Gww, Asm, Cpr
EBF (evergreen broadleaf forest) - Cum, Rob, Wac, Whr, Wom, Tum
GRA (grassland) - Dap, Emr, Rig, Stp, TTE, Ync
WET (permanent wetland) - Fog
DBF (deciduous broadleaf forest) - Lox

```{r}
#Add in site identifier to each site (did this earlier, need to improve this)

Ade_gs <- Ade_gs %>% mutate(site = "Ade")
Gin_gs <- Gin_gs %>% mutate(site = "Gin")
How_gs <- How_gs %>% mutate(site = "How")
Rdf_gs <- Rdf_gs %>% mutate(site = "Rdf")
Das_gs <- Das_gs %>% mutate(site = "Das")
Dry_gs <- Dry_gs %>% mutate(site = "Dry")
Gww_gs <- Gww_gs %>% mutate(site = "Gww")
Asm_gs <- Asm_gs %>% mutate(site = "Asm")
Cum_gs <- Cum_gs %>% mutate(site = "Cum")
Rob_gs <- Rob_gs %>% mutate(site = "Rob")
Wac_gs <- Wac_gs %>% mutate(site = "Wac")
Whr_gs <- Whr_gs %>% mutate(site = "Whr")
Wom_gs <- Wom_gs %>% mutate(site = "Wom")
Dap_gs <- Dap_gs %>% mutate(site = "Dap")
Emr_gs <- Emr_gs %>% mutate(site = "Emr")
Rig_gs <- Rig_gs %>% mutate(site = "Rig")
Stp_gs <- Stp_gs %>% mutate(site = "Stp")
TTE_gs <- TTE_gs %>% mutate(site = "TTE")
Ync_gs <- Ync_gs %>% mutate(site = "Ync")
Fog_gs <- Fog_gs %>% mutate(site = "Fog")
Lox_gs <- Lox_gs %>% mutate(site = "Lox")
Tum_gs <- Tum_gs %>% mutate(site = "Tum")
Cpr_gs <- Cpr_gs %>% mutate(site = "Cpr")
```

```{r}
WSA <- bind_rows(Ade_gs, Gin_gs, How_gs, Rdf_gs) %>% group_by(site)

SAV <- bind_rows(Das_gs, Dry_gs, Gww_gs, Asm_gs, Cpr_gs) %>% group_by(site)

EBF <- bind_rows(Cum_gs, Rob_gs, Wac_gs, Whr_gs, Wom_gs, Tum_gs) %>% group_by(site)

GRA <- bind_rows(Dap_gs, Emr_gs, Rig_gs, Stp_gs, TTE_gs, Ync_gs) %>% group_by(site)

WET <- Fog_gs %>% group_by(site)

DBF <- Lox_gs %>% group_by(site)

```

```{r}
WSA$PFT <- "WSA"
SAV$PFT <- "SAV"
EBF$PFT <- "EBF"
GRA$PFT <- "GRA"
WET$PFT <- "WET"
DBF$PFT <- "DBF"

```

```{r}
PFT_all <- bind_rows(WSA, SAV, EBF, GRA, WET, DBF)

```


```{r}
library(ggplot2)

ggplot(PFT_all, aes(x = VPD, y = GPP, colour = PFT)) +

  geom_point(alpha = 0.05) +

  geom_smooth(method = "gam",
              formula = y ~ s(x, k = 10),
              se = FALSE,
              linewidth = 1.2) +

  theme_minimal(base_size = 14) +

  labs(
    x = "VPD",
    y = "GPP",
    colour = "PFT",
    title = "GPP–VPD relationships across PFTs"
  )

```
Mean functional response of PFT to atmospheric drying using GAM to characterise non-linear responses of GPP to VPD. 

**Need to bin VPD because these lines look too smooth**


Fit GAM for each PFT
1. No VPD
2. VPD only
3. Full with VPD

What is the additive predictive value of VPD across plant functional groups in Australia?


```{r}
PFT_groups <- c("WSA", "SAV", "EBF", "GRA", "WET", "DBF")

```

Create train/test within each PFT

```{r}
for(group in PFT_groups){
  
  df <- get(group)
  
  idx <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  test <- test %>% drop_na()
  
  train$cumulative_cwd_7day <- rollapply(train$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  assign(paste0(group, "_train"), train, envir = .GlobalEnv)
  assign(paste0(group, "_test"),  test,  envir = .GlobalEnv)
}


```

```{r}
PFT_all <- c("GRA", "EBF", "SAV", "WSA", "WET", "DBF")

all_metrics <- map_dfr(PFT_all, function(PFT) {
  
  train_data <- get(paste0(PFT, "_train"))
  test_data  <- get(paste0(PFT, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(PFT = PFT)
})

```



```{r}
all_metrics %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```

```{r}
library(tidyr)
library(dplyr)

vpd_additive <- all_metrics %>%
  pivot_wider(names_from = Model, values_from = c(R2, RMSE)) %>%
  mutate(
    R2_gain_VPD = R2_Full_VPD - R2_No_VPD,
    RMSE_drop_VPD = RMSE_No_VPD - RMSE_Full_VPD
  )

```

```{r}
library(ggplot2)

ggplot(vpd_additive, aes(x = PFT, y = R2_gain_VPD, fill = PFT)) +
  geom_col() +
  theme_minimal(base_size = 14) +
  labs(
    y = "R² gain from VPD",
    x = "Plant Functional Type",
    title = "Additive predictive value of VPD on GPP"
  )

```

```{r}
ggplot(all_metrics %>% filter(Model %in% c("No_VPD", "Full_VPD")),
       aes(x = PFT, y = R2, colour = Model)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 14) +
  labs(
    y = "R²",
    x = "Plant Functional Type",
    title = "Predictive value of VPD across PFTs"
  )

```
```{r}
full_model_only <- all_metrics %>%
  filter(Model == "Full_VPD")
```

```{r}
library(ggplot2)

ggplot(full_model_only, aes(x = PFT, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Plant Functional Type",
    y = expression(R^2),
    title = "Full GAM predictability of GPP across PFTs"
  )

```
**Predictability decreases when sites are grouped by PFT**
- PFTs are heterogeneous in their GPP-VPD sensitivity
- Plant functional groups do not behave as single functional units 

Predictive performance decreased when sites were aggregated by plant functional type, reflecting the substantial within-PFT heterogeneity in GPP responses to atmospheric and hydrological drivers. This suggests that PFT-level grouping conceals site-specific controls.

This supports emerging literature that PFTs are perhaps too broad - model fit was particularly low compared to site-specific GAMs. 

What if we grouped sites based on their annual precipitation?

Separate sites into very wet, wet, dry, and very dry? Is this better than separating into PFT in terms of predictability?

```{r}
#Look at annual precip again
annual_summary <- annual_precip %>%
  group_by(site) %>%
  summarise(mean_precip = mean(Precip, na.rm = TRUE))

annual_summary <- annual_summary %>%
  arrange(desc(mean_precip)) %>%
  mutate(site = factor(site, levels = site))

ggplot(annual_summary, aes(x = site, y = mean_precip)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "OzFlux site",
    y = "Mean annual precipitation (mm)",
    title = "Mean annual precipitation across Australian Fluxnet sites"
  )
```
23 sites 
4 groups, very wet, wet, dry and very dry
Grouped into 4 groups based on quantile to make it even (although a bit arbitrary), could improve this by finding reference in the literature for what is considered 'wet', 'dry' etc

```{r}
#Separate into quantiles of annual precipitation
site_precip_class <- annual_summary %>%
  mutate(
    precip_quantile = ntile(annual_summary, 4),
    precip_class = case_when(
      precip_quantile == 1 ~ "Very wet",
      precip_quantile == 2 ~ "Wet",
      precip_quantile == 3 ~ "Dry",
      precip_quantile == 4 ~ "Very Dry"
    )
  )

```

```{r}
Very_wet <- bind_rows(Wac, Rob, How, Das, Fog, Rdf)
Wet <- bind_rows(Dap, Ade, Tum, Wom, Dry, Stp)
Dry <- bind_rows(Ync, Emr, Gin, Cum, Rig, Asm)
Very_dry <- bind_rows(Whr, Gww, Cpr, TTE, Lox)
```

```{r}
precip_groups <- c("Very_dry", "Dry", "Wet", "Very_wet")
```


```{r}
#Add in cumulative water deficit and lagged climate variables for each precipitation class
for(group in precip_groups){
  
  df <- get(group)
  
  idx <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train <- df[idx, ]
  test  <- df[-idx, ]
  
  test <- test %>% drop_na()
  
  train$cumulative_cwd_7day <- rollapply(train$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  test$cumulative_cwd_7day  <- rollapply(test$cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  
  train <- train %>% drop_na()
  test  <- test %>% drop_na()
  
  train <- train %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  test <- test %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  assign(paste0(group, "_train"), train, envir = .GlobalEnv)
  assign(paste0(group, "_test"),  test,  envir = .GlobalEnv)
}


```

Fit GAMs for each precip class
```{r}
all_metrics_precip <- map_dfr(precip_groups, function(group) {
  
  train_data <- get(paste0(group, "_train"))
  test_data  <- get(paste0(group, "_test"))
  
  fit_gam_models(train_data, test_data) %>%
    mutate(precip_class = group)
})

```

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

```

```{r}
all_metrics_precip %>%
  filter(Model == "Full_VPD") %>%
  ggplot(aes(x = precip_class, y = R2)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation regime",
    y = expression(R^2),
    title = "Full GAM predictability across precipitation classes"
  )

```
Full model predictability is better for precipitation classes than for plant functional groups, however, it is not as good as site-level sensitivity. 

VPD additive value:

```{r}
vpd_additive_precip <- all_metrics_precip %>%
  select(precip_class, Model, R2) %>%
  pivot_wider(names_from = Model, values_from = R2) %>%
  mutate(
    VPD_additive = Full_VPD - No_VPD
  )

ggplot(vpd_additive_precip,
       aes(x = precip_class, y = VPD_additive)) +
  geom_col(fill = "firebrick") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Precipitation class",
    y = expression(Delta*R^2~"(VPD contribution)"),
    title = "Additive contribution of VPD to GPP predictability"
  )

```
The addition of VPD improved model fit for all precipitation classes, more in dry classes than in wet classes. 

VPD improves predictability more on a wider, climate scale than at an individual site or PFT level scale. PFTs assume all forests/savannas/woody savannas behave similarly, however, within PFTs VPD-GPP relationships are heterogenous.

Precipitation class performs nearly as well as site-specific GAMS, highlighting the importance of hydrological footprint on ecosystem functioning and response. 

Summary:
- VPD needs to be binned before plotting
- site, PFT and precipitation class predictability across Australia
- PFT models perform worse than precip class models 
- VPD has greater additional value in precip class models than PFT and site-level models