---
title: "RF_extreme_events"
output: html_document
date: "2026-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Random Forest
- normal summer conditions versus extreme VPD summer conditions
- Compare driver importance
- Compare site predictability
- Removed LAI (not really a driver of GPP)

Predictors = VPD, Tair, SW_rad, Precip, rolling 7-day cwd
Response = GPP
Method = Leave-one-year out cross validation

Approach 1: VPD and VPD lag 3 90th percentile
Approach 2: Compound drought (VPD + cwd 7 day) 75th percentile

```{r}
library(tidyverse)
library(randomForest)
library(ggplot2)

# =============================================================================
# 1. PARAMETERS
# =============================================================================

SUMMER_MONTHS  <- 6:8   # JJA
MIN_SUMMERS    <- 2     # minimum summers required per site
RF_NTREE       <- 500
MIN_OBS        <- 20    # minimum observations to fit a model

# Approach 1 thresholds
A1_PERCENTILE  <- 0.90  # 90th percentile for sustained VPD

# Approach 2 thresholds
A2_VPD_PCTILE  <- 0.75  # 75th percentile for compound VPD
A2_CWD_PCTILE  <- 0.75  # 75th percentile for compound CWD

PREDICTORS     <- c("VPD", "Tair", "SW_rad", "Precip", "cwd_7day")

# =============================================================================
# 2. PREPARE BASE DATA
# Assumes sites_daily has: site, date, GPP, VPD, Tair, SW_rad, Precip, cwd
# =============================================================================

data_prepared <- bind_rows(sites_daily) %>%
  mutate(date = as.Date(date)) %>%
  arrange(site, date) %>%
  group_by(site) %>%
  mutate(
    cwd_7day  = zoo::rollapply(cwd, width = 7, FUN = sum, align = "right", fill = NA),
    VPD_lag3  = lag(VPD, 3),
    month     = month(date),
    year      = year(date),
    is_summer = month %in% SUMMER_MONTHS
  ) %>%
  ungroup()

# Filter to sites with at least MIN_SUMMERS summers
valid_sites <- data_prepared %>%
  filter(is_summer) %>%
  group_by(site) %>%
  summarise(n_summers = n_distinct(year), .groups = "drop") %>%
  filter(n_summers >= MIN_SUMMERS) %>%
  pull(site)

message("Excluding: ", paste(setdiff(unique(data_prepared$site), valid_sites), collapse = ", "))

data_prepared <- data_prepared %>% filter(site %in% valid_sites)

# =============================================================================
# 3. DEFINE REGIMES
# =============================================================================

# --- Approach 1: Sustained VPD ---
# Extreme = today AND 3 days ago both above site-specific 90th percentile
# Normal  = all remaining summer days

thresholds_a1 <- data_prepared %>%
  filter(is_summer, !is.na(VPD)) %>%
  group_by(site) %>%
  summarise(vpd_90 = quantile(VPD, A1_PERCENTILE), .groups = "drop")

data_a1 <- data_prepared %>%
  left_join(thresholds_a1, by = "site") %>%
  mutate(
    regime = case_when(
      !is_summer                          ~ NA_character_,
      VPD >= vpd_90 & VPD_lag3 >= vpd_90 ~ "extreme",
      TRUE                                ~ "normal"
    )
  ) %>%
  filter(is_summer, !is.na(regime))

# --- Approach 2: Compound drought ---
# Extreme = VPD AND cwd_7day both above site-specific 75th percentile
# Normal  = all remaining summer days (including middle ground)

thresholds_a2 <- data_prepared %>%
  filter(is_summer, !is.na(VPD), !is.na(cwd_7day)) %>%
  group_by(site) %>%
  summarise(
    vpd_75 = quantile(VPD,      A2_VPD_PCTILE),
    cwd_75 = quantile(cwd_7day, A2_CWD_PCTILE),
    .groups = "drop"
  )

data_a2 <- data_prepared %>%
  left_join(thresholds_a2, by = "site") %>%
  mutate(
    regime = case_when(
      !is_summer                          ~ NA_character_,
      VPD >= vpd_75 & cwd_7day >= cwd_75 ~ "extreme",
      TRUE                                ~ "normal"
    )
  ) %>%
  filter(is_summer, !is.na(regime))

# =============================================================================
# 4. VISUALISE REGIMES — check before fitting
# =============================================================================

sites_list   <- unique(data_a1$site)
site_groups  <- split(sites_list, ceiling(seq_along(sites_list) / 6))

# Approach 1 — GPP
for (grp in site_groups) {
  p <- data_a1 %>%
    filter(site %in% grp) %>%
    ggplot(aes(x = date, y = GPP, colour = regime)) +
    geom_point(size = 0.8, alpha = 0.6) +
    scale_colour_manual(values = c("normal" = "steelblue", "extreme" = "red")) +
    facet_wrap(~ site, scales = "free", ncol = 2) +
    labs(title = "Approach 1 — Sustained VPD: Summer GPP", colour = "regime") +
    theme_bw()
  print(p)
}

# Approach 2 — GPP
for (grp in site_groups) {
  p <- data_a2 %>%
    filter(site %in% grp) %>%
    ggplot(aes(x = date, y = GPP, colour = regime)) +
    geom_point(size = 0.8, alpha = 0.6) +
    scale_colour_manual(values = c("normal" = "steelblue", "extreme" = "red")) +
    facet_wrap(~ site, scales = "free", ncol = 2) +
    labs(title = "Approach 2 — Compound Drought: Summer GPP", colour = "regime") +
    theme_bw()
  print(p)
}

# Observation counts — good to check before fitting
message("--- Approach 1 observation counts ---")
print(data_a1 %>% count(site, regime) %>% pivot_wider(names_from = regime, values_from = n))

message("--- Approach 2 observation counts ---")
print(data_a2 %>% count(site, regime) %>% pivot_wider(names_from = regime, values_from = n))

```

```{r}

# =============================================================================
# 5. FIT RANDOM FOREST WITH LEAVE-ONE-YEAR-OUT CV
# =============================================================================

fit_rf_loyo_cv <- function(site_df, regime_label) {

  df <- site_df %>%
    filter(regime == regime_label) %>%
    select(year, GPP, all_of(PREDICTORS)) %>%
    drop_na()

  years <- unique(df$year)
  n_obs <- nrow(df)

  if (length(years) < 2) {
    message("  Skipping ", regime_label, ": fewer than 2 years with data")
    return(NULL)
  }

  if (n_obs < MIN_OBS) {
    message("  Skipping ", regime_label, ": only ", n_obs, " obs (min = ", MIN_OBS, ")")
    return(NULL)
  }

  # Leave-one-year-out CV
  cv_results <- map_dfr(years, function(test_year) {
    train <- df %>% filter(year != test_year)
    test  <- df %>% filter(year == test_year)

    if (nrow(train) < 10 || nrow(test) < 5) return(NULL)

    rf    <- randomForest(x = train[, PREDICTORS], y = train$GPP,
                          ntree = RF_NTREE, importance = TRUE)
    preds <- predict(rf, newdata = test[, PREDICTORS])

    ss_res <- sum((test$GPP - preds)^2)
    ss_tot <- sum((test$GPP - mean(test$GPP))^2)

    tibble(test_year = test_year, r2 = 1 - ss_res / ss_tot, n_test = nrow(test))
  })

  if (nrow(cv_results) == 0 || !"r2" %in% names(cv_results)) {
    message("  Skipping ", regime_label, ": no valid CV folds")
    return(NULL)
  }

  # Final model on all data for importance
  final_rf <- randomForest(x = df[, PREDICTORS], y = df$GPP,
                           ntree = RF_NTREE, importance = TRUE)

  importance_df <- importance(final_rf, type = 1) %>%
    as.data.frame() %>%
    rownames_to_column("predictor") %>%
    rename(importance = `%IncMSE`)

  list(
    cv_results = cv_results,
    importance = importance_df,
    n_obs      = n_obs,
    mean_r2    = mean(cv_results$r2, na.rm = TRUE),
    sd_r2      = sd(cv_results$r2,   na.rm = TRUE)
  )
}

run_all_sites <- function(df, approach_label) {
  sites <- unique(df$site)
  results <- map(sites, function(s) {
    message("Fitting ", approach_label, " — site: ", s)
    site_df <- df %>% filter(site == s)
    list(
      site    = s,
      normal  = fit_rf_loyo_cv(site_df, "normal"),
      extreme = fit_rf_loyo_cv(site_df, "extreme")
    )
  })
  names(results) <- sites
  results
}

# =============================================================================
# 6. RUN BOTH APPROACHES
# =============================================================================

message("========== Approach 1: Sustained VPD ==========")
results_a1 <- run_all_sites(data_a1, "A1")

message("========== Approach 2: Compound Drought ==========")
results_a2 <- run_all_sites(data_a2, "A2")

# =============================================================================
# 7. SUMMARISE & EXTRACT IMPORTANCE
# =============================================================================

summarise_results <- function(results, approach) {
  map_dfr(results, function(res) {
    bind_rows(
      if (!is.null(res$normal))
        tibble(approach = approach, site = res$site, regime = "normal",
               mean_r2 = res$normal$mean_r2, sd_r2 = res$normal$sd_r2,
               n_obs   = res$normal$n_obs),
      if (!is.null(res$extreme))
        tibble(approach = approach, site = res$site, regime = "extreme",
               mean_r2 = res$extreme$mean_r2, sd_r2 = res$extreme$sd_r2,
               n_obs   = res$extreme$n_obs)
    )
  })
}

extract_importance <- function(results, approach) {
  map_dfr(results, function(res) {
    bind_rows(
      if (!is.null(res$normal))
        res$normal$importance %>%
          mutate(approach = approach, site = res$site, regime = "normal"),
      if (!is.null(res$extreme))
        res$extreme$importance %>%
          mutate(approach = approach, site = res$site, regime = "extreme")
    )
  })
}

summary_df <- bind_rows(
  summarise_results(results_a1, "A1: Sustained VPD"),
  summarise_results(results_a2, "A2: Compound Drought")
)

importance_df <- bind_rows(
  extract_importance(results_a1, "A1: Sustained VPD"),
  extract_importance(results_a2, "A2: Compound Drought")
)

print(summary_df)

# =============================================================================
# 8. PLOT RESULTS
# =============================================================================

# --- R² comparison by site and approach ---
ggplot(summary_df, aes(x = reorder(site, mean_r2), y = mean_r2, colour = regime)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = mean_r2 - sd_r2, ymax = mean_r2 + sd_r2),
                width = 0.2, position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("normal" = "steelblue", "extreme" = "red")) +
  facet_wrap(~ approach) +
  coord_flip() +
  labs(title = "Model R² by site and regime",
       x = "Site", y = "Mean R² (LOYO-CV)", colour = "Regime") +
  theme_bw()

# --- Mean predictor importance across sites ---
importance_df %>%
  group_by(approach, regime, predictor) %>%
  summarise(mean_imp = mean(importance, na.rm = TRUE),
            sd_imp   = sd(importance,   na.rm = TRUE),
            .groups  = "drop") %>%
  ggplot(aes(x = reorder(predictor, mean_imp), y = mean_imp, fill = regime)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_imp - sd_imp, ymax = mean_imp + sd_imp),
                position = position_dodge(width = 0.7), width = 0.25) +
  scale_fill_manual(values = c("normal" = "steelblue", "extreme" = "red")) +
  facet_wrap(~ approach) +
  coord_flip() +
  labs(title = "Mean predictor importance: normal vs extreme",
       x = "Predictor", y = "Mean % increase in MSE", fill = "Regime") +
  theme_bw()

# --- Importance by site ---
importance_df %>%
  ggplot(aes(x = predictor, y = importance, fill = regime)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("normal" = "steelblue", "extreme" = "red")) +
  facet_wrap(~ site, scales = "free_x") +
  coord_flip() +
  labs(title = "Predictor importance by site",
       x = "Predictor", y = "% increase in MSE", fill = "Regime") +
  theme_bw() +
  theme(legend.position = "top")
```
Is VPD becoming more important in recent years in long-term sites (How and Tum)?

```{r}
library(tidyverse)
library(randomForest)

# Extract How and Tum
how_df <- sites_daily[["How"]] %>% mutate(site = "How")
tum_df <- sites_daily[["Tum"]] %>% mutate(site = "Tum")

long_sites <- bind_rows(how_df, tum_df) %>%
  mutate(
    date      = as.Date(date),
    month     = month(date),
    year      = year(date),
    is_summer = month %in% 6:8
  ) %>%
  filter(is_summer) %>%
  mutate(
    cumulative_cwd_7day = zoo::rollapply(cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  ) %>%
  drop_na(GPP, Tair, SW_rad, VPD, Precip, cumulative_cwd_7day)

# Fit RF per site per year and extract VPD importance
predictors <- c("Tair", "SW_rad", "VPD", "Precip", "cumulative_cwd_7day")

yearly_results <- long_sites %>%
  group_by(site, year) %>%
  group_map(~ {
    df   <- .x
    site <- .y$site
    yr   <- .y$year

    # Need enough observations to fit a meaningful model
    if (nrow(df) < 20) {
      message("Skipping ", site, " ", yr, ": only ", nrow(df), " obs")
      return(NULL)
    }

    # Simple train/test split within year
    set.seed(123)
    idx        <- sample(1:nrow(df), size = floor(0.8 * nrow(df)))
    train_data <- df[idx, ]
    test_data  <- df[-idx, ]

    rf <- randomForest(
      x          = train_data[, predictors],
      y          = train_data$GPP,
      ntree      = 500,
      importance = TRUE
    )

    # R² on test set
    preds  <- predict(rf, newdata = test_data[, predictors])
    ss_res <- sum((test_data$GPP - preds)^2)
    ss_tot <- sum((test_data$GPP - mean(test_data$GPP))^2)
    r2     <- 1 - ss_res / ss_tot

    # Variable importance
    imp <- importance(rf, type = 1) %>%
      as.data.frame() %>%
      rownames_to_column("variable") %>%
      rename(importance = `%IncMSE`) %>%
      mutate(site = site, year = yr, r2 = r2)

    imp
  }) %>%
  bind_rows()

# ---- Plot 1: VPD importance over time ----
yearly_results %>%
  filter(variable == "VPD") %>%
  ggplot(aes(x = year, y = importance, colour = site)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_colour_manual(values = c("How" = "#2E86AB", "Tum" = "#A23B72")) +
  facet_wrap(~ site) +
  labs(title = "VPD importance over time (summer RF)",
       x = "Year", y = "VPD % increase in MSE") +
  theme_classic() +
  theme(legend.position = "none")

# ---- Plot 2: All predictor importance over time ----
yearly_results %>%
  ggplot(aes(x = year, y = importance, colour = variable)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ site) +
  labs(title = "Predictor importance over time (summer RF)",
       x = "Year", y = "% increase in MSE", colour = "Predictor") +
  theme_classic()

# ---- Plot 3: R² over time ----
yearly_results %>%
  distinct(site, year, r2) %>%
  ggplot(aes(x = year, y = r2, colour = site)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_colour_manual(values = c("How" = "#2E86AB", "Tum" = "#A23B72")) +
  facet_wrap(~ site) +
  labs(title = "Model R² over time (summer RF)",
       x = "Year", y = "R²") +
  theme_classic() +
  theme(legend.position = "none")
```
```{r}
plot(sites_daily$How$date, sites_daily$How$GPP)
```
```{r}
plot(sites_daily$Tum$date, sites_daily$Tum$GPP)
```
```{r}
library(tidyverse)

# Extract and label sites
how_df <- sites_daily[["How"]] %>% mutate(site = "Howard Springs")
tum_df <- sites_daily[["Tum"]] %>% mutate(site = "Tumbarumba")

drought_df <- bind_rows(how_df, tum_df) %>%
  mutate(
    date  = as.Date(date),
    year  = year(date),
    month = month(date)
  ) %>%
  filter(year >= 2001 & year <= 2012)

# Define periods
drought_df <- drought_df %>%
  mutate(period = case_when(
    year >= 2001 & year <= 2005 ~ "Shared drought (2001-2005)",
    year >= 2006 & year <= 2007 ~ "Divergence (2006-2007)",
    year >= 2008 & year <= 2009 ~ "Continued drought (2008-2009)",
    year >= 2010 & year <= 2011 ~ "La Nina recovery (2010-2011)",
    TRUE ~ "Other"
  ))

# Shading df for background periods
period_shading <- tibble(
  xmin   = as.Date(c("2001-01-01", "2006-01-01", "2008-01-01", "2010-01-01")),
  xmax   = as.Date(c("2005-12-31", "2007-12-31", "2009-12-31", "2011-12-31")),
  period = c("Shared drought (2001-2005)", "Divergence (2006-2007)",
             "Continued drought (2008-2009)", "La Nina recovery (2010-2011)"),
  fill   = c("#D4E6F1", "#FADBD8", "#FDEBD0", "#D5F5E3")
)

period_colours <- c(
  "Shared drought (2001-2005)"    = "#D4E6F1",
  "Divergence (2006-2007)"        = "#FADBD8",
  "Continued drought (2008-2009)" = "#FDEBD0",
  "La Nina recovery (2010-2011)"  = "#D5F5E3"
)

# Helper to add period shading to any plot
add_shading <- function(p) {
  for (i in 1:nrow(period_shading)) {
    p <- p + annotate("rect",
                      xmin  = period_shading$xmin[i],
                      xmax  = period_shading$xmax[i],
                      ymin  = -Inf, ymax = Inf,
                      fill  = period_shading$fill[i],
                      alpha = 0.4)
  }
  p
}

# ---- Annual means per site ----
annual_df <- drought_df %>%
  group_by(site, year) %>%
  summarise(
    mean_GPP   = mean(GPP,   na.rm = TRUE),
    mean_VPD   = mean(VPD,   na.rm = TRUE),
    mean_Precip = mean(Precip, na.rm = TRUE),
    mean_Tair  = mean(Tair,  na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  mutate(date = as.Date(paste0(year, "-07-01")))  # mid-year date for plotting

site_colours <- c("Howard Springs" = "#2E86AB", "Tumbarumba" = "#A23B72")

# ---- Plot 1: GPP ----
p_gpp <- ggplot(annual_df, aes(x = date, y = mean_GPP, colour = site)) +
  annotate("rect", xmin = period_shading$xmin, xmax = period_shading$xmax,
           ymin = -Inf, ymax = Inf, fill = period_shading$fill, alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_colour_manual(values = site_colours) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Mean daily GPP", colour = NULL,
       title = "GPP during the Millennium Drought") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_gpp)

# ---- Plot 2: VPD ----
p_vpd <- ggplot(annual_df, aes(x = date, y = mean_VPD, colour = site)) +
  annotate("rect", xmin = period_shading$xmin, xmax = period_shading$xmax,
           ymin = -Inf, ymax = Inf, fill = period_shading$fill, alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_colour_manual(values = site_colours) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Mean VPD (kPa)", colour = NULL,
       title = "VPD during the Millennium Drought") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_vpd)

# ---- Plot 3: Precipitation ----
p_precip <- ggplot(annual_df, aes(x = date, y = mean_Precip, colour = site)) +
  annotate("rect", xmin = period_shading$xmin, xmax = period_shading$xmax,
           ymin = -Inf, ymax = Inf, fill = period_shading$fill, alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_colour_manual(values = site_colours) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Mean daily precipitation (mm)", colour = NULL,
       title = "Precipitation during the Millennium Drought") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_precip)

# ---- Plot 4: Temperature ----
p_tair <- ggplot(annual_df, aes(x = date, y = mean_Tair, colour = site)) +
  annotate("rect", xmin = period_shading$xmin, xmax = period_shading$xmax,
           ymin = -Inf, ymax = Inf, fill = period_shading$fill, alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_colour_manual(values = site_colours) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = "Year", y = "Mean temperature (°C)", colour = NULL,
       title = "Temperature during the Millennium Drought") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_tair)
```
```{r}
# ---- RF importance by drought period ----

predictors <- c("Tair", "SW_rad", "VPD", "Precip", "cumulative_cwd_7day")

period_df <- drought_df %>%
  filter(period != "Other") %>%
  arrange(site, date) %>%
  group_by(site) %>%
  mutate(
    cumulative_cwd_7day = zoo::rollapply(cwd, width = 7, FUN = sum,
                                         align = "right", fill = NA, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  drop_na(GPP, all_of(predictors))

# Fit RF per site per period
period_results <- period_df %>%
  group_by(site, period) %>%
  group_map(~ {
    df     <- .x
    site   <- .y$site
    period <- .y$period

    if (nrow(df) < 50) {
      message("Skipping ", site, " ", period, ": only ", nrow(df), " obs")
      return(NULL)
    }

    set.seed(123)
    idx        <- sample(1:nrow(df), size = floor(0.8 * nrow(df)))
    train_data <- df[idx, ]
    test_data  <- df[-idx, ]

    rf <- randomForest(
      x          = train_data[, predictors],
      y          = train_data$GPP,
      ntree      = 500,
      importance = TRUE
    )

    preds  <- predict(rf, newdata = test_data[, predictors])
    ss_res <- sum((test_data$GPP - preds)^2)
    ss_tot <- sum((test_data$GPP - mean(test_data$GPP))^2)
    r2     <- 1 - ss_res / ss_tot

    imp <- importance(rf, type = 1) %>%
      as.data.frame() %>%
      rownames_to_column("variable") %>%
      rename(importance = `%IncMSE`) %>%
      mutate(site = site, period = period, r2 = r2, n_obs = nrow(df))

    imp
  }) %>%
  bind_rows()

# Set period order for plotting
period_order <- c(
  "Shared drought (2001-2005)",
  "Divergence (2006-2007)",
  "Continued drought (2008-2009)",
  "La Nina recovery (2010-2011)"
)

period_results <- period_results %>%
  mutate(period = factor(period, levels = period_order))

# ---- Plot 1: VPD importance by period at each site ----
period_results %>%
  filter(variable == "VPD") %>%
  ggplot(aes(x = period, y = importance, fill = site)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("Howard Springs" = "#2E86AB",
                                "Tumbarumba"     = "#A23B72")) +
  labs(title = "VPD importance by drought period",
       x = NULL, y = "% increase in MSE", fill = NULL) +
  theme_classic() +
  theme(legend.position  = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))

# ---- Plot 2: All predictor importance by period, faceted by site ----
period_results %>%
  ggplot(aes(x = variable, y = importance, fill = period)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c(
    "Shared drought (2001-2005)"    = "#AED6F1",
    "Divergence (2006-2007)"        = "#F1948A",
    "Continued drought (2008-2009)" = "#FAD7A0",
    "La Nina recovery (2010-2011)"  = "#A9DFBF"
  )) +
  facet_wrap(~ site) +
  coord_flip() +
  labs(title = "Predictor importance by drought period",
       x = "Predictor", y = "% increase in MSE", fill = "Period") +
  theme_classic() +
  theme(legend.position = "top")

# ---- Plot 3: R² by period and site ----
period_results %>%
  distinct(site, period, r2, n_obs) %>%
  ggplot(aes(x = period, y = r2, fill = site)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0("n=", n_obs)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Howard Springs" = "#2E86AB",
                                "Tumbarumba"     = "#A23B72")) +
  labs(title = "Model predictability (R²) by drought period",
       x = NULL, y = "R²", fill = NULL) +
  theme_classic() +
  theme(legend.position  = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

