---
title: "future_envelope_final"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1) *Frequency vs magnitude*
Is GPP more sensitive to increases in frequency of VPD extremes or magnitude of VPD extremes?

15, 30 and 50% increase in the number of days above 90th percentile in the summer (Extreme VPD events become more common). Replace summer days with 90th percentile value. Replace days that are close to the threshold already

15, 30 and 50% increase in magnitude of the 90th percentile VPD extremes (Extremes get more intense). Summer VPD values equal to or above 90th percentile get increased in magnitude. Whilst the other days stay the same.

Random Forest is leave-one-year out trained and tested for each scenario. Trained on all years except one, tested on one year. Repeated for all years and concatenated together. 

**Increase in frequency**

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

#Setup
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(0, 0.15, 0.30, 0.50)
scenario_labels <- c("Observed", "+15% extreme days", "+30% extreme days", "+50% extreme days")

scenario_colours <- c(
  "Observed"           = "#4682B4",
  "+15% extreme days"  = "#F28C28",
  "+30% extreme days"  = "#E06C4B",
  "+50% extreme days"  = "#8B1A1A"
)
```

```{r}
run_loyo_summer_extreme <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  #90th percentile VPD from summer obs across all years
  vpd_90 <- df %>%
    filter(summer) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)
  
  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }
  
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(hydro_year != test_year)
    
    # ── Full held-out year 
    test_nonsummer <- df %>% filter(hydro_year == test_year, !summer)
    test_summer    <- df %>% filter(hydro_year == test_year,  summer)
    
    if (nrow(train_df) < 50 || nrow(test_summer) < 10) return(NULL)
    
    # Standardise on full training set
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)
    
    # Identify extreme vs normal days in test summer
    extreme_days <- test_summer %>% filter(VPD >= vpd_90)
    normal_days  <- test_summer %>% filter(VPD <  vpd_90)
    n_normal     <- nrow(normal_days)
    n_extreme    <- nrow(extreme_days)
    
    map2_dfr(vpd_scalars, scenario_labels, function(extra_frac, label) {
      
      # Manipulate summer days only 
      if (extra_frac == 0) {
        summer_scenario <- test_summer
        
      } else {
        n_replace <- round(n_normal * extra_frac)
        
        if (n_replace == 0 || n_extreme == 0) {
          summer_scenario <- test_summer
        } else {
          extra_extreme <- extreme_days %>%
            slice_sample(n = n_replace, replace = TRUE)
          
          normal_kept <- normal_days %>%
            slice_sample(n = max(0, n_normal - n_replace), replace = FALSE)
          
          summer_scenario <- bind_rows(normal_kept, extreme_days, extra_extreme) %>%
            arrange(date)
        }
      }
      
      # Combine manipulated summer + unchanged non-summer
      full_test <- bind_rows(test_nonsummer, summer_scenario) %>%
        arrange(date)
      
      full_scaled <- scale_with_params(full_test, scale_means, scale_sds)
      preds       <- predict(rf_model, full_scaled)
      
      data.frame(
        site           = site_name,
        date           = full_test$date,
        year           = test_year,
        scenario       = label,
        summer         = full_test$summer,
        VPD_raw        = full_test$VPD,
        n_extreme_days = sum(full_test$summer & full_test$VPD >= vpd_90),
        n_total_days   = nrow(full_test),
        GPP_pred       = preds,
        GPP_obs        = full_test$GPP
      )
    })
  })
  
  loyo_results
}

#Run across all sites
loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_summer_extreme)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggrepel)

#Plot 1: R-squared per site (full year only) 

r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

#Plot 2: % change in annual GPP per scenario per site

scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15% extreme days",
                                           "+30% extreme days",
                                           "+50% extreme days")),
    site = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = c(
    "+15% extreme days" = "#F28C28",
    "+30% extreme days" = "#E06C4B",
    "+50% extreme days" = "#8B1A1A"
  )) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean annual GPP relative to observed",
    fill = "Extreme VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

#Plot 3: R-squared vs % change coloured by cluster group

# Use summer R² as most relevant for the scenario
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15% extreme days",
                                                 "+30% extreme days",
                                                 "+50% extreme days")))

cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

2) *Baseline and extremes*

Sort sites into tropical, arid and temperate
Tropical (+0.524 kPa), Arid (+0.644 kPa) and Temperate (+0.344 Kpa) increase to ALL daily VPD values (Fang et al., 2022).

Biome-specific background increases, accounts for the fact that arid systems will face larger increases in VPD stress than temperate ones. Applying a uniform shift to every daily value in the whole dataset.

Background rising VPD for each climate AND +28% magnitude of VPD values equal to and above the 90th percentile of VPD in the summer (Zhou et al., 2019)

Random Forest is leave-one-year out trained and tested for each scenario. Trained on all years except one, tested on one year. Repeated for all years and concatenated together. 

Tropical - Ade, Dap, Das, Dry, Fog, Gin, How, Rdf, Stp
Arid - Asm, Cpr, Emr Gww, Ync
Temperate - Boy, Cum, Rig, Tum, Wac, Whr, Wom

Beck et al., 2018

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggrepel)

#Setup 

predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

scenario_labels  <- c("Observed", "Future VPD")
scenario_colours <- c("Observed" = "#4682B4", "Future VPD" = "#C0392B")

biome_vpd_increase <- c(
  Ade = 0.524, Dap = 0.524, Das = 0.524, Dry = 0.524,
  Fog = 0.524, Gin = 0.524, How = 0.524, Rdf = 0.524, Stp = 0.524,
  Asm = 0.644, Cpr = 0.644, Emr = 0.644, Gww = 0.644, Ync = 0.644,
  Boy = 0.344, Cum = 0.344, Rig = 0.344, Tum = 0.344,
  Wac = 0.344, Whr = 0.344, Wom = 0.344
)

biome_labels <- c(
  Ade = "Tropical", Dap = "Tropical", Das = "Tropical", Dry = "Tropical",
  Fog = "Tropical", Gin = "Tropical", How = "Tropical", Rdf = "Tropical", Stp = "Tropical",
  Asm = "Arid", Cpr = "Arid", Emr = "Arid", Gww = "Arid", Ync = "Arid",
  Boy = "Temperate", Cum = "Temperate", Rig = "Temperate", Tum = "Temperate",
  Wac = "Temperate", Whr = "Temperate", Wom = "Temperate"
)

biome_colours <- c(
  "Tropical"  = "#2E8B57",
  "Arid"      = "#C8860A",
  "Temperate" = "#4682B4"
)

# LOYO function

run_loyo_biome_vpd <- function(df, site_name) {

  message("Processing: ", site_name)

  vpd_delta <- biome_vpd_increase[site_name]
  if (is.na(vpd_delta)) {
    warning("Skipping ", site_name, ": no biome VPD delta defined")
    return(NULL)
  }

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)

  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }

  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }

  loyo_results <- map_dfr(all_years, function(test_year) {

    train_df <- df %>% filter(hydro_year != test_year)
    test_df  <- df %>% filter(hydro_year == test_year)

    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    # Standardise on training set parameters
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)

    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }

    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)

    map_dfr(scenario_labels, function(label) {

      test_scenario <- if (label == "Observed") {
        test_df
      } else {
        test_df %>% mutate(VPD = VPD + vpd_delta)
      }

      test_scaled <- scale_with_params(test_scenario, scale_means, scale_sds)
      preds       <- predict(rf_model, test_scaled)

      data.frame(
        site        = site_name,
        date        = test_scenario$date,
        year        = test_year,
        scenario    = label,
        summer      = test_scenario$summer,
        VPD_raw     = test_scenario$VPD,
        n_total_days = nrow(test_scenario),
        GPP_pred    = preds,
        GPP_obs     = test_scenario$GPP
      )
    })
  })

  loyo_results
}

# Run across all sites 

loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_biome_vpd)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    biome    = biome_labels[site]
  )

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))

```

```{r}
#Plot 1: R-squared per site coloured by biome 

r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    biome = biome_labels[site],
    site  = factor(site, levels = wet_to_dry_order)
  )

ggplot(r2_loyo, aes(x = site, y = R2, fill = biome)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  scale_fill_manual(values = biome_colours) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)", fill = "Biome") +
  theme_classic() +
  theme(legend.position = "top", axis.text.y = element_text(size = 10))

#Plot 2: % change in GPP per site, faceted by biome

scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change = (`Future VPD` - Observed) / Observed * 100,
    biome      = biome_labels[site],
    site       = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = biome)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = biome_colours) +
  coord_flip() +
  facet_wrap(~ biome, scales = "free_y", ncol = 1) +
  labs(
    x    = NULL,
    y    = "% change in mean annual GPP relative to observed",
    fill = "Biome"
  ) +
  theme_classic() +
  theme(legend.position = "none", axis.text.y = element_text(size = 10),
        strip.text = element_text(face = "bold"))

#Plot 3: R-squared vs % change coloured by cluster group, faceted by biome

r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2, .groups = "drop")

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(pct_change = (`Future VPD` - Observed) / Observed * 100)

cluster_colours <- c("Low" = "steelblue", "Medium" = "orange", "High" = "firebrick")

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(
    cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")),
    biome         = biome_labels[site]
  )

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30", max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ biome, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

```

Now need to do +28% magnitude 

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggrepel)

#Setup

predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

scenario_labels  <- c("Observed", "+28% extreme VPD")
scenario_colours <- c("Observed" = "#4682B4", "+28% extreme VPD" = "#C0392B")

biome_labels <- c(
  Ade = "Tropical", Dap = "Tropical", Das = "Tropical", Dry = "Tropical",
  Fog = "Tropical", Gin = "Tropical", How = "Tropical", Rdf = "Tropical", Stp = "Tropical",
  Asm = "Arid", Cpr = "Arid", Emr = "Arid", Gww = "Arid", Ync = "Arid",
  Boy = "Temperate", Cum = "Temperate", Rig = "Temperate", Tum = "Temperate",
  Wac = "Temperate", Whr = "Temperate", Wom = "Temperate"
)

biome_colours <- c(
  "Tropical"  = "#2E8B57",
  "Arid"      = "#C8860A",
  "Temperate" = "#4682B4"
)

vpd_increase <- 0.28  # 28% multiplicative increase on extreme days

#LOYO function

run_loyo_vpd_magnitude <- function(df, site_name) {

  message("Processing: ", site_name)

  if (!site_name %in% names(biome_labels)) {
    warning("Skipping ", site_name, ": no biome defined")
    return(NULL)
  }

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)

  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }

  # 90th percentile VPD from summer obs across all years, per site
  vpd_90 <- df %>%
    filter(summer) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)

  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }

  loyo_results <- map_dfr(all_years, function(test_year) {

    train_df <- df %>% filter(hydro_year != test_year)
    test_df  <- df %>% filter(hydro_year == test_year)

    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    # Standardise on training set parameters
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)

    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }

    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)

    map_dfr(scenario_labels, function(label) {

      test_scenario <- if (label == "Observed") {
        test_df
      } else {
        # Increase VPD by 28% on summer extreme days only
        test_df %>%
          mutate(VPD = ifelse(summer & VPD >= vpd_90, VPD * (1 + vpd_increase), VPD))
      }

      test_scaled <- scale_with_params(test_scenario, scale_means, scale_sds)
      preds       <- predict(rf_model, test_scaled)

      data.frame(
        site             = site_name,
        date             = test_scenario$date,
        year             = test_year,
        scenario         = label,
        summer           = test_scenario$summer,
        VPD_raw          = test_scenario$VPD,
        n_extreme_days   = sum(test_scenario$summer & test_scenario$VPD >= vpd_90),
        n_total_days     = nrow(test_scenario),
        GPP_pred         = preds,
        GPP_obs          = test_scenario$GPP
      )
    })
  })

  loyo_results
}

#Run across all sites

loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_vpd_magnitude)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    biome    = biome_labels[site]
  )

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))

#Plot 1: R² per site coloured by biome

r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    biome = biome_labels[site],
    site  = factor(site, levels = wet_to_dry_order)
  )

ggplot(r2_loyo, aes(x = site, y = R2, fill = biome)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  scale_fill_manual(values = biome_colours) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)", fill = "Biome") +
  theme_classic() +
  theme(legend.position = "top", axis.text.y = element_text(size = 10))

#Plot 2: % change in GPP per site, by biome

scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change = (`+28% extreme VPD` - Observed) / Observed * 100,
    biome      = biome_labels[site],
    site       = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = biome)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = biome_colours) +
  coord_flip() +
  facet_wrap(~ biome, scales = "free_y", ncol = 1) +
  labs(
    x    = NULL,
    y    = "% change in mean annual GPP relative to observed",
    fill = "Biome"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.y     = element_text(size = 10),
    strip.text      = element_text(face = "bold")
  )

#Plot 3: R² vs % change coloured by cluster group and grouped by biome

r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2, .groups = "drop")

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(pct_change = (`+28% extreme VPD` - Observed) / Observed * 100)

cluster_colours <- c("Low" = "steelblue", "Medium" = "orange", "High" = "firebrick")

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(
    cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")),
    biome         = biome_labels[site]
  )

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30", max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ biome, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```


