---
title: "explore_ozflux_03"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Monday 2nd Feb:

- Add vcmax and LLC to GAMs

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      LE = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn", "LE")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}
```
```{r}
AU_How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv") 

AU_Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

AU_Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

AU_Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")
```

```{r}
flux_vars <- c("GPP", "LE", "Tair", "SW_rad", "Rn", "Precip", "Pa")

AU_How <- AU_How %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Dap <- AU_Dap %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Das <- AU_Das %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Dry <- AU_Dry %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))


AU_Stp <- AU_Stp %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Asm <- AU_Asm %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = LE * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
      Precip = Precip * SEC_2_HALF_HOUR
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(site, date) %>%
    summarise(
      Precip = ifelse(
        sum(!is.na(Precip)) > 0,
        sum(Precip, na.rm = TRUE),
        NA
      ),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      LE = mean(LE, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(df)
  
}
```

```{r}
How_daily <- convert_to_daily(AU_How)
Dap_daily <- convert_to_daily(AU_Dap)
Das_daily <- convert_to_daily(AU_Das)
Dry_daily <- convert_to_daily(AU_Dry)
Stp_daily <- convert_to_daily(AU_Stp)
Asm_daily <- convert_to_daily(AU_Asm)
```

```{r}
How_daily <- How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

Dap_daily <- Dap_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

Das_daily <- Das_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

Dry_daily <- Dry_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

Stp_daily <- Stp_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

Asm_daily <- Asm_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))
```

```{r}
How_daily <- How_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)

Dap_daily <- Dap_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)

Das_daily <- Das_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)

Dry_daily <- Dry_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)

Stp_daily <- Stp_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)

Asm_daily <- Asm_daily %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, SW_rad, LE)
```

Add in doy and year again:

```{r}
How_daily <- How_daily %>%
  mutate(
    year = lubridate::year(date),
    doy  = lubridate::yday(date)
  )

Dap_daily <- Dap_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Das_daily <- Das_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
    )

Dry_daily <- Dry_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Stp_daily <- Stp_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )

Asm_daily <- Asm_daily %>%
  mutate(
    year = lubridate::year(date),
    doy = lubridate::yday(date)
  )
```


```{r}
install.packages("ncdf4")
```

Load in Vcmax data:

```{r}
library(terra)
library(ncdf4)

#AU How 
latitude <- -12.4943
longitude <- 131.1523

#
## Vcmax from GOME-2 SIF
#
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/GOME2_VcmaxTg_05deg.tif"

vcmax_gome_raster <- rast(gome_file)
coords <- data.frame(lon = longitude, lat = latitude)
vcmax_gome <- terra::extract(vcmax_gome_raster, coords)
vcmax_gome <- data.frame(lat = latitude, lon = longitude, 
                           Vcmax = vcmax_gome)
vcmax_gome <- vcmax_gome$Vcmax.GOME2_VcmaxTg_05deg

print(vcmax_gome)
```

```{r}
# Vcmax data -> https://zenodo.org/records/14848098

library(terra)
library(ncdf4)

# AU How
latitude <- -12.4943
longitude <- 131.1523



latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
 
  lat_res <- 180. / nlines  
  line <- round((90. - lat) / lat_res)
  
  lon_res <- 360. / nsamples
  sample <- round((lon + 180.) / lon_res)
  
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  return(list(line = line, sample = sample))
}

#
## global leaf chlorophyll content
#
nc <- nc_open("/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc")

# Pixel coordinates
line_idx   <- 500   # 1..1800
sample_idx <- 1000  # 1..3600

coords <- latlon_to_line_sample(latitude, longitude)

pixel_time_series <- ncvar_get(nc, "chl_all",
                               start = c(coords$line, coords$sample, 1, 1),
                               count = c(1, 1, -1, -1))
fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
pixel_time_series[pixel_time_series == fillvalue] <- NA

dim(pixel_time_series)  # 52 x 9 (weeks x years)
pixel_time_series

LCC <- pixel_time_series / 100 # ug/cm²

# LCC: 52 weeks × 9 years
weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
weeks <- 1:52

plot(1:52, weekly_climatology, type='l', xlab='Week', ylab='LCC (μg/cm²)')

# Create a sequence days
daily_weeks <- seq(1, 52, by = 1/7) 

# Interpolate linearly
daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y

# Plot
plot(daily_weeks, daily_climatology, type='l', xlab='Week', ylab='LCC (μg/cm²)')
points(weeks, weekly_climatology, col='red')  # show original weekly points

nc_close(nc)
```
Right now:
- Vcmax is a single constant per site
- LCC = weekly climatology 
- GPP = daily, year-specific

Need to:
- Attach Vcmax to daily flux data
- Turn LCC climatology into a daily series

```{r}

How_daily <- How_daily %>%
  mutate(Vcmax = vcmax_gome)
```

Day-of-year values of LCC and attach to flux data

```{r}
lcc_day <- data.frame(
  doy = round(seq(1, 365, length.out = length(daily_climatology))), 
  LCC = daily_climatology
)
```

```{r}
How_daily <- How_daily %>%
  mutate(doy = yday(date)) %>%
  left_join(lcc_day, by = "doy")
```

GAMs:
- Start simple
- Does the addition of plant traits improve GPP predictability across Australian sites?

Compare GAM within site AU_How

1. GAM - SW_rad, VPD, Tair

```{r}

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

idx <- createDataPartition(How_daily$GPP, p=0.8, list=FALSE)
train_data <- How_daily[idx, ]
test_data <- How_daily[-idx, ]

gam_basic <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad), data=train_data, 
                 method = "REML")

predictions <- predict(gam_basic, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```
RMSE: 1.753954
R-squared: 0.5600302
NSE: 0.5565368

2. GAM - SW_rad, VPD, Tair, LCC

```{r}
How_daily <- How_daily %>%
  drop_na(LCC)

```


```{r}
idx <- createDataPartition(How_daily$GPP, p=0.8, list=FALSE)
train_data2 <- How_daily[idx, ]
test_data2 <- How_daily[-idx, ]

gam_2 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(LCC), data=train_data2, 
                 method = "REML")

predictions <- predict(gam_2, newdata=test_data2)
results <- data.frame(observed=test_data2$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

Only slightly improved with the addition of LCC

Now let's compare across sites and see if Vcmax or LCC can improve cross-site prediction:

Let's compare AU_How to AU_Asm and AU_Das, Alice Springs is very dry whereas Daly Savanna is more similar to AU_How (seasonally wet)

Vcmax and LCC data for Asm:

```{r}
library(terra)
library(ncdf4)

#AU Asm 
latitude <- -22.2830
longitude <- 133.2490

#
## Vcmax from GOME-2 SIF
#
gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/GOME2_VcmaxTg_05deg.tif"

vcmax_gome_raster <- rast(gome_file)
coords <- data.frame(lon = longitude, lat = latitude)
vcmax_gome <- terra::extract(vcmax_gome_raster, coords)
vcmax_gome <- data.frame(lat = latitude, lon = longitude, 
                           Vcmax = vcmax_gome)
vcmax_gome <- vcmax_gome$Vcmax.GOME2_VcmaxTg_05deg

print(vcmax_gome)

```

```{r}
# Vcmax data -> https://zenodo.org/records/14848098

library(terra)
library(ncdf4)

# AU Asm
latitude <- -22.2830
longitude <- 133.2490



latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
 
  lat_res <- 180. / nlines  
  line <- round((90. - lat) / lat_res)
  
  lon_res <- 360. / nsamples
  sample <- round((lon + 180.) / lon_res)
  
  line <- pmin(pmax(line, 1), nlines)
  sample <- pmin(pmax(sample, 1), nsamples)
  
  return(list(line = line, sample = sample))
}

#
## global leaf chlorophyll content
#
nc <- nc_open("/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc")

# Pixel coordinates
line_idx   <- 500   # 1..1800
sample_idx <- 1000  # 1..3600

coords <- latlon_to_line_sample(latitude, longitude)

pixel_time_series <- ncvar_get(nc, "chl_all",
                               start = c(coords$line, coords$sample, 1, 1),
                               count = c(1, 1, -1, -1))
fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
pixel_time_series[pixel_time_series == fillvalue] <- NA

dim(pixel_time_series)  # 52 x 9 (weeks x years)
pixel_time_series

LCC <- pixel_time_series / 100 # ug/cm²

# LCC: 52 weeks × 9 years
weekly_climatology <- rowMeans(LCC, na.rm = TRUE)
weeks <- 1:52

plot(1:52, weekly_climatology, type='l', xlab='Week', ylab='LCC (μg/cm²)')

# Create a sequence days
daily_weeks <- seq(1, 52, by = 1/7) 

# Interpolate linearly
daily_climatology <- approx(x = weeks, y = weekly_climatology, xout = daily_weeks)$y

# Plot
plot(daily_weeks, daily_climatology, type='l', xlab='Week', ylab='LCC (μg/cm²)')
points(weeks, weekly_climatology, col='red')  # show original weekly points

nc_close(nc)
```

Attach Vcmax to AU_Asm

```{r}
Asm_daily <- Asm_daily %>%
  mutate(Vcmax = vcmax_gome)
```

```{r}
lcc_day <- data.frame(
  doy = round(seq(1, 365, length.out = length(daily_climatology))), 
  LCC = daily_climatology
)
```

```{r}
Asm_daily <- Asm_daily %>%
  mutate(doy = yday(date)) %>%
  left_join(lcc_day, by = "doy")
```

Get rid of NAs

```{r}
Asm_daily <- Asm_daily %>%
  drop_na(LCC)

```


Now, lets compare AU_How and AU_Asm

```{r}
train_data3 <- How_daily
test_data3 <- Asm_daily

gam_3 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(LCC), data=train_data3, 
                 method = "REML")

predictions <- predict(gam_3, newdata=test_data3)
results <- data.frame(observed=test_data3$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

```{r}
summary(gam_basic)

```

```{r}
summary(gam_2)
```

```{r}
par(mfrow = c(2, 2))
plot(gam_2, pages = 1, shade = TRUE)
```

3. GAM - VPD, Tair, SW_rad, LCC, Vcmax

```{r}
idx <- createDataPartition(How_daily$GPP, p=0.8, list=FALSE)
train_data4 <- How_daily[idx, ]
test_data4 <- How_daily[-idx, ]

gam_4 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(LCC) + Vcmax, data=train_data4, 
                 method = "REML")

predictions <- predict(gam_4, newdata=test_data4)
results <- data.frame(observed=test_data4$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

VPD stress days:

```{r}
How_stress_days <- How_daily %>% filter(VPD > quantile(VPD, 0.9))

Asm_stress_days <- Asm_daily %>% filter(VPD > quantile(VPD, 0.9))
```

Let's compare the predictability of AU_How high VPD days versus Asm high VPD days

```{r}
train_data5 <- How_stress_days
test_data5 <- Asm_stress_days

gam_5 <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(LCC) + Vcmax, data=train_data5, 
                 method = "REML")

predictions <- predict(gam_5, newdata=test_data5)
results <- data.frame(observed=test_data5$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

```{r}
plot(How_daily$VPD)
```


