---
title: "cpr_diagnostic"
output: html_document
date: "2026-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Site Calperum has extremely low RF model performance despite 15 years of observation. Violates assumption that longer records will have higher model fit/performance as the model is able to learn climate-GPP relationships across time. 

Fire at Calperum - See Sun & Marschner 2024, here we are going to test whether pre-fire, during fire/recovery and post-recovery contribute to model fit.

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

set.seed(42)

predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

compute_metrics <- function(obs, pred) {
  data.frame(
    R2   = round(cor(obs, pred, use = "complete.obs")^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2, na.rm = TRUE)), 3),
    NSE  = round(1 - sum((obs - pred)^2, na.rm = TRUE) /
                     sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE), 3)
  )
}

# Prepare the site
cpr <- sites_daily[["Cpr"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, Precip, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),
    cumulative_cwd_7day = rollapply(
      cwd, width = 14, FUN = sum,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors), GPP)

formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# Define periods
# Pre-fire:            2010-2013 
# During-fire/recovery:2014-2019 
# Post-recovery:       2020-2025 
cpr_pre    <- cpr %>% filter(year >= 2010 & year <= 2013)
cpr_during <- cpr %>% filter(year >= 2014 & year <= 2019)
cpr_post   <- cpr %>% filter(year >= 2020 & year <= 2025)
cpr_full   <- cpr  # full record for baseline LOYO

cat("Pre-fire rows:              ", nrow(cpr_pre),    " | years:", paste(unique(cpr_pre$year),    collapse = ", "), "\n")
cat("During-fire/recovery rows:  ", nrow(cpr_during), " | years:", paste(unique(cpr_during$year), collapse = ", "), "\n")
cat("Post-recovery rows:         ", nrow(cpr_post),   " | years:", paste(unique(cpr_post$year),   collapse = ", "), "\n")
cat("Full record rows:           ", nrow(cpr_full),   " | years:", paste(unique(cpr_full$year),   collapse = ", "), "\n")

# Helper: LOYO within a data subset
run_loyo_subset <- function(df, label) {
  years <- unique(df$year)
  if (length(years) < 2) {
    message("Skipping ", label, ": fewer than 2 years")
    return(NULL)
  }

  map_dfr(years, function(yr) {
    train_df <- df %>% filter(year != yr)
    test_df  <- df %>% filter(year == yr)
    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    rf <- randomForest(formula, data = train_df,
                       ntree = 500, mtry = mtry_val, importance = FALSE)

    data.frame(
      label    = label,
      year     = yr,
      date     = test_df$date,
      GPP_obs  = test_df$GPP,
      GPP_pred = predict(rf, test_df)
    )
  })
}

# Test 1: Full record LOYO 
loyo_full <- run_loyo_subset(cpr_full, "Full record LOYO\n(2010-2025)")

# Test 2: Pre-fire LOYO (2010-2013)
loyo_pre <- run_loyo_subset(cpr_pre, "Pre-fire LOYO\n(2010-2013)")

# Test 3: During-fire/recovery LOYO (2014-2019) 
loyo_during <- run_loyo_subset(cpr_during, "During-fire/recovery LOYO\n(2014-2019)")

# Test 4: Post-recovery LOYO (2020-2025) 
loyo_post <- run_loyo_subset(cpr_post, "Post-recovery LOYO\n(2020-2025)")

# Test 5: Train pre-fire, test post-recovery
# Direct test of regime shift — if fire changed GPP-climate relationship,
# pre-fire training should predict post-recovery poorly
rf_pre <- randomForest(formula, data = cpr_pre,
                       ntree = 500, mtry = mtry_val, importance = FALSE)

cross_fire <- data.frame(
  label    = "Train pre-fire /\nTest post-recovery",
  year     = cpr_post$year,
  date     = cpr_post$date,
  GPP_obs  = cpr_post$GPP,
  GPP_pred = predict(rf_pre, cpr_post)
)

# Metrics 
all_results <- bind_rows(loyo_full, loyo_pre, loyo_during, loyo_post, cross_fire) %>%
  mutate(label = factor(label, levels = c(
    "Full record LOYO\n(2010-2025)",
    "Pre-fire LOYO\n(2010-2013)",
    "During-fire/recovery LOYO\n(2014-2019)",
    "Post-recovery LOYO\n(2020-2025)",
    "Train pre-fire /\nTest post-recovery"
  )))

metrics <- all_results %>%
  group_by(label) %>%
  summarise(compute_metrics(GPP_obs, GPP_pred), .groups = "drop")

cat("\n── AU-Cpr fire regime shift diagnostic ──\n")
print(metrics)

```
```{r}
# Step 1: Run LOYO predictions for each period and combine 
# Already have these:
# loyo_pre, loyo_during, loyo_post, loyo_full

# Combine all period predictions into one data frame
ts_combined <- bind_rows(
  loyo_pre %>% mutate(period = "Pre-fire"),
  loyo_during %>% mutate(period = "During Fire / Recovery"),
  loyo_post %>% mutate(period = "Post-fire")
) %>%
  mutate(
    period = factor(period, levels = c(
      "Pre-fire",
      "During Fire / Recovery",
      "Post-fire"
    )),
    date = as.Date(date)
  )

```


```{r}

# Step 2: Compute R-squared and RMSE per period
period_metrics <- ts_combined %>%
  group_by(period) %>%
  summarise(
    r2       = round(cor(GPP_obs, GPP_pred, use = "complete.obs")^2, 3),
    rmse     = round(sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)), 3),
    gpp_mean = round(mean(GPP_obs, na.rm = TRUE), 3),
    .groups  = "drop"
  ) %>%
  mutate(nrmse = round(rmse / gpp_mean, 3))

full_r2   <- round(cor(loyo_full$GPP_obs, loyo_full$GPP_pred, use = "complete.obs")^2, 3)
full_rmse <- round(sqrt(mean((loyo_full$GPP_obs - loyo_full$GPP_pred)^2, na.rm = TRUE)), 3)
full_nrmse <- round(full_rmse / mean(loyo_full$GPP_obs, na.rm = TRUE), 3)

# Step 3: Define period bands
period_bands <- tibble(
  period = factor(c("Pre-fire", "During Fire / Recovery", "Post-fire"),
                  levels = c("Pre-fire", "During Fire / Recovery", "Post-fire")),
  xmin   = as.Date(c("2010-08-05", "2014-01-01", "2020-01-01")),
  xmax   = as.Date(c("2013-12-31", "2019-12-31", "2024-12-31"))
)

# Step 4: Build period_labels from period_metrics
period_labels <- period_bands %>%
  mutate(x_mid = xmin + (xmax - xmin) / 2) %>%
  left_join(period_metrics, by = "period")  # <-- period_metrics

# Step 4: Period colours 
period_fill_cols <- c(
  "Pre-fire"               ="darkolivegreen",
  "During Fire / Recovery" = "#BF812D",
  "Post-fire"              = "#01665E" 
)

# Step 5: Build the plot
ggplot() +

  # Background shading per period
  geom_rect(
    data = period_bands,
    aes(xmin = xmin, xmax = xmax,
        ymin = -Inf, ymax = Inf,
        fill = period),
    alpha = 0.08,
    inherit.aes = FALSE
  ) +

  # Vertical dashed dividers between periods
  geom_vline(
    xintercept = as.numeric(as.Date(c("2014-01-01", "2020-01-01"))),
    linetype   = "dashed",
    colour     = "grey40",
    linewidth  = 0.5
  ) +

  # Observed GPP — dark, semi-transparent
  geom_line(
    data   = ts_combined,
    aes(x = date, y = GPP_obs),
    colour = "grey20",
    alpha  = 0.5,
    linewidth = 0.35
  ) +

  # Predicted GPP — coloured by period
  geom_line(
    data = ts_combined,
    aes(x = date, y = GPP_pred, colour = period),
    alpha = 0.85,
    linewidth = 0.5
  ) +

  # Period R-squared annotations — inside each shaded band, top
# R-squared annotation
  geom_text(
    data = period_labels,
    aes(x = x_mid, y = Inf,
        label = paste0("R² = ", r2),
        colour = period),
    vjust       = 1.6,
    fontface    = "bold",
    size        = 3.5,
    inherit.aes = FALSE
  ) +

  # RMSE annotation — just below R-squared
  geom_text(
    data = period_labels,
    aes(x = x_mid, y = Inf,
        label = paste0("nRMSE = ", nrmse),
        colour = period),
    vjust       = 3.2,
    fontface = "bold",
    size        = 3.5,
    inherit.aes = FALSE
  ) +

  # Period name — below RMSE
  geom_text(
    data = period_labels,
    aes(x = x_mid, y = Inf,
        label = as.character(period),
        colour = period),
    vjust       = 4.8,
    size        = 3.9,
    inherit.aes = FALSE
  ) +

  # Full record annotations — bottom left
  annotate(
    "text",
    x        = as.Date("2010-03-01"),
    y        = -Inf,
    label    = paste0("Full record R² = ", full_r2, "  |  nRMSE = ", full_nrmse),
    hjust    = 0,
    vjust    = -0.8,
    size     = 3.2,
    colour   = "grey30",
    fontface = "italic"
  ) +
  scale_colour_manual(values = period_fill_cols) +
  scale_fill_manual(values = period_fill_cols) +

  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y",
    expand      = expansion(mult = c(0.01, 0.01))
  ) +

  labs(
    x      = NULL,
    y      = expression(GPP ~ (g ~ C ~ m^{-2} ~ d^{-1})),
    colour = NULL
  ) +

  theme_classic(base_size = 12) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    legend.position  = "none",
    panel.grid.major.y = element_line(colour = "grey93", linewidth = 0.3),
    strip.background = element_blank()
  )

```


```{r}
# Check GPP range and variance per period to confirm scale explanation
ts_combined %>%
  group_by(period) %>%
  summarise(
    gpp_mean   = round(mean(GPP_obs, na.rm = TRUE), 3),
    gpp_sd     = round(sd(GPP_obs, na.rm = TRUE), 3),
    gpp_range  = round(max(GPP_obs, na.rm = TRUE) - 
                       min(GPP_obs, na.rm = TRUE), 3),
    rmse       = round(sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)), 3),
    r2         = round(cor(GPP_obs, GPP_pred, 
                           use = "complete.obs")^2, 3)
  )
```


```{r}

# Plot 1: Obs vs predicted time series per test
ts_df <- all_results %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(type = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted"))

metric_labels <- metrics %>%
  mutate(ann_label = paste0("R² = ", R2, "  NSE = ", NSE))

p_ts <- ggplot(ts_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_line(linewidth = 0.4) +
  geom_text(
    data        = metric_labels,
    aes(label   = ann_label),
    x           = -Inf, y = Inf,
    hjust       = -0.05, vjust = 1.6,
    inherit.aes = FALSE,
    size = 3, colour = "grey20", fontface = "bold"
  ) +
  # Fire event marker on full record panel only
  geom_vline(
    data = data.frame(
      label = factor("Full record LOYO\n(2010-2025)",
                     levels = levels(ts_df$label)),
      xint  = as.Date("2014-01-15")
    ),
    aes(xintercept = xint),
    linetype = "dashed", colour = "#CC6677", linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ label, ncol = 2, scales = "free_x") +
  scale_colour_manual(values = c("Observed"  = "#2c2c2c",
                                 "Predicted" = "#88CCEE")) +
  scale_alpha_manual(values  = c("Observed"  = 0.45,
                                 "Predicted" = 0.9)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "AU-Cpr: Fire regime shift diagnostic",
    subtitle = "Red dashed line = January 2014 wildfire",
    x        = NULL,
    y        = expression(GPP ~ (g ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    panel.grid.major = element_line(colour = "grey94", linewidth = 0.3),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# Plot 2: R-squared comparison bar chart
# Extract full-record baseline R-squared dynamically
baseline_r2 <- metrics %>%
  filter(grepl("Full record", label)) %>%
  pull(R2)

p_r2 <- ggplot(metrics, aes(x = label, y = R2, fill = R2)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = R2), vjust = -0.4, size = 3.5,
            fontface = "bold", colour = "grey20") +
  geom_hline(yintercept = baseline_r2, linetype = "dashed",
             colour = "#CC6677", linewidth = 0.6) +
  annotate("text", x = 0.6, y = baseline_r2 + 0.01,
           label = paste0("Full record\nbaseline (R²=", baseline_r2, ")"),
           hjust = 0, size = 2.8, colour = "#CC6677", fontface = "italic") +
  scale_fill_gradient(low = "#E06C4B", high = "#4682B4",
                      limits = c(0, 1), oob = scales::squish) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "AU-Cpr: R² by modelling period",
    subtitle = "Improvement over full-record LOYO indicates fire-induced non-stationarity",
    x        = NULL,
    y        = "R²"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 9),
    axis.text.x        = element_text(size = 8),
    legend.position    = "none",
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
  )

print(p_ts)
print(p_r2)
```

Is VPD importance to model fit different pre-fire, during-fire/recovery, and post-fire recovery?

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

set.seed(42)

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors],
                                center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],
                                center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

# Predictor sets — same as main analysis
predictors_with    <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
predictors_without <- c("Tair", "SW_rad",         "cumulative_cwd_7day")

# Prepare Cpr
cpr_full <- sites_daily[["Cpr"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),
    cumulative_cwd_7day = rollapply(
      cwd, width = 7, FUN = sum,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors_with), GPP)

# All three period subsets
cpr_pre    <- cpr_full %>% filter(year >= 2010 & year <= 2013)
cpr_during <- cpr_full %>% filter(year >= 2014 & year <= 2019)
cpr_post   <- cpr_full %>% filter(year >= 2020 & year <= 2025)

# LOYO with and without VPD for a given dataset
run_vpd_importance <- function(df, period_label) {

  fit_loyo <- function(predictors, model_label) {

    years <- unique(df$year)
    all_preds      <- list()
    all_importance <- list()

    for (yr in years) {
      train_df <- df %>% filter(year != yr)
      test_df  <- df %>% filter(year == yr)

      if (nrow(train_df) < 50 || nrow(test_df) < 10) next

      mtry_val <- max(2, floor(sqrt(length(predictors))))
      scaled   <- standardise_train_test(train_df, test_df, predictors)

      formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
      model    <- randomForest(formula, data = scaled$train,
                               ntree = 500, mtry = mtry_val, importance = TRUE)
      preds    <- predict(model, scaled$test)

      all_preds[[as.character(yr)]] <- data.frame(
        Observed  = scaled$test$GPP,
        Predicted = preds,
        year      = yr
      )

      all_importance[[as.character(yr)]] <- data.frame(
        variable   = rownames(importance(model)),
        importance = importance(model, type = 1)[, 1],
        year       = yr
      )
    }

    results <- bind_rows(all_preds)
    if (nrow(results) == 0) return(NULL)

    rsq  <- cor(results$Observed, results$Predicted, use = "complete.obs")^2
    rmse <- sqrt(mean((results$Observed - results$Predicted)^2, na.rm = TRUE))
    nse  <- 1 - sum((results$Observed - results$Predicted)^2, na.rm = TRUE) /
                sum((results$Observed - mean(results$Observed, na.rm = TRUE))^2,
                    na.rm = TRUE)

    imp_df <- bind_rows(all_importance) %>%
      group_by(variable) %>%
      summarise(importance = mean(importance), .groups = "drop") %>%
      mutate(model = model_label, period = period_label)

    list(
      metrics    = data.frame(period = period_label, model = model_label,
                              R2 = round(rsq, 3), RMSE = round(rmse, 3),
                              NSE = round(nse, 3)),
      importance = imp_df
    )
  }

  with_vpd    <- fit_loyo(predictors_with,    "with_VPD")
  without_vpd <- fit_loyo(predictors_without, "without_VPD")

  if (is.null(with_vpd) || is.null(without_vpd)) return(NULL)

  list(
    metrics    = bind_rows(with_vpd$metrics,    without_vpd$metrics),
    importance = bind_rows(with_vpd$importance, without_vpd$importance)
  )
}

# Run for each period 
results_pre    <- run_vpd_importance(cpr_pre,    "Pre-fire (2010-2013)")
results_during <- run_vpd_importance(cpr_during, "During-fire/recovery (2014-2019)")
results_post   <- run_vpd_importance(cpr_post,   "Post-recovery (2020-2025)")
results_full   <- run_vpd_importance(cpr_full,   "Full record (2010-2025)")

# Compile metrics
all_metrics <- bind_rows(
  results_full$metrics,
  results_pre$metrics,
  results_during$metrics,
  results_post$metrics
) %>%
  mutate(period = factor(period, levels = c(
    "Full record (2010-2025)",
    "Pre-fire (2010-2013)",
    "During-fire/recovery (2014-2019)",
    "Post-recovery (2020-2025)"
  )))

# change in R-squared from adding VPD per period 
delta_r2 <- all_metrics %>%
  select(period, model, R2) %>%
  pivot_wider(names_from = model, values_from = R2) %>%
  mutate(delta_R2 = with_VPD - without_VPD)

cat("\n── VPD importance by period at AU-Cpr ──\n")
print(delta_r2)

# Variable importance across periods 
all_importance <- bind_rows(
  results_full$importance,
  results_pre$importance,
  results_during$importance,
  results_post$importance
) %>%
  filter(model == "with_VPD") %>%
  mutate(period = factor(period, levels = c(
    "Full record (2010-2025)",
    "Pre-fire (2010-2013)",
    "During-fire/recovery (2014-2019)",
    "Post-recovery (2020-2025)"
  )))

# Plot 1: change in R-squared from adding VPD by period 
p_delta <- delta_r2 %>%
  ggplot(aes(x = period, y = delta_R2, fill = delta_R2)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = round(delta_R2, 3)),
            vjust = -0.4, size = 3.5, fontface = "bold", colour = "grey20") +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +
  scale_fill_gradient(low = "#E06C4B", high = "#4682B4",
                      limits = c(0, 0.3), oob = scales::squish) +
  scale_y_continuous(limits = c(0, 0.35)) +
  labs(
    title    = "AU-Cpr: VPD importance (ΔR²) by period",
    subtitle = "ΔR² = R² with VPD − R² without VPD | Higher = VPD more important",
    x        = NULL,
    y        = "ΔR² from adding VPD"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 9),
    axis.text.x        = element_text(size = 9),
    legend.position    = "none",
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
  )

# Plot 2: R-squared with and without VPD side by side by period 
p_r2 <- all_metrics %>%
  mutate(
    model = recode(model,
                   with_VPD    = "With VPD",
                   without_VPD = "Without VPD")
  ) %>%
  ggplot(aes(x = period, y = R2, fill = model)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.9) +
  geom_text(aes(label = R2),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3, fontface = "bold", colour = "grey20") +
  scale_fill_manual(values = c("With VPD"    = "#4682B4",
                                "Without VPD" = "#E06C4B")) +
  scale_y_continuous(limits = c(0, 0.6)) +
  labs(
    title    = "AU-Cpr: Model R² with and without VPD by period",
    subtitle = "Comparing VPD contribution across full record, pre-fire, during-fire/recovery and post-recovery",
    x        = NULL,
    y        = "R²",
    fill     = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 9),
    axis.text.x        = element_text(size = 9),
    legend.position    = "bottom",
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
  )

# Plot 3: Variable importance by period 
p_imp <- ggplot(all_importance,
                aes(x = reorder(variable, importance),
                    y = importance, fill = period)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.9) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Full record (2010-2025)"          = "grey60",
    "Pre-fire (2010-2013)"             = "#88CCEE",
    "During-fire/recovery (2014-2019)" = "#DDCC77",
    "Post-recovery (2020-2025)"        = "#CC6677"
  )) +
  labs(
    title    = "AU-Cpr: Variable importance by period (%IncMSE)",
    subtitle = "Did fire alter the relative importance of climate drivers?",
    x        = NULL,
    y        = "% Increase in MSE",
    fill     = "Period"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 9),
    legend.position = "bottom"
  )

print(p_delta)
print(p_r2)
print(p_imp)
```

```{r}
library(tidyverse)

# Compile metrics with clean labels
plot_metrics <- all_metrics %>%
  mutate(
    period_clean = case_when(
      grepl("Full record",   period) ~ "Full Record",
      grepl("Pre-fire",      period) ~ "Pre-fire",
      grepl("During",        period) ~ "During Fire /\nRecovery",
      grepl("Post-recovery", period) ~ "Post-fire"
    ),
    period_clean = factor(period_clean, levels = c(
      "Full Record", "Pre-fire", "During Fire /\nRecovery", "Post-fire"
    ))
  ) %>%
  filter(model == "with_VPD")

# Pull full record R² for baseline line
baseline <- plot_metrics %>%
  filter(period_clean == "Full Record") %>%
  pull(R2)

ggplot(plot_metrics, aes(x = period_clean, y = R2, fill = period_clean)) +
  geom_col(width = 0.55, alpha = 0.92) +
  geom_text(
    aes(label = R2),
    vjust  = -0.5,
    size   = 4,
    fontface = "bold",
    colour = "grey20"
  ) +
  geom_hline(
    yintercept = baseline,
    linetype   = "dashed",
    colour     = "grey50",
    linewidth  = 0.6
  ) +
  annotate(
    "text",
    x     = 0.6,
    y     = baseline + 0.015,
    label = paste0("Full record baseline (R² = ", baseline, ")"),
    hjust = 0,
    size  = 3,
    colour = "grey40",
    fontface = "italic"
  ) +
  scale_fill_manual(values = c(
    "Full Record"           = "grey70",
    "Pre-fire"              = "#88CCEE",
    "During Fire /\nRecovery" = "#DDCC77",
    "Post-fire"             = "#CC6677"
  )) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x        = NULL,
    y        = expression(R^2),
    caption  = "Random forest LOYO framework | Predictors: Tair, SW_rad, VPD, CWD"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(colour = "grey40", size = 10),
    plot.caption       = element_text(colour = "grey55", size = 8),
    axis.text.x        = element_text(size = 11, colour = "grey20"),
    legend.position    = "none",
    panel.grid.major.y = element_line(colour = "grey93", linewidth = 0.35)
  )
```

