---
title: "future_envelope_final"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Future envelope final script

1) *Frequency vs magnitude*
Is GPP more sensitive to increases in frequency of VPD extremes or magnitude of VPD extremes?

15, 30 and 50% increase in the number of days above 90th percentile in the summer (Extreme VPD events become more common). Replace summer days with 90th percentile value. Replace days that are close to the threshold already

15, 30 and 50% increase in magnitude of the 90th percentile VPD extremes (Extremes get more intense). Summer VPD values equal to or above 90th percentile get increased in magnitude. Whilst the other days stay the same.

Random Forest is leave-one-year out trained and tested for each scenario. Trained on all years bar one, tested on one year. Repeated for all years and concatenated together. 

```{r}
library(tidyverse)
library(randomForest)
library(purrr)
library(zoo)
library(lubridate)

# ── Config ────────────────────────────────────────────────────────────────────
FEATURES      <- c("VPD", "Tair", "SW_rad", "cwd_7day")
TARGET        <- "GPP"
SUMMER_MONTHS <- c(12, 1, 2)
FREQ_INCREASES <- c(0.15, 0.30, 0.50)
MAG_INCREASES  <- c(0.15, 0.30, 0.50)

# ── Step 1: Prep — rolling CWD sum, hydro year, month ────────────────────────
sites_daily <- map(sites_daily, function(df) {
  df %>%
    arrange(date) %>%
    mutate(
      month    = month(date),
      year     = year(date),
      # Hydrological year: December counts toward the following year
      # so that DJF summer is never split across two folds
      hydro_year = ifelse(month == 12, year + 1, year),
      # Cumulative sum over 7 days (min 3 obs required), not rolling mean
      cwd_7day = rollapply(cwd, width = 7,
                           FUN = function(x) {
                             if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
                           },
                           align = "right", fill = NA)
    )
})

# ── Step 2: Scenario builders ─────────────────────────────────────────────────

get_vpd90 <- function(df) {
  df %>%
    filter(month %in% SUMMER_MONTHS) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)
}

make_freq_scenario <- function(df, pct_increase) {
  vpd90         <- get_vpd90(df)
  summer_idx    <- which(df$month %in% SUMMER_MONTHS)
  n_extreme_now <- sum(df$VPD[summer_idx] >= vpd90, na.rm = TRUE)
  n_to_add      <- round(n_extreme_now * pct_increase)

  below_thresh        <- summer_idx[!is.na(df$VPD[summer_idx]) & df$VPD[summer_idx] < vpd90]
  below_thresh_sorted <- below_thresh[order(df$VPD[below_thresh], decreasing = TRUE)]
  to_replace          <- head(below_thresh_sorted, n_to_add)

  df_new <- df
  df_new$VPD[to_replace] <- vpd90
  df_new$scenario <- paste0("freq_+", pct_increase * 100, "%")
  df_new
}

make_mag_scenario <- function(df, pct_increase) {
  vpd90       <- get_vpd90(df)
  summer_idx  <- which(df$month %in% SUMMER_MONTHS)
  extreme_idx <- summer_idx[!is.na(df$VPD[summer_idx]) & df$VPD[summer_idx] >= vpd90]

  df_new <- df
  df_new$VPD[extreme_idx] <- df_new$VPD[extreme_idx] * (1 + pct_increase)
  df_new$scenario <- paste0("mag_+", pct_increase * 100, "%")
  df_new
}

make_baseline <- function(df) {
  df %>% mutate(scenario = "baseline")
}

# ── Step 3: Scaling helper ────────────────────────────────────────────────────
scale_with_params <- function(data, means, sds) {
  for (v in names(means)) {
    data[[v]] <- (data[[v]] - means[v]) / sds[v]
  }
  data
}

# ── Step 4: LOYO RF for a single scenario dataframe ──────────────────────────
run_loyo_rf <- function(df) {
  df_clean <- df %>%
    select(all_of(c(TARGET, FEATURES, "year", "hydro_year", "date", "scenario", "site"))) %>%
    drop_na()

  hydro_years <- unique(df_clean$hydro_year)

  results <- map_dfr(hydro_years, function(yr) {
    train <- df_clean %>% filter(hydro_year != yr)
    test  <- df_clean %>% filter(hydro_year == yr)

    if (nrow(train) < 30 || nrow(test) < 1) return(NULL)

    # Compute scaling params from training data only
    scale_means <- colMeans(train[, FEATURES], na.rm = TRUE)
    scale_sds   <- apply(train[, FEATURES], 2, sd, na.rm = TRUE)

    train_scaled <- scale_with_params(train, scale_means, scale_sds)
    test_scaled  <- scale_with_params(test,  scale_means, scale_sds)

    rf <- randomForest(
      x         = train_scaled[, FEATURES],
      y         = train_scaled[[TARGET]],
      ntree     = 500,
      mtry      = 3,
      importance = FALSE
    )

    test %>%
      mutate(GPP_pred = predict(rf, newdata = test_scaled[, FEATURES]))
  })

  results
}

# ── Step 5: Run all scenarios across all sites ────────────────────────────────
all_results <- imap_dfr(sites_daily, function(df, site_name) {
  message("Processing site: ", site_name)

  scenarios <- c(
    list(make_baseline(df)),
    map(FREQ_INCREASES, ~ make_freq_scenario(df, .x)),
    map(MAG_INCREASES,  ~ make_mag_scenario(df, .x))
  )

  map_dfr(scenarios, function(scen_df) {
    run_loyo_rf(scen_df)
  })
})

# ── Step 6: Summary stats per site per scenario ───────────────────────────────
summary_stats <- all_results %>%
  group_by(site, scenario) %>%
  summarise(
    mean_GPP_pred = mean(GPP_pred, na.rm = TRUE),
    mean_GPP_obs  = mean(GPP,      na.rm = TRUE),
    ss_res    = sum((GPP - GPP_pred)^2,                  na.rm = TRUE),
    ss_tot    = sum((GPP - mean(GPP, na.rm = TRUE))^2,   na.rm = TRUE),
    r_squared = 1 - (ss_res / ss_tot),
    .groups = "drop"
  ) %>%
  left_join(
    filter(., scenario == "baseline") %>%
      select(site, baseline_mean = mean_GPP_pred),
    by = "site"
  ) %>%
  mutate(pct_change_GPP = ((mean_GPP_pred - baseline_mean) / baseline_mean) * 100)
```

```{r}

#Plots

scenario_colours <- c(
  "baseline"   = "grey40",
  "freq_+15%"  = "#fee8c8",
  "freq_+30%"  = "#fd8d3c",
  "freq_+50%"  = "#b30000",
  "mag_+15%"   = "#d4e8fb",
  "mag_+30%"   = "#3182bd",
  "mag_+50%"   = "#08306b"
)

# Plot 1: % change in mean GPP per site per scenario
p1 <- summary_stats %>%
  filter(scenario != "baseline") %>%
  mutate(scenario = factor(scenario, levels = names(scenario_colours))) %>%
  ggplot(aes(x = reorder(site, pct_change_GPP), y = pct_change_GPP, fill = scenario)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = scenario_colours) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  labs(
    title = "% Change in Mean GPP vs Baseline by Scenario",
    x = "Site", y = "% Change in Mean GPP", fill = "Scenario"
  ) +
  theme_bw()

print(p1)

# Plot 2: Time series observed vs predicted per scenario (faceted by site)
# Scenario options: "baseline", "freq_+15%", "freq_+30%", "freq_+50%",
#                   "mag_+15%",  "mag_+30%",  "mag_+50%"
scenario_to_plot <- "freq_+30%"

p2 <- all_results %>%
  filter(scenario == scenario_to_plot) %>%
  pivot_longer(cols = c(GPP, GPP_pred), names_to = "type", values_to = "value") %>%
  mutate(type = recode(type, GPP = "Observed", GPP_pred = "Predicted")) %>%
  ggplot(aes(x = date, y = value, colour = type)) +
  geom_line(alpha = 0.7, linewidth = 0.4) +
  scale_colour_manual(values = c("Observed" = "grey30", "Predicted" = "#fd8d3c")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(
    title = paste("Observed vs Predicted GPP —", scenario_to_plot),
    x = "Date", y = "GPP", colour = NULL
  ) +
  theme_bw(base_size = 9)

print(p2)

# Plot 3: R-squared per site per scenario
p3 <- summary_stats %>%
  filter(scenario != "baseline") %>%
  mutate(scenario = factor(scenario, levels = names(scenario_colours))) %>%
  ggplot(aes(x = reorder(site, r_squared), y = r_squared, fill = scenario)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = scenario_colours) +
  coord_flip() +
  labs(
    title = "R² of Random Forest per Site per Scenario",
    x = "Site", y = "R²", fill = "Scenario"
  ) +
  theme_bw()

print(p3)

# Plot 4: R² vs % GPP change, coloured by cluster
p4 <- summary_stats %>%
  filter(scenario != "baseline") %>%
  left_join(site_slopes %>% select(site, cluster_multi), by = "site") %>%
  mutate(scenario_type = if_else(str_detect(scenario, "freq"), "Frequency", "Magnitude")) %>%
  ggplot(aes(x = r_squared, y = pct_change_GPP, colour = cluster_multi, shape = scenario_type)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_colour_manual(values = c("Low" = "#2196F3", "Medium" = "#FF9800", "High" = "#4CAF50")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "R² vs % Change in GPP, by Cluster and Scenario Type",
    x = "R²", y = "% Change in Mean GPP",
    colour = "Cluster", shape = "Scenario type"
  ) +
  theme_bw()

print(p4)

# Plot 5: Frequency vs magnitude % GPP change scatter per site
freq_mag_wide <- summary_stats %>%
  filter(scenario != "baseline") %>%
  mutate(
    scenario_type = if_else(str_detect(scenario, "freq"), "freq", "mag"),
    pct_level     = str_extract(scenario, "\\d+")
  ) %>%
  select(site, scenario_type, pct_level, pct_change_GPP) %>%
  pivot_wider(names_from = scenario_type, values_from = pct_change_GPP)

p5 <- freq_mag_wide %>%
  left_join(site_slopes %>% select(site, cluster_multi), by = "site") %>%
  ggplot(aes(x = freq, y = mag, colour = cluster_multi, label = site)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.7, size = 2.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Low" = "#2196F3", "Medium" = "#FF9800", "High" = "#4CAF50")) +
  facet_wrap(~ pct_level, labeller = label_both) +
  labs(
    title = "Frequency vs Magnitude: % Change in GPP per Site",
    x = "% Change GPP (Frequency increase)",
    y = "% Change GPP (Magnitude increase)",
    colour = "Cluster"
  ) +
  theme_bw()

print(p5)
```
```{r}
library(ggrepel)

cluster_labels <- site_slopes %>%
  select(site, cluster_multi) %>%
  mutate(
    site          = as.character(site),
    cluster_multi = as.character(cluster_multi)
  )

cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

plot_data_freqmag <- summary_stats %>%
  filter(scenario != "baseline") %>%
  mutate(site = as.character(site)) %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))

ggplot(plot_data_freqmag, aes(x = r_squared, y = pct_change_GPP)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to baseline",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

```{r}
# Frequency scenarios
p_freq <- plot_data_freqmag %>%
  filter(str_detect(scenario, "freq")) %>%
  ggplot(aes(x = r_squared, y = pct_change_GPP)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    title  = "Frequency scenarios",
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to baseline",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

# Magnitude scenarios
p_mag <- plot_data_freqmag %>%
  filter(str_detect(scenario, "mag")) %>%
  ggplot(aes(x = r_squared, y = pct_change_GPP)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    title  = "Magnitude scenarios",
    x      = "R² (leave-one-year-out)",
    y      = "% change in mean GPP relative to baseline",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

print(p_freq)
print(p_mag)
```


Lets try this again:
- Increase in frequency

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)

#Setup
predictors      <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
vpd_scalars     <- c(0, 0.15, 0.30, 0.50)
scenario_labels <- c("Observed", "+15% extreme days", "+30% extreme days", "+50% extreme days")

scenario_colours <- c(
  "Observed"           = "#4682B4",
  "+15% extreme days"  = "#F28C28",
  "+30% extreme days"  = "#E06C4B",
  "+50% extreme days"  = "#8B1A1A"
)
```

```{r}
run_loyo_summer_extreme <- function(df, site_name) {
  
  message("Processing: ", site_name)
  
  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)
  
  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }
  
  # 90th percentile VPD from summer obs across all years
  vpd_90 <- df %>%
    filter(summer) %>%
    pull(VPD) %>%
    quantile(0.90, na.rm = TRUE)
  
  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }
  
  loyo_results <- map_dfr(all_years, function(test_year) {
    
    train_df <- df %>% filter(hydro_year != test_year)
    
    # ── Full held-out year ────────────────────────────────────────────
    test_nonsummer <- df %>% filter(hydro_year == test_year, !summer)
    test_summer    <- df %>% filter(hydro_year == test_year,  summer)
    
    if (nrow(train_df) < 50 || nrow(test_summer) < 10) return(NULL)
    
    # Standardise on full training set
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
    
    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }
    
    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)
    
    # Identify extreme vs normal days in test summer
    extreme_days <- test_summer %>% filter(VPD >= vpd_90)
    normal_days  <- test_summer %>% filter(VPD <  vpd_90)
    n_normal     <- nrow(normal_days)
    n_extreme    <- nrow(extreme_days)
    
    map2_dfr(vpd_scalars, scenario_labels, function(extra_frac, label) {
      
      # ── Manipulate summer days only ───────────────────────────────
      if (extra_frac == 0) {
        summer_scenario <- test_summer
        
      } else {
        n_replace <- round(n_normal * extra_frac)
        
        if (n_replace == 0 || n_extreme == 0) {
          summer_scenario <- test_summer
        } else {
          extra_extreme <- extreme_days %>%
            slice_sample(n = n_replace, replace = TRUE)
          
          normal_kept <- normal_days %>%
            slice_sample(n = max(0, n_normal - n_replace), replace = FALSE)
          
          summer_scenario <- bind_rows(normal_kept, extreme_days, extra_extreme) %>%
            arrange(date)
        }
      }
      
      # ── Combine manipulated summer + unchanged non-summer ─────────
      full_test <- bind_rows(test_nonsummer, summer_scenario) %>%
        arrange(date)
      
      full_scaled <- scale_with_params(full_test, scale_means, scale_sds)
      preds       <- predict(rf_model, full_scaled)
      
      data.frame(
        site           = site_name,
        date           = full_test$date,
        year           = test_year,
        scenario       = label,
        summer         = full_test$summer,
        VPD_raw        = full_test$VPD,
        n_extreme_days = sum(full_test$summer & full_test$VPD >= vpd_90),
        n_total_days   = nrow(full_test),
        GPP_pred       = preds,
        GPP_obs        = full_test$GPP
      )
    })
  })
  
  loyo_results
}

# ── Run across all sites ──────────────────────────────────────────────────────
loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_summer_extreme)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(scenario = factor(scenario, levels = scenario_labels))

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggrepel)

# ── Plot 1: R² per site (full year only) ─────────────────────────────────────

r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = wet_to_dry_order))

ggplot(r2_loyo, aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)") +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 2: % change in annual GPP per scenario per site ─────────────────────

scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(
    scenario = factor(scenario, levels = c("+15% extreme days",
                                           "+30% extreme days",
                                           "+50% extreme days")),
    site = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = c(
    "+15% extreme days" = "#F28C28",
    "+30% extreme days" = "#E06C4B",
    "+50% extreme days" = "#8B1A1A"
  )) +
  coord_flip() +
  labs(
    x    = NULL,
    y    = "% change in mean annual GPP relative to observed",
    fill = "Extreme VPD scenario"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 10)
  )

# ── Plot 3: R² vs % change coloured by cluster group ─────────────────────────

# Use summer R² as most relevant for the scenario
r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(
    R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    .groups = "drop"
  )

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    "+15% extreme days" = (`+15% extreme days` - Observed) / Observed * 100,
    "+30% extreme days" = (`+30% extreme days` - Observed) / Observed * 100,
    "+50% extreme days" = (`+50% extreme days` - Observed) / Observed * 100
  ) %>%
  select(site, `+15% extreme days`, `+30% extreme days`, `+50% extreme days`) %>%
  pivot_longer(cols = ends_with("days"),
               names_to  = "scenario",
               values_to = "pct_change") %>%
  mutate(scenario = factor(scenario, levels = c("+15% extreme days",
                                                 "+30% extreme days",
                                                 "+50% extreme days")))

cluster_colours <- c(
  "Low"    = "steelblue",
  "Medium" = "orange",
  "High"   = "firebrick"
)

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")))

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ scenario, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

Increase in magnitude:

Need to do






2) *Baseline and extremes*

Sort sites into tropical, arid and temperate
Tropical (+0.524 kPa), Arid (+0.644 kPa) and Temperate (+0.344 Kpa) increase to ALL daily VPD values (Fang et al., 2022)

Biome-specific background increases, accounts for the fact that arid systems will face larger increases in VPD stress than temperate ones.
Make sure site biome assignment is unambiguous
Clearly state you are applying a uniform shift to every daily value in the whole dataset

Background rising VPD for each climate AND +28% magnitude of VPD values equal to and above the 90th percentile of VPD in the summer (Zhou et al., 2019)

Random Forest is leave-one-year out trained and tested for each scenario. Trained on all years bar one, tested on one year. Repeated for all years and concatenated together. 

Tropical - Ade, Dap, Das, Dry, Fog, Gin, How, Rdf, Stp
Arid - Asm, Cpr, Emr Gww, Ync
Temperate - Boy, Cum, Rig, Tum, Wac, Whr, Wom

Beck et al., 2018

```{r}
library(dplyr)
library(purrr)
library(randomForest)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggrepel)

# ── Setup ─────────────────────────────────────────────────────────────────────

predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

scenario_labels  <- c("Observed", "Future VPD")
scenario_colours <- c("Observed" = "#4682B4", "Future VPD" = "#C0392B")

biome_vpd_increase <- c(
  Ade = 0.524, Dap = 0.524, Das = 0.524, Dry = 0.524,
  Fog = 0.524, Gin = 0.524, How = 0.524, Rdf = 0.524, Stp = 0.524,
  Asm = 0.644, Cpr = 0.644, Emr = 0.644, Gww = 0.644, Ync = 0.644,
  Boy = 0.344, Cum = 0.344, Rig = 0.344, Tum = 0.344,
  Wac = 0.344, Whr = 0.344, Wom = 0.344
)

biome_labels <- c(
  Ade = "Tropical", Dap = "Tropical", Das = "Tropical", Dry = "Tropical",
  Fog = "Tropical", Gin = "Tropical", How = "Tropical", Rdf = "Tropical", Stp = "Tropical",
  Asm = "Arid", Cpr = "Arid", Emr = "Arid", Gww = "Arid", Ync = "Arid",
  Boy = "Temperate", Cum = "Temperate", Rig = "Temperate", Tum = "Temperate",
  Wac = "Temperate", Whr = "Temperate", Wom = "Temperate"
)

biome_colours <- c(
  "Tropical"  = "#2E8B57",
  "Arid"      = "#C8860A",
  "Temperate" = "#4682B4"
)

# ── LOYO function ─────────────────────────────────────────────────────────────

run_loyo_biome_vpd <- function(df, site_name) {

  message("Processing: ", site_name)

  vpd_delta <- biome_vpd_increase[site_name]
  if (is.na(vpd_delta)) {
    warning("Skipping ", site_name, ": no biome VPD delta defined")
    return(NULL)
  }

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year       = year(date),
      hydro_year = ifelse(month(date) == 12, year(date) + 1, year(date)),
      summer     = month(date) %in% c(12, 1, 2),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = function(x) {
        if (sum(!is.na(x)) >= 3) sum(x, na.rm = TRUE) else NA
      }, align = "right", fill = NA)
    ) %>%
    drop_na(all_of(predictors), GPP)

  if (nrow(df) < 50) {
    warning("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }

  all_years <- unique(df$hydro_year)
  if (length(all_years) < 2) {
    warning("Skipping ", site_name, ": insufficient years")
    return(NULL)
  }

  loyo_results <- map_dfr(all_years, function(test_year) {

    train_df <- df %>% filter(hydro_year != test_year)
    test_df  <- df %>% filter(hydro_year == test_year)

    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    # Standardise on training set parameters
    scale_means <- colMeans(train_df[predictors], na.rm = TRUE)
    scale_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)

    scale_with_params <- function(data, means, sds) {
      for (v in names(means)) data[[v]] <- (data[[v]] - means[v]) / sds[v]
      data
    }

    train_scaled <- scale_with_params(train_df, scale_means, scale_sds)
    formula      <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
    rf_model     <- randomForest(formula, data = train_scaled,
                                 ntree = 500, mtry = 3, importance = FALSE)

    map_dfr(scenario_labels, function(label) {

      test_scenario <- if (label == "Observed") {
        test_df
      } else {
        test_df %>% mutate(VPD = VPD + vpd_delta)
      }

      test_scaled <- scale_with_params(test_scenario, scale_means, scale_sds)
      preds       <- predict(rf_model, test_scaled)

      data.frame(
        site        = site_name,
        date        = test_scenario$date,
        year        = test_year,
        scenario    = label,
        summer      = test_scenario$summer,
        VPD_raw     = test_scenario$VPD,
        n_total_days = nrow(test_scenario),
        GPP_pred    = preds,
        GPP_obs     = test_scenario$GPP
      )
    })
  })

  loyo_results
}

# ── Run across all sites ──────────────────────────────────────────────────────

loyo_results <- map2(sites_daily, names(sites_daily), run_loyo_biome_vpd)
loyo_results <- Filter(Negate(is.null), loyo_results)

all_scenarios <- bind_rows(loyo_results) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_labels),
    biome    = biome_labels[site]
  )

message("Sites retained: ", length(unique(all_scenarios$site)))
message("Total rows: ", nrow(all_scenarios))

# ── Plot 1: R² per site coloured by biome ────────────────────────────────────

r2_loyo <- all_scenarios %>%
  filter(scenario == "Observed") %>%
  group_by(site) %>%
  summarise(
    R2   = cor(GPP_obs, GPP_pred, use = "complete.obs")^2,
    RMSE = sqrt(mean((GPP_obs - GPP_pred)^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    biome = biome_labels[site],
    site  = factor(site, levels = wet_to_dry_order)
  )

ggplot(r2_loyo, aes(x = site, y = R2, fill = biome)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(R2, 2)), hjust = -0.2, size = 3, colour = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.1)) +
  scale_fill_manual(values = biome_colours) +
  labs(x = NULL, y = "R² — leave-one-year-out (out-of-sample)", fill = "Biome") +
  theme_classic() +
  theme(legend.position = "top", axis.text.y = element_text(size = 10))

# ── Plot 2: % change in GPP per site, faceted by biome ───────────────────────

scenario_summary_long <- all_scenarios %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(
    pct_change = (`Future VPD` - Observed) / Observed * 100,
    biome      = biome_labels[site],
    site       = factor(site, levels = wet_to_dry_order)
  )

ggplot(scenario_summary_long, aes(x = site, y = pct_change, fill = biome)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_fill_manual(values = biome_colours) +
  coord_flip() +
  facet_wrap(~ biome, scales = "free_y", ncol = 1) +
  labs(
    x    = NULL,
    y    = "% change in mean annual GPP relative to observed",
    fill = "Biome"
  ) +
  theme_classic() +
  theme(legend.position = "none", axis.text.y = element_text(size = 10),
        strip.text = element_text(face = "bold"))

# ── Plot 3: R² vs % change coloured by cluster group, faceted by biome ───────

r2_sites <- all_scenarios %>%
  filter(scenario == "Observed", summer) %>%
  group_by(site) %>%
  summarise(R2 = cor(GPP_obs, GPP_pred, use = "complete.obs")^2, .groups = "drop")

pct_change_sites <- all_scenarios %>%
  filter(summer) %>%
  group_by(site, scenario) %>%
  summarise(mean_GPP = mean(GPP_pred, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = scenario, values_from = mean_GPP) %>%
  mutate(pct_change = (`Future VPD` - Observed) / Observed * 100)

cluster_colours <- c("Low" = "steelblue", "Medium" = "orange", "High" = "firebrick")

plot_data <- pct_change_sites %>%
  mutate(site = as.character(site)) %>%
  left_join(r2_sites, by = "site") %>%
  left_join(cluster_labels, by = "site") %>%
  mutate(
    cluster_multi = factor(cluster_multi, levels = c("Low", "Medium", "High")),
    biome         = biome_labels[site]
  )

ggplot(plot_data, aes(x = R2, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = cluster_multi), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "grey30", fill = "grey80",
              linewidth = 0.7, linetype = "dotted") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30", max.overlaps = 15) +
  scale_colour_manual(values = cluster_colours, na.value = "grey50") +
  facet_wrap(~ biome, ncol = 3) +
  labs(
    x      = "R² (leave-one-year-out, summer)",
    y      = "% change in mean summer GPP relative to observed",
    colour = "VPD sensitivity"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

```




