---
title: "future_envelope_03"
output: html_document
date: "2026-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Train once on observed full site record for each site
Predict on observed data
Predict on scenario 1, 2, 3
Compare each scenario to baseline predicted GPP

Scenario 1: Aridity-index-specific uniform VPD shift applied to all daily VPD
  Humid        +0.255 kPa
  Dry sub-humid +0.340 kPa
  Semi-Arid    +0.4675 kPa
  Arid         +0.680 kPa

Scenario 2: +28% magnitude increase on days at/above site-specific 90th percentile VPD
Scenario 3: Combined — uniform shift first, then 28% magnitude on days above
            the shifted 90th percentile threshold

REMOVE - Cpr, Emr, Gww, Wac and Ync

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)
library(purrr)

set.seed(42)

predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

# ── Aridity index site groupings — fill in manually ──────────────────────────
humid_sites      <- c("Fog", "How", "Tum", "Wom")   
dry_subhumid_sites <- c("Ade", "Cum", "Dap") 
semi_arid_sites  <- c("Boy", "Das", "Dry", "Gin", "Rdf", "Rig", "Stp", "Whr")   
arid_sites       <- c("Asm")

# Aridity-index-specific VPD shifts
aridity_vpd_shift <- c(
  humid        = 0.255,
  dry_subhumid = 0.340,
  semi_arid    = 0.4675,
  arid         = 0.680
)

extreme_magnitude_increase <- 1.28  # Hermann et al. 2024

prepare_site_envelope <- function(df, site_name) {

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
    arrange(date) %>%
    mutate(
      year = year(date),
      cumulative_cwd_7day = rollapply(
        cwd, width = 7, FUN = sum,
        align = "right", fill = NA, na.rm = TRUE
      )
    ) %>%
    drop_na(all_of(predictors), GPP)

  if (nrow(df) < 100) {
    message("Skipping ", site_name, ": insufficient data")
    return(NULL)
  }

  message("Prepared ", site_name, ": ", nrow(df), " rows")
  df
}


# FUTURE ENVELOPE — train once on observed, predict on all scenarios

run_future_envelope <- function(df, site_name) {

  message("Running future envelope: ", site_name)

  vpd_shift <- case_when(
    site_name %in% humid_sites       ~ aridity_vpd_shift["humid"],
    site_name %in% dry_subhumid_sites ~ aridity_vpd_shift["dry_subhumid"],
    site_name %in% semi_arid_sites   ~ aridity_vpd_shift["semi_arid"],
    site_name %in% arid_sites        ~ aridity_vpd_shift["arid"],
    TRUE ~ NA_real_
  )

  if (is.na(vpd_shift)) {
    warning(site_name, " not assigned to an aridity class — skipping")
    return(NULL)
  }

  # ── Train RF once on full observed record
  formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  mtry_val <- max(2, floor(sqrt(length(predictors))))

  rf_model <- randomForest(
    formula, data = df,
    ntree = 500, mtry = mtry_val, importance = FALSE
  )

  # ── Site-specific 90th percentile VPD threshold
  vpd_90 <- quantile(df$VPD, 0.90, na.rm = TRUE)

  # ── Scenario datasets — only VPD changes, all else held at observed

  # Scenario 1: uniform aridity-specific mean VPD shift on all days
  df_s1 <- df %>% mutate(VPD = VPD + vpd_shift)

  # Scenario 2: 28% magnitude increase on days at/above 90th percentile only
  df_s2 <- df %>%
    mutate(VPD = ifelse(VPD >= vpd_90, VPD * extreme_magnitude_increase, VPD))

  # Scenario 3: combined — uniform shift first, then extreme magnitude
  # threshold shifts with the mean so extremes are defined relative to future baseline
  vpd_90_shifted <- vpd_90 + vpd_shift
  df_s3 <- df %>%
    mutate(
      VPD = VPD + vpd_shift,
      VPD = ifelse(VPD >= vpd_90_shifted, VPD * extreme_magnitude_increase, VPD)
    )

  # ── Predict — delta always vs baseline predicted not observed
  GPP_baseline <- predict(rf_model, df)
  GPP_s1       <- predict(rf_model, df_s1)
  GPP_s2       <- predict(rf_model, df_s2)
  GPP_s3       <- predict(rf_model, df_s3)

  df %>%
    mutate(
      site              = site_name,
      aridity_class     = case_when(
        site_name %in% humid_sites        ~ "Humid",
        site_name %in% dry_subhumid_sites ~ "Dry sub-humid",
        site_name %in% semi_arid_sites    ~ "Semi-Arid",
        site_name %in% arid_sites         ~ "Arid"
      ),
      vpd_shift         = vpd_shift,
      vpd_90_threshold  = vpd_90,
      GPP_obs           = GPP,
      GPP_baseline      = GPP_baseline,
      GPP_s1            = GPP_s1,
      GPP_s2            = GPP_s2,
      GPP_s3            = GPP_s3,
      delta_s1_abs      = GPP_s1 - GPP_baseline,
      delta_s2_abs      = GPP_s2 - GPP_baseline,
      delta_s3_abs      = GPP_s3 - GPP_baseline,
      delta_s1_pct      = (delta_s1_abs / GPP_baseline) * 100,
      delta_s2_pct      = (delta_s2_abs / GPP_baseline) * 100,
      delta_s3_pct      = (delta_s3_abs / GPP_baseline) * 100
    )
}


# RUN ACROSS ALL SITES

prepared_envelope <- map2(sites_daily, names(sites_daily),
                          prepare_site_envelope)
prepared_envelope <- Filter(Negate(is.null), prepared_envelope)

envelope_results <- map2_dfr(prepared_envelope, names(prepared_envelope),
                              run_future_envelope)

message("\nSites completed: ", n_distinct(envelope_results$site))


# ANNUAL SUMMARIES

annual_summary <- envelope_results %>%
  group_by(site, aridity_class, year) %>%
  summarise(
    GPP_baseline_annual = sum(GPP_baseline, na.rm = TRUE),
    GPP_s1_annual       = sum(GPP_s1,       na.rm = TRUE),
    GPP_s2_annual       = sum(GPP_s2,       na.rm = TRUE),
    GPP_s3_annual       = sum(GPP_s3,       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    delta_s1_abs = GPP_s1_annual - GPP_baseline_annual,
    delta_s2_abs = GPP_s2_annual - GPP_baseline_annual,
    delta_s3_abs = GPP_s3_annual - GPP_baseline_annual,
    delta_s1_pct = (delta_s1_abs / GPP_baseline_annual) * 100,
    delta_s2_pct = (delta_s2_abs / GPP_baseline_annual) * 100,
    delta_s3_pct = (delta_s3_abs / GPP_baseline_annual) * 100
  )

site_summary <- annual_summary %>%
  group_by(site, aridity_class) %>%
  summarise(
    mean_delta_s1_pct = round(mean(delta_s1_pct, na.rm = TRUE), 2),
    mean_delta_s2_pct = round(mean(delta_s2_pct, na.rm = TRUE), 2),
    mean_delta_s3_pct = round(mean(delta_s3_pct, na.rm = TRUE), 2),
    mean_delta_s1_abs = round(mean(delta_s1_abs, na.rm = TRUE), 3),
    mean_delta_s2_abs = round(mean(delta_s2_abs, na.rm = TRUE), 3),
    mean_delta_s3_abs = round(mean(delta_s3_abs, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(mean_delta_s3_pct)

cat("\n── Mean annual GPP change by site and scenario ──\n")
print(site_summary)


# SHARED PLOT ELEMENTS

scenario_colours <- c(
  "Mean VPD shift"          = "#88CCEE",
  "+28% extreme magnitude"  = "#CC6677",
  "Combined"                = "#332288"
)

scenario_levels <- c("Mean VPD shift", "+28% extreme magnitude", "Combined")

aridity_levels <- c("Humid", "Dry sub-humid", "Semi-Arid", "Arid")

# Helper to reshape site_summary for plotting
make_plot_df <- function(data, pct_or_abs) {
  if (pct_or_abs == "pct") {
    data %>%
      pivot_longer(
        cols      = c(mean_delta_s1_pct, mean_delta_s2_pct, mean_delta_s3_pct),
        names_to  = "scenario", values_to = "delta"
      ) %>%
      mutate(scenario = recode(scenario,
        mean_delta_s1_pct = "Mean VPD shift",
        mean_delta_s2_pct = "+28% extreme magnitude",
        mean_delta_s3_pct = "Combined"
      ))
  } else {
    data %>%
      pivot_longer(
        cols      = c(mean_delta_s1_abs, mean_delta_s2_abs, mean_delta_s3_abs),
        names_to  = "scenario", values_to = "delta"
      ) %>%
      mutate(scenario = recode(scenario,
        mean_delta_s1_abs = "Mean VPD shift",
        mean_delta_s2_abs = "+28% extreme magnitude",
        mean_delta_s3_abs = "Combined"
      ))
  }
}
```

```{r}

# PLOT 1: Mean % GPP change per site — wet to dry order

p_pct <- make_plot_df(site_summary, "pct") %>%
  mutate(
    scenario = factor(scenario, levels = scenario_levels),
    site     = factor(site, levels = wet_to_dry_order)
  ) %>%
  ggplot(aes(x = site, y = delta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  scale_fill_manual(values = scenario_colours) +
  labs(
    title    = "Projected GPP change under future VPD scenarios",
    subtitle = "Mean annual % change vs baseline predicted GPP",
    x        = "Site (Wet to Dry)",
    y        = "Mean annual ΔGPP (%)",
    fill     = "Scenario"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    legend.position = "bottom"
  )


# PLOT 2: Mean absolute GPP change per site — wet to dry order

p_abs <- make_plot_df(site_summary, "abs") %>%
  mutate(
    scenario = factor(scenario, levels = scenario_levels),
    site     = factor(site, levels = wet_to_dry_order)
  ) %>%
  ggplot(aes(x = site, y = delta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  scale_fill_manual(values = scenario_colours) +
  labs(
    title    = "Projected GPP change under future VPD scenarios",
    subtitle = "Mean annual absolute change vs baseline predicted GPP",
    x        = "Site (Wet to Dry)",
    y        = expression(Mean ~ annual ~ Delta*GPP ~ (g ~ C ~ m^{-2} ~ yr^{-1})),
    fill     = "Scenario"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    legend.position = "bottom"
  )


# Aridity label helper for faceted boxplots
aridity_label_df <- data.frame(
  aridity_class = aridity_levels,
  y             = Inf
) %>%
  mutate(aridity_class = factor(aridity_class, levels = aridity_levels))


# PLOT 3: Boxplot of annual % change by aridity class and scenario

p_aridity_pct <- annual_summary %>%
  pivot_longer(
    cols      = c(delta_s1_pct, delta_s2_pct, delta_s3_pct),
    names_to  = "scenario", values_to = "delta_pct"
  ) %>%
  mutate(
    scenario      = recode(scenario,
      delta_s1_pct = "Mean VPD shift",
      delta_s2_pct = "+28% extreme magnitude",
      delta_s3_pct = "Combined"
    ),
    scenario      = factor(scenario, levels = scenario_levels),
    aridity_class = factor(aridity_class, levels = aridity_levels)
  ) %>%
  ggplot(aes(x = scenario, y = delta_pct, fill = scenario)) +
  geom_boxplot(alpha = 0.85, outlier.shape = NA, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30",
             linewidth = 0.5) +
  geom_text(
    data        = aridity_label_df,
    aes(x = 2, y = Inf, label = aridity_class),
    inherit.aes = FALSE,
    vjust       = 1.5, size = 4, fontface = "bold", colour = "grey20"
  ) +
  facet_wrap(~ aridity_class, ncol = 4) +
  scale_fill_manual(values = scenario_colours) +
  scale_x_discrete(labels = NULL) +
  labs(
    title    = "Annual GPP change by aridity class and scenario",
    subtitle = "Distribution across site-years — % change vs baseline predicted GPP",
    x        = NULL,
    y        = "Annual ΔGPP (%)",
    fill     = "Scenario"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 10),
    strip.background   = element_blank(),
    strip.text         = element_blank(),
    legend.position    = "bottom",
    axis.ticks.x       = element_blank(),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
  )


# PLOT 4: Boxplot of annual absolute change by aridity class and scenario

p_aridity_abs <- annual_summary %>%
  pivot_longer(
    cols      = c(delta_s1_abs, delta_s2_abs, delta_s3_abs),
    names_to  = "scenario", values_to = "delta_abs"
  ) %>%
  mutate(
    scenario      = recode(scenario,
      delta_s1_abs = "Mean VPD shift",
      delta_s2_abs = "+28% extreme magnitude",
      delta_s3_abs = "Combined"
    ),
    scenario      = factor(scenario, levels = scenario_levels),
    aridity_class = factor(aridity_class, levels = aridity_levels)
  ) %>%
  ggplot(aes(x = scenario, y = delta_abs, fill = scenario)) +
  geom_boxplot(alpha = 0.85, outlier.shape = NA, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30",
             linewidth = 0.5) +
  geom_text(
    data        = aridity_label_df,
    aes(x = 2, y = Inf, label = aridity_class),
    inherit.aes = FALSE,
    vjust       = 1.5, size = 4, fontface = "bold", colour = "grey20"
  ) +
  facet_wrap(~ aridity_class, ncol = 4) +
  scale_fill_manual(values = scenario_colours) +
  scale_x_discrete(labels = NULL) +
  labs(
    title    = "Annual GPP change by aridity class and scenario",
    subtitle = "Distribution across site-years — absolute change vs baseline predicted GPP",
    x        = NULL,
    y        = expression(Annual ~ Delta*GPP ~ (g ~ C ~ m^{-2} ~ yr^{-1})),
    fill     = "Scenario"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 10),
    strip.background   = element_blank(),
    strip.text         = element_blank(),
    legend.position    = "bottom",
    axis.ticks.x       = element_blank(),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
  )

print(p_pct)
print(p_abs)
print(p_aridity_pct)
print(p_aridity_abs)
```

```{r}
# Check how many future VPD values exceed the training range per site
vpd_range_check <- map_dfr(names(prepared_envelope), function(s) {
  
  df <- prepared_envelope[[s]]
  vpd_shift <- case_when(
    s %in% humid_sites        ~ aridity_vpd_shift["humid"],
    s %in% dry_subhumid_sites ~ aridity_vpd_shift["dry_subhumid"],
    s %in% semi_arid_sites    ~ aridity_vpd_shift["semi_arid"],
    s %in% arid_sites         ~ aridity_vpd_shift["arid"]
  )
  
  vpd_max_observed <- max(df$VPD, na.rm = TRUE)
  vpd_max_s1       <- max(df$VPD + vpd_shift, na.rm = TRUE)
  vpd_90           <- quantile(df$VPD, 0.90, na.rm = TRUE)
  vpd_max_s2       <- max(ifelse(df$VPD >= vpd_90,
                                 df$VPD * 1.28, df$VPD), na.rm = TRUE)
  
  data.frame(
    site               = s,
    vpd_max_observed   = round(vpd_max_observed, 3),
    vpd_max_s1         = round(vpd_max_s1, 3),
    vpd_max_s2         = round(vpd_max_s2, 3),
    s1_exceeds_obs     = vpd_max_s1 > vpd_max_observed,
    s2_exceeds_obs     = vpd_max_s2 > vpd_max_observed,
    pct_days_s1_beyond = round(mean((df$VPD + vpd_shift) > vpd_max_observed,
                                    na.rm = TRUE) * 100, 1),
    pct_days_s2_beyond = round(mean(ifelse(df$VPD >= vpd_90,
                                           df$VPD * 1.28,
                                           df$VPD) > vpd_max_observed,
                                    na.rm = TRUE) * 100, 1)
  )
})

print(vpd_range_check)
```

```{r}
# How much do scenario VPDs exceed the observed maximum?
vpd_exceedance <- map_dfr(names(prepared_envelope), function(s) {
  
  df <- prepared_envelope[[s]]
  vpd_shift <- case_when(
    s %in% humid_sites        ~ aridity_vpd_shift["humid"],
    s %in% dry_subhumid_sites ~ aridity_vpd_shift["dry_subhumid"],
    s %in% semi_arid_sites    ~ aridity_vpd_shift["semi_arid"],
    s %in% arid_sites         ~ aridity_vpd_shift["arid"]
  )
  
  vpd_max_obs <- max(df$VPD, na.rm = TRUE)
  vpd_90      <- quantile(df$VPD, 0.90, na.rm = TRUE)
  
  vpd_s1       <- df$VPD + vpd_shift
  vpd_s2       <- ifelse(df$VPD >= vpd_90, df$VPD * 1.28, df$VPD)
  vpd_s3_step1 <- df$VPD + vpd_shift
  vpd_s3       <- ifelse(vpd_s3_step1 >= vpd_90 + vpd_shift,
                         vpd_s3_step1 * 1.28, vpd_s3_step1)
  
  data.frame(
    site              = s,
    vpd_max_obs       = round(vpd_max_obs, 3),
    s1_max_exceedance = round(max(vpd_s1, na.rm = TRUE) - vpd_max_obs, 3),
    s2_max_exceedance = round(max(vpd_s2, na.rm = TRUE) - vpd_max_obs, 3),
    s3_max_exceedance = round(max(vpd_s3, na.rm = TRUE) - vpd_max_obs, 3),
    s1_pct_beyond     = round(mean(vpd_s1 > vpd_max_obs, na.rm = TRUE) * 100, 1),
    s2_pct_beyond     = round(mean(vpd_s2 > vpd_max_obs, na.rm = TRUE) * 100, 1),
    s3_pct_beyond     = round(mean(vpd_s3 > vpd_max_obs, na.rm = TRUE) * 100, 1)
  )
})

print(vpd_exceedance)
```

```{r}
library(tidyverse)

# ── Build site order from most negative slope to least negative ───────────────
sensitivity_order <- site_slopes %>%
  filter(!is.na(slope)) %>%
  arrange(slope) %>%
  pull(site)
```

```{r}
print(sensitivity_order)
```

```{r}
library(tidyverse)

# ── Hardcoded sensitivity order — most negative slope first (Fog) to least (Tum)
sensitivity_order <- c(
  "Fog", "Ade", "Dap", "Rdf", "Das", "Asm", "How", "Dry",
  "Rig", "Ync", "Stp", "Emr", "Gin", "Cpr", "Boy", "Wac",
  "Wom", "Whr", "Cum", "Gww", "Tum"
)

p_pct_sensitivity <- make_plot_df(site_summary, "pct") %>%
  mutate(
    scenario = factor(scenario, levels = scenario_levels),
    site     = factor(site, levels = rev(sensitivity_order))
  ) %>%
  filter(!is.na(site)) %>%
  ggplot(aes(x = site, y = delta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  scale_fill_manual(values = scenario_colours) +
  labs(
    title    = "Projected GPP change under future VPD scenarios",
    subtitle = "Sites ordered from highest to lowest GPP–VPD sensitivity (Fog = steepest slope)",
    x        = "Site (High → Low Sensitivity)",
    y        = "Mean annual ΔGPP (%)",
    fill     = "Scenario"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    legend.position = "bottom"
  )

print(p_pct_sensitivity)
```

