---
title: "site_predictability_final"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Model without and with VPD to get VPD importance
2. Total site predictability with full model of VPD

Need to ensure all models are leave-one-year out validation

Compare model without VPD and model with VPD

meta data of MAP, MAT and MAL stored in site_meta

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)

set.seed(42) # Added for reproducibility

standardise_train_test <- function(train_df, test_df, predictors) {
  # Fix: compute mean/SD from training data only to avoid data leakage
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  
  train_df[predictors] <- scale(train_df[predictors],
                                 center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],
                                 center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}
```


```{r}
run_rf_with_without_vpd <- function(df, site_name) {
  
  # Removed Precip from drop_na and everywhere below
  df <- df %>% drop_na(GPP, Tair, SW_rad, VPD, cwd, date)
  
  if (nrow(df) < 100) {
    warning(paste("Site", site_name, "has insufficient data"))
    return(NULL)
  }
  
  df <- df %>%
    arrange(date) %>%
    mutate(
      year = lubridate::year(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    )
  
  # Removed Precip from both predictor sets
  predictors_with    <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
  predictors_without <- c("Tair", "SW_rad", "cumulative_cwd_7day")
  
  fit_rf_loyo <- function(predictors, label) {
    
    model_df <- df %>% drop_na(GPP, all_of(predictors))
    
    if (nrow(model_df) < 100) {
      message("  Skipping ", label, " at ", site_name, ": insufficient data")
      return(NULL)
    }
    
    # Removed global standardise() call — now done per fold inside the loop
    
    years <- unique(model_df$year)
    
    all_preds      <- list()
    all_importance <- list()
    
    for (year_i in years) {
      
      train_data <- model_df %>% filter(year != year_i)
      test_data  <- model_df %>% filter(year == year_i)
      
      if (nrow(test_data) < 10 || nrow(train_data) < 50) next
      
      # Fix: standardise using training fold statistics only (no leakage)
      scaled      <- standardise_train_test(train_data, test_data, predictors)
      train_data  <- scaled$train
      test_data   <- scaled$test
      
      # Tune mtry: floor(sqrt(p)) instead of hardcoded 3
      mtry_val <- max(2, floor(sqrt(length(predictors))))
      
      formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
      
      model <- randomForest(formula, data = train_data,
                            ntree = 500, mtry = mtry_val, importance = TRUE)
      
      preds <- predict(model, test_data)
      
      all_preds[[as.character(year_i)]] <- data.frame(
        Observed  = test_data$GPP,
        Predicted = preds,
        year      = year_i
      )
      
      all_importance[[as.character(year_i)]] <- data.frame(
        variable   = rownames(importance(model)),
        importance = importance(model, type = 1)[, 1],
        year       = year_i
      )
    }
    
    results <- bind_rows(all_preds)
    
    if (nrow(results) == 0) return(NULL)
    
    rmse     <- sqrt(mean((results$Observed - results$Predicted)^2))
    rsq      <- cor(results$Observed, results$Predicted)^2
    mean_obs <- mean(results$Observed)
    nse      <- 1 - sum((results$Observed - results$Predicted)^2) /
                     sum((results$Observed - mean_obs)^2)
    
    importance_df <- bind_rows(all_importance) %>%
      group_by(variable) %>%
      summarise(importance = mean(importance), .groups = "drop") %>%
      mutate(model = label, site = site_name)
    
    list(
      metrics     = data.frame(site = site_name, model = label,
                               RMSE = rmse, R2 = rsq, NSE = nse),
      importance  = importance_df,
      predictions = results %>% mutate(site = site_name, model = label)
    )
  }
  
  with_vpd    <- fit_rf_loyo(predictors_with,    "with_VPD")
  without_vpd <- fit_rf_loyo(predictors_without, "without_VPD")
  
  if (is.null(with_vpd) || is.null(without_vpd)) return(NULL)
  
  list(
    site        = site_name,
    metrics     = bind_rows(with_vpd$metrics,     without_vpd$metrics),
    importance  = bind_rows(with_vpd$importance,  without_vpd$importance),
    predictions = bind_rows(with_vpd$predictions, without_vpd$predictions)
  )
}
```

Run model

```{r}
rf_results_vpd <- purrr::map2(
  sites_daily,
  names(sites_daily),
  run_rf_with_without_vpd
)
```

```{r}
all_metrics_vpd <- purrr::map_df(rf_results_vpd, ~ .x$metrics)
```

Set.seed(42) at the top - ensure reproducibility
New standardise train and test takes train and test separately
Precip removed
mtry will now be 2 for without VPD and 2 for with VPD



Plots:
1. Total R-squared across sites
2. Change in R-squared when adding VPD
3. MAP, MAT and MAL with R-squared
4. No. of years of observation against R-squared

```{r}


#1) Change in R-squared across sites

all_metrics_vpd %>%
  select(site, model, R2) %>%
  pivot_wider(names_from = model, values_from = R2) %>%
  mutate(site    = factor(site, levels = wet_to_dry_order),
         R2_gain = with_VPD - without_VPD) %>%
  ggplot(aes(x = site, y = R2_gain, fill = R2_gain > 0)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = thesis_cols["main"], "FALSE" = thesis_cols["extra"]),
                    labels = c("TRUE" = "VPD improves R²", "FALSE" = "VPD reduces R²")) +
  coord_flip() +
  labs(title = "ΔR² from adding VPD — Dry to Wet order", x = NULL, y = "ΔR² (with − without)", fill = NULL) +
  theme_classic() +
  theme(legend.position = "top")

library(ggrepel)

#2) R² (with_VPD only) — wet to dry order 
all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  mutate(site = factor(site, levels = wet_to_dry_order)) %>%
  ggplot(aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  coord_flip() +
  labs(
    title = "Model R² (with VPD) — Wet to Dry",
    x     = NULL,
    y     = "R²"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.y = element_text(size = 10))

#3) Mean annual precip, mean annual temp and maximum LAI and R-squared across sites

ggplot(plot_data, aes(x = value, y = R2)) +
  geom_point(size = 3, colour = "#4682B4") +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  labs(y = "R² (RF with VPD)", x = NULL) +
  theme_bw(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 11),
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

#4) Number of years of observation

ggplot(plot_data, aes(x = n_years, y = R2)) +
  geom_point(size = 3, colour = "#4682B4") +
  geom_text(aes(label = site), vjust = -0.8, size = 2.8, colour = "grey30") +
  labs(x = "No. of Years of Observations", y = "R2") +
  theme_classic()
```




