---
title: "site_predictability_final"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Model without and with VPD to get VPD importance
2. Total site predictability with full model of VPD

Need to ensure all models are leave-one-year out validation

Compare model without VPD and model with VPD

meta data of MAP, MAT and MAL stored in site_meta

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)

set.seed(42) # Added for reproducibility

standardise_train_test <- function(train_df, test_df, predictors) {
  # Fix: compute mean/SD from training data only to avoid data leakage
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  
  train_df[predictors] <- scale(train_df[predictors],
                                 center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],
                                 center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}
```


```{r}
run_rf_with_without_vpd <- function(df, site_name) {
  
  # Removed Precip from drop_na and everywhere below
  df <- df %>% drop_na(GPP, Tair, SW_rad, VPD, cwd, date)
  
  if (nrow(df) < 100) {
    warning(paste("Site", site_name, "has insufficient data"))
    return(NULL)
  }
  
  df <- df %>%
    arrange(date) %>%
    mutate(
      year = lubridate::year(date),
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE)
    )
  
  # Removed Precip from both predictor sets
  predictors_with    <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")
  predictors_without <- c("Tair", "SW_rad", "cumulative_cwd_7day")
  
  fit_rf_loyo <- function(predictors, label) {
    
    model_df <- df %>% drop_na(GPP, all_of(predictors))
    
    if (nrow(model_df) < 100) {
      message("  Skipping ", label, " at ", site_name, ": insufficient data")
      return(NULL)
    }
    
    # Removed global standardise() call — now done per fold inside the loop
    
    years <- unique(model_df$year)
    
    all_preds      <- list()
    all_importance <- list()
    
    for (year_i in years) {
      
      train_data <- model_df %>% filter(year != year_i)
      test_data  <- model_df %>% filter(year == year_i)
      
      if (nrow(test_data) < 10 || nrow(train_data) < 50) next
      
      # Fix: standardise using training fold statistics only (no leakage)
      test_dates <- test_data$date  
     
      scaled      <- standardise_train_test(train_data, test_data, predictors)
      train_data  <- scaled$train
      test_data   <- scaled$test
      
      # Tune mtry: floor(sqrt(p)) instead of hardcoded 3
      mtry_val <- max(2, floor(sqrt(length(predictors))))
      
      formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
      
      model <- randomForest(formula, data = train_data,
                            ntree = 500, mtry = mtry_val, importance = TRUE)
      
      preds <- predict(model, test_data)
      
      all_preds[[as.character(year_i)]] <- data.frame(
        Observed  = test_data$GPP,
        Predicted = preds,
        year      = year_i,
        date      = test_data$date
      )
      
      all_importance[[as.character(year_i)]] <- data.frame(
        variable   = rownames(importance(model)),
        importance = importance(model, type = 1)[, 1],
        year       = year_i
      )
    }
    
    results <- bind_rows(all_preds)
    
    if (nrow(results) == 0) return(NULL)
    
    rmse     <- sqrt(mean((results$Observed - results$Predicted)^2))
    rsq      <- cor(results$Observed, results$Predicted)^2
    mean_obs <- mean(results$Observed)
    nse      <- 1 - sum((results$Observed - results$Predicted)^2) /
                     sum((results$Observed - mean_obs)^2)
    
    importance_df <- bind_rows(all_importance) %>%
      group_by(variable) %>%
      summarise(importance = mean(importance), .groups = "drop") %>%
      mutate(model = label, site = site_name)
    
    list(
      metrics     = data.frame(site = site_name, model = label,
                               RMSE = rmse, R2 = rsq, NSE = nse),
      importance  = importance_df,
      predictions = results %>% mutate(site = site_name, model = label)
    )
  }
  
  with_vpd    <- fit_rf_loyo(predictors_with,    "with_VPD")
  without_vpd <- fit_rf_loyo(predictors_without, "without_VPD")
  
  if (is.null(with_vpd) || is.null(without_vpd)) return(NULL)
  
  list(
    site        = site_name,
    metrics     = bind_rows(with_vpd$metrics,     without_vpd$metrics),
    importance  = bind_rows(with_vpd$importance,  without_vpd$importance),
    predictions = bind_rows(with_vpd$predictions, without_vpd$predictions)
  )
}
```

Run model

```{r}
rf_results_vpd <- purrr::map2(
  sites_daily,
  names(sites_daily),
  run_rf_with_without_vpd
)
```

```{r}
all_metrics_vpd <- purrr::map_df(rf_results_vpd, ~ .x$metrics)
```

Set.seed(42) at the top - ensure reproducibility
New standardise train and test takes train and test separately
Precip removed
mtry will now be 2 for without VPD and 2 for with VPD

Plots:
1. Total R-squared across sites
2. Change in R-squared when adding VPD
3. MAP, MAT and MAL with R-squared
4. No. of years of observation against R-squared


```{r}
# --- Compute number of years per site ---
n_years_df <- purrr::map_df(rf_results_vpd, function(res) {
  if (is.null(res)) return(NULL)
  data.frame(
    site    = res$site,
    n_years = length(unique(res$predictions$year))
  )
})

# --- Add n_years to site_meta ---
site_meta <- site_meta %>%
  left_join(n_years_df, by = "site")

```

```{r}

# --- Prepare plot_data for MAP, MAT, MAL vs R² ---
plot_data <- all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  select(site, R2) %>%
  left_join(
    site_meta %>%
      select(site, MAP, MAT, MAL, n_years),
    by = "site"
  ) %>%
  pivot_longer(
    cols = c(MAP, MAT, MAL),
    names_to = "variable",
    values_to = "value"
  )

# 1) Change in R² across sites
all_metrics_vpd %>%
  select(site, model, R2) %>%
  pivot_wider(names_from = model, values_from = R2) %>%
  mutate(
    site    = factor(site, levels = wet_to_dry_order),
    R2_gain = with_VPD - without_VPD
  ) %>%
  ggplot(aes(x = site, y = R2_gain)) +
  geom_col(width = 0.7, fill = thesis_cols["main"]) +   
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
  coord_flip() +
  labs(
    title = "ΔR² from adding VPD",
    y = "ΔR² (with − without)",
    x = "Sites (Wet to Dry)"   
  ) +
  theme_classic() +
  theme(legend.position = "none")

# 2) R² (with_VPD only) - wet to dry order
all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  mutate(site = factor(site, levels = wet_to_dry_order)) %>%
  ggplot(aes(x = site, y = R2)) +
  geom_col(width = 0.7, fill = thesis_cols["main"]) +
  coord_flip() +
  labs(
    title = "Model R² (with VPD)",
    y = "R²",
    x = "Sites (Wet to Dry)") +
  theme_classic(base_size = 13) +
  theme(axis.text.y = element_text(size = 10))

# 3) MAP, MAT, MAL vs R²
library(ggrepel)
ggplot(plot_data, aes(x = value, y = R2)) +
  geom_point(size = 3, colour = thesis_cols["main"]) +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  labs(y = "R² (RF with VPD)", x = NULL) +
  var_labels <- c(
  "MAP" = "Mean Annual Precipitation (mm)",
  "MAT" = "Mean Annual Temperature (°C)",
  "MAL" = "Maximum Annual LAI"
) +
# add labeller = labeller(variable = var_labels) to facet_wrap
facet_wrap(~ variable, scales = "free_x", ncol = 3,
           labeller = labeller(variable = var_labels))
  theme_bw(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 11),
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )

# 4) Number of years vs R² 
ggplot(plot_data %>% distinct(site, R2, n_years), aes(x = n_years, y = R2)) +
  geom_point(size = 3, colour = thesis_cols["main"]) +
  geom_text(aes(label = site), vjust = -0.8, size = 2.8, colour = "grey30") +
  labs(x = "No. of Years of Observations", y = "R²") +
  theme_classic()
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggrepel)

# ── Shared theme and palette ──────────────────────────────────────────────────

thesis_theme <- theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 9),
    axis.title       = element_text(size = 11),
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 11),
    panel.grid.major = element_line(colour = "grey94", linewidth = 0.3),
    legend.position  = "none"
  )

main_col <- thesis_cols["main"]   # your existing single colour

# ── Plot 1: ΔR² from adding VPD ──────────────────────────────────────────────

delta_r2_data <- all_metrics_vpd %>%
  select(site, model, R2) %>%
  pivot_wider(names_from = model, values_from = R2) %>%
  mutate(
    site    = factor(site, levels = wet_to_dry_order),
    R2_gain = with_VPD - without_VPD
  )

mean_delta <- mean(delta_r2_data$R2_gain, na.rm = TRUE)

p1 <- ggplot(delta_r2_data, aes(x = R2_gain, y = site)) +
  geom_col(width = 0.7, fill = main_col) +
  geom_vline(
    xintercept = mean_delta,
    linetype   = "dashed",
    colour     = "grey30",
    linewidth  = 0.5
  ) +
  annotate(
    "text",
    x      = mean_delta + 0.005,
    y      = 1,
    label  = paste0("Mean = ", round(mean_delta, 3)),
    hjust  = 0,
    size   = 3,
    colour = "grey30"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.05))) +
  labs(
    x = expression(Delta * R^2 ~ "(with − without VPD)"),
    y = "Site (Wet to Dry)"
  ) +
  thesis_theme

print(p1)

# ── Plot 2: R² (with VPD) ordered by GPP variability ─────────────────────────

median_r2 <- median(
  all_metrics_vpd %>% filter(model == "with_VPD") %>% pull(R2),
  na.rm = TRUE
)

p2 <- all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  mutate(site = factor(site, levels = wet_to_dry_order)) %>%
  ggplot(aes(x = R2, y = site)) +
  geom_col(width = 0.7, fill = main_col) +
  geom_vline(
    xintercept = median_r2,
    linetype   = "dashed",
    colour     = "grey30",
    linewidth  = 0.5
  ) +
  annotate(
    "text",
    x      = median_r2 + 0.01,
    y      = 1,
    label  = paste0("Median = ", round(median_r2, 2)),
    hjust  = 0,
    size   = 3,
    colour = "grey30"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = expression(R^2 ~ "(with VPD)"),
    y = "Site (Wet to Dry)"
  ) +
  thesis_theme

print(p2)

# ── Plot 3: MAP, MAT, MAL vs R² ──────────────────────────────────────────────

var_labels <- c(
  "MAP" = "MAP (mm / year)",
  "MAT" = "MAT (°C)",
  "MAL" = "MAL (m² / m²)"
)

plot_data <- all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  select(site, R2) %>%
  left_join(
    site_meta %>% select(site, MAT, MAP, MAL, n_years),
    by = "site"
  ) %>%
  pivot_longer(
    cols      = c(MAT, MAP, MAL),
    names_to  = "variable",
    values_to = "value"
  ) %>%
  mutate(variable = factor(variable, levels = c("MAT", "MAP", "MAL")))


p3 <- ggplot(plot_data, aes(x = value, y = R2)) +
  geom_smooth(
    method    = "loess",
    se        = TRUE,
    colour    = "grey40",
    fill      = "grey85",
    linewidth = 0.6,
    span      = 1.3       # increase if the line is too wiggly, decrease for more flex
  ) +
  geom_point(size = 3, colour = main_col) +
  geom_text_repel(
    aes(label = site),
    size         = 2.8,
    colour       = "grey40",
    max.overlaps = 15
  ) +
  facet_wrap(
    ~ variable,
    scales   = "free_x",
    ncol     = 3,
    labeller = labeller(variable = var_labels)
  ) +
  scale_y_continuous(
  limits = c(0, 0.8),
  expand = expansion(mult = c(0.02, 0.05))
) +
  labs(
    y       = expression(R^2 ~ "(RF with VPD)"),
    x       = NULL,
    caption = "MAP = Mean Annual Precipitation; MAT = Mean Annual Temperature; MAL = Maximum Annual LAI"
  ) +
  thesis_theme +
  theme(
    panel.border = element_rect(colour = "black", fill = NA),
    plot.caption = element_text(size = 8, colour = "grey40", hjust = 0)
  )

print(p3)

# ── Plot 4: Record length vs R² (null result — include only if needed) ────────

record_data <- plot_data %>%
  distinct(site, R2, n_years)

# fit lm to extract p-value for annotation
lm_fit <- lm(R2 ~ n_years, data = record_data)
lm_p   <- summary(lm_fit)$coefficients[2, 4]
lm_r2  <- summary(lm_fit)$r.squared

p4 <- ggplot(record_data, aes(x = n_years, y = R2)) +
  geom_smooth(
    method    = "lm",
    se        = TRUE,
    colour    = "grey50",
    fill      = "grey85",
    linewidth = 0.6
  ) +
  geom_point(size = 3, colour = main_col) +
  geom_text_repel(
    aes(label = site),
    size   = 2.8,
    colour = "grey40"
  ) +
  annotate(
    "text",
    x      = Inf, y = Inf,
    label  = paste0("p = ", round(lm_p, 3), "  R² = ", round(lm_r2, 3)),
    hjust  = 1.1, vjust = 1.5,
    size   = 3.2,
    colour = "grey30"
  ) +
  labs(
    x = "Years of observation",
    y = expression(R^2)
  ) +
  thesis_theme

print(p4)

```



Units on the MAP, MAT and MAL? 
Also a line of best fit to show the trend?

```{r}
metrics_wide <- all_metrics_vpd %>%
  select(site, model, R2, RMSE, NSE) %>%
  pivot_wider(
    names_from  = model,
    values_from = c(R2, RMSE, NSE)
  ) %>%
  mutate(
    delta_R2   = R2_with_VPD   - R2_without_VPD,
    delta_RMSE = RMSE_without_VPD - RMSE_with_VPD,
    delta_NSE  = NSE_with_VPD  - NSE_without_VPD
  )

summary(metrics_wide)
```

New MAL, MAT and MAP plot with Loess line and units
```{r}
ggplot(plot_data, aes(x = value, y = R2)) +
  geom_point(size = 3, colour = thesis_cols["main"]) +
  geom_smooth(
    method    = "loess",
    span      = 1.0,
    se        = TRUE,
    colour    = "grey40",
    fill      = "grey70",
    linewidth = 0.7
  ) +
  geom_text_repel(aes(label = site), size = 2.8, colour = "grey30",
                  max.overlaps = 15) +
  facet_wrap(~ variable, scales = "free_x", ncol = 3,
             labeller = labeller(variable = c(
               MAP = "MAP (mm/year)",
               MAT = "MAT(°C)",
               MAL = "MAL"
             ))) +
  coord_cartesian(ylim = c(0, 1)) +   # clips display without dropping data
  labs(y = "R² (RF with VPD)", x = NULL) +
  theme_bw(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 11),
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA)
  )
```

```{r}
years_r2_df <- plot_data %>% 
  distinct(site, R2, n_years)

model_years <- lm(R2 ~ n_years, data = years_r2_df)

summary(model_years)
```
NEED TO FIGURE OUT: 
Time series of observed and predicted GPP:

```{r}
# ── Build ts_long ─────────────────────────────────────────────────────────────
ts_long <- purrr::map_dfr(rf_results_vpd, function(res) {
  if (is.null(res)) return(NULL)
  res$predictions %>%
    filter(model == "with_VPD") %>%
    select(site, date, Observed, Predicted)
}) %>%
  pivot_longer(
    cols      = c(Observed, Predicted),
    names_to  = "type",
    values_to = "GPP"
  ) %>%
  mutate(date = as.Date(date))

# ── Build site R² labels ──────────────────────────────────────────────────────
site_r2 <- all_metrics_vpd %>%
  filter(model == "with_VPD") %>%
  select(site, R2) %>%
  mutate(label = paste0("R² = ", round(R2, 2)))

# ── Order sites wet to dry, 4 per page ───────────────────────────────────────
ordered_sites <- wet_to_dry_order[wet_to_dry_order %in% unique(ts_long$site)]
site_pages    <- split(ordered_sites, ceiling(seq_along(ordered_sites) / 4))

# ── Plot function ─────────────────────────────────────────────────────────────
plot_ts_page <- function(site_group) {

  plot_df <- ts_long %>%
    filter(site %in% site_group) %>%
    mutate(site = factor(site, levels = site_group))

  label_df <- site_r2 %>%
    filter(site %in% site_group) %>%
    mutate(site = factor(site, levels = site_group))

  ggplot(plot_df, aes(x = date, y = GPP, colour = type)) +
    geom_line(aes(alpha = type), linewidth = 0.4) +
    geom_text(
      data        = label_df,
      aes(label   = label),
      x           = -Inf, y = Inf,
      hjust       = -0.08, vjust = 1.6,
      inherit.aes = FALSE,
      size        = 3,
      colour      = "grey20",
      fontface    = "bold"
    ) +
    facet_wrap(~ site, ncol = 2, scales = "free_y") +
    scale_colour_manual(
      values = c("Observed" = "#2c2c2c", "Predicted" = "#4682B4"),
      name   = NULL
    ) +
    scale_alpha_manual(
      values = c("Observed" = 0.5, "Predicted" = 0.9),
      guide  = "none"
    ) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(
      x = NULL,
      y = expression(GPP ~ (g ~ C ~ m^{-2} ~ d^{-1}))
    ) +
    theme_classic(base_size = 11) +
    theme(
      strip.background   = element_blank(),
      strip.text         = element_text(face = "bold", size = 10),
      legend.position    = "bottom",
      legend.key.width   = unit(1.2, "cm"),
      panel.grid.major.y = element_line(colour = "grey94", linewidth = 0.25),
      axis.text.x        = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y        = element_text(size = 8),
      panel.spacing      = unit(1, "lines")
    )
}

# ── Print all pages ───────────────────────────────────────────────────────────
purrr::walk(site_pages, ~ print(plot_ts_page(.x)))

```



