---
title: "supplementary_materials"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Supplementary materials and data:
- Create site metadata - consistent use of this
- Latitude and longitude for each site
- PFT for each site
- Climate for each site
- Mean annual precip
- Mean annual temp
- Maximum LAI across all years


Need to ensure that plots are colour-blind friendly and consistent across the thesis - one main accent colour followed by consistent supporting colours.


```{r}
library(khroma)
tol_muted <- khroma::colour("muted")(9)
tol_muted
```

```{r}
thesis_cols <- c(
  main = "#44AA99",
  secondary = "#88CCEE",
  highlight = "#CC6677",
  extra = "#999933",
  neutral = "grey60"
)
```

Supplementary materials:

```{r}

library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)

# Site Metadata - fill in lat/lon values here

site_meta <- tibble(
  site      = c("How", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr",
                "Fog", "Gin", "Gww", "Rdf", "Rig", "Wac", "Whr", "Wom",
                "Ync", "Cpr", "Tum", "Ade", "Boy"),
  lat       = c(-12.4943, -14.1593, -14.0633, -15.2588, -17.1507, -22.2830, -33.6152, -23.8587, -12.5452,  -31.3764, -30.1913, -14.5636, -36.6499, -37.4259, -36.6732, -37.4222, -34.9893, -34.0021, -35.6566, -13.0769, -32.4771),
  lon       = c(131.1523, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746, 131.3072, 115.7138, 120.6541, 132.4776, 145.5759, 145.1878, 145.0294, 144.0944, 146.2907, 140.5891, 148.1517, 131.1178, 116.9386),
PFT              = c("WSA", "SAV", "GRA", "SAV", "GRA", "SAV", "EBF", "GRA",
                       "WET", "WSA", "SAV", "WSA", "GRA", "EBF", "EBF", "EBF",
                       "GRA", "SAV", "EBF", "WSA", "WSA"),

  koppen           = c("Tropical", "Tropical", "Tropical", "Tropical", "Grassland", "Desert", "Temperate", "Grassland", "Tropical", "Subtropical", "Grassland", "Tropical", "Temperate", "Temperate", "Temperate", "Temperate", "Grassland", "Grassland", "Temperate", "Tropical", "Temperate")
)

```

```{r}

# Map

australia <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

sites_sf <- site_meta %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = australia, fill = "grey95", colour = "grey50") +
  geom_sf(data = sites_sf, colour = thesis_cols["highlight"], size = 2.5) +
  geom_text_repel(
    data = site_meta,
    aes(x = lon, y = lat, label = site),
    size = 3,
    segment.colour = "grey40",
    segment.size = 0.4,
    box.padding = 0.5,
    point.padding = 0.3,
    max.overlaps = Inf        # forces all labels to be drawn
  ) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  labs(
    title = "OzFlux study sites",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw()
```
Can I add in precipitation/PFT?


MAP, MAT and MAL:

```{r}
site_summary <- map_dfr(sites_daily, function(df) {
  df %>%
    group_by(site) %>%
    summarise(
      MAT = mean(Tair, na.rm = TRUE),
      MAP = sum(Precip, na.rm = TRUE) / n_distinct(year),
      MAL = max(LAI_4day, na.rm = TRUE),
      .groups = "drop"
    )
})

site_meta <- site_meta %>%
  left_join(site_summary, by = "site")
```

```{r}
library(ggplot2)

ggplot(site_meta %>% arrange(desc(MAP)),
       aes(x = reorder(site, MAP), y = MAP)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAP (mm/year)", title = "Mean Annual Precipitation") +
  theme_bw()

ggplot(site_meta %>% arrange(desc(MAT)),
       aes(x = reorder(site, MAT), y = MAT)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAT (°C)", title = "Mean Annual Temperature") +
  theme_bw()

ggplot(site_meta %>% arrange(desc(MAL)),
       aes(x = reorder(site, MAL), y = MAL)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAL (m²/m²)", title = "Maximum Annual LAI") +
  theme_bw()
```
```{r}
LAI_order <- c("Tum", "Wom", "Wac", "Rig", "Dap", "How", "Fog", "Das", "Cum", "Ade", "Dry", "Rdf", "Whr", "Emr", "Boy", "Stp", "Gin", "Ync", "Cpr", "Asm", "Gww")

wet_to_dry_order <- c("Wac", "How", "Das", "Fog", "Ade", "Dap", "Rdf", "Tum", "Dry", "Cum", "Wom", "Stp", "Gin", "Emr", "Boy", "Rig", "Whr", "Gww", "Ync", "Cpr", "Asm")
```

```{r}
ggplot(site_meta %>% arrange(desc(MAP)),
       aes(x = reorder(site, -MAP), y = MAP, fill = MAP)) +
  geom_col() +
  scale_fill_gradient(
    low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Mean Annual Precipitation (mm)",
    title = "Sites by Mean Annual Precipitation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplot(site_meta %>% arrange(desc(MAT)),
       aes(x = reorder(site, -MAT), y = MAT, fill = MAT)) +
  geom_col() +
  scale_fill_gradient(
    low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Mean Annual Temperature (°C)",
    title = "Sites by Mean Annual Temperature"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplot(site_meta %>% arrange(desc(MAL)),
       aes(x = reorder(site, -MAL), y = MAL, fill = MAL)) +
  geom_col() +
  scale_fill_gradient(
  low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Maximum LAI",
    title = "Sites by Leaf Area Index (LAI)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```


