---
title: "supplementary_materials"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Supplementary materials and data:
- Create site metadata - consistent use of this
- Latitude and longitude for each site
- PFT for each site
- Climate for each site
- Mean annual precip
- Mean annual temp
- Maximum LAI across all years


Need to ensure that plots are colour-blind friendly and consistent across the thesis - one main accent colour followed by consistent supporting colours.


```{r}
library(khroma)
tol_muted <- khroma::colour("muted")(9)
tol_muted
```

```{r}
thesis_cols <- c(
  main = "#44AA99",
  secondary = "#88CCEE",
  highlight = "#CC6677",
  extra = "#999933",
  neutral = "grey60"
)
```

Supplementary materials:

```{r}

library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)

site_meta <- tibble(
  site = c("How", "Das", "Dap", "Dry", "Stp", "Asm", "Cum", "Emr",
           "Fog", "Gin", "Rdf", "Rig", "Wac", "Whr", "Wom",
           "Ync", "Cpr", "Tum", "Ade", "Boy"),
  lat  = c(-12.4943, -14.1593, -14.0633, -15.2588, -17.1507, -22.2830, -33.6152, -23.8587,
           -12.5452, -31.3764, -14.5636, -36.6499, -37.4259, -36.6732, -37.4222,
           -34.9893, -34.0021, -35.6566, -13.0769, -32.4771),
  lon  = c(131.1523, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 132.4776, 145.5759, 145.1878, 145.0294, 144.0944,
           146.2907, 140.5891, 148.1517, 131.1178, 116.9386),
  PFT  = c("WSA", "SAV", "GRA", "SAV", "GRA", "SAV", "EBF", "GRA",
           "WET", "WSA", "WSA", "GRA", "EBF", "EBF", "EBF",
           "GRA", "SAV", "EBF", "WSA", "WSA"),
  koppen = c("Tropical", "Tropical", "Tropical", "Tropical", "Grassland", "Desert",
             "Temperate", "Grassland", "Tropical", "Subtropical", "Tropical", "Temperate",
             "Temperate", "Temperate", "Temperate", "Grassland", "Grassland",
             "Temperate", "Tropical", "Temperate"),
  AI = c(0.754, 0.482, 0.502, 0.34, 0.233, 0.122, 0.522, 0.272, 0.684, 0.321,
         0.396, 0.365, 1.118, 0.349, 0.854, 0.23, 0.122, 1.221, 0.62, 0.227)
)

sites_sf <- site_meta |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

```

```{r}

australia <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf")

```

```{r}
#Aridity index sites map

ggplot() +
  geom_sf(data = australia, fill = "grey95", colour = "grey60", linewidth = 0.4) +
  geom_sf(
    data = sites_sf,
    aes(colour = AI),
    size  = 3,
    shape = 19
  ) +
  geom_text_repel(
    data          = site_meta,
    aes(x = lon, y = lat, label = site),
    size          = 2.8,
    fontface      = "bold",
    colour        = "grey20",
    segment.size  = 0.35,
    segment.color = "grey50",
    box.padding   = 0.5,
    point.padding = 0.3,
    max.overlaps  = Inf,
    show.legend   = FALSE
  ) +
 scale_colour_gradientn(
    colours = c("#8C510A", "#BF812D", "#F6E8C3", "#80CDC1", "#01665E"),
    values  = scales::rescale(c(0, 0.2, 0.5, 0.85, 1.3), to = c(0, 1)),
    limits  = c(0, 1.3),
    name    = "Aridity index (AI)",
    guide   = guide_colorbar(
      barwidth       = 10,
      barheight      = 0.5,
      title.position = "top",
      title.hjust    = 0.5
    )
  ) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  labs(
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position   = "bottom",
    legend.direction  = "horizontal",
    legend.margin = ggplot2::margin(t = 8, r = 12, b = 8, l = 12),
    legend.background = element_rect(fill = "white", colour = "grey80", linewidth = 0.3),
    legend.key        = element_blank(),
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(size = 9, colour = "grey40"),
    axis.title        = element_text(size = 9)
  )

```
MAP, MAT and MAL:

```{r}
site_summary <- map_dfr(sites_daily, function(df) {
  
  # Calculate annual summaries first
  annual <- df %>%
    group_by(site, year) %>%
    summarise(
      annual_MAT = mean(Tair,   na.rm = TRUE),
      annual_MAP = sum(Precip,  na.rm = TRUE),
      annual_MAL = max(LAI_4day, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(annual_MAL = ifelse(is.infinite(annual_MAL), NA, annual_MAL))
  
  # Then average across years
  annual %>%
    group_by(site) %>%
    summarise(
      MAT = mean(annual_MAT, na.rm = TRUE),  # mean of annual means
      MAP = mean(annual_MAP, na.rm = TRUE),  # mean of annual totals
      MAL = mean(annual_MAL, na.rm = TRUE),  # mean of annual maximums
      .groups = "drop"
    )
})

# Remove old MAT/MAP/MAL columns before joining to avoid duplicates
site_meta <- site_meta %>%
  select(-any_of(c("MAT", "MAP", "MAL", "MAT.x", "MAP.x", "MAL.x",
                               "MAT.y", "MAP.y", "MAL.y"))) %>%
  left_join(site_summary, by = "site")


```

```{r}
library(ggplot2)

ggplot(site_meta %>% arrange(desc(MAP)),
       aes(x = reorder(site, MAP), y = MAP)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAP (mm/year)", title = "Mean Annual Precipitation") +
  theme_bw()

ggplot(site_meta %>% arrange(desc(MAT)),
       aes(x = reorder(site, MAT), y = MAT)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAT (°C)", title = "Mean Annual Temperature") +
  theme_bw()

ggplot(site_meta %>% arrange(desc(MAL)),
       aes(x = reorder(site, MAL), y = MAL)) +
  geom_bar(stat = "identity", fill = thesis_cols["neutral"]) +
  coord_flip() +
  labs(x = NULL, y = "MAL (m²/m²)", title = "Maximum Annual LAI") +
  theme_bw()
```

```{r}
LAI_order <- c("Tum", "Wom", "Wac", "Dap", "Rig", "Fog", "How", "Das", "Ade", "Cum", "Dry", "Rdf", "Emr", "Whr", "Boy", "Gin", "Stp", "Ync", "Asm", "Cpr")

wet_to_dry_order <- c("Wac", "How", "Das", "Fog", "Ade", "Dap", "Rdf", "Dry", "Tum", "Cum", "Wom", "Stp", "Gin", "Emr", "Boy", "Whr", "Rig", "Ync", "Cpr", "Asm")

aridity_index_order <- c("Tum", "Wac", "Wom", "How", "Fog", "Ade", "Cum", "Dap", "Das", "Rdf", "Rig", "Whr", "Dry", "Gin", "Emr", "Stp", "Ync", "Boy", "Asm", "Cpr")
```

```{r}
ggplot(site_meta %>% arrange(desc(MAP)),
       aes(x = reorder(site, -MAP), y = MAP, fill = MAP)) +
  geom_col() +
  scale_fill_gradient(
    low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Mean Annual Precipitation (mm)",
    title = "Sites by Mean Annual Precipitation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplot(site_meta %>% arrange(desc(MAT)),
       aes(x = reorder(site, -MAT), y = MAT, fill = MAT)) +
  geom_col() +
  scale_fill_gradient(
    low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Mean Annual Temperature (°C)",
    title = "Sites by Mean Annual Temperature"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplot(site_meta %>% arrange(desc(MAL)),
       aes(x = reorder(site, -MAL), y = MAL, fill = MAL)) +
  geom_col() +
  scale_fill_gradient(
  low  = thesis_cols["neutral"],
    high = thesis_cols["extra"],
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "Maximum LAI",
    title = "Sites by Mean Max. Annual Leaf Area Index (LAI)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```
Which sites have the most inter-annual GPP variability?

```{r}
iav_df <- purrr::map_df(names(sites_daily), function(s) {
  sites_daily[[s]] %>%
    drop_na(GPP, date) %>%
    mutate(year = lubridate::year(date)) %>%
    group_by(year) %>%
    summarise(annual_GPP = mean(GPP, na.rm = TRUE), .groups = "drop") %>%
    summarise(
      site     = s,
      mean_GPP = mean(annual_GPP),
      sd_GPP   = sd(annual_GPP),
      cv_GPP   = sd(annual_GPP) / mean(annual_GPP),  # relative variability
      n_years  = n()
    )
}) %>%
  arrange(desc(cv_GPP))

print(iav_df)
```


```{r}
# Annual means per site for the dot strip
annual_means <- purrr::map_df(names(sites_daily), function(s) {
  sites_daily[[s]] %>%
    drop_na(GPP, date) %>%
    mutate(year = lubridate::year(date)) %>%
    group_by(year) %>%
    summarise(annual_GPP = mean(GPP, na.rm = TRUE), site = s, .groups = "drop")
})

site_order_cv <- iav_df %>% arrange(cv_GPP) %>% pull(site)

ggplot() +
  # individual year dots
  geom_point(
    data    = annual_means %>% mutate(site = factor(site, levels = site_order_cv)),
    aes(x = annual_GPP, y = site),
    colour = "grey60", size = 1.8, alpha = 0.7
  ) +
  # CV as error-bar style segment (mean ± 1 SD)
  geom_segment(
    data = iav_df %>% mutate(site = factor(site, levels = site_order_cv)),
    aes(x = mean_GPP - sd_GPP, xend = mean_GPP + sd_GPP, y = site, yend = site),
    colour = thesis_cols["main"], linewidth = 1.1
  ) +
  geom_point(
    data = iav_df %>% mutate(site = factor(site, levels = site_order_cv)),
    aes(x = mean_GPP, y = site),
    colour = thesis_cols["main"], size = 2.5
  ) +
  labs(
    title   = "Interannual variability of GPP by site",
    subtitle = "Dots = individual years | Segment = mean ± 1 SD | Ordered by CV",
    x = expression(Mean ~ annual ~ GPP ~ (g ~ C ~ m^{-2} ~ d^{-1})),
    y = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```


```{r}
GPP_variability_order <- c("Ync", "Asm", "Stp", "Cpr", "Rdf", "Dap", "Whr", "Ade", "Boy", "How", "Gin", "Gww", "Dry", "Cum", "Emr", "Rig", "Fog", "Tum", "Das", "Wom", "Wac")
```



