---
title: "Model_diagnosis"
output: html_document
date: "2026-03-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
plot(sites_daily$Tum$date, sites_daily$Tum$cwd)
```




```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day")

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

#Extract AU-Tum, drops rows with missing values, calculates 7 day rolling cwd sum
tum <- sites_daily[["Tum"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),
    cumulative_cwd_7day = rollapply(
      cwd, width = 7, FUN = sum,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors), GPP)

formula <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# LOYO Random Forest - for each year in the record it holds that year out, trains on all other years and then predicts on the held-out year
years <- unique(tum$year)
loyo_preds <- map_dfr(years, function(yr) {
  
  train_df <- tum %>% filter(year != yr)
  test_df  <- tum %>% filter(year == yr)
  
  if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)
  
  scaled     <- standardise_train_test(train_df, test_df, predictors)
  rf_model   <- randomForest(formula, data = scaled$train,
                             ntree = 500, mtry = mtry_val, importance = FALSE)
  preds      <- predict(rf_model, scaled$test)
  
  data.frame(
    date      = test_df$date,
    GPP_obs   = test_df$GPP,
    GPP_pred  = preds,
    method    = "LOYO"
  )
})

# 80/20 split model - first 80% of rows chronologically as training and the last 20% as the test set
split_idx  <- floor(0.8 * nrow(tum))
train_8020 <- tum[1:split_idx, ]
test_8020  <- tum[(split_idx + 1):nrow(tum), ]

scaled_8020  <- standardise_train_test(train_8020, test_8020, predictors)
rf_8020      <- randomForest(formula, data = scaled_8020$train,
                             ntree = 500, mtry = mtry_val, importance = FALSE)
preds_8020   <- predict(rf_8020, scaled_8020$test)

split_preds <- data.frame(
  date     = test_8020$date,
  GPP_obs  = test_8020$GPP,
  GPP_pred = preds_8020,
  method   = "80/20 Split"
)

#Metrics
compute_metrics <- function(df) {
  obs  <- df$GPP_obs
  pred <- df$GPP_pred
  r2   <- cor(obs, pred)^2
  rmse <- sqrt(mean((obs - pred)^2))
  nse  <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
  data.frame(R2 = round(r2, 3), RMSE = round(rmse, 3), NSE = round(nse, 3))
}

loyo_metrics  <- compute_metrics(loyo_preds)
split_metrics <- compute_metrics(split_preds)

cat("── LOYO metrics ──\n");  print(loyo_metrics)
cat("── 80/20 metrics ──\n"); print(split_metrics)

#Plot
plot_df <- bind_rows(loyo_preds, split_preds) %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(
    method = factor(method, levels = c("LOYO", "80/20 Split")),
    type   = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted")
  )

# Label data for facet annotations
label_df <- data.frame(
  method = factor(c("LOYO", "80/20 Split"), levels = c("LOYO", "80/20 Split")),
  label  = c(
    paste0("R² = ", loyo_metrics$R2,  "  NSE = ", loyo_metrics$NSE),
    paste0("R² = ", split_metrics$R2, "  NSE = ", split_metrics$NSE)
  )
)

p <- ggplot(plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_line(linewidth = 0.5) +
  geom_text(
    data    = label_df,
    aes(label = label),
    x       = -Inf, y = Inf,
    hjust   = -0.05, vjust = 1.5,
    inherit.aes = FALSE,
    size    = 3.5, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ method, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.5,      "Predicted" = 0.9)) +
  labs(
    title    = "AU-Tum: Observed vs Predicted GPP",
    subtitle = "Comparing Leave-One-Year-Out vs 80/20 chronological split",
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    strip.text      = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

ggsave("AU_Tum_RF_comparison.png", plot = p,
       width = 12, height = 7, dpi = 200, bg = "white")

print(p)
```
Removed cwd and added in precip

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors <- c("Tair", "SW_rad", "VPD", "Precip")

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

# ── Extract AU-Tum ────────────────────────────────────────────────────────────
tum <- sites_daily[["Tum"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, Precip, date) %>%
  arrange(date) %>%
  drop_na(all_of(predictors), GPP)

formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# ── 1. LOYO Random Forest ─────────────────────────────────────────────────────
years <- unique(tum$year)
loyo_preds <- map_dfr(years, function(yr) {
  
  train_df <- tum %>% filter(year != yr)
  test_df  <- tum %>% filter(year == yr)
  
  if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)
  
  scaled   <- standardise_train_test(train_df, test_df, predictors)
  rf_model <- randomForest(formula, data = scaled$train,
                           ntree = 500, mtry = mtry_val, importance = FALSE)
  preds    <- predict(rf_model, scaled$test)
  
  data.frame(
    date     = test_df$date,
    GPP_obs  = test_df$GPP,
    GPP_pred = preds,
    method   = "LOYO",
    period   = "Test (held-out year)"
  )
})

# ── 2. 80/20 Chronological Train/Test Split ───────────────────────────────────
split_idx  <- floor(0.8 * nrow(tum))
train_8020 <- tum[1:split_idx, ]
test_8020  <- tum[(split_idx + 1):nrow(tum), ]

scaled_8020 <- standardise_train_test(train_8020, test_8020, predictors)
rf_8020     <- randomForest(formula, data = scaled_8020$train,
                            ntree = 500, mtry = mtry_val, importance = FALSE)

# Predict on both train and test periods for plotting
preds_train_8020 <- predict(rf_8020, scaled_8020$train)
preds_test_8020  <- predict(rf_8020, scaled_8020$test)

split_preds <- bind_rows(
  data.frame(
    date     = train_8020$date,
    GPP_obs  = train_8020$GPP,
    GPP_pred = preds_train_8020,
    method   = "80/20 Split",
    period   = "Train (in-sample)"
  ),
  data.frame(
    date     = test_8020$date,
    GPP_obs  = test_8020$GPP,
    GPP_pred = preds_test_8020,
    method   = "80/20 Split",
    period   = "Test (out-of-sample)"
  )
)

# ── Metrics (test periods only) ───────────────────────────────────────────────
compute_metrics <- function(df) {
  obs  <- df$GPP_obs
  pred <- df$GPP_pred
  data.frame(
    R2   = round(cor(obs, pred)^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2)), 3),
    NSE  = round(1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2), 3)
  )
}

loyo_metrics  <- compute_metrics(loyo_preds)
split_metrics <- compute_metrics(split_preds %>% filter(period == "Test (out-of-sample)"))
train_metrics <- compute_metrics(split_preds %>% filter(period == "Train (in-sample)"))

cat("── LOYO metrics (all years, out-of-sample) ──\n");        print(loyo_metrics)
cat("── 80/20 Test metrics (out-of-sample) ──\n");             print(split_metrics)
cat("── 80/20 Train metrics (in-sample, for reference) ──\n"); print(train_metrics)

# ── Plot ──────────────────────────────────────────────────────────────────────
plot_df <- bind_rows(loyo_preds, split_preds) %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(
    method = factor(method, levels = c("LOYO", "80/20 Split")),
    type   = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted")
  )

# Shading to show train region in 80/20 panel only
split_date <- tum$date[split_idx]

shade_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xmin   = min(tum$date),
  xmax   = split_date,
  ymin   = -Inf,
  ymax   = Inf
)

# Vertical split line for 80/20 panel only
vline_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xint   = split_date
)

# Metric annotations per panel
label_df <- data.frame(
  method = factor(c("LOYO", "80/20 Split"), levels = c("LOYO", "80/20 Split")),
  label  = c(
    paste0("Test R² = ", loyo_metrics$R2, "  |  NSE = ", loyo_metrics$NSE),
    paste0("Train R² = ", train_metrics$R2,
           "  |  Test R² = ", split_metrics$R2,
           "  |  Test NSE = ", split_metrics$NSE)
  )
)

p <- ggplot(plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_rect(
    data        = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill        = "steelblue", alpha = 0.07
  ) +
  geom_vline(
    data        = vline_df,
    aes(xintercept = xint),
    inherit.aes = FALSE,
    linetype    = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_line(linewidth = 0.45) +
  geom_text(
    data        = label_df,
    aes(label   = label),
    x           = -Inf, y = Inf,
    hjust       = -0.04, vjust = 1.6,
    inherit.aes = FALSE,
    size        = 3.3, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ method, ncol = 1) +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.85)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title    = "AU-Tum: Observed vs Predicted GPP",
    subtitle = "LOYO (fully out-of-sample) vs 80/20 chronological split — shaded region = training period",
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL,
    alpha    = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    strip.text       = element_text(face = "bold", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

ggsave("AU_Tum_RF_comparison.png", plot = p,
       width = 14, height = 8, dpi = 200, bg = "white")

print(p)
```

Added in a VPD lag 3 and a 30 day cwd, removed precip

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors <- c("Tair", "SW_rad", "VPD", "VPD_lag3", "cumulative_cwd_30day")

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

# ── Extract AU-Tum and engineer features ──────────────────────────────────────
tum <- sites_daily[["Tum"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),

    # 3-day lagged VPD — captures carry-over stomatal / hydraulic stress
    VPD_lag3 = lag(VPD, 3),

    # 30-day raw rolling sum of daily CWD (same method as before, longer window)
    cumulative_cwd_30day = rollapply(
      cwd, width = 30, FUN = sum,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors), GPP)

formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# ── 1. LOYO Random Forest ─────────────────────────────────────────────────────
years <- unique(tum$year)

loyo_preds <- map_dfr(years, function(yr) {

  train_df <- tum %>% filter(year != yr)
  test_df  <- tum %>% filter(year == yr)

  if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

  scaled   <- standardise_train_test(train_df, test_df, predictors)
  rf_model <- randomForest(formula, data = scaled$train,
                           ntree = 500, mtry = mtry_val, importance = FALSE)
  preds    <- predict(rf_model, scaled$test)

  data.frame(
    date     = test_df$date,
    GPP_obs  = test_df$GPP,
    GPP_pred = preds,
    method   = "LOYO",
    period   = "Test (held-out year)"
  )
})

# ── 2. 80/20 Chronological Train/Test Split ───────────────────────────────────
split_idx  <- floor(0.8 * nrow(tum))
train_8020 <- tum[1:split_idx, ]
test_8020  <- tum[(split_idx + 1):nrow(tum), ]

scaled_8020      <- standardise_train_test(train_8020, test_8020, predictors)
rf_8020          <- randomForest(formula, data = scaled_8020$train,
                                 ntree = 500, mtry = mtry_val, importance = TRUE)
preds_train_8020 <- predict(rf_8020, scaled_8020$train)
preds_test_8020  <- predict(rf_8020, scaled_8020$test)

split_preds <- bind_rows(
  data.frame(
    date     = train_8020$date,
    GPP_obs  = train_8020$GPP,
    GPP_pred = preds_train_8020,
    method   = "80/20 Split",
    period   = "Train (in-sample)"
  ),
  data.frame(
    date     = test_8020$date,
    GPP_obs  = test_8020$GPP,
    GPP_pred = preds_test_8020,
    method   = "80/20 Split",
    period   = "Test (out-of-sample)"
  )
)

# ── Metrics ───────────────────────────────────────────────────────────────────
compute_metrics <- function(df) {
  obs  <- df$GPP_obs
  pred <- df$GPP_pred
  data.frame(
    R2   = round(cor(obs, pred)^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2)), 3),
    NSE  = round(1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2), 3)
  )
}

loyo_metrics  <- compute_metrics(loyo_preds)
split_metrics <- compute_metrics(split_preds %>% filter(period == "Test (out-of-sample)"))
train_metrics <- compute_metrics(split_preds %>% filter(period == "Train (in-sample)"))

cat("── LOYO metrics (all years, out-of-sample) ──\n");        print(loyo_metrics)
cat("── 80/20 Test metrics (out-of-sample) ──\n");             print(split_metrics)
cat("── 80/20 Train metrics (in-sample, for reference) ──\n"); print(train_metrics)

# ── Variable importance from 80/20 model ─────────────────────────────────────
importance_df <- data.frame(
  variable   = rownames(importance(rf_8020)),
  importance = importance(rf_8020, type = 1)[, 1]  # %IncMSE
) %>% arrange(desc(importance))

cat("\n── Variable importance (80/20 model, %IncMSE) ──\n")
print(importance_df)

# ── Plot: time series ─────────────────────────────────────────────────────────
plot_df <- bind_rows(loyo_preds, split_preds) %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(
    method = factor(method, levels = c("LOYO", "80/20 Split")),
    type   = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted")
  )

split_date <- tum$date[split_idx]

shade_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xmin   = min(tum$date),
  xmax   = split_date,
  ymin   = -Inf,
  ymax   = Inf
)

vline_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xint   = split_date
)

label_df <- data.frame(
  method = factor(c("LOYO", "80/20 Split"), levels = c("LOYO", "80/20 Split")),
  label  = c(
    paste0("Test R² = ", loyo_metrics$R2, "  |  NSE = ", loyo_metrics$NSE),
    paste0("Train R² = ", train_metrics$R2,
           "  |  Test R² = ", split_metrics$R2,
           "  |  Test NSE = ", split_metrics$NSE)
  )
)

p_ts <- ggplot(plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE, fill = "steelblue", alpha = 0.07
  ) +
  geom_vline(
    data = vline_df, aes(xintercept = xint),
    inherit.aes = FALSE, linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_line(linewidth = 0.45) +
  geom_text(
    data = label_df, aes(label = label),
    x = -Inf, y = Inf, hjust = -0.04, vjust = 1.6,
    inherit.aes = FALSE, size = 3.3, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ method, ncol = 1) +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.85)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title    = "AU-Tum: Observed vs Predicted GPP",
    subtitle = "Predictors: Tair, SW_rad, VPD, VPD_lag3, 30-day CWD  |  Shaded = training period (80/20)",
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    strip.text       = element_text(face = "bold", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# ── Plot: variable importance ─────────────────────────────────────────────────
p_imp <- ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#E06C4B", alpha = 0.85, width = 0.6) +
  coord_flip() +
  labs(
    title    = "Variable Importance (80/20 model)",
    subtitle = "%IncMSE — how much model error increases when variable is permuted",
    x        = NULL,
    y        = "% Increase in MSE"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )

ggsave("AU_Tum_timeseries.png",  plot = p_ts,  width = 14, height = 8,  dpi = 200, bg = "white")
ggsave("AU_Tum_importance.png",  plot = p_imp, width = 8,  height = 5,  dpi = 200, bg = "white")

print(p_ts)
print(p_imp)
```
Now, bounded 30 day cwd, 3 day VPD lag

Before, cwd summed daily values over 30 days.Wet days negatively cancel out dry days. So the site could have 25 dry days and 5 very wet days and would end up with a near zero 30 day cwd, even though the vegetation experienced prolonged stress.

bounded cwd runs a running total that is floored at zero

(positive cwd = demand exceeds supply (dry)
negative cwd = supply exceeds demand (wet))

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors <- c("Tair", "SW_rad", "VPD", "VPD_lag3", "cwd_bounded_30day")

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

# ── Extract AU-Tum and engineer features ──────────────────────────────────────
tum <- sites_daily[["Tum"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),

    # 3-day lagged VPD
    VPD_lag3 = lag(VPD, 3),

    # Bounded cumulative CWD — deficit accumulates but resets toward 0 on recharge
    # pmax(..., 0) ensures the running total never goes negative
    cwd_bounded = accumulate(cwd, ~ pmax(.x + .y, 0)),

    # 30-day rolling mean of the bounded cumulative deficit
    # using mean rather than sum so the scale is interpretable (avg daily deficit)
    cwd_bounded_30day = rollapply(
      cwd_bounded, width = 30, FUN = mean,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors), GPP)

formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# ── 1. LOYO Random Forest ─────────────────────────────────────────────────────
years <- unique(tum$year)

loyo_preds <- map_dfr(years, function(yr) {

  train_df <- tum %>% filter(year != yr)
  test_df  <- tum %>% filter(year == yr)

  if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

  scaled   <- standardise_train_test(train_df, test_df, predictors)
  rf_model <- randomForest(formula, data = scaled$train,
                           ntree = 500, mtry = mtry_val, importance = FALSE)
  preds    <- predict(rf_model, scaled$test)

  data.frame(
    date     = test_df$date,
    GPP_obs  = test_df$GPP,
    GPP_pred = preds,
    method   = "LOYO",
    period   = "Test (held-out year)"
  )
})

# ── 2. 80/20 Chronological Train/Test Split ───────────────────────────────────
split_idx  <- floor(0.8 * nrow(tum))
train_8020 <- tum[1:split_idx, ]
test_8020  <- tum[(split_idx + 1):nrow(tum), ]

scaled_8020      <- standardise_train_test(train_8020, test_8020, predictors)
rf_8020          <- randomForest(formula, data = scaled_8020$train,
                                 ntree = 500, mtry = mtry_val, importance = TRUE)
preds_train_8020 <- predict(rf_8020, scaled_8020$train)
preds_test_8020  <- predict(rf_8020, scaled_8020$test)

split_preds <- bind_rows(
  data.frame(
    date     = train_8020$date,
    GPP_obs  = train_8020$GPP,
    GPP_pred = preds_train_8020,
    method   = "80/20 Split",
    period   = "Train (in-sample)"
  ),
  data.frame(
    date     = test_8020$date,
    GPP_obs  = test_8020$GPP,
    GPP_pred = preds_test_8020,
    method   = "80/20 Split",
    period   = "Test (out-of-sample)"
  )
)

# ── Metrics ───────────────────────────────────────────────────────────────────
compute_metrics <- function(df) {
  obs  <- df$GPP_obs
  pred <- df$GPP_pred
  data.frame(
    R2   = round(cor(obs, pred)^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2)), 3),
    NSE  = round(1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2), 3)
  )
}

loyo_metrics  <- compute_metrics(loyo_preds)
split_metrics <- compute_metrics(split_preds %>% filter(period == "Test (out-of-sample)"))
train_metrics <- compute_metrics(split_preds %>% filter(period == "Train (in-sample)"))

cat("── LOYO metrics (all years, out-of-sample) ──\n");        print(loyo_metrics)
cat("── 80/20 Test metrics (out-of-sample) ──\n");             print(split_metrics)
cat("── 80/20 Train metrics (in-sample, for reference) ──\n"); print(train_metrics)

# ── Variable importance from 80/20 model ─────────────────────────────────────
importance_df <- data.frame(
  variable   = rownames(importance(rf_8020)),
  importance = importance(rf_8020, type = 1)[, 1]  # %IncMSE
) %>% arrange(desc(importance))

cat("\n── Variable importance (80/20 model, %IncMSE) ──\n")
print(importance_df)

# ── Quick sanity check: plot raw vs bounded CWD to verify the fix ─────────────
p_cwd <- tum %>%
  select(date, cwd, cwd_bounded) %>%
  pivot_longer(cols = c(cwd, cwd_bounded), names_to = "type", values_to = "value") %>%
  mutate(type = recode(type,
                       cwd          = "Daily CWD increment (raw)",
                       cwd_bounded  = "Bounded cumulative CWD")) %>%
  ggplot(aes(x = date, y = value, colour = type)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  facet_wrap(~ type, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("Daily CWD increment (raw)" = "#4682B4",
                                 "Bounded cumulative CWD"    = "#E06C4B")) +
  labs(
    title    = "AU-Tum: Raw vs Bounded CWD",
    subtitle = "Sanity check — bounded CWD should never go below 0",
    x = NULL, y = "CWD (mm/day)", colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# ── Plot: time series ─────────────────────────────────────────────────────────
plot_df <- bind_rows(loyo_preds, split_preds) %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(
    method = factor(method, levels = c("LOYO", "80/20 Split")),
    type   = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted")
  )

split_date <- tum$date[split_idx]

shade_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xmin = min(tum$date), xmax = split_date, ymin = -Inf, ymax = Inf
)

vline_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xint   = split_date
)

label_df <- data.frame(
  method = factor(c("LOYO", "80/20 Split"), levels = c("LOYO", "80/20 Split")),
  label  = c(
    paste0("Test R² = ", loyo_metrics$R2, "  |  NSE = ", loyo_metrics$NSE),
    paste0("Train R² = ", train_metrics$R2,
           "  |  Test R² = ", split_metrics$R2,
           "  |  Test NSE = ", split_metrics$NSE)
  )
)

p_ts <- ggplot(plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE, fill = "steelblue", alpha = 0.07
  ) +
  geom_vline(
    data = vline_df, aes(xintercept = xint),
    inherit.aes = FALSE, linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_line(linewidth = 0.45) +
  geom_text(
    data = label_df, aes(label = label),
    x = -Inf, y = Inf, hjust = -0.04, vjust = 1.6,
    inherit.aes = FALSE, size = 3.3, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ method, ncol = 1) +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.85)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title    = "AU-Tum: Observed vs Predicted GPP",
    subtitle = "Predictors: Tair, SW_rad, VPD, VPD_lag3, bounded 30-day CWD  |  Shaded = training period (80/20)",
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    strip.text       = element_text(face = "bold", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# ── Plot: variable importance ─────────────────────────────────────────────────
p_imp <- ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#E06C4B", alpha = 0.85, width = 0.6) +
  coord_flip() +
  labs(
    title    = "Variable Importance (80/20 model)",
    subtitle = "%IncMSE — how much model error increases when variable is permuted",
    x = NULL, y = "% Increase in MSE"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )

ggsave("AU_Tum_timeseries.png",  plot = p_ts,  width = 14, height = 8, dpi = 200, bg = "white")
ggsave("AU_Tum_importance.png",  plot = p_imp, width = 8,  height = 5, dpi = 200, bg = "white")
ggsave("AU_Tum_cwd_check.png",   plot = p_cwd, width = 12, height = 6, dpi = 200, bg = "white")

print(p_cwd)
print(p_ts)
print(p_imp)
```
This time - cwd is bounded and 90 days, VPD lag 3, Tair lag 3, and precipitation added back in. 
Precip captures pulse rainfall responses

Low R-squared:
- Genuine interannual variability - disturbance, insects, drought events
- Measurement uncertainty within the GPP itself
- GPP ~ climate relationship itself can shift over time, meaning training on previous years doesn't translate to future years

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)

# ── Setup ─────────────────────────────────────────────────────────────────────
predictors <- c("Tair", "SW_rad", "VPD", "VPD_lag3", "Tair_lag3",
                "Precip", "cwd_bounded_90day")

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

# ── Extract AU-Tum and engineer features ──────────────────────────────────────
tum <- sites_daily[["Tum"]] %>%
  drop_na(GPP, Tair, SW_rad, VPD, cwd, Precip, date) %>%
  arrange(date) %>%
  mutate(
    year = year(date),

    # Lagged meteorology — carry-over stomatal / heat stress
    VPD_lag3  = lag(VPD,  3),
    Tair_lag3 = lag(Tair, 3),

    # Bounded cumulative CWD — running total floored at zero
    cwd_bounded = accumulate(cwd, ~ pmax(.x + .y, 0)),

    # 90-day rolling mean of bounded CWD
    cwd_bounded_90day = rollapply(
      cwd_bounded, width = 90, FUN = mean,
      align = "right", fill = NA, na.rm = TRUE
    )
  ) %>%
  drop_na(all_of(predictors), GPP)

formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
mtry_val <- max(2, floor(sqrt(length(predictors))))

# ── 1. LOYO Random Forest ─────────────────────────────────────────────────────
years <- unique(tum$year)

loyo_preds <- map_dfr(years, function(yr) {

  train_df <- tum %>% filter(year != yr)
  test_df  <- tum %>% filter(year == yr)

  if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

  scaled   <- standardise_train_test(train_df, test_df, predictors)
  rf_model <- randomForest(formula, data = scaled$train,
                           ntree = 500, mtry = mtry_val, importance = FALSE)
  preds    <- predict(rf_model, scaled$test)

  data.frame(
    date     = test_df$date,
    GPP_obs  = test_df$GPP,
    GPP_pred = preds,
    method   = "LOYO",
    period   = "Test (held-out year)"
  )
})

# ── 2. 80/20 Chronological Train/Test Split ───────────────────────────────────
split_idx  <- floor(0.8 * nrow(tum))
train_8020 <- tum[1:split_idx, ]
test_8020  <- tum[(split_idx + 1):nrow(tum), ]

scaled_8020      <- standardise_train_test(train_8020, test_8020, predictors)
rf_8020          <- randomForest(formula, data = scaled_8020$train,
                                 ntree = 500, mtry = mtry_val, importance = TRUE)
preds_train_8020 <- predict(rf_8020, scaled_8020$train)
preds_test_8020  <- predict(rf_8020, scaled_8020$test)

split_preds <- bind_rows(
  data.frame(
    date     = train_8020$date,
    GPP_obs  = train_8020$GPP,
    GPP_pred = preds_train_8020,
    method   = "80/20 Split",
    period   = "Train (in-sample)"
  ),
  data.frame(
    date     = test_8020$date,
    GPP_obs  = test_8020$GPP,
    GPP_pred = preds_test_8020,
    method   = "80/20 Split",
    period   = "Test (out-of-sample)"
  )
)

# ── Metrics ───────────────────────────────────────────────────────────────────
compute_metrics <- function(df) {
  obs  <- df$GPP_obs
  pred <- df$GPP_pred
  data.frame(
    R2   = round(cor(obs, pred)^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2)), 3),
    NSE  = round(1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2), 3)
  )
}

loyo_metrics  <- compute_metrics(loyo_preds)
split_metrics <- compute_metrics(split_preds %>% filter(period == "Test (out-of-sample)"))
train_metrics <- compute_metrics(split_preds %>% filter(period == "Train (in-sample)"))

cat("── LOYO metrics (all years, out-of-sample) ──\n");        print(loyo_metrics)
cat("── 80/20 Test metrics (out-of-sample) ──\n");             print(split_metrics)
cat("── 80/20 Train metrics (in-sample, for reference) ──\n"); print(train_metrics)

# ── Variable importance from 80/20 model ─────────────────────────────────────
importance_df <- data.frame(
  variable   = rownames(importance(rf_8020)),
  importance = importance(rf_8020, type = 1)[, 1]  # %IncMSE
) %>% arrange(desc(importance))

cat("\n── Variable importance (80/20 model, %IncMSE) ──\n")
print(importance_df)

# ── Sanity check: raw vs bounded CWD ─────────────────────────────────────────
p_cwd <- tum %>%
  select(date, cwd, cwd_bounded) %>%
  pivot_longer(cols = c(cwd, cwd_bounded), names_to = "type", values_to = "value") %>%
  mutate(type = recode(type,
                       cwd         = "Daily CWD increment (raw)",
                       cwd_bounded = "Bounded cumulative CWD")) %>%
  ggplot(aes(x = date, y = value, colour = type)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  facet_wrap(~ type, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("Daily CWD increment (raw)" = "#4682B4",
                                 "Bounded cumulative CWD"    = "#E06C4B")) +
  labs(
    title    = "AU-Tum: Raw vs Bounded CWD",
    subtitle = "Bounded CWD should never go below 0 and should build during dry periods",
    x = NULL, y = "CWD (mm/day)", colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# ── Plot: time series ─────────────────────────────────────────────────────────
plot_df <- bind_rows(loyo_preds, split_preds) %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(
    method = factor(method, levels = c("LOYO", "80/20 Split")),
    type   = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted")
  )

split_date <- tum$date[split_idx]

shade_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xmin = min(tum$date), xmax = split_date, ymin = -Inf, ymax = Inf
)

vline_df <- data.frame(
  method = factor("80/20 Split", levels = c("LOYO", "80/20 Split")),
  xint   = split_date
)

label_df <- data.frame(
  method = factor(c("LOYO", "80/20 Split"), levels = c("LOYO", "80/20 Split")),
  label  = c(
    paste0("Test R² = ", loyo_metrics$R2, "  |  NSE = ", loyo_metrics$NSE),
    paste0("Train R² = ", train_metrics$R2,
           "  |  Test R² = ", split_metrics$R2,
           "  |  Test NSE = ", split_metrics$NSE)
  )
)

p_ts <- ggplot(plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE, fill = "steelblue", alpha = 0.07
  ) +
  geom_vline(
    data = vline_df, aes(xintercept = xint),
    inherit.aes = FALSE, linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_line(linewidth = 0.45) +
  geom_text(
    data = label_df, aes(label = label),
    x = -Inf, y = Inf, hjust = -0.04, vjust = 1.6,
    inherit.aes = FALSE, size = 3.3, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ method, ncol = 1) +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.85)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title    = "AU-Tum: Observed vs Predicted GPP",
    subtitle = "Predictors: Tair, SW_rad, VPD, VPD_lag3, Tair_lag3, Precip, bounded 90-day CWD  |  Shaded = training period (80/20)",
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    strip.text       = element_text(face = "bold", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

# ── Plot: variable importance ─────────────────────────────────────────────────
p_imp <- ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#E06C4B", alpha = 0.85, width = 0.6) +
  coord_flip() +
  labs(
    title    = "Variable Importance (80/20 model)",
    subtitle = "%IncMSE — how much model error increases when variable is permuted",
    x = NULL, y = "% Increase in MSE"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )

ggsave("AU_Tum_timeseries.png", plot = p_ts,  width = 14, height = 8, dpi = 200, bg = "white")
ggsave("AU_Tum_importance.png", plot = p_imp, width = 8,  height = 5, dpi = 200, bg = "white")
ggsave("AU_Tum_cwd_check.png",  plot = p_cwd, width = 12, height = 6, dpi = 200, bg = "white")

print(p_cwd)
print(p_ts)
print(p_imp)
```
```{r}
# How many years does each site have?
map_df(sites_daily, ~ data.frame(n_years = n_distinct(year(.$date))),
       .id = "site") %>% arrange(n_years)

# Are any of the short-record sites also the worst performing?
# Compare n_years against your metrics table
```

Now let's do all my sites:

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)
library(purrr)

set.seed(42)

# ══════════════════════════════════════════════════════════════════════════════
# SETUP
# ══════════════════════════════════════════════════════════════════════════════
predictors <- c("Tair", "SW_rad", "VPD", "VPD_lag3", "Tair_lag3",
                "Precip", "cwd_bounded_90day")

min_years <- 4  # sites with fewer years than this are excluded

standardise_train_test <- function(train_df, test_df, predictors) {
  train_means <- colMeans(train_df[predictors], na.rm = TRUE)
  train_sds   <- apply(train_df[predictors], 2, sd, na.rm = TRUE)
  train_df[predictors] <- scale(train_df[predictors], center = train_means, scale = train_sds)
  test_df[predictors]  <- scale(test_df[predictors],  center = train_means, scale = train_sds)
  list(train = train_df, test = test_df)
}

compute_metrics <- function(obs, pred) {
  data.frame(
    R2   = round(cor(obs, pred, use = "complete.obs")^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2, na.rm = TRUE)), 3),
    NSE  = round(1 - sum((obs - pred)^2, na.rm = TRUE) /
                     sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE), 3)
  )
}

# ══════════════════════════════════════════════════════════════════════════════
# FEATURE ENGINEERING — applied per site before modelling
# ══════════════════════════════════════════════════════════════════════════════
prepare_site <- function(df, site_name) {

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, cwd, Precip, date) %>%
    arrange(date) %>%
    mutate(
      year      = year(date),
      VPD_lag3  = lag(VPD,  3),
      Tair_lag3 = lag(Tair, 3),
      cwd_bounded = accumulate(cwd, ~ pmax(.x + .y, 0)),
      cwd_bounded_90day = rollapply(
        cwd_bounded, width = 90, FUN = mean,
        align = "right", fill = NA, na.rm = TRUE
      )
    ) %>%
    drop_na(all_of(predictors), GPP)

  n_yrs <- n_distinct(df$year)

  if (n_yrs < min_years) {
    message("Skipping ", site_name, ": only ", n_yrs, " years (min = ", min_years, ")")
    return(NULL)
  }

  message("Prepared ", site_name, ": ", nrow(df), " rows, ", n_yrs, " years")
  df
}

# ══════════════════════════════════════════════════════════════════════════════
# LOYO MODEL — run for a single site
# ══════════════════════════════════════════════════════════════════════════════
run_loyo <- function(df, site_name) {

  formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  mtry_val <- max(2, floor(sqrt(length(predictors))))
  years    <- unique(df$year)

  loyo_preds <- map_dfr(years, function(yr) {

    train_df <- df %>% filter(year != yr)
    test_df  <- df %>% filter(year == yr)

    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    scaled   <- standardise_train_test(train_df, test_df, predictors)
    rf_model <- randomForest(formula, data = scaled$train,
                             ntree = 500, mtry = mtry_val, importance = FALSE)
    preds    <- predict(rf_model, scaled$test)

    data.frame(
      site     = site_name,
      date     = test_df$date,
      year     = yr,
      GPP_obs  = test_df$GPP,
      GPP_pred = preds,
      VPD      = test_df$VPD
    )
  })

  if (nrow(loyo_preds) == 0) return(NULL)
  loyo_preds
}

# ══════════════════════════════════════════════════════════════════════════════
# RUN ACROSS ALL SITES
# ══════════════════════════════════════════════════════════════════════════════
prepared  <- map2(sites_daily, names(sites_daily), prepare_site)
prepared  <- Filter(Negate(is.null), prepared)

all_preds <- map2_dfr(prepared, names(prepared), run_loyo)

message("\nSites retained: ", n_distinct(all_preds$site))
message("Total rows:     ", nrow(all_preds))

# ══════════════════════════════════════════════════════════════════════════════
# METRICS
# ══════════════════════════════════════════════════════════════════════════════

# Overall per-site metrics
site_metrics <- all_preds %>%
  group_by(site) %>%
  summarise(compute_metrics(GPP_obs, GPP_pred), .groups = "drop") %>%
  mutate(n_years = map_int(prepared[site], ~ n_distinct(.x$year)))

# Per-year metrics within each site
year_metrics <- all_preds %>%
  group_by(site, year) %>%
  summarise(compute_metrics(GPP_obs, GPP_pred), .groups = "drop")

cat("\n── Per-site LOYO metrics ──\n")
print(site_metrics %>% arrange(R2))

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 1: Time series — observed vs predicted, one panel per site
# ══════════════════════════════════════════════════════════════════════════════
ts_plot_df <- all_preds %>%
  pivot_longer(cols = c(GPP_obs, GPP_pred),
               names_to = "type", values_to = "GPP") %>%
  mutate(type = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted"))

# R² label per site for facet annotation
r2_labels <- site_metrics %>%
  mutate(label = paste0("R² = ", R2, "  NSE = ", NSE))

p_ts <- ggplot(ts_plot_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
  geom_line(linewidth = 0.35) +
  geom_text(
    data        = r2_labels,
    aes(label   = label),
    x           = -Inf, y = Inf,
    hjust       = -0.05, vjust = 1.6,
    inherit.aes = FALSE,
    size        = 2.8, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ site, scales = "free_x", ncol = 3) +
  scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
  scale_alpha_manual(values  = c("Observed" = 0.4,      "Predicted" = 0.85)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "LOYO Random Forest: Observed vs Predicted GPP — All Sites",
    subtitle = paste0("Predictors: ", paste(predictors, collapse = ", ")),
    x        = NULL,
    y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    colour   = NULL, alpha = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 7)
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 2: R² vs number of years — does record length affect model fit?
# ══════════════════════════════════════════════════════════════════════════════
p_nyears <- ggplot(site_metrics, aes(x = n_years, y = R2)) +
  geom_point(aes(colour = R2), size = 3.5, alpha = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey40",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(aes(label = site), vjust = -0.9, size = 2.8, colour = "grey30") +
  scale_colour_gradient(low = "#E06C4B", high = "#4682B4") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Model Fit vs Record Length",
    subtitle = "Does more training data improve LOYO R²?",
    x        = "Number of years of data",
    y        = "LOYO R²",
    colour   = "R²"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 3: Per-year R² within each site — which years drive poor performance?
# ══════════════════════════════════════════════════════════════════════════════
p_yr_r2 <- ggplot(year_metrics, aes(x = year, y = R2, colour = R2)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_line(linewidth = 0.35, alpha = 0.5, colour = "grey60") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ site, ncol = 3, scales = "free_x") +
  scale_colour_gradient(low = "#E06C4B", high = "#4682B4") +
  scale_y_continuous(limits = c(-0.5, 1)) +
  labs(
    title    = "Per-Year LOYO R² by Site",
    subtitle = "Identifies anomalous years dragging down overall performance",
    x        = NULL, y = "R² (held-out year)", colour = "R²"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    legend.position  = "right",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 7)
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 4: Residuals vs VPD — is there systematic bias at high VPD?
# ══════════════════════════════════════════════════════════════════════════════
residual_df <- all_preds %>%
  mutate(residual = GPP_obs - GPP_pred)

p_resid_vpd <- ggplot(residual_df, aes(x = VPD, y = residual)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "#4682B4") +
  geom_smooth(method = "loess", span = 0.4, se = TRUE,
              colour = "#E06C4B", fill = "#E06C4B", alpha = 0.2, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ site, ncol = 3, scales = "free_x") +
  labs(
    title    = "Residuals vs VPD by Site",
    subtitle = "Positive residual = model underpredicts GPP; negative = overpredicts. Key check for high-VPD bias",
    x        = "VPD (kPa)",
    y        = expression(GPP[obs] - GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 5: Overall observed vs predicted scatter per site
# ══════════════════════════════════════════════════════════════════════════════
p_scatter <- ggplot(all_preds, aes(x = GPP_obs, y = GPP_pred)) +
  geom_point(alpha = 0.2, size = 0.5, colour = "#4682B4") +
  geom_abline(slope = 1, intercept = 0, colour = "#E06C4B",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(
    data        = r2_labels,
    aes(label   = paste0("R² = ", R2)),
    x           = -Inf, y = Inf,
    hjust       = -0.1, vjust = 1.6,
    inherit.aes = FALSE,
    size        = 2.8, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ site, ncol = 3, scales = "free") +
  labs(
    title    = "Observed vs Predicted GPP — All Sites (LOYO)",
    subtitle = "Dashed line = 1:1. Points above line = underprediction; below = overprediction",
    x        = expression(GPP[obs] ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    y        = expression(GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

# ══════════════════════════════════════════════════════════════════════════════
# SAVE ALL PLOTS
# ══════════════════════════════════════════════════════════════════════════════
n_sites  <- n_distinct(all_preds$site)
ts_h     <- max(8, ceiling(n_sites / 3) * 3.5)
diag_h   <- max(8, ceiling(n_sites / 3) * 3.0)

ggsave("LOYO_timeseries_all_sites.png",  plot = p_ts,        width = 16, height = ts_h,   dpi = 200, bg = "white")
ggsave("LOYO_R2_vs_nyears.png",          plot = p_nyears,    width = 9,  height = 6,      dpi = 200, bg = "white")
ggsave("LOYO_peryear_R2.png",            plot = p_yr_r2,     width = 16, height = diag_h, dpi = 200, bg = "white")
ggsave("LOYO_residuals_vs_VPD.png",      plot = p_resid_vpd, width = 16, height = diag_h, dpi = 200, bg = "white")
ggsave("LOYO_obs_vs_pred_scatter.png",   plot = p_scatter,   width = 16, height = diag_h, dpi = 200, bg = "white")

message("All plots saved.")

print(p_ts)
print(p_nyears)
print(p_yr_r2)
print(p_resid_vpd)
print(p_scatter)
```
Could i look at years with a low R-squared and see if that correlates with drought etc?

```{r}
dir.create("site_timeseries", showWarnings = FALSE)
 
walk(unique(all_preds$site), function(s) {
 
  site_df <- all_preds %>%
    filter(site == s) %>%
    pivot_longer(cols = c(GPP_obs, GPP_pred),
                 names_to = "type", values_to = "GPP") %>%
    mutate(type = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted"))
 
  m <- site_metrics %>% filter(site == s)
  label <- paste0("R² = ", m$R2, "  |  NSE = ", m$NSE, "  |  RMSE = ", m$RMSE)
 
  p <- ggplot(site_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
    geom_line(linewidth = 0.5) +
    annotate("text", x = min(site_df$date), y = Inf,
             label = label, hjust = -0.05, vjust = 1.6,
             size = 3.5, colour = "grey20", fontface = "bold") +
    scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
    scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.9)) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(
      title    = paste0(s, ": Observed vs Predicted GPP (LOYO)"),
      subtitle = paste0("Predictors: ", paste(predictors, collapse = ", ")),
      x        = NULL,
      y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
      colour   = NULL, alpha = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey40", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1)
    )
 
  ggsave(paste0("site_timeseries/", s, "_timeseries.png"),
         plot = p, width = 12, height = 5, dpi = 200, bg = "white")
})
 
message("Time series plots saved to site_timeseries/")
 
# ══════════════════════════════════════════════════════════════════════════════
# PLOT 2: Per-site per-year R² — one PNG per site
# ══════════════════════════════════════════════════════════════════════════════
dir.create("site_year_r2", showWarnings = FALSE)
 
walk(unique(year_metrics$site), function(s) {
 
  yr_df <- year_metrics %>% filter(site == s)
  overall_r2 <- site_metrics %>% filter(site == s) %>% pull(R2)
 
  p <- ggplot(yr_df, aes(x = year, y = R2, fill = R2)) +
    geom_col(width = 0.7, alpha = 0.9) +
    geom_hline(yintercept = overall_r2, linetype = "dashed",
               colour = "grey30", linewidth = 0.7) +
    geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
    annotate("text",
             x    = max(yr_df$year),
             y    = overall_r2 + 0.03,
             label = paste0("Overall R² = ", overall_r2),
             hjust = 1, size = 3.3, colour = "grey30", fontface = "italic") +
    geom_text(aes(label = R2,
                  vjust = ifelse(R2 >= 0, -0.4, 1.3)),
              size = 3, colour = "grey20") +
    scale_fill_gradient(low = "#E06C4B", high = "#4682B4", limits = c(-0.5, 1),
                        oob = scales::squish) +
    scale_x_continuous(breaks = yr_df$year) +
    scale_y_continuous(limits = c(min(-0.1, min(yr_df$R2) - 0.1), 1.05)) +
    labs(
      title    = paste0(s, ": Per-Year LOYO R²"),
      subtitle = "Each bar = R² for that held-out year. Dashed line = overall site R²",
      x        = NULL,
      y        = "R²",
      fill     = "R²"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey40", size = 10),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1),
      legend.position  = "none"
    )
 
  ggsave(paste0("site_year_r2/", s, "_year_r2.png"),
         plot = p, width = 10, height = 5, dpi = 200, bg = "white")
})
 
message("Per-year R² plots saved to site_year_r2/")
 
# ══════════════════════════════════════════════════════════════════════════════
# PLOT 3: R² vs number of years (all sites on one plot — stays combined)
# ══════════════════════════════════════════════════════════════════════════════
p_nyears <- ggplot(site_metrics, aes(x = n_years, y = R2)) +
  geom_point(aes(colour = R2), size = 3.5, alpha = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey40",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(aes(label = site), vjust = -0.9, size = 2.8, colour = "grey30") +
  scale_colour_gradient(low = "#E06C4B", high = "#4682B4") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Model Fit vs Record Length",
    subtitle = "Does more training data improve LOYO R²?",
    x        = "Number of years of data",
    y        = "LOYO R²",
    colour   = "R²"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )
 
# ══════════════════════════════════════════════════════════════════════════════
# PLOT 4: Residuals vs VPD (all sites combined, faceted)
# ══════════════════════════════════════════════════════════════════════════════
residual_df <- all_preds %>%
  mutate(residual = GPP_obs - GPP_pred)
 
p_resid_vpd <- ggplot(residual_df, aes(x = VPD, y = residual)) +
  geom_point(alpha = 0.12, size = 0.5, colour = "#4682B4") +
  geom_smooth(method = "loess", span = 0.4, se = TRUE,
              colour = "#E06C4B", fill = "#E06C4B", alpha = 0.2, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ site, ncol = 3, scales = "free_x") +
  labs(
    title    = "Residuals vs VPD by Site",
    subtitle = "Positive = model underpredicts GPP; negative = overpredicts. Key check for high-VPD bias",
    x        = "VPD (kPa)",
    y        = expression(GPP[obs] - GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )
 
# ══════════════════════════════════════════════════════════════════════════════
# PLOT 5: Obs vs pred scatter (all sites combined, faceted)
# ══════════════════════════════════════════════════════════════════════════════
p_scatter <- ggplot(all_preds, aes(x = GPP_obs, y = GPP_pred)) +
  geom_point(alpha = 0.2, size = 0.5, colour = "#4682B4") +
  geom_abline(slope = 1, intercept = 0, colour = "#E06C4B",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(
    data        = site_metrics,
    aes(label   = paste0("R² = ", R2)),
    x           = -Inf, y = Inf,
    hjust       = -0.1, vjust = 1.6,
    inherit.aes = FALSE,
    size        = 2.8, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ site, ncol = 3, scales = "free") +
  labs(
    title    = "Observed vs Predicted GPP — All Sites (LOYO)",
    subtitle = "Dashed line = 1:1. Points above = underprediction; below = overprediction",
    x        = expression(GPP[obs] ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    y        = expression(GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
 
# ══════════════════════════════════════════════════════════════════════════════
# SAVE COMBINED PLOTS
# ══════════════════════════════════════════════════════════════════════════════
n_sites <- n_distinct(all_preds$site)
diag_h  <- max(8, ceiling(n_sites / 3) * 3.2)
 
ggsave("LOYO_R2_vs_nyears.png",        plot = p_nyears,    width = 9,  height = 6,      dpi = 200, bg = "white")
ggsave("LOYO_residuals_vs_VPD.png",    plot = p_resid_vpd, width = 16, height = diag_h, dpi = 200, bg = "white")
ggsave("LOYO_obs_vs_pred_scatter.png", plot = p_scatter,   width = 16, height = diag_h, dpi = 200, bg = "white")
 
message("All combined plots saved.")
```

```{r}
map_df(sites_daily, function(df) {
  summarise(df, across(c(GPP, Tair, SW_rad, VPD, cwd, Precip),
                       ~ mean(is.na(.)) * 100, .names = "{.col}_pct_NA"))
}, .id = "site")
```

```{r}
# Check if cwd gaps are from ET or Rn specifically
sites_daily[["Asm"]] %>%
  summarise(
    ET_pct_NA  = mean(is.na(ET))  * 100,
    Rn_pct_NA  = mean(is.na(Rn))  * 100,
    cwd_pct_NA = mean(is.na(cwd)) * 100
  )
```

A lot of missingness is coming from cwd due to Radiation (Rn) being missing 

A few options:
- Remove cwd and replace with precip
- Calculate cwd using something else entirely?

Here, I have removed cwd entirely and replaced it with a rolling sum of precipitation (90 days) as well as normal precip

```{r}
library(tidyverse)
library(randomForest)
library(zoo)
library(lubridate)
library(purrr)

set.seed(42)

# ══════════════════════════════════════════════════════════════════════════════
# SETUP
# ══════════════════════════════════════════════════════════════════════════════

# cwd_bounded_90day replaced with precip_roll90 — Precip has 0% missingness
# across all sites, avoiding the large data gaps caused by missing Rn/ET
predictors <- c("Tair", "SW_rad", "VPD",
                "Precip", "precip_roll90")

min_years <- 2

compute_metrics <- function(obs, pred) {
  data.frame(
    R2   = round(cor(obs, pred, use = "complete.obs")^2, 3),
    RMSE = round(sqrt(mean((obs - pred)^2, na.rm = TRUE)), 3),
    NSE  = round(1 - sum((obs - pred)^2, na.rm = TRUE) /
                     sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE), 3)
  )
}

# ══════════════════════════════════════════════════════════════════════════════
# FEATURE ENGINEERING
# ══════════════════════════════════════════════════════════════════════════════
prepare_site <- function(df, site_name) {

  df <- df %>%
    drop_na(GPP, Tair, SW_rad, VPD, Precip, date) %>%  # cwd removed from drop_na
    arrange(date) %>%
    mutate(
      year      = year(date),
      VPD_lag3  = lag(VPD,  3),
      Tair_lag3 = lag(Tair, 3),

      # 90-day rolling sum of precipitation — antecedent water availability proxy
      # sum rather than mean so it represents total water input over the window
      precip_roll90 = rollapply(
        Precip, width = 90, FUN = sum,
        align = "right", fill = NA, na.rm = TRUE
      )
    ) %>%
    drop_na(all_of(predictors), GPP)

  n_yrs <- n_distinct(df$year)

  if (n_yrs < min_years) {
    message("Skipping ", site_name, ": only ", n_yrs, " years (min = ", min_years, ")")
    return(NULL)
  }

  message("Prepared ", site_name, ": ", nrow(df), " rows, ", n_yrs, " years")
  df
}

# ══════════════════════════════════════════════════════════════════════════════
# LOYO MODEL
# ══════════════════════════════════════════════════════════════════════════════
run_loyo <- function(df, site_name) {

  formula  <- as.formula(paste("GPP ~", paste(predictors, collapse = " + ")))
  mtry_val <- max(2, floor(sqrt(length(predictors))))
  years    <- unique(df$year)

  loyo_preds <- map_dfr(years, function(yr) {

    train_df <- df %>% filter(year != yr)
    test_df  <- df %>% filter(year == yr)

    if (nrow(train_df) < 50 || nrow(test_df) < 10) return(NULL)

    # No standardisation needed — RF is scale-invariant
    rf_model <- randomForest(formula, data = train_df,
                             ntree = 500, mtry = mtry_val, importance = FALSE)
    preds    <- predict(rf_model, test_df)

    data.frame(
      site     = site_name,
      date     = test_df$date,
      year     = yr,
      GPP_obs  = test_df$GPP,
      GPP_pred = preds,
      VPD      = test_df$VPD
    )
  })

  if (nrow(loyo_preds) == 0) return(NULL)
  loyo_preds
}

# ══════════════════════════════════════════════════════════════════════════════
# RUN ACROSS ALL SITES
# ══════════════════════════════════════════════════════════════════════════════
prepared  <- map2(sites_daily, names(sites_daily), prepare_site)
prepared  <- Filter(Negate(is.null), prepared)

all_preds <- map2_dfr(prepared, names(prepared), run_loyo)

message("\nSites retained: ", n_distinct(all_preds$site))
message("Total rows:     ", nrow(all_preds))

# ══════════════════════════════════════════════════════════════════════════════
# METRICS
# ══════════════════════════════════════════════════════════════════════════════
site_metrics <- all_preds %>%
  group_by(site) %>%
  summarise(compute_metrics(GPP_obs, GPP_pred), .groups = "drop") %>%
  mutate(n_years = map_int(prepared[site], ~ n_distinct(.x$year)))

year_metrics <- all_preds %>%
  group_by(site, year) %>%
  summarise(compute_metrics(GPP_obs, GPP_pred), .groups = "drop")

cat("\n── Per-site LOYO metrics ──\n")
print(site_metrics %>% arrange(R2))

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 1: Per-site time series — one PNG per site
# ══════════════════════════════════════════════════════════════════════════════
dir.create("site_timeseries", showWarnings = FALSE)

walk(unique(all_preds$site), function(s) {

  site_df <- all_preds %>%
    filter(site == s) %>%
    pivot_longer(cols = c(GPP_obs, GPP_pred),
                 names_to = "type", values_to = "GPP") %>%
    mutate(type = recode(type, GPP_obs = "Observed", GPP_pred = "Predicted"))

  m     <- site_metrics %>% filter(site == s)
  label <- paste0("R² = ", m$R2, "  |  NSE = ", m$NSE, "  |  RMSE = ", m$RMSE)

  p <- ggplot(site_df, aes(x = date, y = GPP, colour = type, alpha = type)) +
    geom_line(linewidth = 0.5) +
    annotate("text",
             x     = min(site_df$date), y = Inf,
             label = label,
             hjust = -0.05, vjust = 1.6,
             size  = 3.5, colour = "grey20", fontface = "bold") +
    scale_colour_manual(values = c("Observed" = "#2c2c2c", "Predicted" = "#E06C4B")) +
    scale_alpha_manual(values  = c("Observed" = 0.45,     "Predicted" = 0.9)) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(
      title    = paste0(s, ": Observed vs Predicted GPP (LOYO)"),
      subtitle = paste0("Predictors: ", paste(predictors, collapse = ", ")),
      x        = NULL,
      y        = expression(GPP ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
      colour   = NULL, alpha = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey40", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1)
    )

  ggsave(paste0("site_timeseries/", s, "_timeseries.png"),
         plot = p, width = 12, height = 5, dpi = 200, bg = "white")
})

message("Time series plots saved to site_timeseries/")

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 2: Per-site per-year R² — one PNG per site
# ══════════════════════════════════════════════════════════════════════════════
dir.create("site_year_r2", showWarnings = FALSE)

walk(unique(year_metrics$site), function(s) {

  yr_df      <- year_metrics %>% filter(site == s)
  overall_r2 <- site_metrics %>% filter(site == s) %>% pull(R2)

  p <- ggplot(yr_df, aes(x = year, y = R2, fill = R2)) +
    geom_col(width = 0.7, alpha = 0.9) +
    geom_hline(yintercept = overall_r2, linetype = "dashed",
               colour = "grey30", linewidth = 0.7) +
    geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
    annotate("text",
             x     = max(yr_df$year),
             y     = overall_r2 + 0.03,
             label = paste0("Overall R² = ", overall_r2),
             hjust = 1, size = 3.3, colour = "grey30", fontface = "italic") +
    geom_text(aes(label = R2,
                  vjust = ifelse(R2 >= 0, -0.4, 1.3)),
              size = 3, colour = "grey20") +
    scale_fill_gradient(low = "#E06C4B", high = "#4682B4",
                        limits = c(-0.5, 1), oob = scales::squish) +
    scale_x_continuous(breaks = yr_df$year) +
    scale_y_continuous(limits = c(min(-0.1, min(yr_df$R2) - 0.1), 1.05)) +
    labs(
      title    = paste0(s, ": Per-Year LOYO R²"),
      subtitle = "Each bar = R² for that held-out year. Dashed line = overall site R²",
      x        = NULL, y = "R²", fill = "R²"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title         = element_text(face = "bold", size = 13),
      plot.subtitle      = element_text(colour = "grey40", size = 10),
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.text.x        = element_text(angle = 45, hjust = 1),
      legend.position    = "none"
    )

  ggsave(paste0("site_year_r2/", s, "_year_r2.png"),
         plot = p, width = 10, height = 5, dpi = 200, bg = "white")
})

message("Per-year R² plots saved to site_year_r2/")

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 3: R² vs number of years (combined)
# ══════════════════════════════════════════════════════════════════════════════
p_nyears <- ggplot(site_metrics, aes(x = n_years, y = R2)) +
  geom_point(aes(colour = R2), size = 3.5, alpha = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey40",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(aes(label = site), vjust = -0.9, size = 2.8, colour = "grey30") +
  scale_colour_gradient(low = "#E06C4B", high = "#4682B4") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Model Fit vs Record Length",
    subtitle = "Does more training data improve LOYO R²?",
    x        = "Number of years of data",
    y        = "LOYO R²",
    colour   = "R²"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 4: Residuals vs VPD (combined, faceted)
# ══════════════════════════════════════════════════════════════════════════════
residual_df <- all_preds %>%
  mutate(residual = GPP_obs - GPP_pred)

p_resid_vpd <- ggplot(residual_df, aes(x = VPD, y = residual)) +
  geom_point(alpha = 0.12, size = 0.5, colour = "#4682B4") +
  geom_smooth(method = "loess", span = 0.4, se = TRUE,
              colour = "#E06C4B", fill = "#E06C4B", alpha = 0.2, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ site, ncol = 3, scales = "free_x") +
  labs(
    title    = "Residuals vs VPD by Site",
    subtitle = "Positive = model underpredicts GPP; negative = overpredicts. Key check for high-VPD bias",
    x        = "VPD (kPa)",
    y        = expression(GPP[obs] - GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

# ══════════════════════════════════════════════════════════════════════════════
# PLOT 5: Obs vs pred scatter (combined, faceted)
# ══════════════════════════════════════════════════════════════════════════════
r2_labels <- site_metrics %>%
  mutate(label = paste0("R² = ", R2, "  NSE = ", NSE))

p_scatter <- ggplot(all_preds, aes(x = GPP_obs, y = GPP_pred)) +
  geom_point(alpha = 0.2, size = 0.5, colour = "#4682B4") +
  geom_abline(slope = 1, intercept = 0, colour = "#E06C4B",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(
    data        = r2_labels,
    aes(label   = paste0("R² = ", R2)),
    x           = -Inf, y = Inf,
    hjust       = -0.1, vjust = 1.6,
    inherit.aes = FALSE,
    size        = 2.8, colour = "grey20", fontface = "bold"
  ) +
  facet_wrap(~ site, ncol = 3, scales = "free") +
  labs(
    title    = "Observed vs Predicted GPP — All Sites (LOYO)",
    subtitle = "Dashed line = 1:1. Points above = underprediction; below = overprediction",
    x        = expression(GPP[obs] ~ (µmol ~ C ~ m^{-2} ~ d^{-1})),
    y        = expression(GPP[pred] ~ (µmol ~ C ~ m^{-2} ~ d^{-1}))
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 9),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

# ══════════════════════════════════════════════════════════════════════════════
# SAVE COMBINED PLOTS
# ══════════════════════════════════════════════════════════════════════════════
n_sites <- n_distinct(all_preds$site)
diag_h  <- max(8, ceiling(n_sites / 3) * 3.2)

ggsave("LOYO_R2_vs_nyears.png",        plot = p_nyears,    width = 9,  height = 6,      dpi = 200, bg = "white")
ggsave("LOYO_residuals_vs_VPD.png",    plot = p_resid_vpd, width = 16, height = diag_h, dpi = 200, bg = "white")
ggsave("LOYO_obs_vs_pred_scatter.png", plot = p_scatter,   width = 16, height = diag_h, dpi = 200, bg = "white")

message("All plots saved.")
```

