---
title: "explore_ozflux_02"
output: html_document
date: "2026-01-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Thu 29th Jan:

Going to look at the North Australian Tropical Transect flux sites and identifying a drought/VPD event. 

1. Howard Springs = AU-How (2001-2014) 13 years
2. Daly River Pasture = AU-DaP (2007-2013) 6 years
3. Daly River Savanna = AU-DaS (2008-2014) 6 years
4. Dry River = AU-Dry (2008-2014) 6 years
5. Sturt Plains = AU-Stp (2008-2014) 6 years
6. Alice Springs = AU-ASM (2010-2014) 4 years

These flux sites are located along a precipitation gradient. This rainfall gradient is a 'natural laboratory' with Howard Springs as the wettest and Alice Springs the driest. 

1. Read in flux files:

```{r}
library(tidyverse)

read_flux <- function(fname) {
  
  #Ensure it reads one file at a time
  if (length(fname) != 1) {
    stop("read_flux() expects a single filename")
  }
  
  df <- read.csv(fname)
  
  #extract site code from file name
  base <- basename(fname)
  no_ext <- gsub("\\.csv$", "", base)
  
 site <- regmatches(
  no_ext,
  regexpr("[A-Z]{2}-[A-Za-z]{2,3}[0-9]?", no_ext)
)
 
 if(length(site) == 0) {
   stop(paste("Could not extract site code from filename:", base))
 }
 
  #add site column
  df$site <- site
  
   # Convert TIMESTAMP_START to datetime and set as index
  df <- df %>%
    mutate(date = ymd_hm(TIMESTAMP_START)) %>%
    select(-TIMESTAMP_START, -TIMESTAMP_END) %>%
    relocate(date)

  df <- df %>%
    mutate(
      year = year(date),
      doy = yday(date),
      hod = hour(date)
    )
  
#renaming some variables
  
   df <- df %>%
    rename(
      GPP = GPP_NT_VUT_REF,
      LE = LE_F_MDS,
      Precip = P_ERA,
      Tair = TA_F_MDS,
      SW_rad = SW_IN_F_MDS,
      Rn = NETRAD,
      Pa = PA_F,
      VPD = VPD_F_MDS,
      Rn = NETRAD
    )

  vars_to_keep <- c("GPP", "Tair", "SW_rad", "Rn", "VPD", "Precip", "Pa", "Rn", "LE")
  df <- df %>% select(date, all_of(vars_to_keep), year, doy, hod, site)
  
  return(df)
  
}

```

```{r}
AU_How <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-How_FLUXNET2015_FULLSET_HH_2001-2014_1-4.csv") 

AU_Dap <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaP_FLUXNET2015_FULLSET_HH_2007-2013_2-4.csv")

AU_Das <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-DaS_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Dry <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Dry_FLUXNET2015_FULLSET_HH_2008-2014_2-4.csv")

AU_Stp <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-Stp_FLUXNET2015_FULLSET_HH_2008-2014_1-4.csv")

AU_Asm <- read_flux("/Users/alicerolfe/year_4_project/fluxnet_project/FLUX/halfhourly/FLX_AU-ASM_FLUXNET2015_FULLSET_HH_2010-2014_2-4.csv")
```

2. -9999 -> NA

```{r}
flux_vars <- c("GPP", "LE", "Tair", "SW_rad", "Rn", "Precip", "Pa")

AU_How_1 <- AU_How %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Dap_1 <- AU_Dap %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Das_1 <- AU_Das %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Dry_1 <- AU_Dry %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))


AU_Stp_1 <- AU_Stp %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))

AU_Asm_1 <- AU_Asm %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999)
  )) %>%
  mutate(across(
    all_of(flux_vars),
    ~na_if(., -9999.9)
  ))
```


3. Half-hourly to daily 

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = LE * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
      Precip = Precip * SEC_2_HALF_HOUR
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(site, date) %>%
    summarise(
      Precip = ifelse(
        sum(!is.na(Precip)) > 0,
        sum(Precip, na.rm = TRUE),
        NA
      ),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      LE = mean(LE, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(df)
  
}
```

```{r}
AU_How_daily <- convert_to_daily(AU_How_1)
AU_Dap_daily <- convert_to_daily(AU_Dap_1)
AU_Das_daily <- convert_to_daily(AU_Das_1)
AU_Dry_daily <- convert_to_daily(AU_Dry_1)
AU_Stp_daily <- convert_to_daily(AU_Stp_1)
AU_Asm_daily <- convert_to_daily(AU_Asm_1)
```
Convert NaN back to NA

```{r}
AU_How_daily2 <- AU_How_daily %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

AU_Dap_daily2 <- AU_Dap_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

AU_Das_daily2 <- AU_Das_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

AU_Dry_daily2 <- AU_Dry_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

AU_Stp_daily2 <- AU_Stp_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

AU_Asm_daily2 <- AU_Asm_daily %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), NA, .)))

```

4. Daily to monthly


```{r}
#Avoid months built from 1-2 days of data

monthly_flux <- function(df,
                         min_days_flux = 15,
                         min_days_precip = 1) {
  df %>%
    mutate(
      year = lubridate::year(date),
      month = lubridate::month(date),
    ) %>%
    
    group_by(site, year, month) %>%
    summarise(
      n_days = n_distinct(date[!is.na(GPP)]),
      
      # ---- Fluxes / states (means) ----
      GPP = ifelse(
        sum(!is.na(GPP)) >= min_days_flux,
        mean(GPP, na.rm = TRUE),
        NA_real_
      ),
      
      Tair = ifelse(
        sum(!is.na(Tair)) >= min_days_flux,
        mean(Tair, na.rm = TRUE),
        NA_real_
      ),
      
      ET = ifelse(
        sum(!is.na(ET)) >=min_days_flux,
        mean(ET, na.rm = TRUE),
        NA_real_
      ),
      
      VPD = ifelse(
        sum(!is.na(VPD)) >= min_days_flux,
        mean(VPD, na.rm = TRUE),
        NA_real_
      ),
      
      SW_rad = ifelse(
        sum(!is.na(SW_rad)) >= min_days_flux,
        mean(SW_rad, na.rm = TRUE),
        NA_real_
      ),
      
      Rn = ifelse(
        sum(!is.na(Rn)) >= min_days_flux,
        mean(Rn, na.rm = TRUE),
        NA_real_
      ),
      
      Qle = ifelse(
        sum(!is.na(Qle)) >= min_days_flux,
        mean(Qle, na.rm = TRUE),
        NA_real_
      ),
      
      Pa = ifelse(
        sum(!is.na(Pa)) >= min_days_flux,
        mean(Pa, na.rm = TRUE),
        NA_real_
      ),
      
      # ---- Precip (sum) ----
      Precip = ifelse(
        sum(!is.na(Precip)) >= min_days_precip,
        sum(Precip, na.rm = TRUE),
        NA_real_
      ),
      
      .groups = "drop"
    )
}
  
```


```{r}
How_monthly <- monthly_flux(AU_How_daily2)
Dap_monthly <- monthly_flux(AU_Dap_daily2)
Das_monthly <- monthly_flux(AU_Das_daily2)
Dry_monthly <- monthly_flux(AU_Dry_daily2)
Stp_monthly <- monthly_flux(AU_Stp_daily2)
Asm_monthly <- monthly_flux(AU_Asm_daily2)
```


5. Selective NA removal

Drop NAs

```{r}
How_monthly <- How_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)

Dap_monthly <- Dap_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)

Das_monthly <- Das_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)

Dry_monthly <- Dry_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)

Stp_monthly <- Stp_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)

Asm_monthly <- Asm_monthly %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, Qle, SW_rad)
```

Sanity check

```{r}
# Check missingness
colSums(is.na(How_monthly))

# Plot to eyeball drought signals
ggplot(How_monthly, aes(year + month/12, GPP)) +
  geom_line() +
  labs(title = "Monthly GPP – AU-How")

```


6. Growing season

```{r}
filter_gs <- function(df, rad_thresh = 100) {
  df %>%
    filter(SW_rad > rad_thresh, GPP > 0)
}
```

```{r}
How_final <- filter_gs(How_monthly)
Dap_final <- filter_gs(Dap_monthly)
Das_final <- filter_gs(Das_monthly)
Dry_final <- filter_gs(Dry_monthly)
Stp_final <- filter_gs(Stp_monthly)
Asm_final <- filter_gs(Asm_monthly)

```

Now, we have monthly, growing season, cleaned flux data. 
- last time, I used half-hourly data that was not cleaned in this way, let's see if this improves/decreases the predictability of GPP...
- Last time, I found that for drier/arid sites, lagged climate variables and the addition of cwd/pet improved R-squared
- However, for wetter sites, this lag effect and soil water estimation actually decreased the fit of the model

What if I included SW_rad, as well as lagged climate effects and water estimation?

1. Adding water stress component:

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
```

```{r}
priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}
```

Alice Springs: ASM

```{r}
Asm_final$pet <- priestley_taylor_pet(Asm_final$Rn, Asm_final$Tair, Asm_final$Pa)
Asm_final$cwd <- Asm_final$pet - Asm_final$Qle
```

Sturt Plains: Stp

```{r}
Stp_final$pet <- priestley_taylor_pet(Stp_final$Rn, Stp_final$Tair, Stp_final$Pa)
Stp_final$cwd <- Stp_final$pet - Stp_final$Qle
```

Dry River: Dry

```{r}
Dry_final$pet <- priestley_taylor_pet(Dry_final$Rn, Dry_final$Tair, Dry_final$Pa)
Dry_final$cwd <- Dry_final$pet - Dry_final$Qle
```

Daly River Savanna: DaS

```{r}
Das_final$pet <- priestley_taylor_pet(Das_final$Rn, Das_final$Tair, Das_final$Pa)
Das_final$cwd <- Das_final$pet - Das_final$Qle
```

Daly River Pasture: DaP

```{r}
Dap_final$pet <- priestley_taylor_pet(Dap_final$Rn, Dap_final$Tair, Dap_final$Pa)
Dap_final$cwd <- Dap_final$pet - Dap_final$Qle
```

Howard Springs: How

```{r}
How_final$pet <- priestley_taylor_pet(How_final$Rn, How_final$Tair, How_final$Pa)
How_final$cwd <- How_final$pet - How_final$Qle
```

Split by training and testing:

```{r}
idx <- createDataPartition(Asm_final$GPP, p=0.8, list=FALSE)
train_data <- Asm_final[idx, ]
test_data <- Asm_final[-idx, ]
```


```{r}

#Drop any NAs after this

train_data <- train_data %>% drop_na()
test_data <- test_data %>% drop_na()
```

Fit model

```{r}
model <- gam(GPP ~ s(Tair) + s(VPD) + s(SW_rad) + s(cwd), data=train_data, 
                 method = "REML")

predictions <- predict(model, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")


```

```{r}
summary(model)
```

Predictability of GPP increases with monthly data, and variables Tair, VPD, SW_rad and cwd. 

Cumulative precipitation running total
- Compare each year to this and diagnose the drought by the lower anomaly period

```{r}
AU_How_ordered <- AU_How_final %>%
  arrange(site, year, month)
```

```{r}
AU_How_cum <- AU_How_ordered %>%
  group_by(site, year) %>%
  mutate(
    precip_cumsum = cumsum(Precip)
  ) %>%
  ungroup()

```

```{r}
ggplot(AU_How_cum,
       aes(x = month,
           y = precip_cumsum,
           group = year,
           colour = factor(year))) +
  geom_line() +
  labs(
    title = "Cumulative precipitation by year – AU-How",
    y = "Cumulative precipitation (mm)",
    colour = "Year"
  ) +
  theme_minimal()

```

```{r}
annual_precip <- AU_How_cum %>%
  group_by(site, year) %>%
  summarise(
    annual_precip = max(precip_cumsum, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
threshold <- quantile(annual_precip$annual_precip, 0.20, na.rm = TRUE)

drought_years <- annual_precip %>%
  mutate(
    drought_year = annual_precip < threshold
  )

```

```{r}
AU_How_labeled <- AU_How_cum %>%
  left_join(drought_years, by = c("site", "year"))

```

```{r}
ggplot(AU_How_labeled,
       aes(year + month/12, GPP)) +
  geom_line() +
  geom_point(aes(colour = drought_year)) +
  scale_colour_manual(values = c("black", "red")) +
  labs(title = "GPP response during low-precipitation years")

```
Drought period in 2002 and 2007. 

What if we trained the model on the drought in 2002 and tested it on the drought in 2007?

Filter out FALSE drought years

```{r}
How_drought <- AU_How_labeled %>%
  filter(drought_year == TRUE)
```


```{r}
train_data <- How_drought %>% filter(year == 2002)
test_data <- How_drought %>% filter(year == 2007)
```

```{r}

model <- gam(GPP ~ s(Tair, k = 4) + s(VPD, k = 4) + s(SW_rad, k = 4), data=train_data, 
                 method = "REML")

predictions <- predict(model, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")


```

```{r}
sapply(train_data[, c("VPD", "Tair", "Precip", "SW_rad")],
       function(x) length(unique(x)))

```

Now, what if i trained the GAM on a drought event in one site and then tested in on another site?
- Do sites with similar precipitation regimes/ecosystems/PFTs show similar drought responses?

Train data:
Howard Springs - Seasonally wet, Dominant Species = Eucalyptus minaita and Eucalyptus tetrodonta woody savanna

Test data:
Daly Uncleared - Seasonally wet, Dominant Species = Eucalyptus tetrodonta and Corymbia terminalis, Savanna, open forest

Alice Springs - Very dry, Dominant species = Acacia aneura, Acacia shrublands

Cumulative running precipitation for AU_DaS and AU_Asm

```{r}
AU_Asm_ordered <- Asm_final %>%
  arrange(site, year, month)
```

```{r}
AU_Asm_cum <- AU_Asm_ordered %>%
  group_by(site, year) %>%
  mutate(
    precip_cumsum = cumsum(Precip)
  ) %>%
  ungroup()
```


```{r}
ggplot(AU_Asm_cum,
       aes(x = month,
           y = precip_cumsum,
           group = year,
           colour = factor(year))) +
  geom_line() +
  labs(
    title = "Cumulative precipitation by year – AU-How",
    y = "Cumulative precipitation (mm)",
    colour = "Year"
  ) +
  theme_minimal()
```
```{r}
annual_precip_2 <- AU_Asm_cum %>%
  group_by(site, year) %>%
  summarise(
    annual_precip_2 = max(precip_cumsum, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
threshold <- quantile(annual_precip_2$annual_precip_2, 0.20, na.rm = TRUE)

drought_years <- annual_precip_2 %>%
  mutate(
    drought_year = annual_precip_2 < threshold
  )

```

```{r}
AU_Asm_labeled <- AU_Asm_cum %>%
  left_join(drought_years, by = c("site", "year"))
```

```{r}
ggplot(AU_Asm_labeled,
       aes(year + month/12, GPP)) +
  geom_line() +
  geom_point(aes(colour = drought_year)) +
  scale_colour_manual(values = c("black", "red")) +
  labs(title = "GPP response during low-precipitation years")
```

```{r}
Asm_drought <- AU_Asm_labeled %>%
  filter(drought_year == TRUE)
```

Let's train on Asm_drought and test on How_drought

```{r}
train_data <- Asm_drought 
test_data <- How_drought
```

```{r}

model <- gam(GPP ~ s(Tair, k = 4) + s(VPD, k = 4) + s(SW_rad, k = 4), data=train_data, 
                 method = "REML")

predictions <- predict(model, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

Not great at predicting drought responses in different ecosystem types. 

Let's try two different sites that are similar climate/PFT

- Create running precipitation for AU_DaS

```{r}
AU_Das_ordered <- Das_final %>%
  arrange(site, year, month)
```

```{r}
AU_Das_cum <- AU_Das_ordered %>%
  group_by(site, year) %>%
  mutate(
    precip_cumsum = cumsum(Precip)
  ) %>%
  ungroup()
```


```{r}
ggplot(AU_Das_cum,
       aes(x = month,
           y = precip_cumsum,
           group = year,
           colour = factor(year))) +
  geom_line() +
  labs(
    title = "Cumulative precipitation by year – AU-Das",
    y = "Cumulative precipitation (mm)",
    colour = "Year"
  ) +
  theme_minimal()
```


```{r}
annual_precip_3 <- AU_Das_cum %>%
  group_by(site, year) %>%
  summarise(
    annual_precip_3 = max(precip_cumsum, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
threshold <- quantile(annual_precip_3$annual_precip_3, 0.20, na.rm = TRUE)

drought_years <- annual_precip_3 %>%
  mutate(
    drought_year = annual_precip_3 < threshold
  )

```

```{r}
AU_Das_labeled <- AU_Das_cum %>%
  left_join(drought_years, by = c("site", "year"))
```

```{r}
ggplot(AU_Das_labeled,
       aes(year + month/12, GPP)) +
  geom_line() +
  geom_point(aes(colour = drought_year)) +
  scale_colour_manual(values = c("black", "red")) +
  labs(title = "GPP response during low-precipitation years")
```
```{r}

Das_drought <- AU_Das_labeled %>%
  filter(drought_year == TRUE)

```

```{r}
train_data <- Das_drought
test_data <- How_drought
```

```{r}

model <- gam(GPP ~ s(Tair, k = 4) + s(VPD, k = 4) + s(SW_rad, k = 4), data=train_data, 
                 method = "REML")

predictions <- predict(model, newdata=test_data)
results <- data.frame(observed=test_data$GPP, predicted=predictions)

rmse_gam <- sqrt(mean((results$observed - results$predicted)^2))
rsq_gam <- cor(results$observed, results$predicted)^2
mean_observed_gam <- mean(results$observed)
nse_gam <- (1 - sum((results$observed - results$predicted)^2) / 
              sum((results$observed - mean_observed_gam)^2))
cat("RMSE:", rmse_gam, "\n")
cat("R-squared:", rsq_gam, "\n")
cat("Nash-Sutcliffe Efficiency (NSE):", nse_gam, "\n")
```

The model is a lot better at predicting two similar sites rather than different sites.

Sunday 1st Feb:
- Daily, growing-season flux data from NATT
- ET:PET ratio - find droughts
- get_vc_max and get_LCC_data to add to the GAM


```{r}
rm(list = ls())
```

Loaded in same sites, converted them to daily
Drop NA from dataframes

```{r}
AU_How_daily2 <- AU_How_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)

AU_Dap_daily2 <- AU_Dap_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)

AU_Das_daily2 <- AU_Das_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)

AU_Dry_daily2 <- AU_Dry_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)

AU_Stp_daily2 <- AU_Stp_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)

AU_Asm_daily2 <- AU_Asm_daily2 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, ET, LE, SW_rad)
```

1. LE to ET

```{r}
lambda <- 2.45e6
secs_day <- 86400

```

```{r}
AU_How_daily2 <- AU_How_daily2 %>%
  mutate(
    ET_mm_day = LE / lambda * secs_day * 1000
  )

```

Make into a function:

```{r}
add_ET <- function(df, LE_col = LE) {
  lambda <- 2.45e6
  secs_day <- 86400
  
  df %>%
    mutate(
      ET_mm_day = {{LE_col}} / lambda * secs_day * 1000
    )
}
```

Apply to datasets

```{r}
AU_Dap_daily2 <- add_ET(AU_Dap_daily2)
AU_Das_daily2 <- add_ET(AU_Das_daily2)
AU_Dry_daily2 <- add_ET(AU_Dry_daily2)
AU_Stp_daily2 <- add_ET(AU_Stp_daily2)
AU_Asm_daily2 <- add_ET(AU_Asm_daily2)
```

ET/PET ratio is a proxy for water stress
- PET = atmospheric demand
- ET = what actually happens

Low ET/PET ratio - plants want to transpire but cannot. A ratio near 1 indicates low stress, but a ratio near 0 indicates high stress when plants are closing stomata to limit water loss.

2. Calculate PET for the sites

```{r}
priestley_taylor_pet <- function(Rn, Tair, P) {

  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales

  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)

  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))

  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)

  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}
```

```{r}
AU_How_daily2$pet <- priestley_taylor_pet(AU_How_daily2$Rn, AU_How_daily2$Tair, AU_How_daily2$Pa)
AU_Das_daily2$pet <- priestley_taylor_pet(AU_Das_daily2$Rn, AU_Das_daily2$Tair, AU_Das_daily2$Pa)
AU_Dap_daily2$pet <- priestley_taylor_pet(AU_Dap_daily2$Rn, AU_Dap_daily2$Tair, AU_Dap_daily2$Pa)
AU_Dry_daily2$pet <- priestley_taylor_pet(AU_Dry_daily2$Rn, AU_Dry_daily2$Tair, AU_Dry_daily2$Pa)
AU_Stp_daily2$pet <- priestley_taylor_pet(AU_Stp_daily2$Rn, AU_Stp_daily2$Tair, AU_Stp_daily2$Pa)
AU_Asm_daily2$pet <- priestley_taylor_pet(AU_Asm_daily2$Rn, AU_Asm_daily2$Tair, AU_Asm_daily2$Pa)
```

Now make a column with ET/PET ratio:


```{r}

AU_How_daily2 <- AU_How_daily2 %>%
  mutate(
    ET_PET = ET_mm_day / pet
  )


```

```{r}
summary(AU_How_daily2$LE)
```

Really large numbers in ET:PET?
- Now going to use the half-hourly data for calculating ET

Drop NA from half-hourly
```{r}
AU_How_1 <- AU_How_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)

AU_Dap_1 <- AU_Dap_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)

AU_Das_1 <- AU_Das_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)

AU_Dry_1 <- AU_Dry_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)

AU_Stp_1 <- AU_Stp_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)

AU_Asm_1 <- AU_Asm_1 %>%
  drop_na(GPP, VPD, Tair, Precip, Pa, Rn, LE, SW_rad)
```

now, calculate ET for the half-hourly data

```{r}
lambda <- 2.45e6
secs_day <- 86400

```

```{r}
AU_How_1 <- AU_How_1 %>%
  mutate(
    ET_mm_day = LE / lambda * secs_day * 1000
  )

```

Make into a function:

```{r}
add_ET <- function(df, LE_col = LE) {
  lambda <- 2.45e6
  secs_day <- 86400
  
  df %>%
    mutate(
      ET_mm_day = {{LE_col}} / lambda * secs_day * 1000
    )
}
```

```{r}
AU_Dap_1 <- add_ET(AU_Dap_1)
AU_Das_1 <- add_ET(AU_Das_1)
AU_Dry_1<- add_ET(AU_Dry_1)
AU_Stp_1 <- add_ET(AU_Stp_1)
AU_Asm_1<- add_ET(AU_Asm_1)
```

Now convert to daily?

```{r}
convert_to_daily <- function(df) {
  
  #Conversion constants
  
  UMOL_TO_MOL <- 1E-6
  MOL_C_TO_GRAMS_C <- 12.0
  SEC_2_HALF_HOUR <- 1800.
  WM2_TO_KG_M2_S <- 1.0 /  ( 2.45 * 1E6)
  PA_2_KPA = 0.001
  HPA_2_KPA = 0.1
  
  df <- df %>%
    mutate(
      GPP = GPP * UMOL_TO_MOL * MOL_C_TO_GRAMS_C * SEC_2_HALF_HOUR,
      ET = LE * WM2_TO_KG_M2_S * SEC_2_HALF_HOUR, 
      VPD = VPD * HPA_2_KPA,
      Precip = Precip * SEC_2_HALF_HOUR
    )
  
  df$date <- ymd_hms(df$date)
  
  #Resample to daily
  
  df <- df %>%
    mutate(date = floor_date(date, "day")) %>%
    group_by(site, date) %>%
    summarise(
      Precip = ifelse(
        sum(!is.na(Precip)) > 0,
        sum(Precip, na.rm = TRUE),
        NA
      ),
      GPP = sum(GPP, na.rm = TRUE),
      ET = sum(ET, na.rm = TRUE), 
      LE = mean(LE, na.rm = TRUE), 
      SW_rad = mean(SW_rad, na.rm = TRUE),
      Tair = mean(Tair, na.rm = TRUE), 
      VPD = mean(VPD, na.rm = TRUE),
      Pa = mean(Pa, na.rm = TRUE),
      Rn = mean(Rn, na.rm = TRUE),
      ET_mm_day = (ET_mm_day),
      .groups = "drop"
    )
  
  return(df)
  
}
```

```{r}

AU_How_daily <- convert_to_daily(AU_How_1)
AU_Dap_daily <- convert_to_daily(AU_Dap_1)
AU_Das_daily <- convert_to_daily(AU_Das_1)
AU_Dry_daily <- convert_to_daily(AU_Dry_1)
AU_Stp_daily <- convert_to_daily(AU_Stp_1)
AU_Asm_daily <- convert_to_daily(AU_Asm_1)

```

Pet
```{r}
AU_How_daily$pet <- priestley_taylor_pet(AU_How_daily$Rn, AU_How_daily$Tair, AU_How_daily$Pa)
AU_Das_daily$pet <- priestley_taylor_pet(AU_Das_daily$Rn, AU_Das_daily$Tair, AU_Das_daily$Pa)
AU_Dap_daily$pet <- priestley_taylor_pet(AU_Dap_daily$Rn, AU_Dap_daily$Tair, AU_Dap_daily$Pa)
AU_Dry_daily$pet <- priestley_taylor_pet(AU_Dry_daily$Rn, AU_Dry_daily$Tair, AU_Dry_daily$Pa)
AU_Stp_daily$pet <- priestley_taylor_pet(AU_Stp_daily$Rn, AU_Stp_daily$Tair, AU_Stp_daily$Pa)
AU_Asm_daily$pet <- priestley_taylor_pet(AU_Asm_daily$Rn, AU_Asm_daily$Tair, AU_Asm_daily$Pa)
```

Column with ratio

```{r}

AU_How_daily <- AU_How_daily %>%
  mutate(
    ET_PET = ET_mm_day / pet
  )

```


Slightly confused? Not sure what I have done wrong for the ET:PET ratio

