---
title: "01_driver_importance"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Subsection 1: VPD across Australia
- **Driver importance across sites**
- GAMs - predictability
- GPP~VPD responses

Fit Random Forests to determine the importance of explanatory variables of GPP across Australian flux
- Compare random forests with GAM analysis

```{r}
####
##  Fit a random forest to predict GPP, here we're testing the use of lags
##
##  author: Martin De Kauwe
##  date: 27th February, 2025
##  email: mdekauwe@gmail.com
####

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

#standardise function
standardise <- function(df, vars_to_scale) {
  df_scaled <- df
  df_scaled[vars_to_scale] <- scale(df[vars_to_scale])
  
  return(df_scaled)
}

#Function to run RF

run_rf_for_site <- function(df, site_name) {
  
  #Drop NAs
  df <- df %>% drop_na()
  
  #check you have enough data
   if (nrow(df) < 100) {
    warning(paste("Site", site_name, "has insufficient data"))
    return(NULL)
   }
  
  #lagged variables before splitting the data
   df <- df %>%
    arrange(date) %>%  # Ensure data is sorted by date
    mutate(
      # 7-day cumulative CWD
      cumulative_cwd_7day = rollapply(cwd, width=7, FUN=sum, 
                                      align="right", fill=NA, na.rm=TRUE),
      # Temperature lags
      Tair_lag1 = lag(Tair, 1),
      Tair_lag2 = lag(Tair, 2),
      Tair_lag3 = lag(Tair, 3),
      # VPD lags
      VPD_lag1 = lag(VPD, 1),
      VPD_lag2 = lag(VPD, 2),
      VPD_lag3 = lag(VPD, 3)
    ) %>%
    drop_na()
   
   #Now split into train/test
   set.seed(123)
  idx <- createDataPartition(df$GPP, p=0.8, list=FALSE)
  train_data <- df[idx, ]
  test_data <- df[-idx, ]
  
  # Standardise predictors
  predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day",
                  "Tair_lag1", "Tair_lag2", "Tair_lag3", 
                  "VPD_lag1", "VPD_lag2", "VPD_lag3")
  
  train_data <- standardise(train_data, predictors)
  test_data <- standardise(test_data, predictors)
  
  #train random forest
   model <- randomForest(GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day + 
                          Tair_lag1 + Tair_lag2 + Tair_lag3 + 
                          VPD_lag1 + VPD_lag2 + VPD_lag3, 
                        data=train_data, ntree=500, mtry=3)
   
   #now predict
    predictions <- predict(model, test_data)
  results <- data.frame(Observed=test_data$GPP, Predicted=predictions)
  
  #calculate metrics
  rmse <- sqrt(mean((results$Observed - results$Predicted)^2))
  rsq <- cor(results$Observed, results$Predicted)^2
  mean_observed <- mean(results$Observed)
  nse <- (1 - sum((results$Observed - results$Predicted)^2) / 
            sum((results$Observed - mean_observed)^2))
  
  #get variable importance
  
importance_df <- data.frame(
    variable = rownames(importance(model)),
    importance = importance(model)[,1]
  )

#return results
return(list(
    site = site_name,
    model = model,
    metrics = data.frame(site=site_name, RMSE=rmse, R2=rsq, NSE=nse),
    importance = importance_df %>% mutate(site = site_name),
    predictions = results %>% mutate(site = site_name)
  ))

}

# Apply to all sites in the list
rf_results <- map2(sites_daily, names(sites_daily), run_rf_for_site)

# Extract metrics for all sites
all_metrics <- map_df(rf_results, ~.x$metrics)
print(all_metrics)

# Extract importance for all sites
all_importance <- map_df(rf_results, ~.x$importance)

```

```{r}
site_names <- unique(all_importance$site)
n_groups <- ceiling(length(site_names) / 6)

for(i in 1:n_groups) {
  start_idx <- (i-1)*6 + 1
  end_idx <- min(i*6, length(site_names))
  sites_subset <- site_names[start_idx:end_idx]
  
  p <- ggplot(all_importance %>% filter(site %in% sites_subset), 
              aes(x=reorder(variable, importance), y=importance)) +
    geom_col(fill="steelblue") +
    facet_wrap(~site, scales="free_x", ncol=3) +
    coord_flip() +
    theme_minimal() +
    theme(
      strip.text = element_text(size=10),
      axis.text = element_text(size=8)
    ) +
    labs(title=paste("Variable Importance - Group", i), 
         x="Variable", y="Importance")
  
  print(p)
  ggsave(paste0("importance_group_", i, ".png"), width=12, height=8, dpi=300)
}
```

```{r}
install.packages("tidytext")
```

Random forest:
Predictors - VPD lag1, lag2, lag3, cwd 7 day, VPD, SW_rad, Tair, Tair lag 1, Tair lag 2, Tair lag 3


```{r}
# Loop through each site and create individual plots
for(site_name in unique(all_importance$site)) {
  
  plot_data <- all_importance %>%
    filter(site == site_name) %>%
    arrange(desc(importance)) %>%
    mutate(rank = row_number(),
           variable = paste0(rank, ". ", variable))
  
  p <- ggplot(plot_data, aes(x=importance, y=reorder(variable, importance))) +
    geom_col(fill="steelblue") +
    theme_minimal() +
    theme(text = element_text(size=12)) +
    labs(title=paste("Variable Importance -", site_name),
         x="Importance (%IncMSE)", 
         y="")
  
  print(p)
  ggsave(paste0("importance_", site_name, ".png"), width=8, height=6, dpi=300)
}
```
Drier sites
- cwd in top predictor, water-limited environments

VPD lags more important at intermediate sites

Wet sites:
- SW rad and VPD dominate

EBF consistently show SW radiation as their most important predictor with VPD 2-3, water abundant so energy and atmospheric demand control photosynthesis

Fog (wettest site) - only VPD lags matter

Grasslands = strong CWD control + temp lag effects
Savannas - mixed cwd and vpd
EBF - consistently SW radiation then VPD then cwd
WSA - mixed
DBF - SW radiaion

2. Random forest updated to predict GPP
- %IncMSE (permutation) - instead of IncNodePurity
- Boruta package for selecting variables
- Test VPD importance via R2 change in held-out test data

```{r}
install.packages("Boruta")
```


```{r}
####
##  Fit a random forest to predict GPP with lags
##  - %IncMSE permutation importance
##  - Boruta variable selection  
##  - Explicit VPD importance test via R2 drop
####

library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(Boruta)
library(RColorBrewer)

# ---------------------------------------------------------------------------
# Helpers
# ---------------------------------------------------------------------------

standardise <- function(df, vars_to_scale, ref_df = NULL) {
  df_scaled <- df
  if (is.null(ref_df)) ref_df <- df
  for (v in vars_to_scale) {
    m <- mean(ref_df[[v]], na.rm = TRUE)
    s <- sd(ref_df[[v]],   na.rm = TRUE)
    df_scaled[[v]] <- (df[[v]] - m) / s
  }
  return(df_scaled)
}

calc_metrics <- function(obs, pred) {
  rmse <- sqrt(mean((obs - pred)^2, na.rm = TRUE))
  rsq  <- cor(obs, pred, use = "complete.obs")^2
  nse  <- 1 - sum((obs - pred)^2, na.rm = TRUE) /
              sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
  data.frame(RMSE = rmse, R2 = rsq, NSE = nse)
}

# ---------------------------------------------------------------------------
# Main function
# ---------------------------------------------------------------------------

run_rf_for_site <- function(df, site_name) {
  
  df <- df %>% drop_na()
  
  if (nrow(df) < 100) {
    warning(paste("Site", site_name, "has insufficient data"))
    return(NULL)
  }
  
  # Lags BEFORE splitting
  df <- df %>%
    arrange(date) %>%
    mutate(
      cumulative_cwd_7day = rollapply(cwd, width = 7, FUN = sum,
                                      align = "right", fill = NA, na.rm = TRUE),
      Tair_lag1 = lag(Tair, 1),
      Tair_lag2 = lag(Tair, 2),
      Tair_lag3 = lag(Tair, 3),
      VPD_lag1  = lag(VPD, 1),
      VPD_lag2  = lag(VPD, 2),
      VPD_lag3  = lag(VPD, 3)
    ) %>%
    drop_na()
  
  predictors <- c("Tair", "SW_rad", "VPD", "cumulative_cwd_7day",
                  "Tair_lag1", "Tair_lag2", "Tair_lag3",
                  "VPD_lag1",  "VPD_lag2",  "VPD_lag3")
  
  # Train / test split
  set.seed(123)
  idx        <- createDataPartition(df$GPP, p = 0.8, list = FALSE)
  train_data <- df[idx, ]
  test_data  <- df[-idx, ]
  
  # Scale test data using training params only
  train_data <- standardise(train_data, predictors)
  test_data  <- standardise(test_data,  predictors, ref_df = train_data)
  
  # =========================================================================
  # 1. BORUTA
  # =========================================================================
  
  cat("\n", site_name, "- Running Boruta...\n")
  
  boruta_result <- Boruta(
    GPP ~ Tair + SW_rad + VPD + cumulative_cwd_7day +
          Tair_lag1 + Tair_lag2 + Tair_lag3 +
          VPD_lag1  + VPD_lag2  + VPD_lag3,
    data    = train_data,
    doTrace = 0,
    maxRuns = 100
  )
  
  boruta_final   <- TentativeRoughFix(boruta_result)
  boruta_summary <- attStats(boruta_final) %>%
    rownames_to_column("variable") %>%
    mutate(site = site_name)
  
  confirmed_vars <- getSelectedAttributes(boruta_final, withTentative = FALSE)
  cat("  Confirmed variables:", paste(confirmed_vars, collapse = ", "), "\n")
  
  # =========================================================================
  # 2. FULL RF with permutation importance (%IncMSE)
  # =========================================================================
  
  full_formula <- as.formula(
    paste("GPP ~", paste(predictors, collapse = " + "))
  )
  
  rf_full <- randomForest(
    full_formula,
    data       = train_data,
    ntree      = 500,
    mtry       = 3,
    importance = TRUE
  )
  
  pred_full    <- predict(rf_full, test_data)
  metrics_full <- calc_metrics(test_data$GPP, pred_full)
  cat("  Full RF R2:", round(metrics_full$R2, 3), "\n")
  
  # %IncMSE = permutation importance (use this, not IncNodePurity)
  importance_df <- data.frame(
    variable      = rownames(importance(rf_full)),
    pct_IncMSE    = importance(rf_full)[, "%IncMSE"],
    IncNodePurity = importance(rf_full)[, "IncNodePurity"],
    site          = site_name
  )
  
  # =========================================================================
  # 3. VPD IMPORTANCE TEST - R2 drop when VPD removed
  # =========================================================================
  
  predictors_no_vpd <- predictors[!grepl("VPD", predictors)]
  
  formula_no_vpd <- as.formula(
    paste("GPP ~", paste(predictors_no_vpd, collapse = " + "))
  )
  
  rf_no_vpd <- randomForest(
    formula_no_vpd,
    data       = train_data,
    ntree      = 500,
    mtry       = 3,
    importance = TRUE
  )
  
  # Use only columns this model needs to avoid NA issues (fixes Ade_HH, Asm_HH)
  test_data_no_vpd <- test_data %>%
    select(GPP, all_of(predictors_no_vpd)) %>%
    drop_na()
  
  pred_no_vpd    <- predict(rf_no_vpd, test_data_no_vpd)
  metrics_no_vpd <- calc_metrics(test_data_no_vpd$GPP, pred_no_vpd)
  
  vpd_r2_drop <- metrics_full$R2 - metrics_no_vpd$R2
  cat("  R2 without VPD:", round(metrics_no_vpd$R2, 3),
      "| VPD R2 contribution:", round(vpd_r2_drop, 3), "\n")
  
  vpd_importance <- data.frame(
    site           = site_name,
    R2_full        = metrics_full$R2,
    R2_no_vpd      = metrics_no_vpd$R2,
    VPD_R2_contrib = vpd_r2_drop
  )
  
  # =========================================================================
  # Return
  # =========================================================================
  
  return(list(
    site           = site_name,
    model_full     = rf_full,
    model_no_vpd   = rf_no_vpd,
    metrics        = data.frame(site = site_name, metrics_full),
    importance     = importance_df,
    boruta         = boruta_summary,
    boruta_raw     = boruta_final,
    vpd_importance = vpd_importance,
    predictions    = data.frame(
                       site      = site_name,
                       Observed  = test_data$GPP,
                       Predicted = pred_full
                     )
  ))
  
}  # <-- closes run_rf_for_site

# ---------------------------------------------------------------------------
# Run across all sites
# ---------------------------------------------------------------------------

rf_results <- map2(sites_daily, names(sites_daily), run_rf_for_site) %>%
  compact()

# ---------------------------------------------------------------------------
# Extract results
# ---------------------------------------------------------------------------

all_metrics    <- map_df(rf_results, ~ .x$metrics)
all_importance <- map_df(rf_results, ~ .x$importance)
all_boruta     <- map_df(rf_results, ~ .x$boruta)
all_vpd        <- map_df(rf_results, ~ .x$vpd_importance)

print(all_metrics)
print(all_vpd)

# ---------------------------------------------------------------------------
# Plots
# ---------------------------------------------------------------------------

# 1. Permutation importance across sites
all_importance %>%
  ggplot(aes(x = reorder(variable, pct_IncMSE), y = pct_IncMSE, fill = site)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "% Increase in MSE (permutation importance)", fill = "Site") +
  theme_classic()

# 2. Boruta decisions across sites
all_boruta %>%
  ggplot(aes(x = reorder(variable, meanImp), y = meanImp, colour = decision)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = minImp, ymax = maxImp), width = 0.2) +
  facet_wrap(~ site) +
  coord_flip() +
  scale_colour_manual(values = c(Confirmed = "#2ca25f",
                                 Rejected  = "#e34a33",
                                 Tentative = "#fdbb84")) +
  labs(x = NULL, y = "Boruta importance (mean ± range)", colour = NULL) +
  theme_classic() +
  theme(strip.background = element_blank())

# 3. VPD R2 contribution across sites
all_vpd %>%
  pivot_longer(cols = c(R2_full, R2_no_vpd),
               names_to = "model", values_to = "R2") %>%
  mutate(model = recode(model,
                        R2_full   = "Full model",
                        R2_no_vpd = "Without VPD")) %>%
  ggplot(aes(x = site, y = R2, fill = model)) +
  geom_col(position = "dodge") +
  geom_text(data = all_vpd,
            aes(x = site, y = R2_full + 0.02,
                label = paste0("\u0394", round(VPD_R2_contrib, 2))),
            inherit.aes = FALSE, size = 3.5) +
  scale_fill_manual(values = brewer.pal(3, "Set2")[1:2]) +
  labs(x = NULL, y = expression(R^2), fill = NULL,
       title = "VPD contribution to GPP predictability") +
  theme_classic() +
  theme(legend.position = "top")
```



