---
title: "load_LAI"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in LAI data
Every 4 days
1 measurements a day per site

1. Convert date column into proper date column
2. Somehow attach the corresponding date from the LAI to the one I have in the flux data - every 4 days?

Load in data

```{r}
library(tidyverse)
```

```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/site_4day_LAI.csv")
```

```{r}
lai_data <- lai_data %>% select(-.geo)
```

Convert date to proper date format
```{r}
lai_data <- lai_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%dT%H:%M:%S"))
```

```{r}
head(lai_data)
```

Make corresponding names from LAI data and Flux data
```{r}
site_mapping <- data.frame(
  flux_site = c("How", "Stp", "Das", "Dry", "Asm", "Tum", "Wom", "TTE"),
  lai_site = c("Howard Springs", "Sturt Plains", "Daly River Uncleared", "Dry River", "Alice Springs", "Tumbarumba", "Wombat", "Ti Tree East")
)
```

Quickly add in site names to flux data
```{r}
How_HH_daily <- How_HH_daily %>%
  mutate(site = "How")

Stp_HH_daily <- Stp_HH_daily %>%
  mutate(site = "Stp")

Das_HH_daily <- Das_HH_daily %>%
  mutate(site = "Das")

Dry_HH_daily <- Dry_HH_daily %>%
  mutate(site = "Dry")

Asm_HH_daily <- Asm_HH_daily %>%
  mutate(site = "Asm")

Tum_HH_daily <- Tum_HH_daily %>%
  mutate(site = "Tum")

Wom_HH_daily <- Wom_HH_daily %>%
  mutate(site = "Wom")

TTE_HH_daily <- TTE_HH_daily %>%
  mutate(site = "TTE")
```

Convert flux data date to match the format of the lai data
```{r}
# Convert flux data date to Date format (remove time component)
How_HH_daily <- How_HH_daily %>%
  mutate(date = as.Date(date))
```

```{r}
# Check it worked
class(How_HH_daily$date)
head(How_HH_daily$date)
```


Apply to rest of sites:
```{r}
Stp_HH_daily <- Stp_HH_daily %>% mutate(date = as.Date(date))
Das_HH_daily <- Das_HH_daily %>% mutate(date = as.Date(date))
Dry_HH_daily <- Dry_HH_daily %>% mutate(date = as.Date(date))
Asm_HH_daily <- Asm_HH_daily %>% mutate(date = as.Date(date))
Tum_HH_daily <- Tum_HH_daily %>% mutate(date = as.Date(date))
Wom_HH_daily <- Wom_HH_daily %>% mutate(date = as.Date(date))
TTE_HH_daily <- TTE_HH_daily %>% mutate(date = as.Date(date))
```

```{r}
How_lai <- How_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Stp_lai <- Stp_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Das_lai <- Das_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Dry_lai <- Dry_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Asm_lai <- Asm_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Tum_lai <- Tum_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Wom_lai <- Wom_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

TTE_lai <- TTE_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)


```

Fill in missing days:
Going to use linear interpolation - assumes gradual change between measurements. 

```{r}
library(zoo)
```

```{r}
How_lai <- How_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)
```

```{r}
Stp_lai <- Stp_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Das_lai <- Das_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Dry_lai<- Dry_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Asm_lai <- Asm_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Tum_lai<- Tum_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Wom_lai <- Wom_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

TTE_lai <- TTE_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)
```

Create filter - max gap parameter tells it to only interpolate gaps of 4 or fewer observations. 4 days = 4 rows

```{r}
How_lai <- How_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Stp_lai<- Stp_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Das_lai<- Das_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Dry_lai<- Dry_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Asm_lai<- Asm_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Tum_lai<- Tum_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Wom_lai<- Wom_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

TTE_lai<- TTE_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

```

```{r}
plot(Tum_lai$LAI_filled)
plot(Asm_lai$LAI_filled)
plot(Dry_lai$LAI_filled)
plot(TTE_lai$LAI_filled)
plot(Das_lai$LAI_filled)
plot(Stp_lai$LAI_filled)
plot(Wom_lai$LAI_filled)
plot(How_lai$LAI_filled)
```

How - Woody savanna
Das - Savanna
Dry - Savanna
Stp - Grasslands, wetlands
Tum - Evergreen broadleaf forest
Wom - EBF
TTE - GRA
Asm - SAV


0-1 - sparse vegetation
1-3 - grassland/deciduous winter
3-6 - crops at peak, temperate forests
6-8 - dense forest
8+ - dense canopy

```{r}
plot(TTE_lai$VPD, TTE_lai$LAI)
```

Now let's add the LAI data back into the "_HH"daily" dataframes so I can use them for analysis:

```{r}
# Add LAI and LAI_filled columns from TTE_lai to TTE_HH_daily

TTE_HH_daily <- TTE_HH_daily %>%
  left_join(TTE_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

How_HH_daily <- How_HH_daily %>%
  left_join(How_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Stp_HH_daily <- Stp_HH_daily %>%
  left_join(Stp_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Dry_HH_daily <- Dry_HH_daily %>%
  left_join(Dry_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Asm_HH_daily <- Asm_HH_daily %>%
  left_join(Asm_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Tum_HH_daily <- Tum_HH_daily %>%
  left_join(Tum_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Wom_HH_daily <- Wom_HH_daily %>%
  left_join(Wom_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

Das_HH_daily <- Das_HH_daily %>%
  left_join(Das_lai %>% select(date, LAI, LAI_filled), 
            by = "date")

```

Now we can put all of the sites into a list for reproducibility, remember here that some of the sites have LAI and some do not
```{r}
names <- c("How_HH_daily", "Ade_HH_daily", "Das_HH_daily", "Dap_HH_daily", "Dry_HH_daily", "Stp_HH_daily", "Asm_HH_daily", "Cum_HH_daily", "Emr_HH_daily", "Fog_HH_daily", "Gin_HH_daily", "Gww_HH_daily", "Lox_HH_daily", "Rdf_HH_daily", "Rig_HH_daily", "Rob_HH_daily", "TTE_HH_daily", "Wac_HH_daily", "Whr_HH_daily", "Wom_HH_daily", "Ync_HH_daily", "Cpr_HH_daily", "Tum_HH_daily")


sites_final <- mget(names)
```

