---
title: "load_LAI"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in LAI data
Every 4 days
1 measurements a day per site

1. Convert date column into proper date column
2. Somehow attach the corresponding date from the LAI to the one I have in the flux data - every 4 days?

Load in data

```{r}
library(tidyverse)
```

```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/site_4day_LAI.csv")
```

```{r}
lai_data <- lai_data %>% select(-.geo)
```

Convert date to proper date format
```{r}
lai_data <- lai_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%dT%H:%M:%S"))
```

```{r}
head(lai_data)
```

Make corresponding names from LAI data and Flux data
```{r}
site_mapping <- data.frame(
  flux_site = c("How", "Stp", "Dry", "Das", "TTE", "Asm", "Tum", "Wom", "Ade", "Cpr", "Cum", "Dap", "Emr", "Fog", "Gin", "Gww", "Lox", "Rdf", "Rig", "Rob", "Wac", "Whr", "Ync"),
  lai_site = c("How", "Stp", "Dry", "Das", "TTE", "Asm", "Tum", "Wom", "Ade", "Cpr", "Cum", "Dap", "Emr", "Fog", "Gin", "Gww", "Lox", "Rdf", "Rig", "Rob", "Wac", "Whr", "Ync")
)
```

Quickly add in site names to flux data
```{r}
sites_daily <- imap(sites_daily, function(df, name) {
  site_code <- str_remove(name, "_HH_daily")
  df %>% mutate(site = site_code)
})
```

Convert flux data date to match the format of the lai data
```{r}
sites_daily <- map(sites_daily, function(df) {
  df %>% mutate(date = as.Date(date))
})
```

Join the LAI data onto the site data
```{r}
sites_lai <- map(sites_daily, function(df) {
  df %>%
    left_join(site_mapping, by = c("site" = "flux_site")) %>%
    left_join(lai_data %>% select(date, site, mean),
              by = c("date" = "date", "lai_site" = "site")) %>%
    rename(LAI = mean)
})
```


Fill in missing days:
Going to use linear interpolation - assumes gradual change between measurements. 

```{r}
library(zoo)

sites_lai <- map(sites_lai, function(df) {
  df %>%
    mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y) %>%
    arrange(date) %>%
    mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))
})
```

```{r}
walk(names(sites_lai), function(name) {
  plot(sites_lai[[name]]$LAI_filled, main = name)
})
```

```{r}
plot(sites_lai$How_HH_daily$LAI)
```


How - Woody savanna
Das - Savanna
Dry - Savanna
Stp - Grasslands, wetlands
Tum - Evergreen broadleaf forest
Wom - EBF
TTE - GRA
Asm - SAV


0-1 - sparse vegetation
1-3 - grassland/deciduous winter
3-6 - crops at peak, temperate forests
6-8 - dense forest
8+ - dense canopy

```{r}
plot(TTE_lai$VPD, TTE_lai$LAI)
```

Now let's add the LAI data back into the "_HH"daily" dataframes so I can use them for analysis:

```{r}
sites_daily <- map2(sites_daily, sites_lai, function(hh_df, lai_df) {
  hh_df %>%
    left_join(lai_df %>% select(date, LAI, LAI_filled),
              by = "date")
})
```

Mean LAI per site from sites_daily

```{r}
# get mean LAI per site from sites_daily
canopy_order <- map_dfr(sites_daily, function(df) {
  tibble(
    site = unique(df$site),
    mean_LAI = mean(df$LAI_filled, na.rm = TRUE)
  )
}) %>%
  arrange(desc(mean_LAI))

print(canopy_order)
```


```{r}
ggplot(canopy_order, aes(x = reorder(site, desc(mean_LAI)), y = mean_LAI)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(x = "Site", y = "Mean LAI") +
  theme_classic()
```

```{r}
all_sites <- bind_rows(sites_daily)

ggplot(all_sites, aes(x = reorder(site, LAI_filled, FUN = median, na.rm = TRUE), 
                      y = LAI_filled)) +
  geom_boxplot() +
  coord_flip() +
  labs(x = "Site", y = "LAI") +
  theme_classic()
```
Plot mean annual temp and mean annual precip

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

df_annual <- df_daily %>%
  group_by(year) %>%
  summarise(
    GPP = sum(GPP, na.rm = TRUE),
    GPP_n_na = sum(is.na(GPP)),
    ET = sum(ET, na.rm = TRUE),
    ET_n_na = sum(is.na(ET)),
    Precip = sum(Precip, na.rm = TRUE),
    Precip_n_na = sum(is.na(Precip)),
    Tair = mean(Tair, na.rm = TRUE),
    VPD = mean(VPD, na.rm = TRUE),
    LCC = mean(LCC_daily, na.rm = TRUE),
    LAI = mean(LAI_filled, na.rm = TRUE),
    .groups = "drop"
  )

  return(df_annual)
}
```

```{r}
# apply to all sites
sites_annual <- map(sites_daily, convert_to_annual)

# calculate mean annual values per site
site_climate <- map_dfr(names(sites_annual), function(site) {
  df <- sites_annual[[site]]
  
  tibble(
    site = str_remove(site, "_HH_daily"),
    mat = mean(df$Tair, na.rm = TRUE),       # mean annual temperature
    map = mean(df$Precip, na.rm = TRUE),      # mean annual precipitation
    mal = mean(df$LAI, na.rm = TRUE),         # mean annual LAI
    malcc = mean(df$LCC, na.rm = TRUE)        # mean annual LCC
  )
})

print(site_climate)
```

```{r}
ggplot(site_climate, aes(x = map, y = mat, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       title = "Climate space across sites") +
  theme_classic()
```
Add in clustering groups:

```{r}
site_slopes <- site_slopes %>%
  left_join(site_climate, by = "site")
```

```{r}
ggplot(site_slopes, aes(x = map, y = mat, colour = sensitivity_group, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       colour = "VPD sensitivity", title = "Climate space across sites") +
  theme_classic()
```
```{r}
ggplot(site_climate, aes(x = map, y = mal, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual precipitation (mm)", y = "Mean Annual Leaf Area Index",
       title = "Mean annual LAI and precipitation") +
  theme_classic()
```
```{r}
ggplot(site_slopes, aes(x = map, y = mal, colour = sensitivity_group, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual LAI",
       colour = "VPD sensitivity", title = "Mean annual LAI and precip") +
  theme_classic()
```


