---
title: "load_LAI"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in LAI data
Every 4 days
1 measurements a day per site
2. Convert date column into proper date column
3. Somehow attach the corresponding date from the LAI to the one I have in the flux data - every 4 days?

Add site names back in

```{r}
library(tidyverse)
library(data.table)
library(zoo)

sites_daily <- imap(sites_daily, function(df, name) {
  df %>% mutate(site = name)
})
```
Load and clean LAI data

```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/site_LAI_MODIS_2000_2024_2.csv") %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))
```

Verify names of LAI data and Flux data match

```{r}
cat("LAI site names:\n")
print(unique(lai_data$site))
cat("\nFlux site names:\n")
print(names(sites_daily))
```
Make sure flux date is correct format 

```{r}
sites_daily <- map(sites_daily, function(df) {
  df %>% mutate(date = as.Date(date))
})
```

Rolling nearest date to join for each site


```{r}
# Step 5: Exact join first to preserve NAs between 4-day obs
sites_daily <- map(sites_daily, function(df) {
  site_name <- df$site[1]
  
  lai_site <- lai_data %>%
    filter(site == site_name) %>%
    select(date, mean) %>%
    rename(LAI_4day = mean) %>%
    arrange(date)
  
  if (nrow(lai_site) == 0) {
    warning(paste("No LAI data found for site:", site_name))
    return(df %>% mutate(LAI_4day = NA_real_, LAI_filled = NA_real_))
  }
  
  # Exact join - non-matching dates get NA, which is what we want
  left_join(df, lai_site, by = "date")
})

# Step 6: Interpolate between the 4-day obs
sites_daily <- map(sites_daily, function(df) {
  df %>%
    arrange(date) %>%
    mutate(LAI_filled = na.approx(LAI_4day, x = date, maxgap = 8, na.rm = FALSE))
})
  
```

```{r}

# Step 7: Check how many valid LAI days each site has
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  n_valid <- sum(!is.na(df$LAI_filled))
  n_total <- nrow(df)
  cat(sprintf("%s - valid LAI days: %d / %d\n", name, n_valid, n_total))
})

# Step 8: Visual check - line = interpolated, red dots = raw 4-day obs
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  plot(df$date, df$LAI_filled,
       main  = name,
       xlab  = "Date", 
       ylab  = "LAI",
       type  = "l", 
       col   = "darkgreen",
       ylim  = range(c(df$LAI_filled, df$LAI_4day), na.rm = TRUE))
  points(df$date, df$LAI_4day, col = "red", pch = 16, cex = 0.5)
  legend("topright", 
         legend = c("Interpolated", "Raw 4-day"),
         col    = c("darkgreen", "red"), 
         lty    = c(1, NA), 
         pch    = c(NA, 16))
})
```

0-1 - sparse vegetation
1-3 - grassland/deciduous winter
3-6 - crops at peak, temperate forests
6-8 - dense forest
8+ - dense canopy

Max LAI per site from sites_daily

```{r}
# get max LAI per site from sites_daily
canopy_order <- map_dfr(sites_daily, function(df) {
  tibble(
    site = unique(df$site),
    max_LAI = max(df$LAI_filled, na.rm = TRUE)
  )
}) %>%
  arrange(desc(max_LAI))

print(canopy_order)
```
```{r}
LAI_order <- c("Tum", "Wom", "Rob", "Wac", "Rig", "Dap", "Fog", "How", "Cum", "Dry", "Ade", "Das", "Rdf", "Emr", "Whr", "Boy", "Gin", "Stp", "Ync", "Asm", "Cpr", "Gww")
```


```{r}
ggplot(canopy_order, aes(x = reorder(site, desc(max_LAI)), y = max_LAI)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(x = "Site", y = "Mean LAI") +
  theme_classic()
```

```{r}
all_sites <- bind_rows(sites_daily)

ggplot(all_sites, aes(x = reorder(site, LAI_filled, FUN = median, na.rm = TRUE), 
                      y = LAI_filled)) +
  geom_boxplot() +
  coord_flip() +
  labs(x = "Site", y = "LAI") +
  theme_classic()
```

Whittaker biome plot

```{r}
library(plotbiomes)
```

```{r}
#convert precipitation to cm from mm

site_slopes$map_cm <- site_slopes$map / 10

library(plotbiomes)
library(ggplot2)

ggplot() +
  whittaker_base_plot() +
  geom_point(data = site_slopes,
             aes(x = mat, y = map_cm),
             size = 3) +
  theme_classic()
```

```{r}

ggplot() +
  whittaker_base_plot() +
  geom_point(data = site_slopes,
             aes(x = mat, y = map_cm, colour = cluster_slope),
             size = 3) +
  theme_classic() +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white", colour = NA),
        plot.background = element_rect(fill = "white", colour = NA))
```
3 - high sensitivity
2 - moderate
1 - low

```{r}
ls("package:plotbiomes")
```

```{r}
ggsave("whittaker_plot.png", width = 10, height = 8, dpi = 300)

```

