---
title: "load_LAI"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in LAI data
Every 4 days
1 measurements a day per site
2. Convert date column into proper date column
3. Somehow attach the corresponding date from the LAI to the one I have in the flux data - every 4 days?

Add site names back in

```{r}
library(tidyverse)
library(data.table)
library(zoo)

sites_daily <- imap(sites_daily, function(df, name) {
  df %>% mutate(site = name)
})
```
Load and clean LAI data

```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/site_4day_LAI.csv") %>%
  select(-.geo) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%dT%H:%M:%S"))
```

Verify names of LAI data and Flux data match

```{r}
cat("LAI site names:\n")
print(unique(lai_data$site))
cat("\nFlux site names:\n")
print(names(sites_daily))
```
Make sure flux date is correct format 

```{r}
sites_daily <- map(sites_daily, function(df) {
  df %>% mutate(date = as.Date(date))
})
```

Rolling nearest date to join for each site


```{r}
# Step 5: Exact join first to preserve NAs between 4-day obs
sites_daily <- map(sites_daily, function(df) {
  site_name <- df$site[1]
  
  lai_site <- lai_data %>%
    filter(site == site_name) %>%
    select(date, mean) %>%
    rename(LAI_4day = mean) %>%
    arrange(date)
  
  if (nrow(lai_site) == 0) {
    warning(paste("No LAI data found for site:", site_name))
    return(df %>% mutate(LAI_4day = NA_real_, LAI_filled = NA_real_))
  }
  
  # Exact join - non-matching dates get NA, which is what we want
  left_join(df, lai_site, by = "date")
})

# Step 6: Interpolate between the 4-day obs
sites_daily <- map(sites_daily, function(df) {
  df %>%
    arrange(date) %>%
    mutate(LAI_filled = na.approx(LAI_4day, x = date, maxgap = 8, na.rm = FALSE))
})
  
```

```{r}

# Step 7: Check how many valid LAI days each site has
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  n_valid <- sum(!is.na(df$LAI_filled))
  n_total <- nrow(df)
  cat(sprintf("%s - valid LAI days: %d / %d\n", name, n_valid, n_total))
})

# Step 8: Visual check - line = interpolated, red dots = raw 4-day obs
walk(names(sites_daily), function(name) {
  df <- sites_daily[[name]]
  plot(df$date, df$LAI_filled,
       main  = name,
       xlab  = "Date", 
       ylab  = "LAI",
       type  = "l", 
       col   = "darkgreen",
       ylim  = range(c(df$LAI_filled, df$LAI_4day), na.rm = TRUE))
  points(df$date, df$LAI_4day, col = "red", pch = 16, cex = 0.5)
  legend("topright", 
         legend = c("Interpolated", "Raw 4-day"),
         col    = c("darkgreen", "red"), 
         lty    = c(1, NA), 
         pch    = c(NA, 16))
})
```

0-1 - sparse vegetation
1-3 - grassland/deciduous winter
3-6 - crops at peak, temperate forests
6-8 - dense forest
8+ - dense canopy

Max LAI per site from sites_daily

```{r}
# get max LAI per site from sites_daily
canopy_order <- map_dfr(sites_daily, function(df) {
  tibble(
    site = unique(df$site),
    max_LAI = max(df$LAI_filled, na.rm = TRUE)
  )
}) %>%
  arrange(desc(max_LAI))

print(canopy_order)
```


```{r}
ggplot(canopy_order, aes(x = reorder(site, desc(max_LAI)), y = max_LAI)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(x = "Site", y = "Mean LAI") +
  theme_classic()
```

```{r}
all_sites <- bind_rows(sites_daily)

ggplot(all_sites, aes(x = reorder(site, LAI_filled, FUN = median, na.rm = TRUE), 
                      y = LAI_filled)) +
  geom_boxplot() +
  coord_flip() +
  labs(x = "Site", y = "LAI") +
  theme_classic()
```
Plot mean annual temp and mean annual precip

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

df_annual <- df_daily %>%
  group_by(year) %>%
  summarise(
    GPP = sum(GPP, na.rm = TRUE),
    GPP_n_na = sum(is.na(GPP)),
    ET = sum(ET, na.rm = TRUE),
    ET_n_na = sum(is.na(ET)),
    Precip = sum(Precip, na.rm = TRUE),
    Precip_n_na = sum(is.na(Precip)),
    Tair = mean(Tair, na.rm = TRUE),
    VPD = mean(VPD, na.rm = TRUE),
    LCC = mean(LCC_daily, na.rm = TRUE),
    LAI = mean(LAI_filled, na.rm = TRUE),
    .groups = "drop"
  )

  return(df_annual)
}
```

```{r}
# apply to all sites
sites_annual <- map(sites_daily, convert_to_annual)

# calculate mean annual values per site
site_climate <- map_dfr(names(sites_annual), function(site) {
  df <- sites_annual[[site]]
  
  tibble(
    site = str_remove(site, "_HH_daily"),
    mat = mean(df$Tair, na.rm = TRUE),       # mean annual temperature
    map = mean(df$Precip, na.rm = TRUE),      # mean annual precipitation
    mal = max(df$LAI, na.rm = TRUE),         # mean annual LAI
    malcc = max(df$LCC, na.rm = TRUE)        # mean annual LCC
  )
})

print(site_climate)
```

```{r}
ggplot(site_climate, aes(x = map, y = mat, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       title = "Climate space across sites") +
  theme_classic()
```
Add in clustering groups:

```{r}
site_slopes <- site_slopes %>%
  left_join(site_climate, by = "site")
```

My own arbritrary groupings:
```{r}
ggplot(site_slopes, aes(x = map, y = mat, colour = sensitivity_group, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       colour = "VPD sensitivity", title = "Climate space across sites") +
  theme_classic()
```

Slope only clustering
```{r}
ggplot(site_slopes, aes(x = map, y = mat, colour = cluster_slope, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       colour = "VPD sensitivity", title = "Climate space across sites") +
  theme_classic()
```
Slope, intercept and threshold clustering

```{r}
ggplot(site_slopes, aes(x = map, y = mat, colour = cluster_multi, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual Temperature (°C)",
       colour = "VPD sensitivity", title = "Climate space across sites") +
  theme_classic()
```



```{r}
ggplot(site_climate, aes(x = map, y = mal, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual precipitation (mm)", y = "Mean Annual Leaf Area Index",
       title = "Mean annual LAI and precipitation") +
  theme_classic()
```

```{r}
ggplot(site_slopes, aes(x = map, y = mal, colour = sensitivity_group, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual LAI",
       colour = "VPD sensitivity", title = "Mean annual LAI and precip") +
  theme_classic()
```
```{r}
ggplot(site_slopes, aes(x = map, y = mal, colour = cluster_slope, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual LAI",
       colour = "VPD sensitivity", title = "Mean annual LAI and precip") +
  theme_classic()
```
```{r}
ggplot(site_slopes, aes(x = map, y = mal, colour = cluster_multi, label = site)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.3, size = 3) +
  labs(x = "Mean Annual Precipitation (mm)", y = "Mean Annual LAI",
       colour = "VPD sensitivity", title = "Mean annual LAI and precip") +
  theme_classic()
```
Whittaker biome plot

```{r}
library(plotbiomes)
```

```{r}
#convert precipitation to cm from mm

site_slopes$map_cm <- site_slopes$map / 10

library(plotbiomes)
library(ggplot2)

ggplot() +
  whittaker_base_plot() +
  geom_point(data = site_slopes,
             aes(x = mat, y = map_cm),
             size = 3) +
  theme_classic()
```

```{r}

ggplot() +
  whittaker_base_plot() +
  geom_point(data = site_slopes,
             aes(x = mat, y = map_cm, colour = cluster_slope),
             size = 3) +
  theme_classic() +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "white", colour = NA),
        plot.background = element_rect(fill = "white", colour = NA))
```
3 - high sensitivity
2 - moderate
1 - low

```{r}
ls("package:plotbiomes")
```

```{r}
ggsave("whittaker_plot.png", width = 10, height = 8, dpi = 300)

```

