---
title: "load_LAI"
output: html_document
date: "2026-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in LAI data
Howard Springs
Sturt plains
Daly River
Dry River
Alice Springs
Tumbarumba 
Wombat

Every 4 days
1 measurements a day per site

1. Convert date column into proper date column
2. Somehow attach the corresponding date from the LAI to the one I have in the flux data - every 4 days?

Load in data

```{r}
library(tidyverse)
```


```{r}
lai_data <- read.csv("/Users/alicerolfe/year_4_project/fluxnet_project/Data/LAI_2.csv")
```

Convert date to proper date format
```{r}
lai_data <- lai_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%dT%H:%M:%S"))
```

```{r}
head(lai_data)
```
Remove unwanted sites 
```{r}
lai_data <- lai_data %>%
  filter(!site %in% c("Litchfield", "Ti Tree East"))
```


Merge data for:
Howard Springs
Sturt plains
Daly River
Dry River
Alice Springs
Tumbarumba 
Wombat

```{r}
site_mapping <- data.frame(
  flux_site = c("How", "Stp", "Das", "Dry", "Asm", "Tum", "Wom"),
  lai_site = c("Howard Springs", "Sturt Plains", "Daly River Uncleared", "Dry River", "Alice Springs", "Tumbarumba", "Wombat")
)
```

```{r}
How_HH_daily <- How_HH_daily %>%
  mutate(site = "How")

Stp_HH_daily <- Stp_HH_daily %>%
  mutate(site = "Stp")

Das_HH_daily <- Das_HH_daily %>%
  mutate(site = "Das")

Dry_HH_daily <- Dry_HH_daily %>%
  mutate(site = "Dry")

Asm_HH_daily <- Asm_HH_daily %>%
  mutate(site = "Asm")

Tum_HH_daily <- Tum_HH_daily %>%
  mutate(site = "Tum")

Wom_HH_daily <- Wom_HH_daily %>%
  mutate(site = "Wom")
```


```{r}
# Convert flux data date to Date format (remove time component)
How_HH_daily <- How_HH_daily %>%
  mutate(date = as.Date(date))
```

```{r}
# Check it worked
class(How_HH_daily$date)
head(How_HH_daily$date)
```

```{r}
# Now do the merge again
How_lai <- How_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

# Check if it worked
summary(How_lai_merge$LAI)

```


```{r}
Stp_HH_daily <- Stp_HH_daily %>% mutate(date = as.Date(date))
Das_HH_daily <- Das_HH_daily %>% mutate(date = as.Date(date))
Dry_HH_daily <- Dry_HH_daily %>% mutate(date = as.Date(date))
Asm_HH_daily <- Asm_HH_daily %>% mutate(date = as.Date(date))
Tum_HH_daily <- Tum_HH_daily %>% mutate(date = as.Date(date))
Wom_HH_daily <- Wom_HH_daily %>% mutate(date = as.Date(date))
```

```{r}
Stp_lai <- Stp_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Das_lai <- Das_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Dry_lai <- Dry_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Asm_lai <- Asm_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Tum_lai <- Tum_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

Wom_lai <- Wom_HH_daily %>%
  left_join(site_mapping, by = c("site" = "flux_site")) %>%
  left_join(lai_data %>% select(date, site, mean),
            by = c("date" = "date", "lai_site" = "site")) %>%
  rename(LAI = mean)

```

Fill in missing days:
Going to use linear interpolation - assumes gradual change between measurements. 

```{r}
library(zoo)
```


```{r}
# Calculate max gap in terms of number of rows (if daily data, 4 days = 4 rows)
How_lai_merge <- How_lai_merge %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)
```

```{r}
Stp_lai <- Stp_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Das_lai <- Das_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Dry_lai<- Dry_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Asm_lai <- Asm_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Tum_lai<- Tum_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)

Wom_lai <- Wom_lai %>%
  mutate(LAI_interpolated = approx(date, LAI, xout = date, rule = 2)$y)
```

Create filter - max gap parameter tells it to only interpolate gaps of 4 or fewer observations. 4 days = 4 rows

```{r}
How_lai_merge <- How_lai_merge%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Stp_lai<- Stp_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Das_lai<- Das_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Dry_lai<- Dry_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Asm_lai<- Asm_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Tum_lai<- Tum_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

Wom_lai<- Wom_lai%>%
  arrange(date) %>%
  mutate(LAI_filled = na.approx(LAI, maxgap = 4, na.rm = FALSE))

```

```{r}
plot(Tum_lai$LAI_filled)
plot(Asm_lai$LAI_filled)
```

