---
title: "GAM_analysis"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

How does predictability vary across ecosystems in Australia, how does VPD affect predictability?

**1. Sequential order**
M1: Climate (precip, Tair, SW rad)
M2: Climate + VPD
M3a: Climate + VPD + lags 3
M3b: Climate + VPD + lags 5
M3c: Climate + VPD + lags 7
M4: Climate + VPD + best lag + cwd7
M5: Climate + VPD + best lag + cwd7 + LAI

**2. Directly compare**
- Model 1 + Model 2 (unique contribution of VPD)

Everything loaded in properly.

1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# List all sites
aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# Split, rolling CWD, lagged variables
for(site in aus_sites){
  
  df <- get(paste0(site, "_daily")) %>%
    arrange(date)  # ensure temporal order
  
  # Create lagged variables BEFORE splitting
  df <- df %>%
    mutate(
      Tair_lag7 = lag(Tair, 7),
      VPD_lag7 = lag(VPD, 7),
      cumulative_cwd_7day = rollapply(cwd, 7, sum, align="right", fill = NA)
    ) 
    #Removed the drop_na() as it was removing lots and lots of rows 
  
  # Temporal split: 80% train, 20% test
  n_train <- floor(0.8 * nrow(df))
  train <- df[1:n_train, ]
  test <- df[(n_train+1):nrow(df), ]
  
  # Save to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"), test, envir = .GlobalEnv)
}

```


```{r}

#Model 1 - climate only (Air temp, Shortwave radiation, precipitation)

fit_gam_m1 <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip)
  
  #Ade and Lox have very low test data - this prevents the GAM from crashing
   if(nrow(train_clean) < 10 | nrow(test_clean) < 10) {
    return(tibble(RMSE = NA, MAE = NA, R2 = NA))
  }
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  s     <- summary(model)                      
 resid <- test_clean$GPP - preds   
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  mae <- mean(abs(resid))
  R2_adj     = s$r.sq
  
  
  tibble(
    RMSE = rmse,
    R2   = rsq,
    MAE = mae,
    R2_adj_train = R2_adj
  )
}


# Run GAM for all sites and combine results

model_1_results <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_m1(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
for(site in aus_sites) {
  train <- get(paste0(site, "_train"))
  test  <- get(paste0(site, "_test"))
  
  train_clean <- train %>% drop_na(GPP, Tair, SW_rad, Precip)
  test_clean  <- test  %>% drop_na(GPP, Tair, SW_rad, Precip)
  
  cat(site, "- train rows:", nrow(train_clean), 
             "test rows:", nrow(test_clean), "\n")
}
```


```{r}
# Define wet to dry order
wet_dry_order <- c("Wac", "Rob", "How", "Das", "Fog", "Ade", "Dap", "Rdf", "Wom", "Tum", "Dry", "Stp", "Gin", "Cum", "Emr", "Rig", "Whr", "Asm", "Gww", "Cpr", "Ync", "Lox", "TTE")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

#highest max LAI to lowest max LAI
lai_order <- c("Tum", "Wac", "Wom", "Rob", "Rig", "Dap", "How", "Fog", "Dry", "Ade", "Rdf", "Das", "Cum", "Whr", "Emr", "Stp", "Gin", "Ync", "Lox", "Cpr", "TTE", "Asm", "Gww")

model_1_results$site <- factor(model_1_results$site, levels = wet_dry_order)

ggplot(model_1_results, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site (wet to dry)",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites"
  ) +
  theme_minimal()

```

```{r}
# visualise NA patterns
library(naniar)
vis_miss(sites_clean$Lox[, c("date", "GPP", "VPD", "Precip", "Tair", "SW_rad", "Rn")])
```
- I was removing NA rows in all of the variables. This meant if there were loads of NA in Rn, it deleted entire rows even when VPD/Tair/SW_rad were complete. This is why Gww was not doing very well.


Seems to be trend of declining R-squared from high precipitation coefficient of variation to low coefficient of variation. 

```{r}
#Model 2 - climate (temp, precip, SW rad) and VPD

fit_gam_m2 <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD)
  
  #Ade and Lox have very low test data - this prevents the GAM from crashing
   if(nrow(train_clean) < 10 | nrow(test_clean) < 10) {
    return(tibble(RMSE = NA, MAE = NA, R2 = NA))
  }
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  s     <- summary(model)                      
 resid <- test_clean$GPP - preds   
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  mae <- mean(abs(resid))
  R2_adj     = s$r.sq
  
  
  tibble(
    RMSE = rmse,
    R2   = rsq,
    MAE = mae,
    R2_adj_train = R2_adj
  )
}


# Run GAM for all sites and combine results

model_2_results <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_m2(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)
```

```{r}
model_2_results$site <- factor(model_2_results$site, levels = wet_dry_order)

ggplot(model_2_results, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites"
  ) +
  theme_minimal()
```

```{r}
#compare the no VPD and climate + VPD model

library(dplyr)

gam_comparison <- model_1_results %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    model_2_results %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_VPD - R2_no_vpd
  )

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```
Potentially here, we have the difference between wet sites and dry sites. In the wetter sites, current VPD improved model R-squared much more than drier sites. Could it be that lags matter more in dry sites? Cumulative water stress? Perhaps in wet sites, plants respond rapidly to VPD, explaining a lot more variation than in drier sites.

Seems to be a really mixed pattern here. 

```{r}
# Check data availability per site
map_dfr(names(sites_daily), function(site) {
  df <- sites_daily[[site]]
  tibble(
    site    = site,
    n_days  = nrow(df),
    n_na_gpp = sum(is.na(df$GPP)),
    n_na_vpd = sum(is.na(df$VPD)),
    pct_complete = mean(!is.na(df$GPP) & !is.na(df$VPD)) * 100
  )
}) %>% arrange(pct_complete)
```




```{r}
#Model 3a - climate (temp, SW rad, precip), VPD, + lags (7 days)

fit_gam_m3 <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, Tair_lag7, VPD_lag7)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, Tair_lag7, VPD_lag7)
  
  #Ade and Lox have very low test data - this prevents the GAM from crashing
   if(nrow(train_clean) < 10 | nrow(test_clean) < 10) {
    return(tibble(RMSE = NA, MAE = NA, R2 = NA))
  }
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) + s(Tair_lag7) + s(VPD_lag7),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  s     <- summary(model)                      
 resid <- test_clean$GPP - preds   
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  mae <- mean(abs(resid))
  R2_adj     = s$r.sq
  
  
  tibble(
    RMSE = rmse,
    R2   = rsq,
    MAE = mae,
    R2_adj_train = R2_adj
  )
}


# Run GAM for all sites and combine results

model_3_results <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_m3(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}

model_3_results$site <- factor(model_3_results$site, levels = wet_dry_order)

ggplot(model_3_results, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}

library(dplyr)

lag_comparison <- model_1_results  %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    model_3_results %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )

lag_comparison <- lag_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

lag_comparison
```

```{r}
ggplot(lag_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 3 day climate lags across OzFlux sites"
  )

```
**For now only going to use Lag 7**

#Model 3b - climate + VPD + climate lags 5 days

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag5, Tair_lag5)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag5, Tair_lag5)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) +s(VPD_lag5) + s(Tair_lag5),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag5_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

lag5_metrics$site <- factor(lag5_metrics$site, levels = wet_dry_order)

ggplot(lag5_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()


#Compare no lags and lags with 5 days

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison


ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 5 day climate lags across OzFlux sites"
  )

#Model 3c - climate + VPD + 7 day climate lags

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, Tair_lag7)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, Tair_lag7)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) +s(VPD_lag7) + s(Tair_lag7),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


lag7_metrics$site <- factor(lag7_metrics$site, levels = wet_dry_order)

ggplot(lag7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()

#compare no lags, 3 day lags, 5 day and 7 day (Model 3a, b, c):

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag7_metrics %>%
      select(site, R2) %>%
      rename(R2_lag7 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag5 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(R2_lag3 = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag3 = R2_lag3 - R2_no_lags,
    delta_R2_lag5 = R2_lag5 - R2_no_lags,
    delta_R2_lag7 = R2_lag7 - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison


Let's use lag 7 - similar performances. 7 days aligns with typical plant physiological memory and maintains consistency with the cumulative water deficit over 7 days.

```{r}
#Model 4: Climate + cwd + lags + VPD

fit_gam_m4 <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, Tair_lag7, VPD_lag7, cumulative_cwd_7day)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, Tair_lag7, VPD_lag7, cumulative_cwd_7day)
  
  #Ade and Lox have very low test data - this prevents the GAM from crashing
   if(nrow(train_clean) < 10 | nrow(test_clean) < 10) {
    return(tibble(RMSE = NA, MAE = NA, R2 = NA))
  }
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag7) + s(Tair_lag7) + s(cumulative_cwd_7day),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  s     <- summary(model)                      
 resid <- test_clean$GPP - preds   
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  mae <- mean(abs(resid))
  R2_adj     = s$r.sq
  
  
  tibble(
    RMSE = rmse,
    R2   = rsq,
    MAE = mae,
    R2_adj_train = R2_adj
  )
}


# Run GAM for all sites and combine results

model_4_results <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_m4(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```


```{r}
model_4_results$site <- factor(model_4_results$site, levels = wet_dry_order)

ggplot(model_4_results, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```


```{r}
#Model 5: Climate + cwd + lags + VPD + LCC

fit_gam_m5 <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, Tair_lag7, VPD_lag7, cumulative_cwd_7day, LAI_step)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, Tair_lag7, VPD_lag7, cumulative_cwd_7day, LAI_step)
  
  #Ade and Lox have very low test data - this prevents the GAM from crashing
   if(nrow(train_clean) < 10 | nrow(test_clean) < 10) {
    return(tibble(RMSE = NA, MAE = NA, R2 = NA))
  }
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag7) + s(Tair_lag7) + s(cumulative_cwd_7day) + s(LAI_step),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  s     <- summary(model)                      
 resid <- test_clean$GPP - preds   
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  mae <- mean(abs(resid))
  R2_adj     = s$r.sq
  
  
  tibble(
    RMSE = rmse,
    R2   = rsq,
    MAE = mae,
    R2_adj_train = R2_adj
  )
}


# Run GAM for all sites and combine results

model_5_results <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_m5(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```


```{r}
model_5_results$site <- factor(model_5_results$site, levels = wet_dry_order)

ggplot(model_5_results, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#table of all models R-squared 

final_comparison <- model_2_results %>%
  select(site, R2) %>%
  rename(model_2 = R2) %>%
  
  inner_join(
    model_3_results %>%
      select(site, R2) %>%
      rename(model_3 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    model_1_results %>%
      select(site, R2) %>%
      rename(model_1 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    model_5_results %>%
      select(site, R2) %>%
      rename(model_5 = R2),
    by = "site"
  ) %>%
  
   inner_join(
    model_4_results %>%
      select(site, R2) %>%
      rename(model_4 = R2),
    by = "site"
  ) 
  
  
final_comparison <- final_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

final_comparison
```
convert wide to long 
```{r}
library(tidyverse)

r2_long <- final_comparison %>%
  pivot_longer(
    cols = starts_with("model_"),
    names_to = "model",
    values_to = "r2"
  )
```


```{r}
r2_long <- r2_long %>%
  mutate(model_num = as.numeric(gsub("model_", "", model)))
```

```{r}
ggplot(r2_long, aes(x = model_num, y = r2, group = site, colour = site)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  labs(
    x = "Model",
    y = expression(R^2),
    colour = "Site"
  ) +
  theme_minimal()
```

```{r}
final_comparison %>%
  pivot_longer(
    cols = starts_with("model_"),
    names_to = "model",
    values_to = "R2"
  ) %>%
  mutate(
    model = factor(model,
                   levels = c("model_1", "model_2", "model_3", 
                             "model_4", "model_5"),
                   labels = c("Climate", "+VPD", "+Lag7", "+CWD", "+LAI"))
  ) %>%
  ggplot(aes(x = model, y = site, fill = R2)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Model",
    y = "Site (Wet → Dry)",
    fill = expression(R^2),
    title = "Model performance across sites and complexity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Create a figure showing best model per site
final_comparison %>%
  pivot_longer(cols = starts_with("model_"), 
               names_to = "model", values_to = "R2") %>%
  group_by(site) %>%
  slice_max(R2, n = 1) %>%
  mutate(
    moisture = case_when(
      site %in% c("Rob", "Wac", "How", "Das", "Fog", "Rdf", "Dap", "Ade") ~ "Wet",
      site %in% c("Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox") ~ "Dry",
      TRUE ~ "Medium"
    )
  ) %>%
  ggplot(aes(x = site, y = model, fill = R2)) +
  geom_tile() +
  facet_grid(~moisture, scales = "free_x", space = "free_x") +
  scale_fill_viridis_c() +
  labs(title = "Best-performing model varies by moisture regime")
```

```{r}
library(dplyr)
library(lubridate)

convert_to_annual <- function(df_daily) {

  # Add year column
  df_daily <- df_daily %>%
    mutate(year = year(date))

df_annual <- df_daily %>%
  group_by(year) %>%
  summarise(
    GPP = sum(GPP, na.rm = TRUE),
    GPP_n_na = sum(is.na(GPP)),
    ET = sum(ET, na.rm = TRUE),
    ET_n_na = sum(is.na(ET)),
    Precip = sum(Precip, na.rm = TRUE),
    Precip_n_na = sum(is.na(Precip)),
    Tair = mean(Tair, na.rm = TRUE),
    VPD = mean(VPD, na.rm = TRUE),
    LCC = mean(LCC_daily, na.rm = TRUE),
    .groups = "drop"
  )

  return(df_annual)
}

# apply to all sites
sites_annual <- map(sites_daily, convert_to_annual)

# calculate mean annual values per site
site_climate <- map_dfr(names(sites_annual), function(site) {
  df <- sites_annual[[site]]
  
  tibble(
    site = str_remove(site, "_HH_daily"),
    mat = mean(df$Tair, na.rm = TRUE),       # mean annual temperature
    map = mean(df$Precip, na.rm = TRUE),      # mean annual precipitation
    mal = mean(df$LAI, na.rm = TRUE),         # mean annual LAI
    malcc = mean(df$LCC, na.rm = TRUE)        # mean annual LCC
  )
})

print(site_climate)
```
```{r}
site_climate %>%
  arrange(map) %>%
  mutate(site = factor(site, levels = site)) %>%
  ggplot(aes(x = site, y = map)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "Site", y = "Mean Annual Precipitation (mm)",
       title = "Precipitation gradient across OzFlux sites") +
  theme_minimal()
```



