---
title: "GAM_analysis"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

How does predictability vary across ecosystems in Australia, how does VPD affect predictability?

**1. Sequential order**
M1: Climate (precip, Tair, SW rad)
M2: Climate + VPD
M3a: Climate + VPD + lags 3
M3b: Climate + VPD + lags 5
M3c: Climate + VPD + lags 7
M4: Climate + VPD + best lag + cwd7
M5: Climate + VPD + best lag + cwd7 + LCC

**2. Directly compare**
- Model 1 + Model 2 (unique contribution of VPD)


Everything loaded in properly.

1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# List all sites
aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# Split, rolling CWD, lagged variables
for(site in aus_sites){
  
  df <- get(paste0(site, "_HH_daily")) %>%
    arrange(date)  # ensure temporal order
  
  # Create lagged variables BEFORE splitting
  df <- df %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      Tair_lag5 = lag(Tair, 5),
      Tair_lag7 = lag(Tair, 7),
      VPD_lag3 = lag(VPD, 3),
      VPD_lag5 = lag(VPD, 5),
      VPD_lag7 = lag(VPD, 7),
      cumulative_cwd_7day = rollapply(cwd, 7, sum, align="right", fill = NA)
    ) %>%
    #Removed the drop_na() as it was removing lots and lots of rows 
  
  # Temporal split: 80% train, 20% test
  n_train <- floor(0.8 * nrow(df))
  train <- df[1:n_train, ]
  test <- df[(n_train+1):nrow(df), ]
  
  # Save to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"), test, envir = .GlobalEnv)
}

```


```{r}

#Model 1 - climate only (Air temp, Shortwave radiation, precipitation)

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```



# Modified function to return predictions for both train and test data
fit_gam_and_return_all_predictions <- function(train_data, test_data, site_name) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip),
    data = train_data,
    method = "REML"
  )
  
  # Predictions for training data
  train_preds <- train_data %>%
    mutate(
      predicted_GPP = predict(model, newdata = train_data),
      site = site_name,
      dataset = "Training"
    )
  
  # Predictions for test data
  test_preds <- test_data %>%
    mutate(
      predicted_GPP = predict(model, newdata = test_data),
      site = site_name,
      dataset = "Testing"
    )
  
  # Combine both
  bind_rows(train_preds, test_preds)
}

# Get predictions for the three sites
sites_to_plot <- "Gww"

all_predictions_df <- bind_rows(
  lapply(sites_to_plot, function(site) {
    fit_gam_and_return_all_predictions(
      get(paste0(site, "_train")),
      get(paste0(site, "_test")),
      site
    )
  })
)


# Shading to show train/test periods
ggplot(all_predictions_df, aes(x = date)) +
  geom_rect(
    data = all_predictions_df %>% 
      group_by(site) %>%
      summarise(
        xmin = min(date[dataset == "Testing"]),
        xmax = max(date[dataset == "Testing"])
      ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "gray90", alpha = 0.5, inherit.aes = FALSE
  ) +
  geom_line(aes(y = GPP, color = "Observed"), linewidth = 0.7) +
  geom_line(aes(y = predicted_GPP, color = "Predicted"), linewidth = 0.7, linetype = "dashed") +
  facet_wrap(~site, ncol = 1, scales = "free_y") +
  scale_color_manual(
    values = c("Observed" = "#2c5f2d", "Predicted" = "#e65100"),
    name = ""
  ) +
  labs(
    x = "Date",
    y = "GPP",
    title = "Observed vs Predicted GPP: Training and Testing Periods",
    subtitle = "Gray shading indicates testing period"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )



```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_metrics$site <- factor(climate_metrics$site, levels = wet_dry_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (least to most daily precip variability)"
  ) +
  theme_minimal()

```

```{r}
# visualise NA patterns
library(naniar)
vis_miss(Gww_HH_daily[, c("date", "GPP", "VPD", "Precip", "Tair", "SW_rad", "Rn")])
```
- I was removing NA rows in all of the variables. This meant if there were loads of NA in Rn, it deleted entire rows even when VPD/Tair/SW_rad were complete. This is why Gww was not doing very well.


Seems to be trend of declining R-squared from high precipitation coefficient of variation to low coefficient of variation. 

```{r}
#Model 2 - climate (temp, precip, SW rad) and VPD

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_vpd_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_vpd_metrics$site <- factor(climate_vpd_metrics$site, levels = wet_dry_order)

ggplot(climate_vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (least to most precip variability)"
  ) +
  theme_minimal()
```

```{r}
mean(climate_vpd_metrics$RMSE)
```


```{r}
#compare the no VPD and climate + VPD model

library(dplyr)

gam_comparison <- climate_metrics %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    climate_vpd_metrics %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_VPD - R2_no_vpd
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```
Potentially here, we have the difference between wet sites and dry sites. In the wetter sites, current VPD improved model R-squared much more than drier sites. Could it be that lags matter more in dry sites? Cumulative water stress? Perhaps in wet sites, plants respond rapidly to VPD, explaining a lot more variation than in drier sites.

Seems to be a really mixed pattern here. 



```{r}
#Model 3a - climate (temp, SW rad, precip), VPD, + lags (3 days)

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag3, Tair_lag3)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag3, Tair_lag3)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) +s(VPD_lag3) +s(Tair_lag3),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag3_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```

```{r}

lag3_metrics$site <- factor(lag3_metrics$site, levels = wet_dry_order)

ggplot(lag3_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}

library(dplyr)

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 3 day climate lags across OzFlux sites"
  )

```


```{r}
#Model 3b - climate + VPD + climate lags 5 days

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag5, Tair_lag5)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag5, Tair_lag5)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) +s(VPD_lag5) + s(Tair_lag5),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag5_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```


```{r}

lag5_metrics$site <- factor(lag5_metrics$site, levels = wet_dry_order)

ggplot(lag5_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#Compare no lags and lags with 5 days

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 5 day climate lags across OzFlux sites"
  )

```

```{r}
#Model 3c - climate + VPD + 7 day climate lags

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, Tair_lag7)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, Tair_lag7)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) +s(VPD_lag7) + s(Tair_lag7),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```


```{r}

lag7_metrics$site <- factor(lag7_metrics$site, levels = wet_dry_order)

ggplot(lag7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#compare no lags, 3 day lags, 5 day and 7 day (Model 3a, b, c):

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag7_metrics %>%
      select(site, R2) %>%
      rename(R2_lag7 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag5 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(R2_lag3 = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag3 = R2_lag3 - R2_no_lags,
    delta_R2_lag5 = R2_lag5 - R2_no_lags,
    delta_R2_lag7 = R2_lag7 - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

Let's use lag 7 - similar performances. 7 days aligns with typical plant physiological memory and maintains consistency with the cumulative water deficit over 7 days.

```{r}
#Model 4: Climate + cwd + lags + VPD

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, cumulative_cwd_7day, Tair_lag7)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, cumulative_cwd_7day, Tair_lag7)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) + s(VPD) +s(VPD_lag7) +s(Tair_lag7) + s(cumulative_cwd_7day),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

cwd7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```


```{r}
cwd7_metrics$site <- factor(cwd7_metrics$site, levels = wet_dry_order)

ggplot(cwd7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```


```{r}
#Model 5: Climate + cwd + lags + VPD + LCC

fit_gam_and_metrics <- function(train_data, test_data) {
  
  #Drop NA only in the predictors of the model
  train_clean <- train_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, cumulative_cwd_7day, Tair_lag7, LCC_daily)
  test_clean <- test_data %>% drop_na(GPP, Tair, SW_rad, Precip, VPD, VPD_lag7, cumulative_cwd_7day, Tair_lag7, LCC_daily)
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) +s(VPD) + s(VPD) +s(VPD_lag7) +s(Tair_lag7) + s(cumulative_cwd_7day) +s(LCC_daily),
    data = train_clean,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_clean)
  
  rmse <- sqrt(mean((test_clean$GPP - preds)^2))
  rsq  <- cor(test_clean$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lcc7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```


```{r}
lcc7_metrics$site <- factor(lcc7_metrics$site, levels = wet_dry_order)

ggplot(lcc7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#table of all models R-squared 

final_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(model_2 = R2) %>%
  
  inner_join(
    lag7_metrics %>%
      select(site, R2) %>%
      rename(model_3c = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(model_3b = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(model_3a = R2),
    by = "site"
  ) %>%
  
   inner_join(
    climate_metrics %>%
      select(site, R2) %>%
      rename(model_1 = R2),
    by = "site"
  ) %>%
  
   inner_join(
    lcc7_metrics %>%
      select(site, R2) %>%
      rename(model_5 = R2),
    by = "site"
  ) %>%
  
   inner_join(
    cwd7_metrics %>%
      select(site, R2) %>%
      rename(model_4 = R2),
    by = "site"
  ) 
  


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

final_comparison <- final_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

final_comparison
```

```{r}
final_comparison %>%
  pivot_longer(
    cols = starts_with("model_"),
    names_to = "model",
    values_to = "R2"
  ) %>%
  mutate(
    model = factor(model,
                   levels = c("model_1", "model_2", "model_3a", 
                             "model_3b", "model_3c", "model_4", "model_5"),
                   labels = c("Climate", "+VPD", "+Lag3", "+Lag5", 
                             "+Lag7", "+CWD", "+LCC"))
  ) %>%
  ggplot(aes(x = model, y = site, fill = R2)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Model",
    y = "Site (Wet → Dry)",
    fill = expression(R^2),
    title = "Model performance across sites and complexity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Create a figure showing best model per site
final_comparison %>%
  pivot_longer(cols = starts_with("model_"), 
               names_to = "model", values_to = "R2") %>%
  group_by(site) %>%
  slice_max(R2, n = 1) %>%
  mutate(
    moisture = case_when(
      site %in% c("Rob", "Wac", "How", "Das", "Fog", "Rdf", "Dap", "Ade") ~ "Wet",
      site %in% c("Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox") ~ "Dry",
      TRUE ~ "Medium"
    )
  ) %>%
  ggplot(aes(x = site, y = model, fill = R2)) +
  geom_tile() +
  facet_grid(~moisture, scales = "free_x", space = "free_x") +
  scale_fill_viridis_c() +
  labs(title = "Best-performing model varies by moisture regime")
```




