---
title: "GAM_analysis_2"
output: html_document
date: "2026-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Subsection 1: VPD across Australia
- Driver importance across sites
- **GAMs - predictability**
- GPP~VPD responses

Now I am going to look at the generalisability of the sites
- Leave one site out (LOSO) across all 23 sites - has the GAM learned something ecologically rather than just site-specific quirks?
- Do complex models (Tair, VPD, SW_rad, Precip, VPD lag, Tair lag, cwd, LCC) overfit to site conditions?

For each of the 23 sites, it trains on the remaining 22 and predicts on the held out site. 
I have done this for a simple model (Tair, SW_rad and VPD) and the more complex 8 predictor version.

```{r}

# Leave-One-Site-Out (LOSO) Cross-Validation for GAM Generalisability
# Compares:
# Simple model:  GPP ~ s(Tair) + s(SW_rad) + s(VPD)
# Complex model: GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) +
#                        s(VPD_lag3) + s(Tair_lag3) + s(cumulative_cwd_7day) +
#                        s(LAI_step)


library(tidyverse)
library(mgcv)

# Site list and PFT lookup


aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww",
               "Rdf", "Rig", "Rob", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# Assign PFTs - adjust if your classification differs
pft_lookup <- tibble(
  site = aus_sites,
  pft  = c("WSA", "WSA", "SAV", "GRA", "SAV", "GRA", "SAV",
            "EBF", "GRA", "WET", "WSA", "SAV",
            "WSA", "GRA", "EBF", "EBF", "EBF",
            "EBF", "GRA", "SAV", "EBF")
)

# -----------------------------------------------------------------------------
# 1. Combine all site data into a single pooled dataframe
#    (assumes each site's processed dataframe is already in the global env,
#     named e.g. How_HH_daily, Ade_HH_daily, etc., with lagged variables
#     and cumulative_cwd_7day already computed as in your existing pipeline)
# -----------------------------------------------------------------------------
# -----------------------------------------------------------------------------
# 1. Combine all site data - NO drop_na() here!
# -----------------------------------------------------------------------------

all_sites_df <- bind_rows(
  lapply(aus_sites, function(site) {
    get(paste0(site, "_daily")) %>%
      arrange(date) %>%
      mutate(
        Tair_lag3           = lag(Tair, 3),
        Tair_lag5           = lag(Tair, 5),
        Tair_lag7           = lag(Tair, 7),
        VPD_lag3            = lag(VPD, 3),
        VPD_lag5            = lag(VPD, 5),
        VPD_lag7            = lag(VPD, 7),
        cumulative_cwd_7day = zoo::rollapply(cwd, 7, sum, align = "right", fill = NA)
      ) %>%
      # NO drop_na() here! Keep all rows with any valid data
      mutate(site = site)
  })
) %>%
  left_join(pft_lookup, by = "site")

# -----------------------------------------------------------------------------
# 2. Model formula definitions
# -----------------------------------------------------------------------------

formula_simple <- GPP ~ s(Tair) + s(SW_rad) + s(VPD)

formula_complex <- GPP ~ s(Tair) + s(SW_rad) + s(Precip) +
  s(VPD) + s(VPD_lag7) + s(Tair_lag7) +
  s(cumulative_cwd_7day) + s(LAI_step)

# -----------------------------------------------------------------------------
# 3. LOSO with model-specific NA removal
# -----------------------------------------------------------------------------

run_loso <- function(data, formula, model_name, required_vars) {

  sites <- unique(data$site)

  results <- map_dfr(sites, function(held_out) {

    # Filter NAs only for variables needed by THIS model
    train <- data %>% 
      filter(site != held_out) %>%
      drop_na(all_of(required_vars))
    
    test  <- data %>% 
      filter(site == held_out) %>%
      drop_na(all_of(required_vars))

    # Fit on all other sites
    model <- tryCatch(
      gam(formula, data = train, method = "REML"),
      error = function(e) {
        message("Model failed for holdout site: ", held_out, " — ", e$message)
        return(NULL)
      }
    )

    if (is.null(model)) {
      return(tibble(
        held_out_site = held_out,
        model         = model_name,
        R2            = NA_real_,
        RMSE          = NA_real_,
        n_test        = nrow(test)
      ))
    }

    preds <- predict(model, newdata = test)
    resid <- test$GPP - preds

    r2   <- cor(test$GPP, preds, use = "complete.obs")^2
    rmse <- sqrt(mean(resid^2, na.rm = TRUE))

    tibble(
      held_out_site = held_out,
      model         = model_name,
      R2            = r2,
      RMSE          = rmse,
      n_test        = nrow(test)
    )
  })

  return(results)
}

# -----------------------------------------------------------------------------
# 4. Run LOSO with specific variable lists for each model
# -----------------------------------------------------------------------------

# Define required variables for each model
vars_simple <- c("GPP", "Tair", "SW_rad", "VPD")

vars_complex <- c("GPP", "Tair", "SW_rad", "Precip", "VPD", 
                  "VPD_lag7", "Tair_lag7", "cumulative_cwd_7day", "LAI_step")

cat("Running LOSO for simple model...\n")
loso_simple  <- run_loso(all_sites_df, formula_simple, "Simple", vars_simple)

cat("Running LOSO for complex model...\n")
loso_complex <- run_loso(all_sites_df, formula_complex, "Complex", vars_complex)

loso_results <- bind_rows(loso_simple, loso_complex) %>%
  left_join(pft_lookup, by = c("held_out_site" = "site")) %>%
  mutate(held_out_site = factor(held_out_site, levels = wet_dry_order))

# -----------------------------------------------------------------------------
# 5. Summary statistics
# -----------------------------------------------------------------------------

loso_summary <- loso_results %>%
  group_by(model) %>%
  summarise(
    mean_R2   = mean(R2,   na.rm = TRUE),
    median_R2 = median(R2, na.rm = TRUE),
    sd_R2     = sd(R2,     na.rm = TRUE),
    mean_RMSE = mean(RMSE, na.rm = TRUE),
    sd_RMSE   = sd(RMSE,   na.rm = TRUE),
    .groups   = "drop"
  )

cat("\n--- LOSO Summary ---\n")
print(loso_summary)

# Summary by PFT
loso_by_pft <- loso_results %>%
  group_by(model, pft) %>%
  summarise(
    mean_R2   = mean(R2,   na.rm = TRUE),
    sd_R2     = sd(R2,     na.rm = TRUE),
    n_sites   = n(),
    .groups   = "drop"
  )

cat("\n--- LOSO Summary by PFT ---\n")
print(loso_by_pft)

# -----------------------------------------------------------------------------
# 6. Delta R² (complex - simple) per site
# -----------------------------------------------------------------------------

loso_wide <- loso_results %>%
  select(held_out_site, model, R2, RMSE, pft) %>%
  pivot_wider(names_from = model, values_from = c(R2, RMSE)) %>%
  mutate(
    delta_R2   = R2_Complex   - R2_Simple,
    delta_RMSE = RMSE_Complex - RMSE_Simple  # negative = complex is better
  )

cat("\n--- Delta R² (Complex - Simple) per site ---\n")
print(loso_wide %>% select(held_out_site, pft, R2_Simple, R2_Complex, delta_R2))

```

```{r}


# -----------------------------------------------------------------------------
# 7. Plots
# -----------------------------------------------------------------------------

# --- 7a. R² by site for both models (wet → dry) ---
p1 <- loso_results %>%
  ggplot(aes(x = held_out_site, y = R2, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  scale_fill_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Held-out site (Wet → Dry)",
    y     = expression(R^2),
    fill  = "Model",
    title = "LOSO Generalisation: Simple vs. Complex GAM"
  )

print(p1)

# --- 7b. Delta R² per site, coloured by PFT ---
p2 <- loso_wide %>%
  ggplot(aes(x = held_out_site, y = delta_R2, fill = pft)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Held-out site (Wet → Dry)",
    y     = expression(Delta * R^2 ~ "(Complex - Simple)"),
    fill  = "PFT",
    title = "Does added complexity help generalisation?",
    subtitle = "Positive = complex model generalises better; negative = overfitting"
  )

print(p2)

# --- 7c. Boxplot comparing model distributions ---
p3 <- loso_results %>%
  ggplot(aes(x = model, y = R2, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(colour = pft), width = 0.15, size = 2.5, alpha = 0.8) +
  scale_fill_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  scale_colour_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    x      = "Model",
    y      = expression(R^2),
    fill   = "Model",
    colour = "PFT",
    title  = "Distribution of LOSO R² across all sites"
  )

print(p3)

# --- 7d. R² by PFT ---
p4 <- loso_results %>%
  ggplot(aes(x = pft, y = R2, fill = model)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.75)) +
  geom_jitter(aes(colour = model),
              position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
              size = 2, alpha = 0.8) +
  scale_fill_manual(values   = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  scale_colour_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Plant Functional Type",
    y     = expression(R^2),
    fill  = "Model",
    colour = "Model",
    title = "LOSO generalisation by PFT"
  )

print(p4)

```

```{r}

# -----------------------------------------------------------------------------
# 8. Save results tables
# -----------------------------------------------------------------------------

write_csv(loso_results, "loso_results_all_sites.csv")
write_csv(loso_wide,    "loso_results_wide.csv")
write_csv(loso_summary, "loso_summary.csv")
write_csv(loso_by_pft,  "loso_by_pft.csv")

cat("\nDone. Results saved to CSV.\n")
```

**Results:**
Gin and Dap consistently perform well, both within site and out of site.
Ade, Wom, Whr also perform well

PFT:
Gin = WSA
Dap = GRA
Ade = WSA
Wom = EBF
Whr = EBF

Precip variability:
- All sites on the lower end of variability, Whr and Wom more variable than Rdf

Precip:
- All these sites are intermediate mean annual precip, Whr is pretty dry

Years active:
Gin = 3
Dap = 6
Ade = 2
Wom = 4
Whr = 3

Not even necessarily the longest running sites.

The common thread here is that these sites have intermediate precipitation, with intermediate precipitation variability. 
Sites that are intermediate are both easily predictable and generalisable

Cpr, Rdf and Gww consistently performed not very well, both for within site and out of site GAM testing.
Cpr = SAV
Gww = SAV
Rdf = WSA

In terms of precipitation variability (daily) - coefficient of variance. Gww is the most variable, Cpr also quite variable. Rdf isn't that variable in terms of daily variability.
Rdf, Gww all in the middle precip and Cpr is on the drier end.

In terms of fluxnet years active:
Cpr = 4
Gww = 1
Rdf = 2

```{r}
loso_wide$held_out_site <- factor(loso_wide$held_out_site, levels = wet_dry_order)

ggplot(loso_wide, aes(x = held_out_site, y = R2_Simple)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site (Wet to Dry)",
    y = expression(R^2),
    title = "Generalisability of sites across OzFlux (simple model)"
  ) +
  theme_minimal()
```

```{r}
ggplot(loso_wide, aes(x = held_out_site, y = R2_Complex)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site (Wet to Dry)",
    y = expression(R^2),
    title = "Generalisability of sites across OzFlux (complex model)"
  ) +
  theme_minimal()
```

These plots confirm that site predictability and generalisability is controlled by precipitation variability. 

Next steps:
- Look at wet/dry season
- 7 day cwd might be too short for high precipitation variability sites
- interaction terms?
- Identify 3 highest CV sites and 3 lowest CV sites, plot actual versus predicted GPP time series - am i missing pulses? overestimating during drought?

```{r}
library(tidyverse)

# 1. Prepare data for plotting
# Add a 'Model' column to each dataset for clarity

loso_simple_plot <- loso_results %>%
  filter(model == "Simple") %>%
  select(site = held_out_site, R2) %>%
  mutate(Model = "LOSO Simple")

loso_complex_plot <- loso_results %>%
  filter(model == "Complex") %>%
  select(site = held_out_site, R2) %>%
  mutate(Model = "LOSO Complex")

full_model_plot <- model_5_results %>%
  select(site, R2) %>%
  mutate(Model = "Within-site model 5")

# Combine into one dataframe
plot_df <- bind_rows(loso_simple_plot, loso_complex_plot, full_model_plot)

# Ensure consistent site ordering
plot_df$site <- factor(plot_df$site, levels = wet_dry_order)

# 2. Plot
ggplot(plot_df, aes(x = site, y = R2, fill = Model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("LOSO Simple" = "steelblue",
                               "LOSO Complex" = "firebrick",
                               "Within-site model 5" = "darkgreen")) +
  theme_minimal(base_size = 13) +
  labs(
    x = "Site (Wet → Dry)",
    y = expression(R^2),
    fill = "Model",
    title = "Comparison of Model Predictive Performance Across Sites"
  )
```
Test whether VPD significantly improves model performance and whether VPD improves generalisation across sites
- LOSO no vpd
- LOSO vpd

```{r}
library(mgcv)
library(dplyr)
library(tidyr)
library(ggplot2)

# ------------------------
# 1. LOSO function with NA handling
# ------------------------
run_loso <- function(data, formula, model_name) {
  
  # Extract variable names from formula (including response)
  vars <- all.vars(formula)
  
  sites <- unique(data$site)
  results <- data.frame(
    held_out_site = character(),
    R2 = numeric(),
    model = character(),
    stringsAsFactors = FALSE
  )
  
  for (s in sites) {
    # Filter data and drop rows with NAs in any variables used in the model
    train <- data %>% filter(site != s) %>% drop_na(all_of(vars))
    test  <- data %>% filter(site == s) %>% drop_na(all_of(vars))
    
    # Check if training or test set is empty
    if(nrow(train) < 5 | nrow(test) < 5) {
      # Not enough data to fit or test the model
      results <- rbind(results, data.frame(held_out_site = s, R2 = NA, model = model_name))
      next
    }
    
    # Fit GAM
    gam_mod <- gam(formula, data = train)
    
    # Predict on test set
    pred <- predict(gam_mod, newdata = test)
    
    # Calculate R² safely, only using complete cases
    complete <- complete.cases(pred, test$GPP)
    r2 <- 1 - sum((test$GPP[complete] - pred[complete])^2) / sum((test$GPP[complete] - mean(test$GPP[complete]))^2)
    
    # Store result
    results <- rbind(results, data.frame(held_out_site = s, R2 = r2, model = model_name))
  }
  
  return(results)
}

# ------------------------
# 2. Formulas
# ------------------------
formula_noVPD <- GPP ~ s(Tair) + s(SW_rad) + s(Precip)
formula_withVPD <- GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD)

# ------------------------
# 3. Run LOSO GAMs
# ------------------------
loso_noVPD <- run_loso(all_sites_df, formula_noVPD, "No VPD")
loso_withVPD <- run_loso(all_sites_df, formula_withVPD, "With VPD")

loso_combined <- bind_rows(loso_noVPD, loso_withVPD)

# ------------------------
# 4. Calculate delta R2 per site
# ------------------------
loso_delta <- loso_combined %>%
  pivot_wider(names_from = model, values_from = R2) %>%
  mutate(delta_R2 = `With VPD` - `No VPD`)

# ------------------------
# 5. Paired statistical test
# ------------------------
t_test_result <- t.test(loso_delta$`With VPD`, loso_delta$`No VPD`, paired = TRUE)
wilcox_test_result <- wilcox.test(loso_delta$`With VPD`, loso_delta$`No VPD`, paired = TRUE)

print(t_test_result)
print(wilcox_test_result)

# ------------------------
# 6. Plot delta R2
# ------------------------
ggplot(loso_delta, aes(x = reorder(held_out_site, delta_R2), y = delta_R2, fill = delta_R2 > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick"), guide = "none") +
  labs(
    x = "Held-out site",
    y = expression(Delta*R^2),
    title = "Effect of VPD on LOSO GAM Generalisability"
  ) +
  theme_minimal()
```

```{r}
loso_results %>%
  filter(is.na(R2)) %>%
  select(held_out_site, model)
```

```{r}

# Ensure your data has factors for ordering
loso_combined$site <- factor(loso_combined$held_out_site, 
                             levels = loso_delta$held_out_site[order(loso_delta$delta_R2, decreasing = TRUE)])

# Plot
ggplot(loso_combined, aes(x = R2, y = site, color = model)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  labs(
    x = expression(R^2),
    y = "Site",
    title = "LOSO GAM R² per site with and without VPD"
  ) +
  scale_color_manual(values = c("No VPD" = "firebrick", "With VPD" = "steelblue")) +
  theme_classic()
```

