---
title: "GAM_analysis_2"
output: html_document
date: "2026-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Subsection 1: VPD across Australia
- Driver importance across sites
- **GAMs - predictability**
- GPP~VPD responses

Now I am going to look at the generalisability of the sites
- Leave one site out (LOSO) across all 23 sites - has the GAM learned something ecologically rather than just site-specific quirks?
- Do complex models (Tair, VPD, SW_rad, Precip, VPD lag, Tair lag, cwd, LCC) overfit to site conditions?

For each of the 23 sites, it trains on the remaining 22 and predicts on the held out site. 
I have done this for a simple model (Tair, SW_rad and VPD) and the more complex 8 predictor version.

```{r}

# Leave-One-Site-Out (LOSO) Cross-Validation for GAM Generalisability
# Compares:
# Simple model:  GPP ~ s(Tair) + s(SW_rad) + s(VPD)
# Complex model: GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) +
#                        s(VPD_lag3) + s(Tair_lag3) + s(cumulative_cwd_7day) +
#                        s(LCC_daily)


library(tidyverse)
library(mgcv)

# Site list and PFT lookup


aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# Assign PFTs - adjust if your classification differs
pft_lookup <- tibble(
  site = aus_sites,
  pft  = c("WSA", "WSA", "SAV", "GRA", "SAV", "GRA", "SAV",
            "EBF", "GRA", "WET", "WSA", "SAV", "DBF",
            "WSA", "GRA", "EBF", "GRA", "EBF", "EBF",
            "EBF", "GRA", "SAV", "EBF")
)

# Wet-to-dry ordering for plots 
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry",
                   "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww",
                   "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

# -----------------------------------------------------------------------------
# 1. Combine all site data into a single pooled dataframe
#    (assumes each site's processed dataframe is already in the global env,
#     named e.g. How_HH_daily, Ade_HH_daily, etc., with lagged variables
#     and cumulative_cwd_7day already computed as in your existing pipeline)
# -----------------------------------------------------------------------------

all_sites_df <- bind_rows(
  lapply(aus_sites, function(site) {
    get(paste0(site, "_HH_daily")) %>%
      arrange(date) %>%
      mutate(
        Tair_lag3           = lag(Tair, 3),
        Tair_lag5           = lag(Tair, 5),
        Tair_lag7           = lag(Tair, 7),
        VPD_lag3            = lag(VPD, 3),
        VPD_lag5            = lag(VPD, 5),
        VPD_lag7            = lag(VPD, 7),
        cumulative_cwd_7day = zoo::rollapply(cwd, 7, sum, align = "right", fill = NA)
      ) %>%
      drop_na(GPP, Tair, SW_rad, Precip, VPD,
              VPD_lag3, Tair_lag3,
              cumulative_cwd_7day, LCC_daily) %>%
      mutate(site = site)
  })
) %>%
  left_join(pft_lookup, by = "site")

cat("Total observations across all sites:", nrow(all_sites_df), "\n")
cat("Sites with data:", n_distinct(all_sites_df$site), "\n\n")

# -----------------------------------------------------------------------------
# 2. Model formula definitions
# -----------------------------------------------------------------------------

formula_simple <- GPP ~ s(Tair) + s(SW_rad) + s(VPD)

formula_complex <- GPP ~ s(Tair) + s(SW_rad) + s(Precip) +
  s(VPD) + s(VPD_lag3) + s(Tair_lag3) +
  s(cumulative_cwd_7day) + s(LCC_daily)

# -----------------------------------------------------------------------------
# 3. LOSO cross-validation function
# -----------------------------------------------------------------------------

run_loso <- function(data, formula, model_name) {

  sites <- unique(data$site)

  results <- map_dfr(sites, function(held_out) {

    train <- data %>% filter(site != held_out)
    test  <- data %>% filter(site == held_out)

    # Fit on all other sites
    model <- tryCatch(
      gam(formula, data = train, method = "REML"),
      error = function(e) {
        message("Model failed for holdout site: ", held_out, " — ", e$message)
        return(NULL)
      }
    )

    if (is.null(model)) {
      return(tibble(
        held_out_site = held_out,
        model         = model_name,
        R2            = NA_real_,
        RMSE          = NA_real_,
        n_test        = nrow(test)
      ))
    }

    preds <- predict(model, newdata = test)
    resid <- test$GPP - preds

    r2   <- cor(test$GPP, preds, use = "complete.obs")^2
    rmse <- sqrt(mean(resid^2, na.rm = TRUE))

    tibble(
      held_out_site = held_out,
      model         = model_name,
      R2            = r2,
      RMSE          = rmse,
      n_test        = nrow(test)
    )
  })

  return(results)
}

# -----------------------------------------------------------------------------
# 4. Run LOSO for both models
# -----------------------------------------------------------------------------

cat("Running LOSO for simple model...\n")
loso_simple  <- run_loso(all_sites_df, formula_simple,  "Simple")

cat("Running LOSO for complex model...\n")
loso_complex <- run_loso(all_sites_df, formula_complex, "Complex")

loso_results <- bind_rows(loso_simple, loso_complex) %>%
  left_join(pft_lookup, by = c("held_out_site" = "site")) %>%
  mutate(held_out_site = factor(held_out_site, levels = wet_dry_order))

# -----------------------------------------------------------------------------
# 5. Summary statistics
# -----------------------------------------------------------------------------

loso_summary <- loso_results %>%
  group_by(model) %>%
  summarise(
    mean_R2   = mean(R2,   na.rm = TRUE),
    median_R2 = median(R2, na.rm = TRUE),
    sd_R2     = sd(R2,     na.rm = TRUE),
    mean_RMSE = mean(RMSE, na.rm = TRUE),
    sd_RMSE   = sd(RMSE,   na.rm = TRUE),
    .groups   = "drop"
  )

cat("\n--- LOSO Summary ---\n")
print(loso_summary)

# Summary by PFT
loso_by_pft <- loso_results %>%
  group_by(model, pft) %>%
  summarise(
    mean_R2   = mean(R2,   na.rm = TRUE),
    sd_R2     = sd(R2,     na.rm = TRUE),
    n_sites   = n(),
    .groups   = "drop"
  )

cat("\n--- LOSO Summary by PFT ---\n")
print(loso_by_pft)

# -----------------------------------------------------------------------------
# 6. Delta R² (complex - simple) per site
# -----------------------------------------------------------------------------

loso_wide <- loso_results %>%
  select(held_out_site, model, R2, RMSE, pft) %>%
  pivot_wider(names_from = model, values_from = c(R2, RMSE)) %>%
  mutate(
    delta_R2   = R2_Complex   - R2_Simple,
    delta_RMSE = RMSE_Complex - RMSE_Simple  # negative = complex is better
  )

cat("\n--- Delta R² (Complex - Simple) per site ---\n")
print(loso_wide %>% select(held_out_site, pft, R2_Simple, R2_Complex, delta_R2))

```

```{r}


# -----------------------------------------------------------------------------
# 7. Plots
# -----------------------------------------------------------------------------

# --- 7a. R² by site for both models (wet → dry) ---
p1 <- loso_results %>%
  ggplot(aes(x = held_out_site, y = R2, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  scale_fill_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Held-out site (Wet → Dry)",
    y     = expression(R^2),
    fill  = "Model",
    title = "LOSO Generalisation: Simple vs. Complex GAM"
  )

print(p1)

# --- 7b. Delta R² per site, coloured by PFT ---
p2 <- loso_wide %>%
  ggplot(aes(x = held_out_site, y = delta_R2, fill = pft)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Held-out site (Wet → Dry)",
    y     = expression(Delta * R^2 ~ "(Complex - Simple)"),
    fill  = "PFT",
    title = "Does added complexity help generalisation?",
    subtitle = "Positive = complex model generalises better; negative = overfitting"
  )

print(p2)

# --- 7c. Boxplot comparing model distributions ---
p3 <- loso_results %>%
  ggplot(aes(x = model, y = R2, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(colour = pft), width = 0.15, size = 2.5, alpha = 0.8) +
  scale_fill_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  scale_colour_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    x      = "Model",
    y      = expression(R^2),
    fill   = "Model",
    colour = "PFT",
    title  = "Distribution of LOSO R² across all sites"
  )

print(p3)

# --- 7d. R² by PFT ---
p4 <- loso_results %>%
  ggplot(aes(x = pft, y = R2, fill = model)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.75)) +
  geom_jitter(aes(colour = model),
              position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
              size = 2, alpha = 0.8) +
  scale_fill_manual(values   = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  scale_colour_manual(values = c("Simple" = "steelblue", "Complex" = "firebrick")) +
  theme_minimal(base_size = 13) +
  labs(
    x     = "Plant Functional Type",
    y     = expression(R^2),
    fill  = "Model",
    colour = "Model",
    title = "LOSO generalisation by PFT"
  )

print(p4)

```

```{r}

# -----------------------------------------------------------------------------
# 8. Save results tables
# -----------------------------------------------------------------------------

write_csv(loso_results, "loso_results_all_sites.csv")
write_csv(loso_wide,    "loso_results_wide.csv")
write_csv(loso_summary, "loso_summary.csv")
write_csv(loso_by_pft,  "loso_by_pft.csv")

cat("\nDone. Results saved to CSV.\n")
```

**Results:**
Gin and Dap consistently perform well, both within site and out of site.
Ade, Wom, Whr also perform well

PFT:
Gin = WSA
Dap = GRA
Ade = WSA
Wom = EBF
Whr = EBF

Precip variability:
- All sites on the lower end of variability, Whr and Wom more variable than Rdf

Precip:
- All these sites are intermediate mean annual precip, Whr is pretty dry

Years active:
Gin = 3
Dap = 6
Ade = 2
Wom = 4
Whr = 3

Not even necessarily the longest running sites.

The common thread here is that these sites have intermediate precipitation, with intermediate precipitation variability. 
Sites that are intermediate are both easily predictable and generalisable

Cpr, Rdf and Gww consistently performed not very well, both for within site and out of site GAM testing.
Cpr = SAV
Gww = SAV
Rdf = WSA

In terms of precipitation variability (daily) - coefficient of variance. Gww is the most variable, Cpr also quite variable. Rdf isn't that variable in terms of daily variability.
Rdf, Gww all in the middle precip and Cpr is on the drier end.

In terms of fluxnet years active:
Cpr = 4
Gww = 1
Rdf = 2

```{r}
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

loso_wide$held_out_site <- factor(loso_wide$held_out_site, levels = precip_variability_order)

ggplot(loso_wide, aes(x = held_out_site, y = R2_Simple)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site (most to least variability in precip CV)",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites"
  ) +
  theme_minimal()
```

```{r}
ggplot(loso_wide, aes(x = held_out_site, y = R2_Complex)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site (most to least variable precipitation CV)",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites"
  ) +
  theme_minimal()
```

These plots confirm that site predictability and generalisability is controlled by precipitation variability. 

Next steps:
- Look at wet/dry season
- 7 day cwd might be too short for high precipitation variability sites
- interaction terms?
- Identify 3 highest CV sites and 3 lowest CV sites, plot actual versus predicted GPP time series - am i missing pulses? overestimating during drought?



