---
title: "GAM_analysis"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

How does predictability vary across ecosystems in Australia, how does VPD affect predictability?

**1. Sequential order**
M1: Climate (precip, Tair, SW rad)
M2: Climate + VPD
M3a: Climate + VPD + lags 3
M3b: Climate + VPD + lags 5
M3c: Climate + VPD + lags 7
M4: Climate + VPD + best lag + cwd7
M5: Climate + VPD + best lag + cwd7 + LCC

**2. Directly compare**
- Model 1 + Model 2 (unique contribution of VPD)

Load in water stress component
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)
library(lattice)

priestley_taylor_pet <- function(Rn, Tair, P) {
  
  alpha = 1.26      # PT constant
  cp <- 1.013e-3    # Specific heat of air (MJ/kg°C)
  epsilon <- 0.622  # Ratio of molecular weights of water vapor to dry air
  lambda <- 2.45    # Latent heat of vaporization (MJ/kg)
  G = 0             # Going to assume no ground heat flux on daily timescales
  
  # Psychrometric constant (kPa/°C)
  gamma <- (cp * P) / (epsilon * lambda)
  
  # Saturation vapor pressure (kPa)
  es <- 0.6108 * exp((17.27 * Tair) / (Tair + 237.3))
  
  # Slope of saturation vapor pressure curve (kPa/°C)
  delta <- (4098 * es) / ((Tair+ 237.3)^2)
  
  # Priestley-Taylor PET calculation (mm/day)
  pet <- alpha * (delta / (delta + gamma)) * ((Rn - G) / (lambda * 1e6)) * 86400

  return(pet)
}

sites_daily <- sites_daily %>%
  map(~ .x %>%
        mutate(
          pet = priestley_taylor_pet(Rn, Tair, Pa),
          cwd = pet - ET
        ))
 
```


Add in day of year and year again for the LCC data:
```{r}
library(dplyr)
library(lubridate)

# Define a function to add year and day-of-year
add_year_doy <- function(df) {
  df %>%
    mutate(
      year = year(date),
      doy  = yday(date)
    )
}

sites_daily <- sites_daily %>%
  map(add_year_doy)

```

```{r}
library(dplyr)
library(lubridate)
library(purrr)
library(terra)
library(ncdf4)

gome_file <- "/Users/alicerolfe/year_4_project/fluxnet_project/Data/GOME2_VcmaxTg_05deg.tif"
vcmax_raster <- rast(gome_file)

lcc_nc <- "/Users/alicerolfe/year_4_project/fluxnet_project/Chl_2003_2011_weeklyAll_01Deg_sf100.nc"

# --- 1. Functions to get VCmax and daily LCC ---
get_vcmax <- function(lat, lon, vcmax_raster) {
  coords <- data.frame(lon = lon, lat = lat)
  vcmax <- terra::extract(vcmax_raster, coords)
  as.numeric(vcmax[1,2])
}

latlon_to_line_sample <- function(lat, lon, nlines = 1800, nsamples = 3600) {
  lat_res <- 180 / nlines  
  line <- round((90 - lat) / lat_res)
  lon_res <- 360 / nsamples
  sample <- round((lon + 180) / lon_res)
  list(
    line   = pmin(pmax(line, 1), nlines),
    sample = pmin(pmax(sample, 1), nsamples)
  )
}

get_daily_LCC <- function(lcc_nc, lat, lon) {
  nc <- nc_open(lcc_nc)
  coords <- latlon_to_line_sample(lat, lon)
  
  pixel <- ncvar_get(nc, "chl_all",
                     start = c(coords$line, coords$sample, 1, 1),
                     count = c(1, 1, -1, -1))
  
  fillvalue <- ncatt_get(nc, "chl_all", "_FillValue")$value
  pixel[pixel == fillvalue] <- NA
  nc_close(nc)
  
  LCC <- pixel / 100
  weekly_clim <- rowMeans(LCC, na.rm = TRUE)
  
  weeks <- 1:length(weekly_clim)
  daily_weeks <- seq(1, max(weeks), by = 1/7)
  approx(weeks, weekly_clim, xout = daily_weeks)$y
}

# --- 2. Function to attach VCmax and LCC to a site dataframe ---
attach_site_data <- function(df, lat, lon, vcmax_raster, lcc_nc) {
  
  vcmax_val <- get_vcmax(lat, lon, vcmax_raster)
  daily_LCC <- get_daily_LCC(lcc_nc, lat, lon)
  daily_LCC_full <- rep(daily_LCC, length.out = nrow(df))
  
  df %>%
    mutate(
      vcmax      = vcmax_val,
      LCC_daily  = daily_LCC_full,
      year       = year(date),
      doy        = yday(date)
    )
}

# --- 3. Map over sites_daily using site metadata ---
site_meta <- tibble(
  name = names(sites_daily),
  lat  = c(-12.4943, -13.0769, -14.1593, -14.0633, -15.2588, -17.1507, -22.283, -33.6152, -23.8587,
           -12.5452, -31.3764, -30.1913, -34.4704, -14.5636, -36.6499, -17.1175, -22.287, -37.4259,
           -36.6732, -37.4222, -34.9893, -34.0021, -35.6566),
  lon  = c(131.1523, 131.1178, 131.3881, 131.3181, 132.3706, 133.3502, 133.2490, 150.7236, 148.4746,
           131.3072, 115.7138, 120.6541, 140.6551, 132.4776, 145.5759, 145.6301, 133.64, 145.1878,
           145.0294, 144.0944, 146.2907, 140.5891, 148.1517)
)

sites_daily <- pmap(
  list(sites_daily, site_meta$lat, site_meta$lon),
  function(df, lat, lon) attach_site_data(df, lat, lon, vcmax_raster, lcc_nc)
)

```


```{r}
list2env(sites_daily, envir = .GlobalEnv)
```


Everything loaded in properly.

1. Fit GAM for each site to find R-squared and compare site predictability across OzFlux sites

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(zoo)
library(mgcv)

set.seed(123)  # for reproducibility

# List all sites
aus_sites <- c("How", "Ade", "Das", "Dap", "Dry", "Stp", "Asm",
               "Cum", "Emr", "Fog", "Gin", "Gww", "Lox",
               "Rdf", "Rig", "Rob", "TTE", "Wac", "Whr",
               "Wom", "Ync", "Cpr", "Tum")

# Split, rolling CWD, lagged variables
for(site in aus_sites){
  
  df <- get(paste0(site, "_HH_daily")) %>%
    arrange(date)  # ensure temporal order
  
  # Create lagged variables BEFORE splitting
  df <- df %>%
    mutate(
      Tair_lag3 = lag(Tair, 3),
      Tair_lag5 = lag(Tair, 5),
      Tair_lag7 = lag(Tair, 7),
      VPD_lag3 = lag(VPD, 3),
      VPD_lag5 = lag(VPD, 5),
      VPD_lag7 = lag(VPD, 7),
      cumulative_cwd_7day = rollapply(cwd, 7, sum, align="right", fill = NA)
    ) %>%
    drop_na()
  
  # Temporal split: 80% train, 20% test
  n_train <- floor(0.8 * nrow(df))
  train <- df[1:n_train, ]
  test <- df[(n_train+1):nrow(df), ]
  
  # Save to global environment
  assign(paste0(site, "_train"), train, envir = .GlobalEnv)
  assign(paste0(site, "_test"), test, envir = .GlobalEnv)
}

```


```{r}

#Model 1 - climate only (Air temp, Shortwave radiation, precipitation)

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```



```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_metrics$site <- factor(climate_metrics$site, levels = wet_dry_order)

ggplot(climate_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()

```

Seems to be trend of declining R-squared from high precipitation coefficient of variation to low coefficient of variation. 

```{r}
#Model 2 - climate (temp, precip, SW rad) and VPD

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

climate_vpd_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}
# Define wet to dry order
wet_dry_order <- c("Rob", "How", "Wac", "Das", "Fog", "Dap", "Ade", "Dry", "Stp", "Wom", "Rdf", "Tum", "Gin", "Emr", "Cum", "Gww", "Rig", "Asm", "Cpr", "Whr", "TTE", "Ync", "Lox")

#most to least variable
precip_variability_order <- c("Gww", "TTE", "Asm", "Cpr", "Ync", "Lox", "Emr","Whr", "Wom", "Tum", "Rig", "Wac", "Stp", "Dry", "Gin", "Cum", "Fog", "Das", "Rob", "Rdf", "Dap", "How", "Ade")

climate_vpd_metrics$site <- factor(climate_vpd_metrics$site, levels = wet_dry_order)

ggplot(climate_vpd_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#compare the no VPD and climate + VPD model

library(dplyr)

gam_comparison <- climate_metrics %>%
  select(site, R2) %>%
  rename(R2_no_vpd = R2) %>%
  
  inner_join(
    climate_vpd_metrics %>%
      select(site, R2) %>%
      rename(R2_VPD = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_VPD = R2_VPD - R2_no_vpd
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_VPD)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Added predictive value of VPD across OzFlux sites"
  )

```
Potentially here, we have the difference between wet sites and dry sites. In the wetter sites, current VPD improved model R-squared much more than drier sites. Could it be that lags matter more in dry sites? Cumulative water stress? Perhaps in wet sites, plants respond rapidly to VPD, explaining a lot more variation than in drier sites.

```{r}
#Model 3a - climate (temp, SW rad, precip), VPD, + lags (3 days)

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag3) + s(Tair_lag3),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag3_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}

lag3_metrics$site <- factor(lag3_metrics$site, levels = wet_dry_order)

ggplot(lag3_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```


```{r}

library(dplyr)

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 3 day climate lags across OzFlux sites"
  )

```


```{r}
#Model 3b - climate + VPD + climate lags 5 days

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag5) + s(Tair_lag5),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag5_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)

```

```{r}

lag5_metrics$site <- factor(lag5_metrics$site, levels = wet_dry_order)

ggplot(lag5_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```
```{r}
#Compare no lags and lags with 5 days

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag = R2_lag - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
ggplot(gam_comparison, aes(x = site, y = delta_R2_lag)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site (Wet -> Dry)",
    y = expression(Delta*R^2),
    title = "Extra explanatory power of 5 day climate lags across OzFlux sites"
  )

```

```{r}
#Model 3c - climate + VPD + 7 day climate lags

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag7) + s(Tair_lag7),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lag7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```

```{r}

lag7_metrics$site <- factor(lag7_metrics$site, levels = wet_dry_order)

ggplot(lag7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```
```{r}
#compare no lags, 3 day lags, 5 day and 7 day (Model 3a, b, c):

gam_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(R2_no_lags = R2) %>%
  
  inner_join(
    lag7_metrics %>%
      select(site, R2) %>%
      rename(R2_lag7 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(R2_lag5 = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(R2_lag3 = R2),
    by = "site"
  ) %>%
  
  mutate(
    delta_R2_lag3 = R2_lag3 - R2_no_lags,
    delta_R2_lag5 = R2_lag5 - R2_no_lags,
    delta_R2_lag7 = R2_lag7 - R2_no_lags
  )


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

gam_comparison <- gam_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

gam_comparison
```

```{r}
# Calculate mean improvement for each lag
lag_summary <- gam_comparison %>%
  summarise(
    mean_lag3 = mean(delta_R2_lag3, na.rm = TRUE),
    mean_lag5 = mean(delta_R2_lag5, na.rm = TRUE),
    mean_lag7 = mean(delta_R2_lag7, na.rm = TRUE)
  )

print(lag_summary)

# Then plot with reference lines
gam_comparison %>%
  pivot_longer(
    cols = c(delta_R2_lag3, delta_R2_lag5, delta_R2_lag7),
    names_to = "lag_type",
    values_to = "delta_R2"
  ) %>%
  mutate(
    lag_type = factor(lag_type, 
                     levels = c("delta_R2_lag3", "delta_R2_lag5", "delta_R2_lag7"),
                     labels = c("3-day", "5-day", "7-day"))
  ) %>%
  ggplot(aes(x = lag_type, y = delta_R2)) +
  geom_boxplot(fill = "firebrick", alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Lag window",
    y = expression(Delta*R^2),
    title = "Distribution of lag window performance across all sites"
  )

```
Let's use lag 7 - similar performances. 7 days aligns with typical plant physiological memory and maintains consistency with the cumulative water deficit over 7 days.

```{r}
#Model 4: Climate + cwd + lags + VPD

fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag7) + s(Tair_lag7) +s(cumulative_cwd_7day),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

cwd7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```

```{r}
cwd7_metrics$site <- factor(cwd7_metrics$site, levels = wet_dry_order)

ggplot(cwd7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```


```{r}
#Model 5: Climate + cwd + lags + VPD + LCC


fit_gam_and_metrics <- function(train_data, test_data) {
  
  model <- gam(
    GPP ~ s(Tair) + s(SW_rad) + s(Precip) + s(VPD) + s(VPD_lag7) + s(Tair_lag7) +s(cumulative_cwd_7day) + s(LCC_daily),
    data = train_data,
    method = "REML"
  )
  
  preds <- predict(model, newdata = test_data)
  
  rmse <- sqrt(mean((test_data$GPP - preds)^2))
  rsq  <- cor(test_data$GPP, preds)^2
  
  tibble(
    RMSE = rmse,
    R2   = rsq
  )
}


# Run GAM for all sites and combine results

lcc7_metrics <- bind_rows(
  lapply(aus_sites, function(site) {
    fit_gam_and_metrics(
      get(paste0(site, "_train")),
      get(paste0(site, "_test"))
    ) %>%
      mutate(site = site)
  })
)


```

```{r}
lcc7_metrics$site <- factor(lcc7_metrics$site, levels = wet_dry_order)

ggplot(lcc7_metrics, aes(x = site, y = R2)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Site",
    y = expression(R^2),
    title = "Predictability of GPP across OzFlux sites (dry to wet)"
  ) +
  theme_minimal()
```

```{r}
#table of all models R-squared 

final_comparison <- climate_vpd_metrics %>%
  select(site, R2) %>%
  rename(model_2 = R2) %>%
  
  inner_join(
    lag7_metrics %>%
      select(site, R2) %>%
      rename(model_3c = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag5_metrics %>%
      select(site, R2) %>%
      rename(model_3b = R2),
    by = "site"
  ) %>%
  
  inner_join(
    lag3_metrics %>%
      select(site, R2) %>%
      rename(model_3a = R2),
    by = "site"
  ) %>%
  
   inner_join(
    climate_metrics %>%
      select(site, R2) %>%
      rename(model_1 = R2),
    by = "site"
  ) %>%
  
   inner_join(
    lcc7_metrics %>%
      select(site, R2) %>%
      rename(model_5 = R2),
    by = "site"
  ) %>%
  
   inner_join(
    cwd7_metrics %>%
      select(site, R2) %>%
      rename(model_4 = R2),
    by = "site"
  ) 


wet_dry_order <- c("Wac", "Rob","How", "Das", "Fog", "Rdf", "Dap", "Ade", "Tum", "Wom", "Dry", "Stp", "Ync", "Emr", "Gin", "Cum", "Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox")

final_comparison <- final_comparison %>%
  mutate(site = factor(site, levels = wet_dry_order)) %>%
  arrange(site)

final_comparison
```

```{r}
final_comparison %>%
  pivot_longer(
    cols = starts_with("model_"),
    names_to = "model",
    values_to = "R2"
  ) %>%
  mutate(
    model = factor(model,
                   levels = c("model_1", "model_2", "model_3a", 
                             "model_3b", "model_3c", "model_4", "model_5"),
                   labels = c("Climate", "+VPD", "+Lag3", "+Lag5", 
                             "+Lag7", "+CWD", "+LCC"))
  ) %>%
  ggplot(aes(x = model, y = site, fill = R2)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Model",
    y = "Site (Wet → Dry)",
    fill = expression(R^2),
    title = "Model performance across sites and complexity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
# Create a figure showing best model per site
final_comparison %>%
  pivot_longer(cols = starts_with("model_"), 
               names_to = "model", values_to = "R2") %>%
  group_by(site) %>%
  slice_max(R2, n = 1) %>%
  mutate(
    moisture = case_when(
      site %in% c("Rob", "Wac", "How", "Das", "Fog", "Rdf", "Dap", "Ade") ~ "Wet",
      site %in% c("Rig", "Asm", "Whr", "Gww", "Cpr", "TTE", "Lox") ~ "Dry",
      TRUE ~ "Medium"
    )
  ) %>%
  ggplot(aes(x = site, y = model, fill = R2)) +
  geom_tile() +
  facet_grid(~moisture, scales = "free_x", space = "free_x") +
  scale_fill_viridis_c() +
  labs(title = "Best-performing model varies by moisture regime")
```




